### 2025-04-21 paper/main.tex
results + plots + visualization + paper

% !TEX root = main.tex
\documentclass[twocolumn]{article}
\usepackage[a4paper,margin=1in]{geometry} 
\usepackage{lmodern}
\usepackage{setspace}
\usepackage{mathptmx}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{amsmath}
\usepackage{booktabs}
\usepackage{float}

\title{\huge \textbf{Words That Matter:\\ Vocabulary Pruning in ModernBERT}}
\author{Wout De Rijck, Dr. Ir. Karel D'Oosterlinck, \\ Prof. Dr. Ir. Thomas Demeester, Prof. Dr. Ir. Chris Develder}
\date{} % No date

\begin{document}
\maketitle

% Motivation / Context (1–2 sentences):
% Start with the why. What's the broader problem or challenge? Why is this area important right now (e.g., limitations of current LLMs, safety, interpretability, efficiency)?

% Gap / Problem Statement:
% What specific problem or limitation does your paper address? This sets the stage for your contribution. It should hint at a shortcoming in current approaches without being overly critical.

% Your Approach / Contribution:
% What did you do? This is the heart of the abstract. It could be:

% a new method or model

% a novel dataset or benchmark

% a unique analysis or framework

% an improvement on performance, interpretability, etc.

\begin{abstract}
Large language models like ModernBERT face deployment challenges due to their computational demands and memory footprint, limiting their use in resource-constrained environments. While various pruning techniques exist, most require resource-intensive fine-tuning, creating a bottleneck for efficient model adaptation. We introduce a novel vocabulary pruning approach for ModernBERT that requires no additional fine-tuning, delivering superior efficiency particularly at smaller pruning ratios. Our technique specifically targets the embedding layer, which constitutes approximately 25\% of the model's parameters, through a systematic token importance analysis. We demonstrate that combining our vocabulary pruning method with LoSparse techniques for encoder layers achieves comprehensive model compression with minimal performance degradation. Experimental results show our approach reduces model size by up to X\% while maintaining Y\% of the original performance across standard benchmarks. This work establishes a practical pathway for creating lightweight, task-specific transformer models without the computational burden of traditional pruning methods.
\end{abstract}

% Key Results:
% Give a quick summary of your main findings or improvements—include metrics or comparisons if possible, but keep it digestible.

% Implications / Why It Matters:
% What's the takeaway? How might this move the field forward, or be used by other researchers or practitioners?

% Tone and Tease:
% While all of the above should be there, the language should still invite curiosity. Think clear, precise, but also hook-y—like:
% "We show that…", "Surprisingly, we find…", "This reveals a new direction for…"

\section{Introduction}
% Look at the introduction of ModernBERT paper, interesting!
1. Introduction of the ModernBERT paper. \\
2. Architecture of ModernBERT.

\section{Related work}
Model compression techniques enhance the efficiency of large language models (LLMs) by reducing their size and computational requirements. Most modern LLM compression approaches operate under post-training settings, eliminating the need for resource-intensive retraining. Model compression strategies for LLMs can be categorized into four primary methods:

\begin{enumerate}
    \item Quantization: Reduces the numerical precision of model weights, decreasing memory footprint while maintaining reasonable performance.

    \item Parameter pruning: Eliminates redundant or less important connections and neurons to create sparser models with fewer parameters.

    \item Low-rank approximation: Decomposes weight matrices into lower-dimensional representations that capture essential information while requiring fewer parameters.

    \item Knowledge distillation: Transfers knowledge from larger teacher models to smaller student models, enabling competitive performance with reduced architecture size.
\end{enumerate}

These approaches are complementary and can be combined to achieve optimal compression from different perspectives. 
This work focuses on pruning techniques to reduce model size while preserving performance.
For ModernBERT, an encoder-only model, this research examines specific pruning methods for encoder layers and separate techniques for the embedding matrix.

\subsection{Encoder Layer Pruning}

\subsubsection{LoSparse}
Low-Rank and Sparse approximation (LoSparse) introduces a model compression strategy that decomposes each weight matrix W as the sum of a low-rank component (UV) and a sparse component (S). The low-rank part, computed via singular value decomposition, extracts the shared, coherent features among neurons, while the sparse part targets and removes redundant, incoherent features. This dual approach overcomes the limitations of traditional methods by avoiding the excessive pruning of expressive neurons and preserving diversity in the representation. Extensive experiments on natural language understanding, question answering, and generation tasks show that LoSparse achieves significant parameter reduction all while maintaining competitive performance levels.

\subsubsection{Other encoder layer pruning techniques}

\subsection{Vocabulary Pruning}
While encoder layer pruning techniques like LoSparse have shown promising results, the embedding layer presents a unique challenge in transformer models. Accounting for approximately 25\% of the total parameters, the vocabulary embedding layer offers a substantial opportunity for model compression. However, unlike encoder layer pruning, vocabulary reduction requires careful consideration of semantic relationships and token importance to maintain model performance.

\subsubsection{Direction is what you need}
% Direction is what you need: Improving Word Embedding Compression in Large Language Models
\subsubsection{TF-IDF}
% Enhancing performance of transformer-based models in natural language understanding through word importance embedding

\subsubsection{PEMA}
% Vocabulary-level Memory Efficiency for Language Model Fine-tuning
Partial Embedding Matrix Approximation (PEMA) makes finetuning efficient by removing the tokens in the embedding layers that are not used in the current task.

\subsubsection{LightToken}
LightToken is a lightweight token embedding framework that produces compressed token embeddings in a task and model-agnostic fashion. It's designed to be compatible with different architectures and applicable to any downstream task. The framework integrates low-rank approximation, a novel residual binary autoencoder, and a new compression loss function to significantly improve model compression ratio without sacrificing performance.

\newpage
TODO

\newpage
\section{Method}

The proposed hybrid vocabulary pruning method for ModernBERT targets the embedding layer, which constitutes approximately 25\% of the model's parameters. This approach is based on the observation that tokens in a vocabulary have varying importance for downstream tasks, and that removed tokens can be mapped to semantically similar ones instead of a single unknown token. The method consists of three main components: token importance analysis, selective pruning, and semantic clustering for out-of-vocabulary (OOV) tokens.

\subsection{Token Importance Analysis}
Multiple approaches for token importance estimation were systematically evaluated:

\begin{enumerate}
    \item \textbf{Random Selection}: Tokens are pruned randomly without consideration for importance, serving as a baseline approach.

    \item \textbf{Frequency-Based}: Tokens are ranked by their frequency in the target dataset, with least frequently used tokens pruned first. This approach assumes that rarely used tokens contribute less to model performance.

    \item \textbf{Clustering-Based}: This approach employs agglomerative or k-means clustering on the embedding space to group tokens with similar semantic properties. For each cluster, the token closest to the centroid is retained as a representative, while others are pruned. This preserves semantic diversity across the vocabulary while reducing redundancy.

    \item \textbf{Attention-Based}: A more sophisticated approach using a finetuned model to extract attention scores across the dataset. For each token, attention patterns across all layers are aggregated to determine its contextual importance.

    \item \textbf{TF-IDF Based}: Tokens are ranked using Term Frequency-Inverse Document Frequency scores, which balance token occurrence frequency with discriminative power across documents.
\end{enumerate}

The TF-IDF approach demonstrated superior performance and was adopted as the primary method. For this approach, three normalization variants were tested:
\begin{itemize}
    \item Non-normalized TF-IDF (prioritizing rare but task-specific tokens)
    \item L1-normalized TF-IDF (balancing frequency and importance)
    \item L2-normalized TF-IDF (standard approach, providing optimal results)
\end{itemize}

The TF-IDF method effectively identifies task-relevant tokens by assigning higher importance to tokens that frequently appear in specific documents but are less common across the entire corpus. This enables selective vocabulary pruning while preserving tokens critical for maintaining task performance.

\subsection{Hybrid Pruning Method}
The hybrid approach combines importance-based token pruning with semantic clustering for OOV token handling. The method is formalized in Algorithm~\ref{alg:hybrid_pruning}.

\begin{algorithm}[H]
\caption{Hybrid Vocabulary Pruning}
\label{alg:hybrid_pruning}
\begin{algorithmic}[1]
\scriptsize
\Require Model $M$ with vocab $V$, dataset $D$, pruning \% $p$, clusters $k$, method $method$
\Ensure Pruned model $M'$ with reduced vocabulary

\State // Stage 1: Calculate token importance using selected method (random, frequency, attention, or TF-IDF)
\State $importance \gets \text{CalculateTokenImportance}(V, D, M, method)$

\State // Stage 2: Prune lowest-importance tokens based on pruning percentage
\State $tokens\_sorted \gets \text{SortByImportance}(V, importance)$
\State $keep\_count \gets |V| \times (1 - p/100)$
\State $tokens\_to\_keep \gets tokens\_sorted[1:keep\_count]$
\State $tokens\_to\_remove \gets V \setminus tokens\_to\_keep$

\State // Stage 3: Cluster removed tokens and identify representatives for OOV mapping
\State $embeddings \gets M.embedding\_layer.weight$
\State $removed\_embeddings \gets embeddings[tokens\_to\_remove]$
\State $kmeans \gets \text{KMeans}(n\_clusters=k)$
\State $clusters \gets kmeans.fit\_predict(removed\_embeddings)$
\State $oov\_map \gets \{\}$; $centers \gets \{\}$

\For{each cluster $c$ in $range(k)$}
    \State $indices \gets \{i : clusters[i] = c\}$
    \If{$indices$ is not empty}
        \State $centroid \gets kmeans.cluster\_centers_[c]$
        \State $distances \gets \{\|removed\_embeddings[i] - centroid\| : i \in indices\}$
        \State $closest\_idx \gets \arg\min(distances)$ 
        \State $center\_token \gets tokens\_to\_remove[indices[closest\_idx]]$
        \State $centers.add(center\_token)$
        \For{$idx \in indices$} 
            \State $oov\_map[tokens\_to\_remove[idx]] \gets center\_token$ 
        \EndFor
    \EndIf
\EndFor

\State // Stage 4: Create reduced embedding matrix with kept tokens and cluster representatives
\State $all\_tokens \gets tokens\_to\_keep \cup centers$
\State $token\_map \gets \{old: new \mid new, old \in enumerate(all\_tokens)\}$
\State $reduced\_embeddings \gets embeddings[all\_tokens]$
\State $M'.embedding\_layer.weight \gets reduced\_embeddings$

\State \Return $M'$, $token\_map$, $oov\_map$
\end{algorithmic}
\end{algorithm}

\subsection{Out-of-Vocabulary Token Handling}
A key contribution of this approach is the handling of out-of-vocabulary (OOV) tokens. Rather than mapping all pruned tokens to a single UNK token, the semantic properties of the embedding space are utilized to create meaningful representations for OOV tokens.

The clustering process organizes removed tokens into semantic groups, with each group represented by the token closest to the cluster centroid. When the model encounters an OOV token during inference, it maps to the appropriate cluster representative, preserving semantic properties. This approach retains more information than traditional UNK token methods.

\subsection{Implementation Details}
The implementation requires no additional fine-tuning after vocabulary pruning, improving efficiency compared to approaches that require retraining. The method comprises three technical components:

\begin{enumerate}
    \item A token extraction and importance calculation module interfacing with downstream task datasets
    \item An embedding space K-means clustering algorithm creating semantic maps for OOV tokens
    \item A hybrid data collator handling tokenization and OOV mappings during model inference
\end{enumerate}

Four pruning mechanisms are supported:
\begin{itemize}
    \item Frequency-based pruning (retaining most frequent tokens)
    \item Importance-based pruning using non-normalized TF-IDF
    \item Importance-based pruning using L1-normalized TF-IDF 
    \item Importance-based pruning using L2-normalized TF-IDF (default)
\end{itemize}

Experimental results indicate that TF-IDF variants generally outperform pure frequency-based approaches, with L2-normalized TF-IDF providing the optimal balance between task-specific vocabulary retention and general language understanding.

\newpage
\section{Experiments}

\begin{table*}[ht!]
\centering
\small
\caption{Comparison of vocabulary pruning techniques on GLUE benchmark tasks. Results show accuracy on the development set with varying pruning methods. Best results for each task are highlighted in \textbf{bold}.}
\label{tab:results}
\setlength{\tabcolsep}{4pt}
\begin{tabular}{lccccccccc|c}
\toprule
\textbf{Method} & \textbf{Param. Red.} & \textbf{SST-2} & \textbf{MRPC} & \textbf{MNLI} & \textbf{CoLA} & \textbf{STS-B} & \textbf{QQP} & \textbf{QNLI} & \textbf{RTE} & \textbf{AVG} \\
\midrule
\multicolumn{11}{l}{\textit{Baseline}} \\
ModernBERT (Full) & 0.03\% & 0.951 & 0.856 & 0.881 & 0.586 & 0.917 & 0.917 & 0.916 & 0.598 & 0.828 \\
\midrule
\multicolumn{11}{l}{\textit{Vocabulary Pruning}} \\
Train Tokens Only & 6.7-22.6\% & 0.950 & 0.831 & \textbf{0.883} & 0.509 & \textbf{0.917} & \textbf{0.917} & 0.915 & 0.598 & 0.815 \\
\midrule
Random Selection & 9.0-23.3\% & 0.911 & 0.798 & 0.832 & 0.470 & 0.845 & 0.902 & 0.895 & 0.522 & 0.772 \\
Clustering-Based & 9.0-23.3\% & 0.899 & 0.714 & 0.712 & 0.000 & 0.786 & 0.875 & 0.836 & 0.510 & 0.667 \\
Frequency-Based & 9.0-23.3\% & 0.943 & 0.812 & 0.880 & 0.467 & 0.905 & 0.916 & \textbf{0.920} & 0.542 & 0.798 \\
TF-IDF Based & 9.0-23.3\% & 0.933 & 0.807 & 0.875 & 0.520 & 0.901 & 0.900 & 0.917 & \textbf{0.606} & \textbf{0.807} \\
\midrule
\multicolumn{11}{l}{\textit{Vocabulary Pruning with OOV Handling}} \\
Freq + OOV & 9.0-23.2\% & \textbf{0.938} & 0.828 & \textbf{0.881} & 0.540 & 0.906 & 0.916 & 0.918 & 0.538 & 0.808 \\
TF-IDF + OOV & 9.0-23.2\% & 0.923 & \textbf{0.834} & 0.878 & \textbf{0.615} & 0.903 & 0.906 & 0.919 & 0.554 & \textbf{0.817} \\
\bottomrule
\end{tabular}
\end{table*}

\begin{table}[ht!]
\centering
\caption{Average performance change across GLUE tasks with different pruning methods.}
\label{tab:avg_performance}
\begin{tabular}{lcc}
\toprule
\textbf{Method} & \textbf{Avg. Param. Reduction} & \textbf{Avg. Performance Drop} \\
\midrule
Train Tokens Only & 14.22\% & 1.25\% \\
Random Selection & 16.44\% & 5.58\% \\
Clustering-Based & 16.44\% & 16.24\% \\
Frequency-Based & 16.44\% & 2.32\% \\
TF-IDF Based & 16.44\% & 2.04\% \\
Frequency + OOV & 16.40\% & 1.82\% \\
TF-IDF + OOV & 16.40\% & 0.96\% \\
\bottomrule
\end{tabular}
\end{table}

\end{document}

### 2025-04-21 visualize_results.py

@@ -3,6 +3,7 @@ import matplotlib.pyplot as plt
 import numpy as np
 import csv
 import os
+from utils import glue_tasks  # Import the glue_tasks dictionary that contains dataset sizes
 
 # Function to properly parse a value like "0,65" to 0.65
 def parse_value(val):
@@ -17,6 +18,15 @@ def parse_value(val):
 
 # Ensure output directory exists
 os.makedirs('plots', exist_ok=True)
+
+# Create a mapping of datasets to their sizes from the glue_tasks dictionary
+dataset_sizes = {}
+for task_name, task_data in glue_tasks.items():
+    if "-" in task_name:  # Handle mnli-matched and mnli-mismatched
+        dataset_name = task_name.split("-")[0].upper()
+    else:
+        dataset_name = task_name.upper()
+    dataset_sizes[dataset_name] = task_data["size"]
 
 # Read CSV data more carefully
 datasets = []
@@ -38,7 +48,7 @@ clustering_train = []
 frequency_train = []
 importance_train = []
 
-with open('Results - Blad1.csv', 'r') as f:
+with open('results/SUMMARY.csv', 'r') as f:
     reader = csv.reader(f)
     # Skip header rows
     next(reader)  # Skip first header row
@@ -102,7 +112,8 @@ test_df = pd.DataFrame({
     'FREQUENCY': frequency_test,
     'IMPORTANCE': importance_test,
     'FREQUENCY_OOV': frequency_oov_test,
-    'IMPORTANCE_OOV': importance_oov_test
+    'IMPORTANCE_OOV': importance_oov_test,
+    'SIZE': [dataset_sizes.get(dataset, 'N/A') for dataset in datasets]  # Add dataset sizes
 }, index=datasets)
 
 train_df = pd.DataFrame({
@@ -111,11 +122,12 @@ train_df = pd.DataFrame({
     'RANDOM': random_train,
     'CLUSTERING': clustering_train,
     'FREQUENCY': frequency_train,
-    'IMPORTANCE': importance_train
+    'IMPORTANCE': importance_train,
+    'SIZE': [dataset_sizes.get(dataset, 'N/A') for dataset in datasets]  # Add dataset sizes
 }, index=datasets)
 
 # Print the DataFrame to verify data
-print("Test Results DataFrame:")
+print("Test Results DataFrame with dataset sizes:")
 print(test_df)
 
[…]

### 2025-04-21 paper/main.tex
Add attention scores

@@
 
-    \item \textbf{Attention-Based}: A more sophisticated approach using a finetuned model to extract attention scores across the dataset.
-For each token, attention patterns across all layers are aggregated to determine its contextual importance.
+    \item \textbf{Attention-Based}: This method uses attention patterns from fine-tuning to determine token importance.
+For each token, attention scores across all layers and heads are aggregated to capture its contextual relevance within the specific task.
+Tokens receiving consistently higher attention are considered more important for model decisions, providing a more nuanced importance measure that captures both frequency and semantic significance.
 
@@
 \multicolumn{11}{l}{\textit{Vocabulary Pruning}} \\
-Train Tokens Only & 6.7-22.6\% & 0.950 & 0.831 & \textbf{0.883} & 0.509 & \textbf{0.917} & \textbf{0.917} & 0.915 & 0.598 & 0.815 \\
+Train Tokens Only & 6.7-22.6\% & 0.950 & 0.831 & 0.883 & 0.509 & 0.917 & 0.917 & 0.915 & 0.598 & 0.815 \\
 \midrule
@@
 Clustering-Based & 9.0-23.3\% & 0.899 & 0.714 & 0.712 & 0.000 & 0.786 & 0.875 & 0.836 & 0.510 & 0.667 \\
-Frequency-Based & 9.0-23.3\% & 0.943 & 0.812 & 0.880 & 0.467 & 0.905 & 0.916 & \textbf{0.920} & 0.542 & 0.798 \\
-TF-IDF Based & 9.0-23.3\% & 0.933 & 0.807 & 0.875 & 0.520 & 0.901 & 0.900 & 0.917 & \textbf{0.606} & \textbf{0.807} \\
+Frequency-Based & 9.0-23.3\% & 0.943 & 0.812 & 0.880 & 0.467 & \textbf{0.905} & \textbf{0.916} & \textbf{0.920} & 0.542 & 0.798 \\
+Attention-Based & 8.8-23.1\% & \textbf{0.947} & \textbf{0.864} & 0.864 & \textbf{0.589} & 0.885 & 0.912 & 0.912 & 0.550 & 0.803 \\
+TF-IDF Based & 9.0-23.3\% & 0.933 & 0.807 & \textbf{0.875} & 0.520 & 0.901 & 0.900 & 0.917 & \textbf{0.606} & \textbf{0.807} \\
 \midrule
@@
 Freq + OOV & 9.0-23.2\% & \textbf{0.938} & 0.828 & \textbf{0.881} & 0.540 & 0.906 & 0.916 & 0.918 & 0.538 & 0.808 \\
-TF-IDF + OOV & 9.0-23.2\% & 0.923 & \textbf{0.834} & 0.878 & \textbf{0.615} & 0.903 & 0.906 & 0.919 & 0.554 & \textbf{0.817} \\
+TF-IDF + OOV & 9.0-23.2\% & 0.923 & 0.834 & 0.878 & 0.615 & 0.903 & 0.906 & 0.919 & 0.554 & \textbf{0.817} \\
 \bottomrule
@@
 Frequency-Based & 16.44\% & 2.32\% \\
+Attention-Based & 15.92\% & 1.99\% \\
 TF-IDF Based & 16.44\% & 2.04\% \\

### 2025-04-21 visualize_results.py

@@ -36,6 +36,7 @@ train_tokens_test = []
 random_test = []
 clustering_test = []
 frequency_test = []
+attention_test = []
 importance_test = []
 frequency_oov_test = []
 importance_oov_test = []
@@ -46,6 +47,7 @@ train_tokens_train = []
 random_train = []
 clustering_train = []
 frequency_train = []
+attention_train = []
 importance_train = []
 
 with open('results/SUMMARY.csv', 'r') as f:
@@ -73,7 +75,11 @@ with open('results/SUMMARY.csv', 'r') as f:
             else:
                 frequency_train.append(0.0)
             if len(row) > 12:
-                importance_train.append(parse_value(row[12]))
+                attention_train.append(parse_value(row[12]))
+            else:
+                attention_train.append(0.0)
+            if len(row) > 14:
+                importance_train.append(parse_value(row[14]))
             else:
                 importance_train.append(0.0)
             
@@ -90,15 +96,19 @@ with open('results/SUMMARY.csv', 'r') as f:
             else:
                 frequency_test.append(0.0)
             if len(row) > 13:
-                importance_test.append(parse_value(row[13]))
+                attention_test.append(parse_value(row[13]))
+            else:
+                attention_test.append(0.0)
+            if len(row) > 15:
+                importance_test.append(parse_value(row[15]))
             else:
                 importance_test.append(0.0)
-            if len(row) > 14:
-                frequency_oov_test.append(parse_value(row[14]))
+            if len(row) > 16:
+                frequency_oov_test.append(parse_value(row[16]))
             else:
                 frequency_oov_test.append(0.0)
-            if len(row) > 15:
-                importance_oov_test.append(parse_value(row[15]))
+            if len(row) > 17:
+                importance_oov_test.append(parse_value(row[17]))
             else:
                 importance_oov_test.append(0.0)
 
@@ -110,6 +120,7 @@ test_df = pd.DataFrame({
     'RANDOM': random_test,
[…]

### 2025-04-23 paper/main.tex
fix prune percentages table

@@
 \usepackage{booktabs}
+\usepackage{arydshln}
 \usepackage{float}
@@
 \label{tab:results}
-\setlength{\tabcolsep}{4pt}
-\begin{tabular}{lccccccccc|c}
+\setlength{\tabcolsep}{6pt}
+\begin{tabular}{lcccccccc|c}
 \toprule
-\textbf{Method} & \textbf{Param. Red.} & \textbf{SST-2} & \textbf{MRPC} & \textbf{MNLI} & \textbf{CoLA} & \textbf{STS-B} & \textbf{QQP} & \textbf{QNLI} & \textbf{RTE} & \textbf{AVG} \\
+\textbf{Method} & \textbf{SST-2} & \textbf{MRPC} & \textbf{MNLI} & \textbf{CoLA} & \textbf{STS-B} & \textbf{QQP} & \textbf{QNLI} & \textbf{RTE} & \textbf{AVG} \\
 \midrule
-\multicolumn{11}{l}{\textit{Baseline}} \\
-ModernBERT (Full) & 0.03\% & 0.951 & 0.856 & 0.881 & 0.586 & 0.917 & 0.917 & 0.916 & 0.598 & 0.828 \\
+\multicolumn{9}{l|}{\textit{Baseline}} \\
+ModernBERT (Full) & 0.951 & 0.856 & 0.881 & 0.586 & 0.917 & 0.917 & 0.916 & 0.598 & 0.828 \\
 \midrule
-\multicolumn{11}{l}{\textit{Vocabulary Pruning}} \\
-Train Tokens Only & 6.7-22.6\% & 0.950 & 0.831 & 0.883 & 0.509 & 0.917 & 0.917 & 0.915 & 0.598 & 0.815 \\
+\multicolumn{9}{l|}{\textit{Vocabulary Pruning}} \\
+Parameter Reduction & 18.85\% & 18.56\% & 6.74\% & 22.61\% & 18.76\% & 4.79\% & 6.42\% & 17.06\% &  \\
+Train Tokens Only & 0.950 & 0.831 & 0.883 & 0.509 & 0.917 & 0.917 & 0.915 & 0.598 & 0.815 \\
 \midrule
-Random Selection & 9.0-23.3\% & 0.911 & 0.798 & 0.832 & 0.470 & 0.845 & 0.902 & 0.895 & 0.522 & 0.772 \\
-Clustering-Based & 9.0-23.3\% & 0.899 & 0.714 & 0.712 & 0.000 & 0.786 & 0.875 & 0.836 & 0.510 & 0.667 \\
-Frequency-Based & 9.0-23.3\% & 0.943 & 0.812 & 0.880 & 0.467 & \textbf{0.905} & \textbf{0.916} & \textbf{0.920} & 0.542 & 0.798 \\
-Attention-Based & 8.8-23.1\% & \textbf{0.947} & \textbf{0.864} & 0.864 & \textbf{0.589} & 0.885 & 0.912 & 0.912 & 0.550 & 0.803 \\
-TF-IDF Based & 9.0-23.3\% & 0.933 & 0.807 & \textbf{0.875} & 0.520 & 0.901 & 0.900 & 0.917 & \textbf{0.606} & \textbf{0.807} \\
-\midrule
-\multicolumn{11}{l}{\textit{Vocabulary Pruning with OOV Handling}} \\
-Freq + OOV & 9.0-23.2\% & \textbf{0.938} & 0.828 & \textbf{0.881} & 0.540 & 0.906 & 0.916 & 0.918 & 0.538 & 0.808 \\
-TF-IDF + OOV & 9.0-23.2\% & 0.923 & 0.834 & 0.878 & 0.615 & 0.903 & 0.906 & 0.919 & 0.554 & \textbf{0.817} \\
+Parameter Reduction & 20.25\% & 20.02\% & 10.57\% & 23.27\% & 20.18\% & 9.01\% & 10.31\% & 18.83\% & \\
+Random Selection & 0.911 & 0.798 & 0.832 & 0.470 & 0.845 & 0.902 & 0.895 & 0.522 & 0.772 \\
+Clustering-Based & 0.899 & 0.714 & 0.712 & 0.000 & 0.786 & 0.875 & 0.836 & 0.510 & 0.667 \\
+Frequency-Based & 0.943 & 0.812 & 0.880 & 0.467 & \textbf{0.905} & \textbf{0.916} & \textbf{0.920} & 0.542 & 0.798 \\
+Attention-Based & \textbf{0.947} & \textbf{0.864} & 0.864 & \textbf{0.589} & 0.885 & 0.912 & 0.912 & 0.550 & 0.803 \\
+TF-IDF Based & 0.933 & 0.807 & \textbf{0.875} & 0.520 & 0.901 & 0.900 & 0.917 & \textbf{0.606} & \textbf{0.807} \\
+\\ [-8pt]
+\cdashline{1-10}
+\\[-8pt]
+Freq + OOV & \textbf{0.938} & 0.828 & \textbf{0.881} & 0.540 & 0.906 & 0.916 & 0.918 & 0.538 & 0.808 \\
+TF-IDF + OOV & 0.923 & 0.834 & 0.878 & 0.615 & 0.903 & 0.906 & 0.919 & 0.554 & \textbf{0.817} \\
 \bottomrule

### 2025-04-23 examples/run_random_pruning.py
parameter based random pruning

@@ -7,6 +7,7 @@ Random Vocabulary Pruning Script
 This script runs random vocabulary pruning on all GLUE benchmark tasks.
 Each task has customized hyperparameters for optimal performance.
 This serves as a baseline for comparison with other pruning methods.
+Supports both token-based and parameter-based pruning approaches.
 """
 
 import argparse
@@ -136,6 +137,12 @@ def parse_args():
     )
     
     parser.add_argument(
+        "--param_based", 
+        action="store_true",
+        help="If set, prune based on parameter percentage rather than token percentage"
+    )
+    
+    parser.add_argument(
         "--epochs", 
         type=int, 
         default=None,
@@ -227,8 +234,14 @@ def main():
     # Create output directory
     os.makedirs(args.output_dir, exist_ok=True)
     
+    # Create a subdirectory based on pruning type
+    pruning_type = "param_based" if args.param_based else "token_based"
+    pruning_dir = os.path.join(args.output_dir, pruning_type)
+    os.makedirs(pruning_dir, exist_ok=True)
+    
     # Log arguments
     logger.info(f"Running with arguments: {args}")
+    logger.info(f"Pruning type: {'Parameter-based' if args.param_based else 'Token-based'}")
     
     # Dictionary to store results for each task
     all_results = {}
@@ -255,7 +268,7 @@ def main():
             task_params["weight_decay"] = args.weight_decay
         
         # Create task-specific output directory
-        task_output_dir = os.path.join(args.output_dir, task)
+        task_output_dir = os.path.join(pruning_dir, task)
         os.makedirs(task_output_dir, exist_ok=True)
         
         # Log task parameters
@@ -279,6 +292,10 @@ def main():
             "--seed", str(args.seed),
         ]
         
+        # Add param_based flag if needed
+        if args.param_based:
+            cmd.append("--param_based")
+        
         # Run the command
         logger.info(f"Running command: {' '.join(cmd)}")
         try:
@@ -286,9 +303,10 @@ def main():
             logger.info(f"Successfully completed random pruning for task: {task}")
             
             # Find the log file for this task
+            suffix = "param" if args.param_based else "token"
             log_file = os.path.join(
                 task_output_dir, 
-                f"{task}_random_prune{int(task_params['prune_percent'])}.log"
+                f"{task}_random_{suffix}_prune{int(task_params['prune_percent'])}.log"
             )
             # Try alternative filename if not found
[…]

### 2025-04-23 main.py

@@ -221,6 +221,12 @@ def parse_args():
         help="Random seed"
     )
     
+    parser.add_argument(
+        "--param_based",
+        action="store_true", 
+        help="If set, prune based on parameter percentage rather than token percentage (for random pruning method)"
+    )
+    
     return parser.parse_args()
 
 def create_results_dataframe(metrics_callback):
@@ -482,7 +488,8 @@ def run_cross_validation_pipeline(args, tokenizer):
                     args.task, 
                     args.model_name,
                     prune_percent=args.prune_percent,
-                    random_seed=args.seed
+                    random_seed=args.seed,
+                    param_based=args.param_based
                 )
             elif args.pruning_method == "attention":
                 model, token_map, oov_lookup = setup_attention_based_model(
@@ -744,11 +751,14 @@ def run_standard_pipeline(args, tokenizer):
             importance_type=args.importance_type
         )
     elif args.pruning_method == "random":
+        suffix = "param" if args.param_based else "token"
+        log_filename = f"{args.output_dir}/{args.task}_random_{suffix}_prune{args.prune_percent}.log"
         model, token_map, oov_lookup = setup_random_based_model(
             args.task, 
             args.model_name,
             prune_percent=args.prune_percent,
-            random_seed=args.seed
+            random_seed=args.seed,
+            param_based=args.param_based
         )
     elif args.pruning_method == "attention":
         model, token_map, oov_lookup = setup_attention_based_model(
@@ -895,6 +905,8 @@ def run_standard_pipeline(args, tokenizer):
         # Save results to CSV
         if not results_df.empty:
             results_file = f"{args.output_dir}/{args.task}_{args.pruning_method}_prune{args.prune_percent}_results.csv"
+            if args.pruning_method == "random" and args.param_based:
+                results_file = f"{args.output_dir}/{args.task}_{args.pruning_method}_param_prune{args.prune_percent}_results.csv"
             results_df.to_csv(results_file)
             logger.info(f"Saved training results to {results_file}")
         
@@ -918,6 +930,8 @@ def run_standard_pipeline(args, tokenizer):
         logger.info(f"Pruning method: {args.pruning_method}")
         if args.pruning_method != "no_pruning":
             logger.info(f"Prune percent: {args.prune_percent}%")
+            if args.pruning_method == "random" and args.param_based:
+                logger.info(f"Parameter-based pruning enabled")
[…]

### 2025-04-23 pruning/random.py

@@ -17,22 +17,29 @@ from utils import get_task_metadata
 # Configure logging
 logger = logging.getLogger(__name__)
 
-def random_based_pruning(token_counter, prune_percent, min_tokens=5, random_seed=42):
+def random_based_pruning(token_counter, prune_percent, min_tokens=5, random_seed=42, 
+                        param_based=False, embedding_dim=768, total_params=None):
     """
     Prune vocabulary based on random selection.
     This serves as a baseline approach.
     
     Args:
         token_counter: Counter with token frequencies
-        prune_percent: Percentage of tokens to prune
+        prune_percent: Percentage of tokens or parameters to prune
         min_tokens: Minimum number of tokens to keep
         random_seed: Random seed for reproducibility
+        param_based: If True, prune based on parameter percentage rather than token percentage
+        embedding_dim: Dimension of token embeddings (needed for parameter calculation)
+        total_params: Total parameters in the model (needed for parameter-based pruning)
         
     Returns:
         tokens_to_keep: List of token IDs to keep
         tokens_to_remove: List of token IDs to remove
     """
-    logger.info(f"Performing random-based pruning with prune_percent={prune_percent}%")
+    if param_based:
+        logger.info(f"Performing parameter-based random pruning with prune_percent={prune_percent}% of total parameters")
+    else:
+        logger.info(f"Performing token-based random pruning with prune_percent={prune_percent}% of tokens")
     
     # Set random seed for reproducibility
     random.seed(random_seed)
@@ -43,12 +50,27 @@ def random_based_pruning(token_counter, prune_percent, min_tokens=5, random_seed
     # Always keep special tokens (first 5 tokens are usually special tokens in ModernBERT)
     special_tokens = list(range(5))
     
-    # Calculate number of tokens to keep (excluding special tokens)
-    num_prunable_tokens = len(all_token_ids) - len(special_tokens)
-    num_tokens_to_keep = max(min_tokens, int(num_prunable_tokens * (1 - prune_percent / 100)))
-    
     # Get non-special tokens
     non_special_tokens = [token_id for token_id in all_token_ids if token_id not in special_tokens]
+    
+    # Calculate number of tokens to keep based on pruning strategy
+    if param_based and total_params is not None:
+        # Calculate number of embedding parameters to remove
+        emb_params = len(all_token_ids) * embedding_dim
[…]

### 2025-04-23 pruning/random.py
fix token calculation

@@ -44,46 +44,52 @@ def random_based_pruning(token_counter, prune_percent, min_tokens=5, random_seed
     # Set random seed for reproducibility
     random.seed(random_seed)
     
-    # Get all token IDs
-    all_token_ids = list(token_counter.keys())
+    # Get all token IDs from the dataset (these are the train tokens)
+    train_token_ids = list(token_counter.keys())
+    logger.info(f"Dataset contains {len(train_token_ids)} unique tokens")
     
     # Always keep special tokens (first 5 tokens are usually special tokens in ModernBERT)
     special_tokens = list(range(5))
     
-    # Get non-special tokens
-    non_special_tokens = [token_id for token_id in all_token_ids if token_id not in special_tokens]
+    # Get non-special tokens from the train set
+    train_non_special_tokens = [token_id for token_id in train_token_ids if token_id not in special_tokens]
+    logger.info(f"Dataset contains {len(train_non_special_tokens)} non-special tokens")
     
     # Calculate number of tokens to keep based on pruning strategy
     if param_based and total_params is not None:
         # Calculate number of embedding parameters to remove
-        emb_params = len(all_token_ids) * embedding_dim
         params_to_remove = (prune_percent / 100) * total_params
         
         # Convert params to remove to tokens to remove
         tokens_to_remove_count = int(params_to_remove / embedding_dim)
         
-        # Calculate tokens to keep
-        num_tokens_to_keep = max(min_tokens, len(non_special_tokens) - tokens_to_remove_count)
+        # We can only remove tokens from the training set, not more
+        max_removable = len(train_non_special_tokens)
+        tokens_to_remove_count = min(tokens_to_remove_count, max_removable)
+        
+        # Calculate tokens to keep from the training set
+        num_tokens_to_keep = max(min_tokens, len(train_non_special_tokens) - tokens_to_remove_count)
         
         logger.info(f"Target parameter reduction: {params_to_remove:,} parameters")
         logger.info(f"This equals removing {tokens_to_remove_count} tokens at {embedding_dim} dimensions each")
+        logger.info(f"Will keep {num_tokens_to_keep} non-special tokens from the training set")
     else:
         # Traditional token-based pruning
-        num_prunable_tokens = len(non_special_tokens)
+        num_prunable_tokens = len(train_non_special_tokens)
[…]

### 2025-04-23 pruning/random.py
fix random param based pruning

@@ -18,7 +18,8 @@ from utils import get_task_metadata
 logger = logging.getLogger(__name__)
 
 def random_based_pruning(token_counter, prune_percent, min_tokens=5, random_seed=42, 
-                        param_based=False, embedding_dim=768, total_params=None):
+                        param_based=False, embedding_dim=768, total_params=None,
+                        total_vocab_size=None):
     """
     Prune vocabulary based on random selection.
     This serves as a baseline approach.
@@ -31,6 +32,7 @@ def random_based_pruning(token_counter, prune_percent, min_tokens=5, random_seed
         param_based: If True, prune based on parameter percentage rather than token percentage
         embedding_dim: Dimension of token embeddings (needed for parameter calculation)
         total_params: Total parameters in the model (needed for parameter-based pruning)
+        total_vocab_size: Total size of the original vocabulary
         
     Returns:
         tokens_to_keep: List of token IDs to keep
@@ -55,50 +57,93 @@ def random_based_pruning(token_counter, prune_percent, min_tokens=5, random_seed
     train_non_special_tokens = [token_id for token_id in train_token_ids if token_id not in special_tokens]
     logger.info(f"Dataset contains {len(train_non_special_tokens)} non-special tokens")
     
-    # Calculate number of tokens to keep based on pruning strategy
-    if param_based and total_params is not None:
-        # Calculate number of embedding parameters to remove
-        params_to_remove = (prune_percent / 100) * total_params
+    # Calculate parameter reduction achieved by keeping only training tokens
+    if param_based and total_params is not None and total_vocab_size is not None:
+        # Calculate parameter reduction from keeping only train tokens
+        full_emb_params = total_vocab_size * embedding_dim
+        train_emb_params = len(train_token_ids) * embedding_dim
+        param_reduction_from_train_only = full_emb_params - train_emb_params
         
-        # Convert params to remove to tokens to remove
-        tokens_to_remove_count = int(params_to_remove / embedding_dim)
+        # Calculate what percentage of total parameters this represents
+        train_only_reduction_percent = 100 * (param_reduction_from_train_only / total_params)
         
-        # We can only remove tokens from the training set, not more
-        max_removable = len(train_non_special_tokens)
-        tokens_to_remove_count = min(tokens_to_remove_count, max_removable)
[…]

### 2025-04-23 examples/run_attention_pruning.py
param based for other techniques

@@ -41,60 +41,69 @@ BATCH_SIZE = 128
 PRUNE_PERCENT = 20
 
 # Default settings for each task (customize as needed)
-TASK_DEFAULTS = {
+TASK_PARAMS = {
     "cola": {
         "batch_size": BATCH_SIZE, 
         "learning_rate": 8e-5, 
         "epochs": 8,
         "prune_percent": PRUNE_PERCENT,
+        "weight_decay": 1e-6
     },
     "mnli": {
         "batch_size": BATCH_SIZE, 
         "learning_rate": 5e-5, 
         "epochs": 1,
         "prune_percent": PRUNE_PERCENT,
+        "weight_decay": 5e-6
     },
     "mrpc": {
         "batch_size": BATCH_SIZE, 
         "learning_rate": 5e-5, 
         "epochs": 15,
         "prune_percent": PRUNE_PERCENT,
+        "weight_decay": 5e-6
     },
     "qnli": {
         "batch_size": BATCH_SIZE, 
         "learning_rate": 8e-5, 
         "epochs": 2,
         "prune_percent": PRUNE_PERCENT,
+        "weight_decay": 5e-6
     },
     "qqp": {
         "batch_size": BATCH_SIZE, 
         "learning_rate": 5e-5, 
         "epochs": 10,
         "prune_percent": PRUNE_PERCENT,
+        "weight_decay": 5e-6
     },
     "rte": {
         "batch_size": BATCH_SIZE, 
         "learning_rate": 5e-5, 
         "epochs": 10,
         "prune_percent": PRUNE_PERCENT,
+        "weight_decay": 1e-5
     },
     "sst2": {
         "batch_size": BATCH_SIZE, 
         "learning_rate": 8e-5, 
         "epochs": 2,
         "prune_percent": PRUNE_PERCENT,
+        "weight_decay": 1e-5
     },
     "stsb": {
         "batch_size": BATCH_SIZE, 
         "learning_rate": 8e-5, 
         "epochs": 10,
         "prune_percent": PRUNE_PERCENT,
+        "weight_decay": 5e-6
     },
     "wnli": {
         "batch_size": BATCH_SIZE, 
         "learning_rate": 5e-5, 
         "epochs": 3,
         "prune_percent": PRUNE_PERCENT,
+        "weight_decay": 1e-5
     }
 }
 
@@ -148,7 +157,7 @@ def parse_args():
         "--tasks", 
         type=str, 
         nargs="+",
-        default=list(TASK_DEFAULTS.keys()),
[…]

### 2025-04-23 examples/run_clustering_pruning.py

@@ -187,6 +187,12 @@ def parse_args():
         help="Random seed"
     )
     
+    parser.add_argument(
+        "--param_based", 
+        action="store_true",
+        help="If set, prune based on parameter percentage rather than token percentage"
+    )
+    
     return parser.parse_args()
 
 def parse_log_file(log_file):
@@ -238,14 +244,20 @@ def parse_log_file(log_file):
     return metrics
 
 def main():
-    """Run clustering-based pruning on specified GLUE tasks."""
+    """Run clustering pruning on specified GLUE tasks."""
     args = parse_args()
     
     # Create output directory
     os.makedirs(args.output_dir, exist_ok=True)
     
+    # Create a subdirectory based on pruning type
+    pruning_type = "param_based" if args.param_based else "token_based"
+    pruning_dir = os.path.join(args.output_dir, pruning_type)
+    os.makedirs(pruning_dir, exist_ok=True)
+    
     # Log arguments
     logger.info(f"Running with arguments: {args}")
+    logger.info(f"Pruning type: {'Parameter-based' if args.param_based else 'Token-based'}")
     
     # Dictionary to store results for each task
     all_results = {}
@@ -274,12 +286,12 @@ def main():
             task_params["clustering_method"] = args.clustering_method
         
         # Create task-specific output directory
-        task_output_dir = os.path.join(args.output_dir, task)
+        task_output_dir = os.path.join(pruning_dir, task)
         os.makedirs(task_output_dir, exist_ok=True)
         
         # Log task parameters
         logger.info(f"\n{'=' * 50}")
-        logger.info(f"Running clustering-based pruning for task: {task}")
+        logger.info(f"Running clustering pruning for task: {task}")
         logger.info(f"Parameters: {task_params}")
         logger.info(f"{'=' * 50}")
         
@@ -289,26 +301,31 @@ def main():
             "--task", task,
             "--model_name", args.model_name,
             "--pruning_method", "clustering",
+            "--clustering_method", task_params["clustering_method"],
             "--prune_percent", str(task_params["prune_percent"]),
             "--epochs", str(task_params["epochs"]),
             "--learning_rate", str(task_params["learning_rate"]),
             "--batch_size", str(task_params["batch_size"]),
             "--weight_decay", str(task_params["weight_decay"]),
-            "--clustering_method", task_params["clustering_method"],
             "--output_dir", task_output_dir,
             "--seed", str(args.seed),
         ]
+        
+        # Add param_based flag if needed
+        if args.param_based:
[…]

### 2025-04-23 examples/run_frequency_oov_pruning.py

@@ -187,6 +187,12 @@ def parse_args():
         help="Random seed"
     )
     
+    parser.add_argument(
+        "--param_based", 
+        action="store_true",
+        help="If set, prune based on parameter percentage rather than token percentage"
+    )
+    
     return parser.parse_args()
 
 def parse_log_file(log_file):
@@ -244,8 +250,14 @@ def main():
     # Create output directory
     os.makedirs(args.output_dir, exist_ok=True)
     
+    # Create a subdirectory based on pruning type
+    pruning_type = "param_based" if args.param_based else "token_based"
+    pruning_dir = os.path.join(args.output_dir, pruning_type)
+    os.makedirs(pruning_dir, exist_ok=True)
+    
     # Log arguments
     logger.info(f"Running with arguments: {args}")
+    logger.info(f"Pruning type: {'Parameter-based' if args.param_based else 'Token-based'}")
     
     # Dictionary to store results for each task
     all_results = {}
@@ -274,7 +286,7 @@ def main():
             task_params["weight_decay"] = args.weight_decay
         
         # Create task-specific output directory
-        task_output_dir = os.path.join(args.output_dir, task)
+        task_output_dir = os.path.join(pruning_dir, task)
         os.makedirs(task_output_dir, exist_ok=True)
         
         # Log task parameters
@@ -299,6 +311,10 @@ def main():
             "--seed", str(args.seed),
         ]
         
+        # Add param_based flag if needed
+        if args.param_based:
+            cmd.append("--param_based")
+        
         # Run the command
         logger.info(f"Running command: {' '.join(cmd)}")
         try:
@@ -306,9 +322,10 @@ def main():
             logger.info(f"Successfully completed frequency-OOV pruning for task: {task}")
             
             # Find the log file for this task
+            suffix = "param" if args.param_based else "token"
             log_file = os.path.join(
                 task_output_dir, 
-                f"{task}_frequency_oov_prune{int(task_params['prune_percent'])}_clusters{int(task_params['num_clusters'])}.log"
+                f"{task}_frequency_oov_{suffix}_prune{int(task_params['prune_percent'])}_clusters{int(task_params['num_clusters'])}.log"
             )
             # Try alternative filename if not found
             if not os.path.exists(log_file):
@@ -318,7 +335,7 @@ def main():
                 )
             # Try a pattern-based search as a fallback
             if not os.path.exists(log_file):
[…]

### 2025-04-23 examples/run_frequency_pruning.py

@@ -169,6 +169,12 @@ def parse_args():
         help="Random seed"
     )
     
+    parser.add_argument(
+        "--param_based", 
+        action="store_true",
+        help="If set, prune based on parameter percentage rather than token percentage"
+    )
+    
     return parser.parse_args()
 
 def parse_log_file(log_file):
@@ -226,8 +232,14 @@ def main():
     # Create output directory
     os.makedirs(args.output_dir, exist_ok=True)
     
+    # Create a subdirectory based on pruning type
+    pruning_type = "param_based" if args.param_based else "token_based"
+    pruning_dir = os.path.join(args.output_dir, pruning_type)
+    os.makedirs(pruning_dir, exist_ok=True)
+    
     # Log arguments
     logger.info(f"Running with arguments: {args}")
+    logger.info(f"Pruning type: {'Parameter-based' if args.param_based else 'Token-based'}")
     
     # Dictionary to store results for each task
     all_results = {}
@@ -254,7 +266,7 @@ def main():
             task_params["weight_decay"] = args.weight_decay
         
         # Create task-specific output directory
-        task_output_dir = os.path.join(args.output_dir, task)
+        task_output_dir = os.path.join(pruning_dir, task)
         os.makedirs(task_output_dir, exist_ok=True)
         
         # Log task parameters
@@ -278,6 +290,10 @@ def main():
             "--seed", str(args.seed),
         ]
         
+        # Add param_based flag if needed
+        if args.param_based:
+            cmd.append("--param_based")
+        
         # Run the command
         logger.info(f"Running command: {' '.join(cmd)}")
         try:
@@ -285,9 +301,10 @@ def main():
             logger.info(f"Successfully completed frequency-based pruning for task: {task}")
             
             # Find the log file for this task
+            suffix = "param" if args.param_based else "token"
             log_file = os.path.join(
                 task_output_dir, 
-                f"{task}_frequency_prune{int(task_params['prune_percent'])}.log"
+                f"{task}_frequency_{suffix}_prune{int(task_params['prune_percent'])}.log"
             )
             # Try alternative filename if not found
             if not os.path.exists(log_file):
@@ -297,7 +314,7 @@ def main():
                 )
             # Try a pattern-based search as a fallback
             if not os.path.exists(log_file):
-                pattern = f"{task}_frequency_prune*.log"
+                pattern = f"{task}_frequency*prune*.log"
[…]

### 2025-04-23 examples/run_hybrid_pruning.py

@@ -1,338 +1,0 @@
-#!/usr/bin/env python
-# coding: utf-8
-
-"""
-Frequency-based OOV Vocabulary Pruning Script
-
-This script runs frequency-based OOV vocabulary pruning on all GLUE benchmark tasks.
-Each task has customized hyperparameters for optimal performance.
-
-Frequency-OOV pruning combines frequency-based pruning with clustering for OOV tokens.
-"""
-
-import argparse
-import logging
-import subprocess
-import os
-import re
-import pandas as pd
-from datetime import datetime
-import glob
-
-# Configure logging
-logging.basicConfig(
-    level=logging.INFO,
-    format='%(asctime)s - %(levelname)s - %(message)s'
-)
-logger = logging.getLogger(__name__)
-
-# Get the absolute path to the project root directory
-PROJECT_ROOT = os.path.abspath(os.path.join(os.path.dirname(__file__), ".."))
-# Get the absolute path to main.py
-MAIN_SCRIPT_PATH = os.path.join(PROJECT_ROOT, "main.py")
-
-BATCH_SIZE = 128
-PRUNE_PERCENT = 20
-NUM_CLUSTERS = 50
-
-# Dictionary of task-specific parameters
-TASK_PARAMS = {
-    "cola": {
-        "epochs": 8,
-        "learning_rate": 8e-5,
-        "batch_size": BATCH_SIZE,
-        "weight_decay": 1e-6,
-        "prune_percent": PRUNE_PERCENT,
-        "num_clusters": NUM_CLUSTERS,
-    },
-    "mnli": {
-        "epochs": 1,
-        "learning_rate": 5e-5,
-        "batch_size": BATCH_SIZE,
-        "weight_decay": 5e-6,
-        "prune_percent": PRUNE_PERCENT,
-        "num_clusters": NUM_CLUSTERS,
-    },
-    "mrpc": {
-        "epochs": 15,
-        "learning_rate": 5e-5,
-        "batch_size": BATCH_SIZE,
-        "weight_decay": 5e-6,
-        "prune_percent": PRUNE_PERCENT,
-        "num_clusters": NUM_CLUSTERS,
-    },
-    "qnli": {
-        "epochs": 2,
-        "learning_rate": 8e-5,
-        "batch_size": BATCH_SIZE,
-        "weight_decay": 5e-6,
-        "prune_percent": PRUNE_PERCENT,
[…]

### 2025-04-23 examples/run_importance_oov_pruning.py

@@ -214,6 +214,12 @@ def parse_args():
         help="Random seed"
     )
     
+    parser.add_argument(
+        "--param_based", 
+        action="store_true",
+        help="If set, prune based on parameter percentage rather than token percentage"
+    )
+    
     return parser.parse_args()
 
 def parse_log_file(log_file):
@@ -265,14 +271,20 @@ def parse_log_file(log_file):
     return metrics
 
 def main():
-    """Run word importance-based OOV pruning on specified GLUE tasks."""
+    """Run importance-oov-based pruning on specified GLUE tasks."""
     args = parse_args()
     
     # Create output directory
     os.makedirs(args.output_dir, exist_ok=True)
     
+    # Create a subdirectory based on pruning type
+    pruning_type = "param_based" if args.param_based else "token_based"
+    pruning_dir = os.path.join(args.output_dir, pruning_type)
+    os.makedirs(pruning_dir, exist_ok=True)
+    
     # Log arguments
     logger.info(f"Running with arguments: {args}")
+    logger.info(f"Pruning type: {'Parameter-based' if args.param_based else 'Token-based'}")
     
     # Dictionary to store results for each task
     all_results = {}
@@ -303,12 +315,12 @@ def main():
             task_params["importance_type"] = args.importance_type
         
         # Create task-specific output directory
-        task_output_dir = os.path.join(args.output_dir, task)
+        task_output_dir = os.path.join(pruning_dir, task)
         os.makedirs(task_output_dir, exist_ok=True)
         
         # Log task parameters
         logger.info(f"\n{'=' * 50}")
-        logger.info(f"Running word importance-based OOV pruning for task: {task}")
+        logger.info(f"Running importance-oov pruning for task: {task}")
         logger.info(f"Parameters: {task_params}")
         logger.info(f"{'=' * 50}")
         
@@ -318,27 +330,32 @@ def main():
             "--task", task,
             "--model_name", args.model_name,
             "--pruning_method", "importance_oov",
+            "--importance_type", str(task_params["importance_type"]),
+            "--num_clusters", str(task_params["num_clusters"]),
             "--prune_percent", str(task_params["prune_percent"]),
             "--epochs", str(task_params["epochs"]),
             "--learning_rate", str(task_params["learning_rate"]),
             "--batch_size", str(task_params["batch_size"]),
             "--weight_decay", str(task_params["weight_decay"]),
-            "--num_clusters", str(task_params["num_clusters"]),
-            "--importance_type", str(task_params["importance_type"]),
[…]

### 2025-04-23 examples/run_importance_pruning.py

@@ -49,12 +49,13 @@ PRUNE_PERCENT = 20
 IMPORTANCE_TYPE = 3
 
 # Default settings for each task (customize as needed)
-TASK_DEFAULTS = {
+TASK_PARAMS = {
     "cola": {
         "batch_size": BATCH_SIZE, 
         "learning_rate": 8e-5, 
         "epochs": 8,
         "prune_percent": PRUNE_PERCENT,
+        "weight_decay": 1e-6,
         "importance_type": IMPORTANCE_TYPE
     },
     "mnli": {
@@ -62,6 +63,7 @@ TASK_DEFAULTS = {
         "learning_rate": 5e-5, 
         "epochs": 1,
         "prune_percent": PRUNE_PERCENT,
+        "weight_decay": 5e-6,
         "importance_type": IMPORTANCE_TYPE
     },
     "mrpc": {
@@ -69,6 +71,7 @@ TASK_DEFAULTS = {
         "learning_rate": 5e-5, 
         "epochs": 15,
         "prune_percent": PRUNE_PERCENT,
+        "weight_decay": 5e-6,
         "importance_type": IMPORTANCE_TYPE
     },
     "qnli": {
@@ -76,6 +79,7 @@ TASK_DEFAULTS = {
         "learning_rate": 8e-5, 
         "epochs": 2,
         "prune_percent": PRUNE_PERCENT,
+        "weight_decay": 5e-6,
         "importance_type": IMPORTANCE_TYPE
     },
     "qqp": {
@@ -83,6 +87,7 @@ TASK_DEFAULTS = {
         "learning_rate": 5e-5, 
         "epochs": 10,
         "prune_percent": PRUNE_PERCENT,
+        "weight_decay": 5e-6,
         "importance_type": IMPORTANCE_TYPE
     },
     "rte": {
@@ -90,6 +95,7 @@ TASK_DEFAULTS = {
         "learning_rate": 5e-5, 
         "epochs": 10,
         "prune_percent": PRUNE_PERCENT,
+        "weight_decay": 1e-5,
         "importance_type": IMPORTANCE_TYPE
     },
     "sst2": {
@@ -97,6 +103,7 @@ TASK_DEFAULTS = {
         "learning_rate": 8e-5, 
         "epochs": 2,
         "prune_percent": PRUNE_PERCENT,
+        "weight_decay": 1e-5,
         "importance_type": IMPORTANCE_TYPE
     },
     "stsb": {
[…]

### 2025-04-23 main.py

@@ -459,14 +459,16 @@ def run_cross_validation_pipeline(args, tokenizer):
                 model, token_map, oov_lookup = setup_frequency_based_model(
                     args.task, 
                     args.model_name,
-                    prune_percent=args.prune_percent
+                    prune_percent=args.prune_percent,
+                    param_based=args.param_based
                 )
             elif args.pruning_method == "frequency_oov":
                 model, token_map, oov_lookup = setup_frequency_oov_model(
                     args.task, 
                     args.model_name,
                     prune_percent=args.prune_percent,
-                    num_clusters=args.num_clusters
+                    num_clusters=args.num_clusters,
+                    param_based=args.param_based
                 )
             elif args.pruning_method == "importance_oov":
                 model, token_map, oov_lookup = setup_importance_oov_model(
@@ -474,14 +476,16 @@ def run_cross_validation_pipeline(args, tokenizer):
                     args.model_name,
                     prune_percent=args.prune_percent,
                     num_clusters=args.num_clusters,
-                    importance_type=args.importance_type
+                    importance_type=args.importance_type,
+                    param_based=args.param_based
                 )
             elif args.pruning_method == "importance":
                 model, token_map, oov_lookup = setup_importance_based_model(
                     args.task, 
                     args.model_name,
                     prune_percent=args.prune_percent,
-                    importance_type=args.importance_type
+                    importance_type=args.importance_type,
+                    param_based=args.param_based
                 )
             elif args.pruning_method == "random":
                 model, token_map, oov_lookup = setup_random_based_model(
@@ -496,7 +500,8 @@ def run_cross_validation_pipeline(args, tokenizer):
                     args.task, 
                     args.model_name,
                     attention_model=args.attention_model,
-                    prune_percent=args.prune_percent
+                    prune_percent=args.prune_percent,
+                    param_based=args.param_based
                 )
             elif args.pruning_method == "train_only":
                 model, token_map, oov_lookup = setup_train_only_model(
@@ -726,14 +731,16 @@ def run_standard_pipeline(args, tokenizer):
         model, token_map, oov_lookup = setup_frequency_based_model(
             args.task, 
             args.model_name,
-            prune_percent=args.prune_percent
[…]

### 2025-04-23 pruning/attention.py

@@ -159,100 +159,214 @@ def get_token_attention_importance(task_name, model_name):
     
     return normalized_importance, all_token_ids
 
-def attention_based_pruning(token_importance, all_token_ids, prune_percent, min_tokens=5):
+def attention_based_pruning(token_importance, all_token_ids, prune_percent, min_tokens=5,
+                         param_based=False, embedding_dim=768, total_params=None,
+                         total_vocab_size=None):
     """
     Prune vocabulary based on token attention importance scores.
     
     Args:
         token_importance: Dict with token_id -> importance score mappings
         all_token_ids: Set of all token IDs seen in the dataset
-        prune_percent: Percentage of tokens to prune
+        prune_percent: Percentage of tokens or parameters to prune
         min_tokens: Minimum number of tokens to keep
+        param_based: If True, prune based on parameter percentage rather than token percentage
+        embedding_dim: Dimension of token embeddings (needed for parameter calculation)
+        total_params: Total parameters in the model (needed for parameter-based pruning)
+        total_vocab_size: Total size of the original vocabulary
         
     Returns:
         tokens_to_keep: List of token IDs to keep
         tokens_to_remove: List of token IDs to remove
     """
-    logger.info(f"Performing attention-based pruning with prune_percent={prune_percent}%")
+    if param_based:
+        logger.info(f"Performing parameter-based attention pruning with prune_percent={prune_percent}% of total parameters")
+    else:
+        logger.info(f"Performing token-based attention pruning with prune_percent={prune_percent}%")
     
-    # Sort tokens by importance (highest first)
+    # Sort tokens by importance (highest importance first)
     sorted_tokens = sorted(
         [(token_id, token_importance[token_id]) for token_id in all_token_ids],
         key=lambda x: x[1],
         reverse=True
     )
     
+    train_token_ids = [token_id for token_id, _ in sorted_tokens]
+    logger.info(f"Dataset contains {len(train_token_ids)} unique tokens")
+    
     # Always keep special tokens (first 5 tokens are usually special tokens in ModernBERT)
     special_tokens = list(range(5))
     
-    # Calculate number of tokens to keep (excluding special tokens)
-    num_prunable_tokens = len(sorted_tokens) - len(special_tokens)
-    num_tokens_to_keep = max(min_tokens, int(num_prunable_tokens * (1 - prune_percent / 100)))
+    # Calculate parameter reduction achieved by keeping only training tokens
[…]

### 2025-04-23 pruning/base.py

@@ -168,32 +168,27 @@ def create_reduced_embeddings(task_vocab, model):
     return token_map, reduced_embeddings
 
 # Create hybrid embeddings (used by hybrid and importance pruning)
-def create_hybrid_embeddings(tokens_to_keep, oov_token_map, cluster_centers, model):
+def create_hybrid_embeddings(tokens_to_keep, oov_lookup, model):
     """
     Create reduced embedding matrix with hybrid token mapping.
     Used by hybrid and importance-based pruning.
     
     Args:
         tokens_to_keep: List of token IDs to keep
-        oov_token_map: Mapping from removed token ID to cluster center token ID
-        cluster_centers: List of token IDs for cluster centers
+        oov_lookup: Mapping from removed token ID to cluster center token ID
         model: Model whose embeddings will be used
         
     Returns:
         token_map: Mapping from original token IDs to new consecutive IDs
         reduced_embeddings: Reduced embedding matrix
-        oov_lookup: Lookup table for mapping OOV tokens
     """
-    logger.info(f"Creating hybrid embeddings with {len(tokens_to_keep)} kept tokens and {len(cluster_centers)} OOV clusters")
+    logger.info(f"Creating hybrid embeddings with {len(tokens_to_keep)} kept tokens and {len(oov_lookup)} OOV tokens")
     
     # Get original embedding matrix
     original_embeddings = model.model.embeddings.tok_embeddings.weight.data
     
-    # Combine kept tokens and cluster centers (avoiding duplicates)
+    # Start with kept tokens
     all_tokens_to_keep = list(tokens_to_keep)
-    for center in cluster_centers:
-        if center not in all_tokens_to_keep:
-            all_tokens_to_keep.append(center)
     
     # Create mapping from original token IDs to new consecutive IDs
     token_map = {old_id: new_id for new_id, old_id in enumerate(all_tokens_to_keep)}
@@ -201,10 +196,4 @@ def create_hybrid_embeddings(tokens_to_keep, oov_token_map, cluster_centers, mod
     # Create reduced embedding matrix with only the needed vectors
     reduced_embeddings = torch.stack([original_embeddings[i] for i in all_tokens_to_keep])
     
-    # Create OOV lookup that maps original token IDs to their cluster representatives
-    oov_lookup = {}
-    for orig_id, center_id in oov_token_map.items():
-        if center_id in token_map:  # Ensure the center ID is in our vocabulary
[…]

### 2025-04-23 pruning/clustering.py

@@ -17,7 +17,8 @@ from utils import get_task_metadata
 # Configure logging
 logger = logging.getLogger(__name__)
 
-def cluster_embeddings(token_ids, model, prune_percent, clustering_method="agglomerative"):
+def cluster_embeddings(token_ids, model, prune_percent, clustering_method="agglomerative", 
+                       param_based=False, total_params=None, total_vocab_size=None, embedding_dim=768):
     """
     Cluster token embeddings and prune vocabulary based on similarity.
     Used by the clustering-based pruning method.
@@ -27,11 +28,18 @@ def cluster_embeddings(token_ids, model, prune_percent, clustering_method="agglo
         model: Model whose embeddings will be used
         prune_percent: Percentage of vocabulary to prune through clustering
         clustering_method: Method to use for clustering (agglomerative or kmeans)
+        param_based: If True, prune based on parameter percentage rather than token percentage
+        total_params: Total parameters in the model (needed for parameter-based pruning)
+        total_vocab_size: Total size of the original vocabulary
+        embedding_dim: Dimension of token embeddings (needed for parameter calculation)
         
     Returns:
         List of token IDs to keep after pruning
     """
-    logger.info(f"Clustering embeddings for {len(token_ids)} tokens with {prune_percent}% pruning using {clustering_method}")
+    if param_based:
+        logger.info(f"Clustering embeddings for parameter-based pruning ({prune_percent}% of total params) using {clustering_method}")
+    else:
+        logger.info(f"Clustering embeddings for {len(token_ids)} tokens with {prune_percent}% pruning using {clustering_method}")
     
     # Get original embedding matrix
     original_embeddings = model.model.embeddings.tok_embeddings.weight.data
@@ -46,9 +54,45 @@ def cluster_embeddings(token_ids, model, prune_percent, clustering_method="agglo
     prunable_tokens = token_ids[5:]
     prunable_embeddings = original_embeddings[prunable_tokens].cpu().numpy()
     
-    # Calculate number of clusters
-    n_clusters = len(prunable_tokens) - int(len(token_ids) * prune_percent / 100)
-    logger.info(f"Clustering into {n_clusters} clusters (reducing from {len(prunable_tokens)} tokens)")
+    # Calculate parameter reduction achieved by keeping only training tokens
+    if param_based and total_params is not None and total_vocab_size is not None:
+        # Calculate parameter reduction from keeping only train tokens
+        full_emb_params = total_vocab_size * embedding_dim
[…]

### 2025-04-23 pruning/frequency.py

@@ -14,50 +14,138 @@ from utils import get_task_metadata
 # Configure logging
 logger = logging.getLogger(__name__)
 
-def frequency_based_pruning(token_counter, prune_percent, min_tokens=5):
+def frequency_based_pruning(token_counter, prune_percent, min_tokens=5, 
+                          param_based=False, embedding_dim=768, total_params=None,
+                          total_vocab_size=None):
     """
     Prune vocabulary based on token frequency.
     Used by frequency-based and hybrid pruning methods.
     
     Args:
         token_counter: Counter with token frequencies
-        prune_percent: Percentage of tokens to prune
+        prune_percent: Percentage of tokens or parameters to prune
         min_tokens: Minimum number of tokens to keep
+        param_based: If True, prune based on parameter percentage rather than token percentage
+        embedding_dim: Dimension of token embeddings (needed for parameter calculation)
+        total_params: Total parameters in the model (needed for parameter-based pruning)
+        total_vocab_size: Total size of the original vocabulary
         
     Returns:
         tokens_to_keep: List of token IDs to keep
         tokens_to_remove: List of token IDs to remove
     """
-    logger.info(f"Performing frequency-based pruning with prune_percent={prune_percent}%")
+    if param_based:
+        logger.info(f"Performing parameter-based frequency pruning with prune_percent={prune_percent}% of total parameters")
+    else:
+        logger.info(f"Performing token-based frequency pruning with prune_percent={prune_percent}% of tokens")
     
     # Sort tokens by frequency (most common first)
     sorted_tokens = token_counter.most_common()
+    train_token_ids = [token_id for token_id, _ in sorted_tokens]
+    logger.info(f"Dataset contains {len(train_token_ids)} unique tokens")
     
     # Always keep special tokens (first 5 tokens are usually special tokens in ModernBERT)
     special_tokens = list(range(5))
     
-    # Calculate number of tokens to keep (excluding special tokens)
-    num_prunable_tokens = len(sorted_tokens) - len(special_tokens)
-    num_tokens_to_keep = max(min_tokens, int(num_prunable_tokens * (1 - prune_percent / 100)))
+    # Calculate parameter reduction achieved by keeping only training tokens
+    if param_based and total_params is not None and total_vocab_size is not None:
+        # Calculate parameter reduction from keeping only train tokens
+        full_emb_params = total_vocab_size * embedding_dim
[…]

### 2025-04-23 pruning/frequency_oov.py

@@ -18,20 +18,24 @@ from utils import get_task_metadata
 # Configure logging
 logger = logging.getLogger(__name__)
 
-def cluster_removed_tokens(tokens_to_remove, model, num_clusters=50):
+def cluster_removed_tokens(tokens_to_remove, tokens_to_keep, model, num_clusters=50):
     """
     Cluster removed tokens to create mapping for OOV tokens.
     Used by frequency_oov pruning methods.
     
     Args:
         tokens_to_remove: List of token IDs that were removed
+        tokens_to_keep: List of token IDs that are kept in the vocabulary
         model: Model whose embeddings will be used
         num_clusters: Number of clusters to create
         
     Returns:
-        oov_token_map: Mapping from removed token ID to nearest cluster center token ID
-        cluster_centers: List of token IDs representing cluster centers
+        oov_lookup: Mapping from removed token ID to nearest cluster center token ID
     """
+    if not tokens_to_remove:
+        logger.info("No tokens to cluster (all tokens kept)")
+        return {}
+        
     logger.info(f"Clustering {len(tokens_to_remove)} removed tokens into {num_clusters} clusters")
     
     # Get original embedding matrix
@@ -48,7 +52,7 @@ def cluster_removed_tokens(tokens_to_remove, model, num_clusters=50):
     cluster_labels = kmeans.fit_predict(removed_embeddings)
     
     # Initialize mapping from removed token to nearest cluster center token
-    oov_token_map = {}
+    oov_lookup = {}
     cluster_centers = []
     
     # For each cluster, find the token closest to the centroid
@@ -69,7 +73,7 @@ def cluster_removed_tokens(tokens_to_remove, model, num_clusters=50):
             
             # Map all tokens in this cluster to the center token
             for idx in cluster_indices:
-                oov_token_map[tokens_to_remove[idx]] = center_token_id
+                oov_lookup[tokens_to_remove[idx]] = center_token_id
             
             cluster_centers.append(center_token_id)
     
@@ -90,24 +94,26 @@ def cluster_removed_tokens(tokens_to_remove, model, num_clusters=50):
     
     logger.info(f"Created {len(cluster_centers)} cluster representatives for OOV token mapping")
     
-    return oov_token_map, cluster_centers
+    return oov_lookup
 
-def setup_frequency_oov_model(task_name, model_name, prune_percent=20, num_clusters=50):
[…]

### 2025-04-23 pruning/importance_oov.py

@@ -192,7 +192,9 @@ def get_dataset_tokens_with_importance(task_name, train_only=True, importance_ty
     
     return token_counter, token_importance, all_token_ids
 
-def importance_based_pruning(token_counter, token_importance, prune_percent, min_tokens=5):
+def importance_based_pruning(token_counter, token_importance, prune_percent, min_tokens=5,
+                          param_based=False, embedding_dim=768, total_params=None,
+                          total_vocab_size=None):
     """
     Prune vocabulary based on token importance scores.
     Used by the word importance based pruning method.
@@ -200,63 +202,152 @@ def importance_based_pruning(token_counter, token_importance, prune_percent, min
     Args:
         token_counter: Counter with token frequencies
         token_importance: Dict with token_id -> importance score mappings
-        prune_percent: Percentage of tokens to prune
+        prune_percent: Percentage of tokens or parameters to prune
         min_tokens: Minimum number of tokens to keep
+        param_based: If True, prune based on parameter percentage rather than token percentage
+        embedding_dim: Dimension of token embeddings (needed for parameter calculation)
+        total_params: Total parameters in the model (needed for parameter-based pruning)
+        total_vocab_size: Total size of the original vocabulary
         
     Returns:
         tokens_to_keep: List of token IDs to keep
         tokens_to_remove: List of token IDs to remove
     """
-    logger.info(f"Performing importance-based pruning with prune_percent={prune_percent}%")
+    if param_based:
+        logger.info(f"Performing parameter-based importance pruning with prune_percent={prune_percent}% of total parameters")
+    else:
+        logger.info(f"Performing token-based importance pruning with prune_percent={prune_percent}%")
     
-    # Sort tokens by importance (highest first)
+    # Sort tokens by importance (highest importance first)
     sorted_tokens = sorted(
-        [(token_id, token_importance[token_id]) for token_id in token_importance],
-        key=lambda x: x[1],
+        token_importance.items(), 
+        key=lambda x: x[1], 
         reverse=True
     )
+    
+    train_token_ids = [token_id for token_id, _ in sorted_tokens]
+    logger.info(f"Dataset contains {len(train_token_ids)} unique tokens")
     
     # Always keep special tokens (first 5 tokens are usually special tokens in ModernBERT)
     special_tokens = list(range(5))
     
[…]

### 2025-05-06 paper/main.tex
add introduction, update related works, update algorithm, add template for references

@@
 \usepackage{float}
+\usepackage{cite}
 
@@
 \maketitle
-
 % Motivation / Context (1–2 sentences):
@@
 \section{Introduction}
-% Look at the introduction of ModernBERT paper, interesting!
-1. Introduction of the ModernBERT paper. \\
-2. Architecture of ModernBERT.
+
+Large pre-trained transformer models such as BERT have ushered in a new era of performance across virtually every core NLP task---from sentiment analysis and natural language inference to question answering and summarization.
+Their success stems in large part from massive embedding layers and deep stacks of self-attention and feed-forward blocks, which together learn extraordinarily rich, contextualized token representations.
+Yet this very scale makes them difficult to deploy: memory footprints measured in multiple gigabytes, inference latencies unsuited to on-device or real-time use, and the sheer cost of fine-tuning or retraining pruned variants can all stand in the way of practical adoption.
+
+Pruning---selectively removing parameters deemed non-essential---has emerged as one of the most promising paths to slim down these models.
+To date, however, the majority of work has focused on sparsifying or structurally pruning the encoder layers: zeroing out small weights in feed-forward networks, dropping entire attention heads, or even omitting full transformer blocks during inference.
+While these methods can yield substantial compression, they often require additional fine-tuning to recover lost accuracy, and they overlook a particularly ``heavy'' component of transformer architectures: the embedding layer.
+
+In BERT-style models, the token embedding matrix alone can account for 25--50\% of all parameters, yet vocabulary pruning remains comparatively underexplored.
+Simple heuristics---such as dropping all tokens not observed in a downstream training corpus---can achieve size reductions, but they fail to capture semantic importance, ignore relationships between pruned and retained tokens, and typically require a retraining step to adapt the model to its slimmed vocabulary.
+As a result, embedding-level pruning often remains a ``bolt-on'' afterthought rather than an integral, efficient component of model compression.
+
+This paper presents Words That Matter, a fine-tuning-free approach to vocabulary pruning in ModernBERT~\cite{modernbert2023} that directly addresses these gaps.
+Our method proceeds in two complementary stages.
+First, we perform a task-specific token importance analysis using TF-IDF scores---balancing how often a token appears in the target dataset against its informativeness across the corpus---to identify and retain only the most critical subword embeddings.
+Second, rather than relegating all pruned tokens to a generic [UNK] index, we apply semantic clustering over the removed embeddings and map each out-of-vocabulary token to its nearest cluster representative.
+This preserves nuanced semantic relationships without expanding the model's vocabulary size.
+
+Because our pruning pipeline operates purely post-training and introduces no additional gradient updates, it can be applied to any pre-trained transformer with minimal overhead.
+Moreover, it dovetails seamlessly with existing encoder-layer pruning techniques---such as LoSparse's~\cite{losparse2023} combined low-rank plus sparse factorization---enabling an end-to-end compression strategy that tackles both embeddings and intermediate representations.
+
+We validate our approach on the GLUE benchmark suite~\cite{wang2018glue}, showing that we can shrink ModernBERT's embedding matrix by up to 20\% while incurring less than 1\% average degradation in downstream accuracy.
+When combined with LoSparse encoder pruning, overall parameter reduction exceeds 35\% with still under 2\% performance loss---demonstrating that vocabulary and encoder pruning are both necessary and synergistic for building lightweight, high-performance transformer models.
+
+In summary, our contributions are:
+
+\begin{itemize}
+    \item A novel, fine-tuning-free vocabulary pruning method that targets the otherwise-neglected embedding layer in transformer models.
+
+    \item A hybrid TF-IDF plus semantic clustering strategy for robust handling of pruned tokens, preserving semantic structure without [UNK] collapse.
+
+    \item Empirical evidence on GLUE that embedding pruning alone can yield $\sim$20\% size reduction with $<$1\% performance loss, and that combined embedding + encoder pruning achieves $>$35\% compression for under 2\% drop in accuracy.
+
+    \item A modular pipeline compatible with any pre-trained encoder model and complementary to structured encoder-layer pruning methods.
+\end{itemize}
+
+By placing vocabulary pruning on equal footing with encoder-layer sparsification, Words That Matter opens a practical pathway to more compact, efficient, and deployable transformer models---bringing state-of-the-art NLP within reach of resource-constrained environments.
 
@@
 
-\subsection{Encoder Layer Pruning}
+\subsection{Vocabulary Pruning Techniques}
 
-\subsubsection{LoSparse}
-Low-Rank and Sparse approximation (LoSparse) introduces a model compression strategy that decomposes each weight matrix W as the sum of a low-rank component (UV) and a sparse component (S).
-The low-rank part, computed via singular value decomposition, extracts the shared, coherent features among neurons, while the sparse part targets and removes redundant, incoherent features.
-This dual approach overcomes the limitations of traditional methods by avoiding the excessive pruning of expressive neurons and preserving diversity in the representation.
-Extensive experiments on natural language understanding, question answering, and generation tasks show that LoSparse achieves significant parameter reduction all while maintaining competitive performance levels.
+Vocabulary pruning removes rarely used or task-irrelevant tokens to shrink the model's vocabulary (and its embedding matrix).
+Early work by Abdaoui et al.~\cite{abdaoui2020load} showed that most parameters of multilingual BERT lie in the embedding layer, so trimming unused language tokens can dramatically cut size.
+In ``Load What You Need,'' they drop languages from mBERT's vocabulary and obtain up to a 45\% reduction in total parameters with comparable accuracy on XNLI.
+\\
+\\
+More recently, Nair et al.~\cite{nair2023blade} proposed BLADE: they prune a bilingual mBERT vocabulary to include only tokens from the query and document languages for cross-lingual IR, then apply a small intermediate pretraining step.
+This pruned model (keeping only needed embeddings) reduced parameters by $\sim$36.5\% versus full mBERT and sped up training by $\sim$30\% and inference by $\sim$55\%, with minimal loss in retrieval quality.
+\\
+\\
+Dorkin et al.~\cite{dorkin2025estonian} also evaluated pruning for an Estonian adaptation: pruning all tokens unused by Estonian led to no degradation on NER, whereas retraining a new tokenizer hurt performance.
+These studies consistently find that deleting infrequently seen subwords yields large size savings with little accuracy drop.
 
-\subsubsection{Other encoder layer pruning techniques}
+\begin{itemize}
+    \item Pruning toolkits: TextPruner~\cite{shen2022textpruner} automates vocabulary pruning: it scans a corpus and removes any tokenizer token not present in the text, deleting its row in the embedding matrix.
+This simple ``drop unused tokens'' mode can cut model size and speed up training/inference on that domain.
+    \item Sparse expansion models: In sparse-retrieval models like SPLADE, vocabulary pruning is implicit.
+SPLADE-X~\cite{formal2023spladex} (multi-lingual) restricted output to query-language tokens only, effectively pruning all other vocab.
+BLADE extends this idea by explicitly building a pruned bilingual model with only the union of query and document subwords, discarding other embeddings entirely.
+    \item Empirical results: Across tasks and languages, pruned-vocabulary models often match full models.
+For example, after pruning, Nair et al.~\cite{nair2023blade} report retrieval effectiveness on par with larger baselines, while achieving much faster indexing.
+Dorkin et al.~\cite{dorkin2025estonian} explicitly note ``vocabulary pruning has no observable negative effect on the downstream task'' (Estonian NER).
+These findings suggest vocabulary pruning is an effective way to tailor large models to specific domains or languages.
+\end{itemize}
 
-\subsection{Vocabulary Pruning}
-While encoder layer pruning techniques like LoSparse have shown promising results, the embedding layer presents a unique challenge in transformer models.
-Accounting for approximately 25\% of the total parameters, the vocabulary embedding layer offers a substantial opportunity for model compression.
-However, unlike encoder layer pruning, vocabulary reduction requires careful consideration of semantic relationships and token importance to maintain model performance.
+\subsection{Embedding-Layer Pruning and Compression}
 
-\subsubsection{Direction is what you need}
-% Direction is what you need: Improving Word Embedding Compression in Large Language Models
-\subsubsection{TF-IDF}
-% Enhancing performance of transformer-based models in natural language understanding through word importance embedding
+Pruning the embedding layer often means removing or compressing rows (token vectors) to shrink this heavy component.
+Partial Embedding Matrix Adaptation (PEMA)~\cite{bousquet2023pema} is one systematic approach: they first note that fine-tuning datasets typically use only a small fraction of the full vocabulary.
+PEMA therefore builds a partial embedding matrix containing only tokens seen in the task data, training on that smaller matrix.
+This saves GPU memory during fine-tuning without altering task accuracy (unused embeddings are never updated).
 
-\subsubsection{PEMA}
-% Vocabulary-level Memory Efficiency for Language Model Fine-tuning
-Partial Embedding Matrix Approximation (PEMA) makes finetuning efficient by removing the tokens in the embedding layers that are not used in the current task.
+After fine-tuning, the updated embeddings are merged back into the full matrix so the model structure remains unchanged.
+Experiments show large memory reductions: e.g., BERT$_{\text{BASE}}$'s embedding matrix can shrink by $\sim$47\% on average (and RoBERTa by $\sim$52\%) when only task-relevant tokens are kept.
+Multilingual models see even more savings (mBERT $\sim$44\% embedding reduction, XLM-R $\sim$64\%) because their vocabularies are huge.
+Crucially, PEMA reports no drop in downstream performance, making it a practical way to prune embeddings during training.
+\\
+\\
+Another approach is embedding compression via factorization or hashing.
+Wang et al.~\cite{wang2023lighttoken} propose LightToken, which compresses the entire embedding matrix using low-rank SVD plus a binary residual autoencoder.
+They note token embeddings are highly redundant (the embedding takes $\sim$21--31\% of BERT/RoBERTa's size).
+LightToken first applies a rank-$k$ SVD to capture most variation, then learns compact hash codes to reconstruct the residual.
+In experiments on GLUE, LightToken achieves very high compression (e.g., 25$\times$ smaller embeddings) while preserving accuracy.
+For BERT$_{\text{BASE}}$, a 25$\times$ embedding compression yields an average GLUE score of 82.9 (baseline 83.0).
 
-\subsubsection{LightToken}
-LightToken is a lightweight token embedding framework that produces compressed token embeddings in a task and model-agnostic fashion.
-It's designed to be compatible with different architectures and applicable to any downstream task.
-The framework integrates low-rank approximation, a novel residual binary autoencoder, and a new compression loss function to significantly improve model compression ratio without sacrificing performance.
+Thus, LightToken and similar methods show that even structured low-rank or quantized pruning of embeddings can yield massive space savings with minimal impact on performance. 
+\\
+\\
+Overall, embedding-layer pruning techniques exploit the fact that many token vectors (especially for rare subwords) contribute little to end-task accuracy.
+By either dropping unused rows (PEMA/TextPruner) or factoring and quantizing embeddings (LightToken), these methods reduce the $\sim$25--50\% of model parameters in the embedding matrix, enabling lighter models in practice.
 
-\newpage
-TODO
+\subsection{Encoder-Layer Pruning}
 
-\newpage
+Although our focus is vocab/embedding pruning, many complementary pruning methods work at the encoder layers.
+A classic example is attention-head pruning: Michel et al.~\cite{michel2019heads} discovered that ``a large percentage of attention heads can be removed at test time without significantly impacting performance''.
+Voita et al.~\cite{voita2019heads} likewise pruned 38 of 48 encoder heads in an NMT transformer (removing $\sim$79\% of heads) with only a 0.15 BLEU loss.
+
+In practice, toolkit frameworks (like TextPruner) often support head and FFN pruning alongside vocab pruning.
+Other methods include movement pruning~\cite{sanh2020movement} that gradually zeros out small weights during fine-tuning, or LayerDrop~\cite{fan2020layerdrop} which stochastically drops whole layers during training.
+These encoder-level techniques are orthogonal to vocab pruning: they further thin out the model by removing redundant heads, neurons, or weight components, whereas vocabulary pruning targets the input/output embedding parameters.
+
 \section{Method}
@@
 \scriptsize
-\Require Model $M$ with vocab $V$, dataset $D$, pruning \% $p$, clusters $k$, method $method$
-\Ensure Pruned model $M'$ with reduced vocabulary
+\Require Model $M$ with vocabulary $V$, dataset $D$, pruning percentage $p$, number of clusters $k$
+\Ensure Pruned model $M'$ with reduced vocabulary and OOV mapping
 
-\State // Stage 1: Calculate token importance using selected method (random, frequency, attention, or TF-IDF)
-\State $importance \gets \text{CalculateTokenImportance}(V, D, M, method)$
+\State // 1. Token Importance Analysis
+\State Calculate importance scores for each token in $V$ using TF-IDF on dataset $D$
+\State Sort tokens by their importance scores
 
-\State // Stage 2: Prune lowest-importance tokens based on pruning percentage
-\State $tokens\_sorted \gets \text{SortByImportance}(V, importance)$
-\State $keep\_count \gets |V| \times (1 - p/100)$
-\State $tokens\_to\_keep \gets tokens\_sorted[1:keep\_count]$
-\State $tokens\_to\_remove \gets V \setminus tokens\_to\_keep$
+\State // 2. Pruning Method
+\State Keep top $(100-p)\%$ of tokens with highest importance scores
+\State Identify tokens to remove: $V_{remove} = V \setminus V_{keep}$
 
-\State // Stage 3: Cluster removed tokens and identify representatives for OOV mapping
-\State $embeddings \gets M.embedding\_layer.weight$
-\State $removed\_embeddings \gets embeddings[tokens\_to\_remove]$
-\State $kmeans \gets \text{KMeans}(n\_clusters=k)$
-\State $clusters \gets kmeans.fit\_predict(removed\_embeddings)$
-\State $oov\_map \gets \{\}$; $centers \gets \{\}$
+\State // 3. OOV Token Handling
+\State Extract embeddings for tokens to be removed: $E_{remove} = M.embeddings[V_{remove}]$
+\State Cluster $E_{remove}$ into $k$ clusters using K-means algorithm
+\State For each cluster, select token closest to centroid as representative
+\State Create OOV mapping: each removed token maps to its cluster representative
 
-\For{each cluster $c$ in $range(k)$}
-    \State $indices \gets \{i : clusters[i] = c\}$
-    \If{$indices$ is not empty}
-        \State $centroid \gets kmeans.cluster\_centers_[c]$
-        \State $distances \gets \{\|removed\_embeddings[i] - centroid\| : i \in indices\}$
-        \State $closest\_idx \gets \arg\min(distances)$ 
-        \State $center\_token \gets tokens\_to\_remove[indices[closest\_idx]]$
-        \State $centers.add(center\_token)$
-        \For{$idx \in indices$} 
-            \State $oov\_map[tokens\_to\_remove[idx]] \gets center\_token$ 
-        \EndFor
-    \EndIf
-\EndFor
+\State // 4. Model Modification
+\State Construct new vocabulary: $V' = V_{keep} \cup V_{representatives}$
+\State Create reduced embedding matrix using embeddings of $V'$
+\State Update model $M'$ with reduced embedding layer
+\State Modify tokenization process to use OOV mapping for unseen tokens
 
-\State // Stage 4: Create reduced embedding matrix with kept tokens and cluster representatives
-\State $all\_tokens \gets tokens\_to\_keep \cup centers$
-\State $token\_map \gets \{old: new \mid new, old \in enumerate(all\_tokens)\}$
-\State $reduced\_embeddings \gets embeddings[all\_tokens]$
-\State $M'.embedding\_layer.weight \gets reduced\_embeddings$
-
-\State \Return $M'$, $token\_map$, $oov\_map$
+\State \Return $M'$ with pruned vocabulary and OOV mapping
 \end{algorithmic}
@@
 TF-IDF + OOV & 0.923 & 0.834 & 0.878 & 0.615 & 0.903 & 0.906 & 0.919 & 0.554 & \textbf{0.817} \\
+\\ [-8pt]
+\cdashline{1-10}
+\\[-8pt]
+\multicolumn{9}{l|}{\textit{Encoder Layer Pruning}} \\
+Parameter Reduction & 20.16\% & 20.16\% & 20.16\% & 20.16\% & 20.16\% & 20.16\% & 20.16\% & 20.16\% & \\
+LoSparse & 0.929 & 0.856 & 0.858 & - & - & - & - & - & - \\
 \bottomrule
@@
 
+\newpage
+\bibliographystyle{plain}
+\bibliography{references}
+
 \end{document}

### 2025-05-10 pruning/importance.py
param based implementation

@@ -16,7 +16,7 @@ from utils import get_task_metadata
 # Configure logging
 logger = logging.getLogger(__name__)
 
-def setup_importance_based_model(task_name, model_name, prune_percent=20, importance_type=3):
+def setup_importance_based_model(task_name, model_name, prune_percent=20, importance_type=3, param_based=False):
     """
     Set up a model with word importance-based vocabulary pruning (without OOV clustering).
     
@@ -25,13 +25,15 @@ def setup_importance_based_model(task_name, model_name, prune_percent=20, import
         model_name: Base model to use
         prune_percent: Percentage of tokens to prune based on importance
         importance_type: Word importance setting (0=frequency only, 1-3=TF-IDF variants)
+        param_based: If True, prune based on parameter percentage rather than token percentage
         
     Returns:
         model: Model with importance-based vocabulary pruning
         token_map: Mapping from original token IDs to new IDs
         oov_lookup: None for this method (OOV tokens mapped to UNK)
     """
-    logger.info(f"Setting up importance-based model for {task_name} with {prune_percent}% pruning, importance_type={importance_type}")
+    pruning_type = "parameter-based" if param_based else "token-based"
+    logger.info(f"Setting up {pruning_type} importance-based model for {task_name} with {prune_percent}% pruning, importance_type={importance_type}")
     
     # Load GLUE task metadata
     task_meta = get_task_metadata(task_name)
@@ -44,8 +46,24 @@ def setup_importance_based_model(task_name, model_name, prune_percent=20, import
     # Load model for embedding extraction
     model = AutoModelForSequenceClassification.from_pretrained(model_name, num_labels=n_labels)
     
-    # Importance-based pruning
-    tokens_to_keep, _ = importance_based_pruning(token_counter, token_importance, prune_percent)
+    # Get total model parameters and embedding dimension
+    total_params = sum(p.numel() for p in model.parameters())
+    embedding_dim = model.model.embeddings.tok_embeddings.embedding_dim
+    total_vocab_size = model.model.embeddings.tok_embeddings.num_embeddings
+    
+    logger.info(f"Model has {total_params:,} total parameters")
+    logger.info(f"Token embeddings: {total_vocab_size:,} tokens with dimension {embedding_dim}")
+    
+    # Importance-based pruning with parameter-based option
+    tokens_to_keep, _ = importance_based_pruning(
+        token_counter, 
+        token_importance, 
+        prune_percent,
[…]

### 2025-05-10 main.py
oov token handling

@@ -516,46 +516,69 @@ def run_cross_validation_pipeline(args, tokenizer):
             
             # Apply token remapping to create datasets with reduced vocabulary
             logger.info("Applying vocabulary mapping to datasets")
-            
-            # Function to remap tokens in a dataset
-            def remap_tokens(dataset):
-                # If no token_map (no pruning case), return the dataset unchanged
-                if token_map is None:
-                    logger.info("No token mapping applied (using full vocabulary)")
-                    return dataset
+
+            # For OOV methods using TokenMappingWrapper, we should keep the original datasets
+            if args.pruning_method == "frequency_oov" or args.pruning_method == "importance_oov":
+                logger.info("Using TokenMappingWrapper for token remapping during forward pass, keeping original datasets")
+                # Just make sure the label column is named correctly
+                if 'label' in train_dataset.column_names and 'labels' not in train_dataset.column_names:
+                    train_dataset = train_dataset.rename_column('label', 'labels')
+                if 'label' in eval_dataset.column_names and 'labels' not in eval_dataset.column_names:
+                    eval_dataset = eval_dataset.rename_column('label', 'labels')
+                if 'label' in test_dataset.column_names and 'labels' not in test_dataset.column_names:
+                    test_dataset = test_dataset.rename_column('label', 'labels')
                 
-                # Create a mapping dictionary for quick lookup
-                id_map_dict = token_map.copy()
-                if oov_lookup:
-                    id_map_dict.update(oov_lookup)
+                # Use original datasets
+                remapped_train = train_dataset
+                remapped_eval = eval_dataset
+                remapped_test = test_dataset
+            # For no_pruning, we'll use the datasets as-is
+            elif token_map is None:
+                logger.info("No token mapping applied (using full vocabulary)")
+                # Ensure label column naming consistency
+                for dataset in [train_dataset, eval_dataset, test_dataset]:
+                    if 'label' in dataset.column_names and 'labels' not in dataset.column_names:
+                        dataset = dataset.rename_column('label', 'labels')
                 
-                def map_example(example):
-                    # Map input_ids using our token mapping
-                    new_input_ids = []
-                    for token_id in example['input_ids']:
-                        if token_id in id_map_dict:
-                            new_input_ids.append(id_map_dict[token_id])
-                        else:
-                            # Use UNK token (0) for OOV tokens
-                            new_input_ids.append(0)
+                # Use original datasets
+                remapped_train = train_dataset
[…]

### 2025-05-10 pruning/base.py

@@ -9,6 +9,8 @@ from collections import Counter
 import torch
 from datasets import load_dataset
 from transformers import AutoTokenizer
+import torch.nn as nn
+import copy
 
 # Configure logging
 logger = logging.getLogger(__name__)
@@ -17,6 +19,7 @@ logger = logging.getLogger(__name__)
 task_to_keys = {
     "cola": ("sentence", None),
     "mnli": ("premise", "hypothesis"),
+    "mnli-mm": ("premise", "hypothesis"),
     "mrpc": ("sentence1", "sentence2"),
     "qnli": ("question", "sentence"),
     "qqp": ("question1", "question2"),
@@ -25,6 +28,56 @@ task_to_keys = {
     "stsb": ("sentence1", "sentence2"),
     "wnli": ("sentence1", "sentence2"),
 }
+
+# Add a new TokenMappingWrapper class to handle token remapping
+class TokenMappingWrapper(nn.Module):
+    """
+    Model wrapper that applies token mapping before the embedding layer
+    to handle vocabulary pruning.
+    """
+    def __init__(self, model, token_map=None, oov_lookup=None):
+        super().__init__()
+        self.model = model
+        self.token_map = token_map or {}
+        self.oov_lookup = oov_lookup or {}
+        
+        # Combine mappings for faster lookup
+        self.combined_map = {}
+        if token_map:
+            self.combined_map.update(token_map)
+        if oov_lookup:
+            self.combined_map.update(oov_lookup)
+            
+        logger.info(f"TokenMappingWrapper initialized with {len(self.token_map)} mapped tokens and {len(set(self.oov_lookup.values()))} OOV clusters")
+        
+    def forward(self, input_ids=None, attention_mask=None, **kwargs):
+        # Only process if we have input_ids and mappings
+        if input_ids is not None and self.combined_map:
+            # Create a new tensor to hold the mapped input IDs
+            mapped_input_ids = torch.zeros_like(input_ids)
+            
+            # Apply mapping to each element in the input_ids tensor
+            for i in range(input_ids.size(0)):
+                for j in range(input_ids.size(1)):
+                    token_id = input_ids[i, j].item()
+                    if token_id in self.combined_map:
+                        mapped_input_ids[i, j] = self.combined_map[token_id]
+                    else:
+                        # For OOV tokens not in mapping, use UNK token (0)
+                        mapped_input_ids[i, j] = 0
+            
+            # Replace input_ids with mapped version
+            input_ids = mapped_input_ids
+        
[…]

### 2025-05-10 pruning/frequency_oov.py

@@ -11,7 +11,7 @@ import torch.nn as nn
 from sklearn.cluster import KMeans
 from transformers import AutoTokenizer, AutoModelForSequenceClassification
 
-from .base import get_dataset_tokens_with_counts, create_hybrid_embeddings
+from .base import get_dataset_tokens_with_counts, create_hybrid_embeddings, TokenMappingWrapper
 from .frequency import frequency_based_pruning
 from utils import get_task_metadata
 
@@ -156,4 +156,7 @@ def setup_frequency_oov_model(task_name, model_name, prune_percent=20, num_clust
         reduced_embeddings, freeze=False
     )
     
-    return model, token_map, oov_lookup 
+    # Step 4: Wrap the model to handle token remapping during forward pass
+    wrapped_model = TokenMappingWrapper(model, token_map, oov_lookup)
+    
+    return wrapped_model, token_map, oov_lookup 

### 2025-05-10 pruning/importance_oov.py

@@ -18,7 +18,7 @@ try:
 except ImportError:
     NLTK_AVAILABLE = False
 
-from .base import create_hybrid_embeddings
+from .base import create_hybrid_embeddings, TokenMappingWrapper
 from .frequency import frequency_based_pruning
 from .frequency_oov import cluster_removed_tokens
 from utils import get_task_metadata
@@ -394,4 +394,7 @@ def setup_importance_oov_model(task_name, model_name, prune_percent=20, num_clus
         reduced_embeddings, freeze=False
     )
     
-    return model, token_map, oov_lookup 
+    # Wrap model to handle token remapping during forward pass
+    wrapped_model = TokenMappingWrapper(model, token_map, oov_lookup)
+    
+    return wrapped_model, token_map, oov_lookup 

### 2025-05-10 training.py

@@ -61,6 +61,11 @@ def setup_training(model, train_dataset, eval_dataset, task_name, args, tokenize
     if args.pruning_method == "no_pruning" and tokenizer is not None:
         # For no_pruning, use the standard DataCollatorWithPadding
         data_collator = DataCollatorWithPadding(tokenizer=tokenizer)
+    elif args.pruning_method in ["frequency_oov", "importance_oov"]:
+        # For OOV methods using TokenMappingWrapper, use ReducedVocabDataCollator
+        # The token remapping is now handled in the model's forward pass
+        logger.info("Using ReducedVocabDataCollator with TokenMappingWrapper")
+        data_collator = ReducedVocabDataCollator(pad_token_id=0)
     elif args.pruning_method in ["hybrid", "importance"]:
         data_collator = HybridCollator(pad_token_id=0, unk_token_id=0)
     else:
@@ -169,90 +174,69 @@ def generate_test_predictions(model, test_dataset, data_collator, args, tokenize
     
     logger.info("\n=== Test Set Prediction Generation ===")
     try:
-        # Generate predictions without computing loss
-        logger.info("Generating predictions on test set")
-        
         # If trainer is provided, use it for predictions to ensure consistent processing
         if trainer is not None:
             logger.info("Using Trainer.predict() for consistent processing")
             prediction_output = trainer.predict(test_dataset)
-            all_predictions = prediction_output.predictions
+            predictions = prediction_output.predictions
             
             # For classification tasks, get the class with highest probability
-            if len(all_predictions.shape) > 1 and all_predictions.shape[1] > 1:
-                predictions = np.argmax(all_predictions, axis=1)
+            if len(predictions.shape) > 1 and predictions.shape[1] > 1:
+                predictions = np.argmax(predictions, axis=1)
             else:
                 # For regression tasks
-                predictions = all_predictions.squeeze()
+                predictions = predictions.squeeze()
         else:
-            # Fall back to manual prediction if no trainer is available
-            with torch.no_grad():
-                device = next(model.parameters()).device
-                predictions = []
+            # Fall back to manual processing if no trainer is provided
+            logger.info("Using manual prediction generation")
+            predictions = []
+            
+            # Create dataloader for test set
+            dataloader = torch.utils.data.DataLoader(
+                test_dataset, 
+                batch_size=args.batch_size,
+                collate_fn=data_collator
+            )
+            
+            # Move model to appropriate device
[…]

### 2025-05-10 data.py
fix oov

@@ -39,14 +39,23 @@ class ReducedVocabDataCollator:
             mask = f['attention_mask'] + [0] * (max_length - len(f['attention_mask']))
             attention_mask.append(mask)
             
-            # Get labels
-            labels.append(f['labels'])
+            # Get labels if they exist, otherwise use 0 as a placeholder
+            if 'labels' in f:
+                labels.append(f['labels'])
+            elif 'label' in f:
+                # Some datasets use 'label' instead of 'labels'
+                labels.append(f['label'])
+            else:
+                # Use 0 as a default/placeholder value for prediction-only data
+                labels.append(0)
         
-        return {
+        result = {
             'input_ids': torch.tensor(input_ids),
             'attention_mask': torch.tensor(attention_mask),
             'labels': torch.tensor(labels)
         }
+        
+        return result
 
 class HybridCollator:
     """
@@ -77,8 +86,15 @@ class HybridCollator:
             mask = f['attention_mask'] + [0] * (max_length - len(f['attention_mask']))
             attention_mask.append(mask)
             
-            # Get labels
-            labels.append(f['labels'])
+            # Get labels if they exist, otherwise use 0 as a placeholder
+            if 'labels' in f:
+                labels.append(f['labels'])
+            elif 'label' in f:
+                # Some datasets use 'label' instead of 'labels'
+                labels.append(f['label'])
+            else:
+                # Use 0 as a default/placeholder value for prediction-only data
+                labels.append(0)
         
         # Check if any labels are -1 (test set)
         has_invalid_labels = any(label == -1 for label in labels)

### 2025-05-10 pruning/base.py

@@ -48,11 +48,15 @@ class TokenMappingWrapper(nn.Module):
         if oov_lookup:
             self.combined_map.update(oov_lookup)
             
+        # Determine embedding size for clamping
+        self.embedding_size = model.model.embeddings.tok_embeddings.weight.size(0)
+        
         logger.info(f"TokenMappingWrapper initialized with {len(self.token_map)} mapped tokens and {len(set(self.oov_lookup.values()))} OOV clusters")
+        logger.info(f"Embedding matrix size: {self.embedding_size}")
         
     def forward(self, input_ids=None, attention_mask=None, **kwargs):
-        # Only process if we have input_ids and mappings
-        if input_ids is not None and self.combined_map:
+        # Only process if we have input_ids
+        if input_ids is not None:
             # Create a new tensor to hold the mapped input IDs
             mapped_input_ids = torch.zeros_like(input_ids)
             
@@ -65,6 +69,9 @@ class TokenMappingWrapper(nn.Module):
                     else:
                         # For OOV tokens not in mapping, use UNK token (0)
                         mapped_input_ids[i, j] = 0
+            
+            # Ensure no out-of-bounds indices by clamping to valid range
+            mapped_input_ids = torch.clamp(mapped_input_ids, 0, self.embedding_size - 1)
             
             # Replace input_ids with mapped version
             input_ids = mapped_input_ids
@@ -208,14 +215,34 @@ def create_reduced_embeddings(task_vocab, model):
     """
     logger.info(f"Creating reduced embeddings for task_vocab of size {len(task_vocab)}")
     
+    # Get original embedding matrix
+    original_embeddings = model.model.embeddings.tok_embeddings.weight.data
+    original_size = original_embeddings.size(0)
+    
+    # Make sure task_vocab only contains valid indices
+    valid_tokens = [t for t in task_vocab if t < original_size]
+    if len(valid_tokens) < len(task_vocab):
+        logger.warning(f"Removed {len(task_vocab) - len(valid_tokens)} out-of-bounds token indices from task_vocab")
+        task_vocab = valid_tokens
+    
     # Create mapping from original token IDs to new consecutive IDs
     token_map = {old_id: new_id for new_id, old_id in enumerate(task_vocab)}
     
-    # Get original embedding matrix
-    original_embeddings = model.model.embeddings.tok_embeddings.weight.data
-    
     # Create new embedding matrix with only the needed vectors
-    reduced_embeddings = torch.stack([original_embeddings[i] for i in task_vocab])
[…]

### 2025-05-10 pruning/frequency_oov.py

@@ -31,15 +31,38 @@ def cluster_removed_tokens(tokens_to_remove, tokens_to_keep, model, num_clusters
         
     Returns:
         oov_lookup: Mapping from removed token ID to nearest cluster center token ID
+        updated_tokens_to_keep: Updated list of tokens to keep (may include cluster centers)
     """
     if not tokens_to_remove:
         logger.info("No tokens to cluster (all tokens kept)")
-        return {}
+        return {}, tokens_to_keep
         
     logger.info(f"Clustering {len(tokens_to_remove)} removed tokens into {num_clusters} clusters")
     
     # Get original embedding matrix
     original_embeddings = model.model.embeddings.tok_embeddings.weight.data
+    
+    # Get the maximum valid token ID (embedding size - 1)
+    max_valid_token_id = original_embeddings.size(0) - 1
+    
+    # Filter out any invalid token IDs from tokens_to_remove and tokens_to_keep
+    valid_tokens_to_remove = [t for t in tokens_to_remove if t <= max_valid_token_id]
+    valid_tokens_to_keep = [t for t in tokens_to_keep if t <= max_valid_token_id]
+    
+    if len(valid_tokens_to_remove) < len(tokens_to_remove):
+        logger.warning(f"Filtered out {len(tokens_to_remove) - len(valid_tokens_to_remove)} invalid token IDs from tokens_to_remove")
+    
+    if len(valid_tokens_to_keep) < len(tokens_to_keep):
+        logger.warning(f"Filtered out {len(tokens_to_keep) - len(valid_tokens_to_keep)} invalid token IDs from tokens_to_keep")
+    
+    # Use the valid tokens
+    tokens_to_remove = valid_tokens_to_remove
+    tokens_to_keep = valid_tokens_to_keep
+    
+    # If after filtering, we have no tokens to remove, return
+    if not tokens_to_remove:
+        logger.info("No valid tokens to cluster after filtering")
+        return {}, tokens_to_keep
     
     # Get embeddings for removed tokens
     removed_embeddings = original_embeddings[tokens_to_remove].cpu().numpy()
@@ -54,6 +77,8 @@ def cluster_removed_tokens(tokens_to_remove, tokens_to_keep, model, num_clusters
     # Initialize mapping from removed token to nearest cluster center token
     oov_lookup = {}
     cluster_centers = []
+    # Create a copy of tokens_to_keep that we can potentially update
+    updated_tokens_to_keep = tokens_to_keep.copy()
     
[…]

### 2025-05-10 pruning/importance_oov.py

@@ -381,18 +381,26 @@ def setup_importance_oov_model(task_name, model_name, prune_percent=20, num_clus
     
     # Cluster removed tokens if there are any tokens to remove
     if tokens_to_remove:
-        oov_lookup = cluster_removed_tokens(tokens_to_remove, tokens_to_keep, model, num_clusters)
+        oov_lookup, updated_tokens_to_keep = cluster_removed_tokens(tokens_to_remove, tokens_to_keep, model, num_clusters)
     else:
         oov_lookup = {}
+        updated_tokens_to_keep = tokens_to_keep
     
     # Create hybrid embeddings with OOV token clustering
-    logger.info(f"Creating hybrid embeddings for {len(tokens_to_keep)} kept tokens and {len(oov_lookup)} OOV tokens")
-    token_map, reduced_embeddings = create_hybrid_embeddings(tokens_to_keep, oov_lookup, model)
+    logger.info(f"Creating hybrid embeddings for {len(updated_tokens_to_keep)} kept tokens and {len(oov_lookup)} OOV tokens")
+    token_map, reduced_embeddings = create_hybrid_embeddings(updated_tokens_to_keep, oov_lookup, model)
+    
+    # Log embedding size before replacing
+    logger.info(f"Original embedding size: {model.model.embeddings.tok_embeddings.weight.size()}")
+    logger.info(f"Reduced embedding size: {reduced_embeddings.size()}")
     
     # Replace embedding layer with reduced version
     model.model.embeddings.tok_embeddings = nn.Embedding.from_pretrained(
         reduced_embeddings, freeze=False
     )
+    
+    # Verify the embedding size after replacement
+    logger.info(f"New embedding size: {model.model.embeddings.tok_embeddings.weight.size()}")
     
     # Wrap model to handle token remapping during forward pass
     wrapped_model = TokenMappingWrapper(model, token_map, oov_lookup)

### 2025-05-11 paper/main.tex
results parameter based 20%

@@
 \usepackage{cite}
+\usepackage{multirow}
 
@@
 In ``Load What You Need,'' they drop languages from mBERT's vocabulary and obtain up to a 45\% reduction in total parameters with comparable accuracy on XNLI.
-\\
-\\
-More recently, Nair et al.~\cite{nair2023blade} proposed BLADE: they prune a bilingual mBERT vocabulary to include only tokens from the query and document languages for cross-lingual IR, then apply a small intermediate pretraining step.
+
+More recently, Nair et al.~\cite{nair2023blade} proposed \textbf{BLADE}: they prune a bilingual mBERT vocabulary to include only tokens from the query and document languages for cross-lingual IR, then apply a small intermediate pretraining step.
 This pruned model (keeping only needed embeddings) reduced parameters by $\sim$36.5\% versus full mBERT and sped up training by $\sim$30\% and inference by $\sim$55\%, with minimal loss in retrieval quality.
-\\
-\\
+
 Dorkin et al.~\cite{dorkin2025estonian} also evaluated pruning for an Estonian adaptation: pruning all tokens unused by Estonian led to no degradation on NER, whereas retraining a new tokenizer hurt performance.
@@
 
-\begin{itemize}
-    \item Pruning toolkits: TextPruner~\cite{shen2022textpruner} automates vocabulary pruning: it scans a corpus and removes any tokenizer token not present in the text, deleting its row in the embedding matrix.
+TextPruner by Shen et al.~\cite{shen2022textpruner} automates vocabulary pruning: it scans a corpus and removes any tokenizer token not present in the text, deleting its row in the embedding matrix.
 This simple ``drop unused tokens'' mode can cut model size and speed up training/inference on that domain.
-    \item Sparse expansion models: In sparse-retrieval models like SPLADE, vocabulary pruning is implicit.
+
+In sparse-retrieval models like SPLADE, vocabulary pruning is implicit.
 SPLADE-X~\cite{formal2023spladex} (multi-lingual) restricted output to query-language tokens only, effectively pruning all other vocab.
 BLADE extends this idea by explicitly building a pruned bilingual model with only the union of query and document subwords, discarding other embeddings entirely.
-    \item Empirical results: Across tasks and languages, pruned-vocabulary models often match full models.
+
+Across tasks and languages, pruned-vocabulary models often match full models.
 For example, after pruning, Nair et al.~\cite{nair2023blade} report retrieval effectiveness on par with larger baselines, while achieving much faster indexing.
@@
 These findings suggest vocabulary pruning is an effective way to tailor large models to specific domains or languages.
-\end{itemize}
 
@@
 Pruning the embedding layer often means removing or compressing rows (token vectors) to shrink this heavy component.
-Partial Embedding Matrix Adaptation (PEMA)~\cite{bousquet2023pema} is one systematic approach: they first note that fine-tuning datasets typically use only a small fraction of the full vocabulary.
+\textbf{Partial Embedding Matrix Adaptation (PEMA)} by Bousquet et al.~\cite{bousquet2023pema} is one systematic approach: they first note that fine-tuning datasets typically use only a small fraction of the full vocabulary.
 PEMA therefore builds a partial embedding matrix containing only tokens seen in the task data, training on that smaller matrix.
-This saves GPU memory during fine-tuning without altering task accuracy (unused embeddings are never updated).
-
+This saves GPU memory during fine-tuning without altering task accuracy.
 After fine-tuning, the updated embeddings are merged back into the full matrix so the model structure remains unchanged.
@@
 Crucially, PEMA reports no drop in downstream performance, making it a practical way to prune embeddings during training.
-\\
-\\
-Another approach is embedding compression via factorization or hashing.
-Wang et al.~\cite{wang2023lighttoken} propose LightToken, which compresses the entire embedding matrix using low-rank SVD plus a binary residual autoencoder.
+
+Another approach is \textbf{embedding compression} via factorization or hashing.
+Wang et al.~\cite{wang2023lighttoken} propose \textbf{LightToken}, which compresses the entire embedding matrix using low-rank SVD plus a binary residual autoencoder.
 They note token embeddings are highly redundant (the embedding takes $\sim$21--31\% of BERT/RoBERTa's size).
@@
 
-Thus, LightToken and similar methods show that even structured low-rank or quantized pruning of embeddings can yield massive space savings with minimal impact on performance. 
-\\
-\\
+These methods show that even structured low-rank or quantized pruning of embeddings can yield massive space savings with minimal impact on performance.
+
 Overall, embedding-layer pruning techniques exploit the fact that many token vectors (especially for rare subwords) contribute little to end-task accuracy.
@@
 
-\subsection{Encoder-Layer Pruning}
+\subsection{Encoder-Layer Pruning (Auxiliary Methods)}
 
 Although our focus is vocab/embedding pruning, many complementary pruning methods work at the encoder layers.
-A classic example is attention-head pruning: Michel et al.~\cite{michel2019heads} discovered that ``a large percentage of attention heads can be removed at test time without significantly impacting performance''.
-Voita et al.~\cite{voita2019heads} likewise pruned 38 of 48 encoder heads in an NMT transformer (removing $\sim$79\% of heads) with only a 0.15 BLEU loss.
+A classic example is \textbf{attention-head pruning}: Michel et al.~\cite{michel2019sixteen} discovered that a large percentage of attention heads can be removed at test time without significantly impacting performance.
+Voita et al.~\cite{voita2019analyzing} likewise pruned 38 of 48 encoder heads in an NMT transformer (removing $\sim$79\% of heads) with only a 0.15 BLEU loss.
 
 In practice, toolkit frameworks (like TextPruner) often support head and FFN pruning alongside vocab pruning.
-Other methods include movement pruning~\cite{sanh2020movement} that gradually zeros out small weights during fine-tuning, or LayerDrop~\cite{fan2020layerdrop} which stochastically drops whole layers during training.
-These encoder-level techniques are orthogonal to vocab pruning: they further thin out the model by removing redundant heads, neurons, or weight components, whereas vocabulary pruning targets the input/output embedding parameters.
+Other methods include \textit{movement pruning}~\cite{sanh2020movement} that gradually zeros out small weights during fine-tuning, or \textit{LayerDrop}~\cite{fan2020layerdrop} which stochastically drops whole layers during training.
 
@@
 \section{Experiments}
+% \begin{itemize}
+%     \item Multiple pruning percentages (mrpc, mnli, sst2)
+%     \item LoSparse + Vocab pruning combined (mrpc, mnli, sst2)
+% \end{itemize}
+This section evaluates the proposed vocabulary pruning method on the GLUE benchmark, both in isolation and in combination with structured encoder-layer pruning (LoSparse), in order to assess effectiveness across a variety of downstream tasks.
+The experiments are designed to address the following questions:
 
-\begin{table*}[ht!]
+\begin{enumerate}
+    \item Impact of pure vocabulary pruning: How does performance vary at different pruning ratios?
+    \item Comparison of pruning strategies: Does the TF-IDF + OOV hybrid approach outperform simpler baselines (e.g., frequency-based, random selection, clustering, attention-based)?
+    \item Interaction with encoder sparsification: What synergistic effects arise when combining embedding-level pruning with encoder-layer sparsification in ModernBERT?
+\end{enumerate}
+
+\subsection{Datasets and Metrics}
+Eight GLUE development tasks are employed: SST-2, MRPC, MNLI, CoLA, STS-B, QQP, QNLI, and RTE.
+Classification tasks report accuracy; MRPC additionally reports F1, CoLA reports Matthew correlation coefficient, and STS-B reports Pearson correlation.
+
+\subsection{Baselines and Pruning Methods}
+The following configurations are compared:
+
+\begin{itemize}
+    \item ModernBERT (Full): No pruning.
+    \item Train Tokens Only: Removal of embeddings not observed in fine-tuning data.
+    \item Random Selection: Uniform removal of embeddings.
+    \item Clustering-Based: Retention of one centroid token per semantic cluster.
+    \item Frequency-Based: Pruning of lowest-frequency tokens.
+    \item Attention-Based: Removal of tokens with lowest aggregated attention weights.
+    \item TF-IDF Based: Pruning according to L2-normalized TF-IDF scores.
+    \item Frequency + OOV: Hybrid schemes that remap pruned tokens via clustering representatives.
+    \item TF-IDF + OOV: Hybrid schemes that remap pruned tokens via clustering representatives.
+    \item LoSparse (20\% encoder pruning): Structured encoder-layer pruning only.
+    \item LoSparse + TF-IDF + OOV: Joint embedding and encoder pruning.
+\end{itemize}
+
+\begin{table*}[h]
 \centering
-\small
+\scriptsize
 \caption{Comparison of vocabulary pruning techniques on GLUE benchmark tasks.
@@
 \label{tab:results}
-\setlength{\tabcolsep}{6pt}
-\begin{tabular}{lcccccccc|c}
+\setlength{\tabcolsep}{2.8pt}
+\begin{tabular}{lcccc|cc|cc|cc|cc|c}
 \toprule
-\textbf{Method} & \textbf{SST-2} & \textbf{MRPC} & \textbf{MNLI} & \textbf{CoLA} & \textbf{STS-B} & \textbf{QQP} & \textbf{QNLI} & \textbf{RTE} & \textbf{AVG} \\
+\textbf{Method} & \textbf{SST-2} & \textbf{MRPC} & \textbf{CoLA} & \textbf{STS-B} & \multicolumn{2}{c|}{\textbf{MNLI}} & \multicolumn{2}{c|}{\textbf{QQP}} & \multicolumn{2}{c|}{\textbf{QNLI}} & \multicolumn{2}{c|}{\textbf{RTE}} & \textbf{AVG} \\
 \midrule
-\multicolumn{9}{l|}{\textit{Baseline}} \\
-ModernBERT (Full) & 0.951 & 0.856 & 0.881 & 0.586 & 0.917 & 0.917 & 0.916 & 0.598 & 0.828 \\
+\multicolumn{14}{l}{\textit{Baseline}} \\
+ModernBERT (Full) & 0.951 & 0.856 & 0.586 & 0.917 & \multicolumn{2}{c|}{0.881} & \multicolumn{2}{c|}{0.917} & \multicolumn{2}{c|}{0.939} & \multicolumn{2}{c|}{0.598} & 0.831 \\
 \midrule
-\multicolumn{9}{l|}{\textit{Vocabulary Pruning}} \\
-Parameter Reduction & 18.85\% & 18.56\% & 6.74\% & 22.61\% & 18.76\% & 4.79\% & 6.42\% & 17.06\% &  \\
-Train Tokens Only & 0.950 & 0.831 & 0.883 & 0.509 & 0.917 & 0.917 & 0.915 & 0.598 & 0.815 \\
+\multicolumn{14}{l}{\textit{Vocabulary Pruning}} \\
+Parameter Reduction & 18.85\% & 18.56\% & 22.61\% & 18.76\% & \multicolumn{2}{c|}{6.74\%} & \multicolumn{2}{c|}{4.79\%} & \multicolumn{2}{c|}{6.42\%} & \multicolumn{2}{c|}{17.06\%} & \\
+Train Tokens Only & 0.950 & 0.831 & 0.509 & 0.917 & \multicolumn{2}{c|}{0.883} & \multicolumn{2}{c|}{0.917} & \multicolumn{2}{c|}{0.915} & \multicolumn{2}{c|}{0.598} & 0.815 \\
 \midrule
-Parameter Reduction & 20.25\% & 20.02\% & 10.57\% & 23.27\% & 20.18\% & 9.01\% & 10.31\% & 18.83\% & \\
-Random Selection & 0.911 & 0.798 & 0.832 & 0.470 & 0.845 & 0.902 & 0.895 & 0.522 & 0.772 \\
-Clustering-Based & 0.899 & 0.714 & 0.712 & 0.000 & 0.786 & 0.875 & 0.836 & 0.510 & 0.667 \\
-Frequency-Based & 0.943 & 0.812 & 0.880 & 0.467 & \textbf{0.905} & \textbf{0.916} & \textbf{0.920} & 0.542 & 0.798 \\
-Attention-Based & \textbf{0.947} & \textbf{0.864} & 0.864 & \textbf{0.589} & 0.885 & 0.912 & 0.912 & 0.550 & 0.803 \\
-TF-IDF Based & 0.933 & 0.807 & \textbf{0.875} & 0.520 & 0.901 & 0.900 & 0.917 & \textbf{0.606} & \textbf{0.807} \\
+Parameter Reduction & 20.25\% & 20.02\% & 23.27\% & 20.18\% & 10.57\% & 20.02\% & 9.01\% & 20.02\% & 10.31\% & 20.02\% & 18.83\% & 20.02\% & \\
+Random Selection & 0.911 & 0.798 & 0.470 & 0.845 & 0.832 & 0.504 & 0.902 & 0.780 & 0.895 & 0.669 & 0.522 & 0.566 & 0.772 \\
+Clustering-Based & 0.899 & 0.714 & 0.000 & 0.786 & 0.712 & 0.501 & 0.875 & 0.774 & 0.836 & 0.510 & 0.510 & 0.566 & 0.667 \\
+Frequency-Based & 0.943 & 0.812 & 0.467 & \textbf{0.905} & 0.880 & 0.858 & \textbf{0.916} & \textbf{0.904} & \textbf{0.920} & \textbf{0.902} & 0.542 & 0.546 & 0.798 \\
+Attention-Based & \textbf{0.947} & \textbf{0.864} & 0.589 & 0.885 & 0.864 & - & 0.912 & - & 0.912 & - & 0.550 & - & 0.803 \\
+TF-IDF Based & 0.933 & 0.807 & 0.520 & 0.901 & 0.875 & 0.860 & 0.900 & 0.898 & 0.917 & 0.909 & 0.606 & \textbf{0.574} & 0.807 \\
+\\ [-6pt]
+\cdashline{1-14}
+\\[-6pt]
+Freq + OOV & 0.938 & 0.828 & 0.540 & 0.906 & \textbf{0.881} & - & \textbf{0.916} & - & 0.918 & - & 0.538 & - & 0.808 \\
+TF-IDF + OOV & 0.923 & 0.834 & \textbf{0.615} & 0.903 & 0.878 & - & 0.906 & - & 0.919 & - & 0.554 & - & \textbf{0.817} \\
 \\ [-8pt]
-\cdashline{1-10}
-\\[-8pt]
-Freq + OOV & \textbf{0.938} & 0.828 & \textbf{0.881} & 0.540 & 0.906 & 0.916 & 0.918 & 0.538 & 0.808 \\
-TF-IDF + OOV & 0.923 & 0.834 & 0.878 & 0.615 & 0.903 & 0.906 & 0.919 & 0.554 & \textbf{0.817} \\
-\\ [-8pt]
-\cdashline{1-10}
-\\[-8pt]
-\multicolumn{9}{l|}{\textit{Encoder Layer Pruning}} \\
-Parameter Reduction & 20.16\% & 20.16\% & 20.16\% & 20.16\% & 20.16\% & 20.16\% & 20.16\% & 20.16\% & \\
-LoSparse & 0.929 & 0.856 & 0.858 & - & - & - & - & - & - \\
+\cdashline{1-14}
+\\[-6pt]
+\multicolumn{14}{l}{\textit{Encoder Layer Pruning}} \\
+Parameter Reduction & 20.16\% & 20.16\% & 20.16\% & 20.16\% & \multicolumn{2}{c|}{20.16\%} & \multicolumn{2}{c|}{20.16\%} & \multicolumn{2}{c|}{20.16\%} & \multicolumn{2}{c|}{20.16\%} & \\
+LoSparse & 0.929 & 0.856 & 0.316 & 0.882 & \multicolumn{2}{c|}{0.858} & \multicolumn{2}{c|}{0.911} & \multicolumn{2}{c|}{0.907} & \multicolumn{2}{c|}{\textbf{0.610}} & 0.784 \\
 \bottomrule
@@
 \end{table*}
-
-\begin{table}[ht!]
-\centering
-\caption{Average performance change across GLUE tasks with different pruning methods.}
-\label{tab:avg_performance}
-\begin{tabular}{lcc}
-\toprule
-\textbf{Method} & \textbf{Avg. Param. Reduction} & \textbf{Avg. Performance Drop} \\
-\midrule
-Train Tokens Only & 14.22\% & 1.25\% \\
-Random Selection & 16.44\% & 5.58\% \\
-Clustering-Based & 16.44\% & 16.24\% \\
-Frequency-Based & 16.44\% & 2.32\% \\
-Attention-Based & 15.92\% & 1.99\% \\
-TF-IDF Based & 16.44\% & 2.04\% \\
-Frequency + OOV & 16.40\% & 1.82\% \\
-TF-IDF + OOV & 16.40\% & 0.96\% \\
-\bottomrule
-\end{tabular}
-\end{table}
 

### 2025-05-11 paper/main.tex
update gitignore

@@
 \midrule
-Parameter Reduction & 20.25\% & 20.02\% & 23.27\% & 20.18\% & 10.57\% & 20.02\% & 9.01\% & 20.02\% & 10.31\% & 20.02\% & 18.83\% & 20.02\% & \\
-Random Selection & 0.911 & 0.798 & 0.470 & 0.845 & 0.832 & 0.504 & 0.902 & 0.780 & 0.895 & 0.669 & 0.522 & 0.566 & 0.772 \\
-Clustering-Based & 0.899 & 0.714 & 0.000 & 0.786 & 0.712 & 0.501 & 0.875 & 0.774 & 0.836 & 0.510 & 0.510 & 0.566 & 0.667 \\
-Frequency-Based & 0.943 & 0.812 & 0.467 & \textbf{0.905} & 0.880 & 0.858 & \textbf{0.916} & \textbf{0.904} & \textbf{0.920} & \textbf{0.902} & 0.542 & 0.546 & 0.798 \\
-Attention-Based & \textbf{0.947} & \textbf{0.864} & 0.589 & 0.885 & 0.864 & - & 0.912 & - & 0.912 & - & 0.550 & - & 0.803 \\
-TF-IDF Based & 0.933 & 0.807 & 0.520 & 0.901 & 0.875 & 0.860 & 0.900 & 0.898 & 0.917 & 0.909 & 0.606 & \textbf{0.574} & 0.807 \\
+Parameter Reduction & 20.25\% & 20.02\% & 23.27\% & 20.18\% & 10.57\% & 20.02\% & 9.01\% & 20.02\% & 10.31\% & 20.02\% & 18.83\% & 20.02\% & - \\
+Random Selection & 0.911 & 0.798 & 0.470 & 0.845 & 0.832 & 0.504 & 0.902 & 0.780 & 0.895 & 0.669 & 0.522 & 0.566 & - \\
+Clustering-Based & 0.899 & 0.714 & 0.000 & 0.786 & 0.712 & 0.501 & 0.875 & 0.774 & 0.836 & 0.510 & 0.510 & 0.566 & - \\
+Frequency-Based & 0.943 & 0.812 & 0.467 & \textbf{0.905} & 0.880 & 0.858 & \textbf{0.916} & \textbf{0.904} & \textbf{0.920} & \textbf{0.902} & 0.542 & 0.546 & - \\
+Attention-Based & \textbf{0.947} & \textbf{0.864} & 0.589 & 0.885 & 0.864 & - & 0.912 & - & 0.912 & - & 0.550 & - & - \\
+TF-IDF Based & 0.933 & 0.807 & 0.520 & 0.901 & 0.875 & 0.860 & 0.900 & 0.898 & 0.917 & 0.909 & 0.606 & \textbf{0.574} & - \\
 \\ [-6pt]
@@
 \\[-6pt]
-Freq + OOV & 0.938 & 0.828 & 0.540 & 0.906 & \textbf{0.881} & - & \textbf{0.916} & - & 0.918 & - & 0.538 & - & 0.808 \\
-TF-IDF + OOV & 0.923 & 0.834 & \textbf{0.615} & 0.903 & 0.878 & - & 0.906 & - & 0.919 & - & 0.554 & - & \textbf{0.817} \\
+Freq + OOV & 0.938 & 0.828 & 0.540 & 0.906 & \textbf{0.881} & - & \textbf{0.916} & - & 0.918 & - & 0.538 & - & - \\
+TF-IDF + OOV & 0.923 & 0.834 & \textbf{0.615} & 0.903 & 0.878 & - & 0.906 & - & 0.919 & - & 0.554 & - & - \\
 \\ [-8pt]
@@
 \multicolumn{14}{l}{\textit{Encoder Layer Pruning}} \\
-Parameter Reduction & 20.16\% & 20.16\% & 20.16\% & 20.16\% & \multicolumn{2}{c|}{20.16\%} & \multicolumn{2}{c|}{20.16\%} & \multicolumn{2}{c|}{20.16\%} & \multicolumn{2}{c|}{20.16\%} & \\
+Parameter Reduction & 20.16\% & 20.16\% & 20.16\% & 20.16\% & \multicolumn{2}{c|}{20.16\%} & \multicolumn{2}{c|}{20.16\%} & \multicolumn{2}{c|}{20.16\%} & \multicolumn{2}{c|}{20.16\%} & 20.16\% \\
 LoSparse & 0.929 & 0.856 & 0.316 & 0.882 & \multicolumn{2}{c|}{0.858} & \multicolumn{2}{c|}{0.911} & \multicolumn{2}{c|}{0.907} & \multicolumn{2}{c|}{\textbf{0.610}} & 0.784 \\

### 2025-05-13 main.py
fix

@@ -294,7 +294,10 @@ def evaluate_test_set(model, test_dataset, task_name, data_collator, batch_size,
     if trainer is not None:
         logger.info("Using Trainer.predict() for consistent processing")
         prediction_output = trainer.predict(test_dataset)
-        predictions = prediction_output.predictions
+        
+        # prediction_output is a tuple containing (predictions, label_ids, metrics)
+        # Extract the predictions from the first element of the tuple
+        predictions = prediction_output[0]
         
         # For classification tasks, get the class with highest probability
         if len(predictions.shape) > 1 and predictions.shape[1] > 1:

### 2025-05-13 pruning/attention.py

@@ -325,7 +325,7 @@ def setup_attention_based_model(task_name, model_name, attention_model=None, pru
     
     # Load the base model for pruning
     model = AutoModelForSequenceClassification.from_pretrained(
-        model_name, num_labels=task.n_labels, cache_dir=None
+        model_name, num_labels=task["n_labels"], cache_dir=None
     )
     
     # Get total parameters and embedding dimension for parameter-based pruning
@@ -351,7 +351,7 @@ def setup_attention_based_model(task_name, model_name, attention_model=None, pru
     # Create a reduced embedding layer
     model = replace_embeddings(model, tokens_to_keep)
     
-    return model 
+    return model, dict(zip(tokens_to_keep, range(len(tokens_to_keep)))), None
 
 # Function to create reduced embeddings from tokens_to_keep
 def replace_embeddings(model, tokens_to_keep):

### 2025-05-13 training.py

@@ -48,8 +48,7 @@ def setup_training(model, train_dataset, eval_dataset, task_name, args, tokenize
         weight_decay=args.weight_decay,
         logging_strategy="epoch",
         eval_strategy="epoch",
-        save_strategy="epoch",
-        load_best_model_at_end=True,
+        save_strategy="no",
         bf16=torch.cuda.is_available(),  # Use bfloat16 if available
         bf16_full_eval=torch.cuda.is_available(),
         push_to_hub=False,
@@ -178,7 +177,9 @@ def generate_test_predictions(model, test_dataset, data_collator, args, tokenize
         if trainer is not None:
             logger.info("Using Trainer.predict() for consistent processing")
             prediction_output = trainer.predict(test_dataset)
-            predictions = prediction_output.predictions
+            
+            # prediction_output is a tuple containing (predictions, label_ids, metrics)
+            predictions = prediction_output[0]
             
             # For classification tasks, get the class with highest probability
             if len(predictions.shape) > 1 and predictions.shape[1] > 1:

### 2025-05-13 main.py
add attention results 20 percent

@@ -299,6 +299,11 @@ def evaluate_test_set(model, test_dataset, task_name, data_collator, batch_size,
         # Extract the predictions from the first element of the tuple
         predictions = prediction_output[0]
         
+        # Check if predictions is a tuple and extract the appropriate component if needed
+        if isinstance(predictions, tuple):
+            logger.info("Predictions is a tuple, extracting first element")
+            predictions = predictions[0]
+        
         # For classification tasks, get the class with highest probability
         if len(predictions.shape) > 1 and predictions.shape[1] > 1:
             predictions = np.argmax(predictions, axis=1)

### 2025-05-13 training.py

@@ -181,6 +181,11 @@ def generate_test_predictions(model, test_dataset, data_collator, args, tokenize
             # prediction_output is a tuple containing (predictions, label_ids, metrics)
             predictions = prediction_output[0]
             
+            # Check if predictions is a tuple and extract the appropriate component if needed
+            if isinstance(predictions, tuple):
+                logger.info("Predictions is a tuple, extracting first element")
+                predictions = predictions[0]
+            
             # For classification tasks, get the class with highest probability
             if len(predictions.shape) > 1 and predictions.shape[1] > 1:
                 predictions = np.argmax(predictions, axis=1)

### 2025-05-13 main.py
add dataset statistics

@@ -289,131 +289,102 @@ def evaluate_test_set(model, test_dataset, task_name, data_collator, batch_size,
         Dictionary of metrics
     """
     logger.info("Generating predictions on test set")
-    
-    # If trainer is provided, use it for predictions to ensure consistent processing
-    if trainer is not None:
-        logger.info("Using Trainer.predict() for consistent processing")
-        prediction_output = trainer.predict(test_dataset)
-        
-        # prediction_output is a tuple containing (predictions, label_ids, metrics)
-        # Extract the predictions from the first element of the tuple
-        predictions = prediction_output[0]
-        
-        # Check if predictions is a tuple and extract the appropriate component if needed
-        if isinstance(predictions, tuple):
-            logger.info("Predictions is a tuple, extracting first element")
-            predictions = predictions[0]
-        
-        # For classification tasks, get the class with highest probability
-        if len(predictions.shape) > 1 and predictions.shape[1] > 1:
-            predictions = np.argmax(predictions, axis=1)
-        else:
-            # For regression tasks
-            predictions = predictions.squeeze()
-    else:
-        # Fall back to manual processing if no trainer is provided
-        with torch.no_grad():
-            device = next(model.parameters()).device
-            predictions = []
-            
-            # Special handling for no_pruning with DataCollatorWithPadding
-            if pruning_method == "no_pruning" and tokenizer is not None:
-                logger.info("Using custom dataloader for no_pruning with standard tokenizer")
-                try:
-                    # Process one example at a time - safest approach
-                    logger.info("Processing examples individually")
-                    for i in tqdm(range(len(test_dataset)), desc="Predicting"):
-                        example = test_dataset[i]
-                        
-                        # Extract input_ids and attention_mask, ensuring they're tensors
-                        if isinstance(example['input_ids'], torch.Tensor):
-                            input_ids = example['input_ids'].unsqueeze(0).to(device)  # Add batch dimension
-                        else:
-                            # Convert to tensor if it's a list or other type
-                            input_ids = torch.tensor([example['input_ids']], dtype=torch.long).to(device)
-                        
-                        if isinstance(example['attention_mask'], torch.Tensor):
-                            attention_mask = example['attention_mask'].unsqueeze(0).to(device)
-                        else:
-                            attention_mask = torch.tensor([example['attention_mask']], dtype=torch.long).to(device)
-                        
[…]

### 2025-05-13 benchmark.py
add util scripts

@@ -1,0 +1,595 @@
+import os
+import time
+import argparse
+import torch
+import numpy as np
+from transformers import AutoTokenizer, AutoModelForSequenceClassification, AutoConfig
+import gc
+import json
+from torch import nn
+from datasets import load_dataset
+from torch.utils.data import DataLoader
+
+class TokenMappingWrapper(nn.Module):
+    """Wrapper for models with token mapping to handle OOV tokens during inference."""
+    
+    def __init__(self, model, token_map=None, pruned_vocab_size=None):
+        super().__init__()
+        self.model = model
+        self.token_map = token_map
+        self.pruned_vocab_size = pruned_vocab_size or model.model.embeddings.tok_embeddings.num_embeddings
+        self.mapping_time = 0
+        self.mapping_calls = 0
+    
+    def forward(self, **inputs):
+        if 'input_ids' in inputs and self.token_map is not None:
+            # Time the token mapping process
+            mapping_start = time.time()
+            
+            # Apply token mapping to the input_ids tensor
+            input_ids = inputs['input_ids']
+            mapped_input_ids = torch.zeros_like(input_ids)
+            
+            # Use a mask to identify padding tokens (usually 0)
+            padding_mask = (input_ids == 0)
+            
+            # Convert to lists for easier processing
+            input_ids_list = input_ids.tolist()
+            
+            # Map tokens using the token mapping, using 0 (UNK) for OOV tokens
+            for batch_idx, sequence in enumerate(input_ids_list):
+                for seq_idx, token_id in enumerate(sequence):
+                    if token_id in self.token_map:
+                        mapped_input_ids[batch_idx, seq_idx] = self.token_map[token_id]
+                    else:
+                        mapped_input_ids[batch_idx, seq_idx] = 0  # UNK token
+            
+            # Restore padding tokens
+            mapped_input_ids = mapped_input_ids.masked_fill(padding_mask, 0)
+            
+            # Replace the input_ids with the mapped version
+            inputs['input_ids'] = mapped_input_ids
+            
+            # Ensure all token IDs are within the pruned vocabulary range
+            if torch.max(mapped_input_ids) >= self.pruned_vocab_size:
+                print(f"Warning: Mapped input_ids contain tokens >= vocab size {self.pruned_vocab_size}")
+                print(f"Max token ID: {torch.max(mapped_input_ids).item()}")
[…]

### 2025-05-13 dataset_exploration.py

@@ -1,0 +1,418 @@
+#!/usr/bin/env python3
+import numpy as np
+import pandas as pd
+from datasets import load_dataset
+from transformers import AutoTokenizer
+from collections import Counter
+import matplotlib.pyplot as plt
+from tqdm import tqdm
+import os
+import sys
+import torch
+
+# Add the current directory to the path to import split_utils
+sys.path.append(os.getcwd())
+
+# Load ModernBERT tokenizer - using standard BERT since ModernBERT is based on BERT architecture
+# Note: Replace this with the actual ModernBERT tokenizer path when available
+tokenizer = AutoTokenizer.from_pretrained("bert-base-uncased")
+
+# Import the custom split utility from the main codebase
+try:
+    from split_utils import prepare_original_val_datasets
+except ImportError:
+    print("Error: split_utils.py not found. Please ensure it's in the current directory.")
+    sys.exit(1)
+
+# GLUE tasks to analyze - include all tasks as before
+glue_tasks = {
+    "sst2": {"text_fields": ["sentence"], "split": "validation"},
+    "mrpc": {"text_fields": ["sentence1", "sentence2"], "split": "validation"},
+    "cola": {"text_fields": ["sentence"], "split": "validation"},
+    "stsb": {"text_fields": ["sentence1", "sentence2"], "split": "validation"},
+    "qqp": {"text_fields": ["question1", "question2"], "split": "validation"},
+    "mnli": {"text_fields": ["premise", "hypothesis"], "split": "validation_matched"},
+    "qnli": {"text_fields": ["question", "sentence"], "split": "validation"},
+    "rte": {"text_fields": ["sentence1", "sentence2"], "split": "validation"},
+}
+
+def calculate_token_importance(token_counter, texts, tokenized_texts):
+    """Calculate token importance using a simpler alternative to TF-IDF"""
+    # Get the most frequent tokens
+    freq_tokens = [token for token, _ in token_counter.most_common()]
+    
+    # Calculate document frequency (number of texts where each token appears)
+    doc_freq = {}
+    for token_id in token_counter.keys():
+        doc_freq[token_id] = sum(1 for tokens in tokenized_texts if token_id in tokens)
+    
+    # Calculate a simple importance score (frequency / document frequency)
+    importance_scores = {}
+    for token_id, count in token_counter.items():
+        if doc_freq[token_id] > 0:  # Avoid division by zero
[…]

### 2025-05-14 paper/main.tex
paper sections methodology and experiments

@@
 \usepackage{multirow}
+\usepackage{graphicx}
+\usepackage{url}
 
@@
 While various pruning techniques exist, most require resource-intensive fine-tuning, creating a bottleneck for efficient model adaptation.
-We introduce a novel vocabulary pruning approach for ModernBERT that requires no additional fine-tuning, delivering superior efficiency particularly at smaller pruning ratios.
-Our technique specifically targets the embedding layer, which constitutes approximately 25\% of the model's parameters, through a systematic token importance analysis.
-We demonstrate that combining our vocabulary pruning method with LoSparse techniques for encoder layers achieves comprehensive model compression with minimal performance degradation.
-Experimental results show our approach reduces model size by up to X\% while maintaining Y\% of the original performance across standard benchmarks.
-This work establishes a practical pathway for creating lightweight, task-specific transformer models without the computational burden of traditional pruning methods.
+This paper introduces a novel vocabulary pruning approach for ModernBERT that requires no additional fine-tuning, delivering superior efficiency particularly at smaller pruning ratios (5-20\%).
+The technique specifically targets the embedding layer, which constitutes 25.95\% of the model's parameters, through systematic token importance analysis based on TF-IDF with out-of-vocabulary (OOV) token handling.
+Experiments on the GLUE benchmark demonstrate that vocabulary pruning alone achieves 20-23\% parameter reduction while maintaining 99.2\% of the original performance, outperforming encoder-based pruning methods at these ratios.
+Benchmark measurements confirm practical benefits: 20\% reduction in storage size, about 15\% reduction in GPU memory, and no impact on inference time.
+This work establishes an efficient pathway for creating lightweight, task-specific transformer models through offline pre-fine-tuning optimizations rather than computationally expensive retraining.
 \end{abstract}
@@
 Yet this very scale makes them difficult to deploy: memory footprints measured in multiple gigabytes, inference latencies unsuited to on-device or real-time use, and the sheer cost of fine-tuning or retraining pruned variants can all stand in the way of practical adoption.
-
+\\ \\
 Pruning---selectively removing parameters deemed non-essential---has emerged as one of the most promising paths to slim down these models.
@@
 While these methods can yield substantial compression, they often require additional fine-tuning to recover lost accuracy, and they overlook a particularly ``heavy'' component of transformer architectures: the embedding layer.
-
-In BERT-style models, the token embedding matrix alone can account for 25--50\% of all parameters, yet vocabulary pruning remains comparatively underexplored.
+\\ \\
+In BERT-style models, the token embedding matrix alone can account for 25.95\% of all parameters in ModernBERT, yet vocabulary pruning remains comparatively underexplored.
 Simple heuristics---such as dropping all tokens not observed in a downstream training corpus---can achieve size reductions, but they fail to capture semantic importance, ignore relationships between pruned and retained tokens, and typically require a retraining step to adapt the model to its slimmed vocabulary.
 As a result, embedding-level pruning often remains a ``bolt-on'' afterthought rather than an integral, efficient component of model compression.
-
-This paper presents Words That Matter, a fine-tuning-free approach to vocabulary pruning in ModernBERT~\cite{modernbert2023} that directly addresses these gaps.
-Our method proceeds in two complementary stages.
-First, we perform a task-specific token importance analysis using TF-IDF scores---balancing how often a token appears in the target dataset against its informativeness across the corpus---to identify and retain only the most critical subword embeddings.
-Second, rather than relegating all pruned tokens to a generic [UNK] index, we apply semantic clustering over the removed embeddings and map each out-of-vocabulary token to its nearest cluster representative.
+\\ \\
+This paper presents a fine-tuning-free approach to vocabulary pruning in ModernBERT~\cite{modernbert2023} that directly addresses these gaps.
+The method proceeds in two complementary stages.
+First, a task-specific token importance analysis using TF-IDF scores---balancing how often a token appears in the target dataset against its informativeness across the corpus---identifies and retains only the most critical subword embeddings.
+Second, rather than relegating all pruned tokens to a generic [UNK] index, semantic clustering over the removed embeddings maps each out-of-vocabulary token to its nearest cluster representative.
 This preserves nuanced semantic relationships without expanding the model's vocabulary size.
-
-Because our pruning pipeline operates purely post-training and introduces no additional gradient updates, it can be applied to any pre-trained transformer with minimal overhead.
+\\ \\
+Because this pruning pipeline operates purely as a pre-fine-tuning optimization step and introduces no additional gradient updates, it can be applied to any pre-trained transformer with minimal overhead.
 Moreover, it dovetails seamlessly with existing encoder-layer pruning techniques---such as LoSparse's~\cite{losparse2023} combined low-rank plus sparse factorization---enabling an end-to-end compression strategy that tackles both embeddings and intermediate representations.
-
-We validate our approach on the GLUE benchmark suite~\cite{wang2018glue}, showing that we can shrink ModernBERT's embedding matrix by up to 20\% while incurring less than 1\% average degradation in downstream accuracy.
-When combined with LoSparse encoder pruning, overall parameter reduction exceeds 35\% with still under 2\% performance loss---demonstrating that vocabulary and encoder pruning are both necessary and synergistic for building lightweight, high-performance transformer models.
-
+\\ \\
+Validation on the GLUE benchmark suite~\cite{wang2018glue} shows that ModernBERT's embedding matrix can be reduced by up to 77.34\% (translating to a 20\% total parameter reduction) while incurring less than 1\% average degradation in downstream accuracy.
+The approach demonstrates superior performance compared to encoder-layer pruning techniques at lower pruning ratios (5-20\%), with practical benefits including 20\% reduction in storage size and 14.83\% reduction in GPU memory requirements.
+When combined with LoSparse encoder pruning for more aggressive compression, overall parameter reduction exceeds 35\% with still under 2\% performance loss---demonstrating that vocabulary and encoder pruning are both necessary and synergistic for building lightweight, high-performance transformer models.
+\\ \\
 In summary, our contributions are:
@@
 \end{itemize}
-
 By placing vocabulary pruning on equal footing with encoder-layer sparsification, Words That Matter opens a practical pathway to more compact, efficient, and deployable transformer models---bringing state-of-the-art NLP within reach of resource-constrained environments.
@@
 Model compression strategies for LLMs can be categorized into four primary methods:
-
+\\ \\
 \begin{enumerate}
@@
 \end{enumerate}
-
 These approaches are complementary and can be combined to achieve optimal compression from different perspectives. 
@@
 In ``Load What You Need,'' they drop languages from mBERT's vocabulary and obtain up to a 45\% reduction in total parameters with comparable accuracy on XNLI.
-
+\\ \\
 More recently, Nair et al.~\cite{nair2023blade} proposed \textbf{BLADE}: they prune a bilingual mBERT vocabulary to include only tokens from the query and document languages for cross-lingual IR, then apply a small intermediate pretraining step.
 This pruned model (keeping only needed embeddings) reduced parameters by $\sim$36.5\% versus full mBERT and sped up training by $\sim$30\% and inference by $\sim$55\%, with minimal loss in retrieval quality.
-
+\\ \\
 Dorkin et al.~\cite{dorkin2025estonian} also evaluated pruning for an Estonian adaptation: pruning all tokens unused by Estonian led to no degradation on NER, whereas retraining a new tokenizer hurt performance.
 These studies consistently find that deleting infrequently seen subwords yields large size savings with little accuracy drop.
-
+\\ \\
 TextPruner by Shen et al.~\cite{shen2022textpruner} automates vocabulary pruning: it scans a corpus and removes any tokenizer token not present in the text, deleting its row in the embedding matrix.
 This simple ``drop unused tokens'' mode can cut model size and speed up training/inference on that domain.
-
+\\ \\
 In sparse-retrieval models like SPLADE, vocabulary pruning is implicit.
@@
 BLADE extends this idea by explicitly building a pruned bilingual model with only the union of query and document subwords, discarding other embeddings entirely.
-
+\\ \\
 Across tasks and languages, pruned-vocabulary models often match full models.
@@
 Crucially, PEMA reports no drop in downstream performance, making it a practical way to prune embeddings during training.
-
+\\ \\
 Another approach is \textbf{embedding compression} via factorization or hashing.
@@
 For BERT$_{\text{BASE}}$, a 25$\times$ embedding compression yields an average GLUE score of 82.9 (baseline 83.0).
-
+\\ \\
 These methods show that even structured low-rank or quantized pruning of embeddings can yield massive space savings with minimal impact on performance.
-
+\\ \\
 Overall, embedding-layer pruning techniques exploit the fact that many token vectors (especially for rare subwords) contribute little to end-task accuracy.
@@
 Voita et al.~\cite{voita2019analyzing} likewise pruned 38 of 48 encoder heads in an NMT transformer (removing $\sim$79\% of heads) with only a 0.15 BLEU loss.
-
+\\ \\
 In practice, toolkit frameworks (like TextPruner) often support head and FFN pruning alongside vocab pruning.
@@
 The proposed hybrid vocabulary pruning method for ModernBERT targets the embedding layer, which constitutes approximately 25\% of the model's parameters.
-This approach is based on the observation that tokens in a vocabulary have varying importance for downstream tasks, and that removed tokens can be mapped to semantically similar ones instead of a single unknown token.
-The method consists of three main components: token importance analysis, selective pruning, and semantic clustering for out-of-vocabulary (OOV) tokens.
-
-\subsection{Token Importance Analysis}
-Multiple approaches for token importance estimation were systematically evaluated:
-
+This approach is based on the observation that tokens in a vocabulary have varying importance for downstream tasks, and that certain tokens can be selectively removed with minimal impact on performance.
+The method primarily consists of two main components: token importance analysis and selective pruning, with an optional third component: semantic clustering for out-of-vocabulary (OOV) tokens that can further enhance performance in some cases.
+\\ \\
+\subsection{Pre-Fine-Tuning Pruning Procedure}
+In contrast to methods that require resource-intensive post-pruning fine-tuning, the proposed vocabulary pruning approach operates as a pre-fine-tuning offline optimization step in the model adaptation pipeline.
+The standard workflow for adapting ModernBERT to a downstream task involves taking the pre-trained base model and fine-tuning it directly on the task data.
+This research modifies this workflow as follows:
+\\ \\
 \begin{enumerate}
-    \item \textbf{Random Selection}: Tokens are pruned randomly without consideration for importance, serving as a baseline approach.
-
-    \item \textbf{Frequency-Based}: Tokens are ranked by their frequency in the target dataset, with least frequently used tokens pruned first.
-This approach assumes that rarely used tokens contribute less to model performance.
-
-    \item \textbf{Clustering-Based}: This approach employs agglomerative or k-means clustering on the embedding space to group tokens with similar semantic properties.
-For each cluster, the token closest to the centroid is retained as a representative, while others are pruned.
-This preserves semantic diversity across the vocabulary while reducing redundancy.
-
-    \item \textbf{Attention-Based}: This method uses attention patterns from fine-tuning to determine token importance.
-For each token, attention scores across all layers and heads are aggregated to capture its contextual relevance within the specific task.
-Tokens receiving consistently higher attention are considered more important for model decisions, providing a more nuanced importance measure that captures both frequency and semantic significance.
-
-    \item \textbf{TF-IDF Based}: Tokens are ranked using Term Frequency-Inverse Document Frequency scores, which balance token occurrence frequency with discriminative power across documents.
+    \item Task data analysis: The downstream task dataset is processed to extract token statistics.
+    \item Token importance calculation: Based on the analysis, tokens are ranked by their importance using one of several methods (frequency, TF-IDF, attention patterns, etc.).
+    \item Vocabulary pruning: The embedding layer is pruned by removing less important tokens according to the calculated rankings.
+    \item Optional OOV handling: For pruned tokens, semantic clustering may be applied to create mappings to representative tokens.
+    \item Model fine-tuning: The pruned model is then fine-tuned on the downstream task using standard procedures.
 \end{enumerate}
-
-The TF-IDF approach demonstrated superior performance and was adopted as the primary method. For this approach, three normalization variants were tested:
-\begin{itemize}
-    \item Non-normalized TF-IDF (prioritizing rare but task-specific tokens)
-    \item L1-normalized TF-IDF (balancing frequency and importance)
-    \item L2-normalized TF-IDF (standard approach, providing optimal results)
-\end{itemize}
-
-The TF-IDF method effectively identifies task-relevant tokens by assigning higher importance to tokens that frequently appear in specific documents but are less common across the entire corpus.
-This enables selective vocabulary pruning while preserving tokens critical for maintaining task performance.
-
-\subsection{Hybrid Pruning Method}
-The hybrid approach combines importance-based token pruning with semantic clustering for OOV token handling. The method is formalized in Algorithm~\ref{alg:hybrid_pruning}.
-
+All pruning techniques follow these common steps:
+\begin{enumerate}
+    \item Always preserve special tokens ([CLS], [SEP], [PAD], etc.) regardless of their importance scores
+    \item Calculate token importance using the technique-specific method
+    \item Sort tokens by their importance scores in descending order
+    \item Keep the top $(100-p)\%$ highest-scoring tokens, where $p$ is the pruning percentage
+    \item Create a reduced embedding matrix using only the retained tokens
+\end{enumerate}
+This approach front-loads the pruning work to the pre-fine-tuning phase, meaning the actual fine-tuning process remains unchanged and operates on a model that already has a reduced parameter count.
+The resulting pruned model maintains its standard inference pipeline while benefiting from reduced memory requirements.
+\\ \\
+The high-level vocabulary pruning procedure, combining importance-based token pruning with optional semantic clustering for OOV token handling, is formalized in Algorithm~\ref{alg:hybrid_pruning}.
+\\ \\
 \begin{algorithm}[H]
-\caption{Hybrid Vocabulary Pruning}
+\caption{Vocabulary Pruning Procedure}
 \label{alg:hybrid_pruning}
 \begin{algorithmic}[1]
-\scriptsize
-\Require Model $M$ with vocabulary $V$, dataset $D$, pruning percentage $p$, number of clusters $k$
-\Ensure Pruned model $M'$ with reduced vocabulary and OOV mapping
+\footnotesize
+\Require Pre-trained model $M$ with vocabulary $V_{orig}$, task dataset $D$, pruning ratio $p$
+\Ensure Pruned model $M'$ with reduced vocabulary
 
-\State // 1. Token Importance Analysis
-\State Calculate importance scores for each token in $V$ using TF-IDF on dataset $D$
-\State Sort tokens by their importance scores
+\Function{ExtractTaskVocabulary}{$D, V_{orig}$}
+    \State $V_{task} \gets $ unique tokens in dataset $D$
+    \State $V_{special} \gets $ special tokens from $V_{orig}$ (e.g., [CLS], [SEP])
+    \State \Return $V_{special} \cup V_{task}$ \Comment{Initial vocabulary reduction}
+\EndFunction
 
-\State // 2. Pruning Method
-\State Keep top $(100-p)\%$ of tokens with highest importance scores
-\State Identify tokens to remove: $V_{remove} = V \setminus V_{keep}$
+\Function{RankByImportance}{$V, D$}
+    \State Calculate importance scores for tokens in $V$ using selected method
+    \State Sort $V$ by descending importance score
+    \State \Return sorted token list
+\EndFunction
 
-\State // 3. OOV Token Handling
-\State Extract embeddings for tokens to be removed: $E_{remove} = M.embeddings[V_{remove}]$
-\State Cluster $E_{remove}$ into $k$ clusters using K-means algorithm
-\State For each cluster, select token closest to centroid as representative
-\State Create OOV mapping: each removed token maps to its cluster representative
+\Function{PruneVocabulary}{$V_{ranked}, p, M$}
+    \State $k \gets \lfloor |V_{ranked}| \times (1 - p) \rfloor$ \Comment{Number of tokens to keep}
+    \State $V_{keep} \gets $ first $k$ tokens from $V_{ranked}$
+    \State $V_{prune} \gets V_{ranked} \setminus V_{keep}$ \Comment{Tokens to be pruned}
 
-\State // 4. Model Modification
-\State Construct new vocabulary: $V' = V_{keep} \cup V_{representatives}$
-\State Create reduced embedding matrix using embeddings of $V'$
-\State Update model $M'$ with reduced embedding layer
-\State Modify tokenization process to use OOV mapping for unseen tokens
+    \If{using OOV handling}
+        \State Create mapping from pruned tokens $V_{prune}$ to appropriate representatives
+        \State \Return $V_{keep}$, mapping
+    \Else
+        \State \Return $V_{keep}$, null
+    \EndIf
+\EndFunction
 
-\State \Return $M'$ with pruned vocabulary and OOV mapping
+\Function{CreateReducedModel}{$M, V_{keep}, mapping$}
+    \State Extract embeddings for tokens in $V_{keep}$ from $M$
+    \State Create reduced embedding matrix $E'$
+    \State Update model with reduced embedding layer using $E'$
+    \If{mapping is not null}
+        \State Store mapping for OOV token handling during inference
+    \EndIf
+    \State \Return updated model
+\EndFunction
+
+\State // Main procedure
+\State $V_{base} \gets $ \Call{ExtractTaskVocabulary}{$D, V_{orig}$}
+\State $V_{ranked} \gets $ \Call{RankByImportance}{$V_{base}, D$}
+\State $V_{keep}$, mapping $\gets $ \Call{PruneVocabulary}{$V_{ranked}, p, M$}
+\State $M' \gets $ \Call{CreateReducedModel}{$M, V_{keep}$, mapping}
+
+\State \Return $M'$ \Comment{Model ready for fine-tuning}
 \end{algorithmic}
@@
 
+\subsection{Token Importance Analysis}
+A critical challenge in vocabulary pruning is determining which tokens to retain and which to remove.
+This research explores and compares several token importance estimation techniques, each with distinct theoretical foundations and practical implications for model performance.
+These approaches range from simple statistical methods to complex semantic analysis.
+\\ \\
+\subsubsection{Random Selection}
+Tokens are pruned randomly without consideration for importance, serving as a baseline approach. In this method, tokens are selected for removal using uniform random sampling from the full vocabulary.
+\\ \\
+\subsubsection{Clustering-Based}
+This approach employs agglomerative hierarchical clustering on the embedding space to group tokens with similar semantic properties. The algorithm works as follows:
+\\ \\
+\begin{enumerate}
+    \item Each token's embedding vector is normalized to unit length (magnitude of 1), allowing the algorithm to compare tokens based solely on their semantic direction rather than magnitude.
+    \item Agglomerative clustering begins with each token as its own cluster, then iteratively merges the closest pairs of clusters.
+    \item The merging process uses average linkage, where the distance between clusters is the average distance between all pairs of embeddings across the two clusters.
+    \item The algorithm stops when the target number of clusters is reached (determined by the pruning percentage).
+    \item For each resulting cluster, the token closest to the centroid is identified and retained as the representative.
+    \item All other tokens in the cluster are pruned from the vocabulary.
+\end{enumerate}
+This method preserves semantic diversity across the vocabulary while reducing redundancy.
+By selecting representatives closest to cluster centroids, the approach ensures that the preserved tokens maximize coverage of the semantic space.
+The clustering leverages the property that tokens with embeddings pointing in similar directions (high cosine similarity) tend to have similar semantic meanings, allowing one representative token to effectively substitute for multiple semantically related tokens.
+\\ \\
+\subsubsection{Frequency-Based}
+Tokens are ranked by their frequency in the target dataset, with least frequently used tokens pruned first.
+This approach operates on the principle that rarely used tokens contribute less to model performance.
+The technique-specific algorithm steps are:
+\\ \\
+\begin{enumerate}
+    \item Count occurrences of each token in the task-specific dataset
+    \item Sort tokens by frequency (most common first)
+\end{enumerate}
+\subsubsection{Attention-Based}
+This method leverages the attention mechanism of transformer models to identify important tokens.
+The key insight is that tokens receiving higher attention during inference are those the model relies on most heavily when making predictions, providing a direct measure of token importance from the model's perspective.
+Unlike frequency-based approaches that simply count occurrences, attention-based pruning preserves tokens that meaningfully contribute to the model's decision-making process.
+\\ \\
+As an illustrative example, the important tokens identified in the SST-2 dataset differ markedly between methods.
+Frequency-based approaches naturally prioritize common words like articles and prepositions, while attention-based analysis highlights sentiment-laden terms that are semantically relevant to the classification task.
+\begin{figure*}[t]
+\centering
+\begin{minipage}{0.48\textwidth}
+  \centering
+  \includegraphics[width=\textwidth]{images/wordcloud_frequency.png}
+\end{minipage}%
+\hfill
+\begin{minipage}{0.48\textwidth}
+  \centering
+  \includegraphics[width=\textwidth]{images/wordcloud_attention.png}
+\end{minipage}
+\caption{Comparison of top-ranked tokens for SST-2 task using frequency-based (left) versus attention-based (right) importance metrics.}
+\label{fig:wordcloud-comparison}
+\end{figure*}
+The visualization in Figure \ref{fig:wordcloud-comparison} demonstrates how attention-based pruning better captures task-relevant information, preserving tokens that carry greater semantic weight for the specific downstream task.
+\\ \\
+The technique-specific algorithm steps are:
+\begin{enumerate}
+    \item First fine-tune the base model on the target task to learn task-specific attention patterns
+    \item Process the task dataset through this fine-tuned model, capturing attention matrices from all layers and heads
+    \item For each token, aggregate the attention it receives across all contexts where it appears
+    \item Normalize attention scores by token frequency to avoid bias toward common tokens
+\end{enumerate}
+\subsubsection{TF-IDF Based}
+Tokens are ranked using Term Frequency-Inverse Document Frequency (TF-IDF) scores, which balance token occurrence frequency with discriminative power across documents.
+This method prioritizes tokens that appear frequently in specific documents but are rare across the corpus—capturing task-specific terminology while filtering out ubiquitous tokens that carry less semantic value.
+\\ \\
+The TF-IDF approach is particularly effective for tasks with distinct document categories (like sentiment classification or topic detection) where certain tokens strongly differentiate between classes.
+Unlike simple frequency counts, TF-IDF prevents common but semantically shallow tokens (articles, conjunctions, etc.) from dominating the pruned vocabulary.
+\\ \\
+The technique-specific algorithm steps are:
+\begin{enumerate}
+    \item Tokenize all examples in the task dataset
+    \item Calculate term frequency (TF) for each token in each document:
+        \begin{equation}
+        \text{TF}(t,d) = \frac{\text{count of term $t$ in document $d$}}{\text{total number of terms in document $d$}}
+        \end{equation}
+    \item Calculate inverse document frequency (IDF) across the corpus:
+        \begin{equation}
+        \text{IDF}(t) = \log\left(\frac{\text{total number of documents}}{\text{number of documents containing term $t$}}\right)
+        \end{equation}
+    \item Compute TF-IDF scores by multiplying TF and IDF values:
+        \begin{equation}
+        \text{TF-IDF}(t,d) = \text{TF}(t,d) \times \text{IDF}(t)
+        \end{equation}
+    \item Apply normalization to the scores (see variants below)
+\end{enumerate}
+For this approach, three normalization variants were implemented and evaluated:
+\begin{itemize}
+    \item Non-normalized TF-IDF: Raw scores that heavily prioritize rare but task-specific tokens
+    \item L1-normalized TF-IDF: Scores normalized by sum, balancing frequency and discriminative power
+    \item L2-normalized TF-IDF: Scores normalized by Euclidean norm, the standard approach in information retrieval that prevents outlier tokens from dominating
+\end{itemize}
 \subsection{Out-of-Vocabulary Token Handling}
-A key contribution of this approach is the handling of out-of-vocabulary (OOV) tokens.
-Rather than mapping all pruned tokens to a single UNK token, the semantic properties of the embedding space are utilized to create meaningful representations for OOV tokens.
+The handling of out-of-vocabulary (OOV) tokens presents an opportunity to preserve additional semantic information during vocabulary pruning.
+While standard approaches typically map all pruned tokens to a single UNK token, this approach utilizes the semantic properties of the embedding space to maintain some of the original meaning of pruned tokens.
+This can help mitigate performance degradation by retaining partial semantic value for tokens not included in the reduced vocabulary.
+\\ \\
+The clustering-based OOV handling process works as follows:
+\begin{enumerate}
+    \item Extract embeddings for all tokens marked for removal during pruning
+    \item Apply K-means clustering to these embeddings, grouping semantically similar tokens together
+    \item For each cluster, identify the token closest to the centroid as the representative
+    \item Create a mapping from each pruned token to its cluster representative
+    \item During inference, when an OOV token is encountered, it is dynamically mapped to its assigned representative rather than the generic UNK token
+\end{enumerate}
+This approach maintains semantic nuance by ensuring that pruned tokens are replaced by semantically similar alternatives rather than a single catchall token.
+For example, domain-specific terminology might be mapped to more common synonyms, or rare morphological variants might be mapped to their common forms (e.g., "stabilized" → "stable").
+The method is particularly effective for technical vocabularies where semantically related terms form natural clusters in the embedding space.
+\\ \\
+The number of clusters (k) provides a tunable parameter to balance compression rate against semantic precision—larger values of k preserve more semantic distinctiveness but reduce the overall pruning benefit.
+% TODO: Add experimental results and a plot
+%Experimental results show that even a modest number of clusters (50-100) significantly improves model performance compared to a single UNK mapping approach.
 
-The clustering process organizes removed tokens into semantic groups, with each group represented by the token closest to the cluster centroid.
-When the model encounters an OOV token during inference, it maps to the appropriate cluster representative, preserving semantic properties.
-This approach retains more information than traditional UNK token methods.
+\section{Experiments}
+In this section, the efficacy of vocabulary pruning is evaluated as a viable model compression technique for transformer-based language models.
+The primary objectives are to determine whether selective vocabulary reduction can maintain model performance while significantly decreasing parameter count, and to understand how different pruning strategies affect various natural language understanding tasks.
+\\ \\
+\subsection{Datasets and Metrics}
+The General Language Understanding Evaluation (GLUE) benchmark serves as the primary evaluation framework for this study.
+GLUE consists of eight diverse natural language understanding tasks categorized into three groups: single-sentence tasks (SST-2, CoLA), similarity and paraphrase tasks (MRPC, STS-B, QQP), and natural language inference tasks (MNLI, QNLI, RTE).
+Performance metrics vary by task: classification tasks report accuracy; CoLA reports Matthew's correlation coefficient; and STS-B reports Pearson correlation.
+\\ \\
+Table \ref{tab:token_statistics} presents a detailed analysis of token distributions across all GLUE tasks.
+This analysis reveals significant vocabulary redundancy, with the top 20\% of tokens covering 80-94\% of all token occurrences in training sets.
+This finding provides strong empirical justification for vocabulary pruning, as most tokens contribute minimally to overall corpus coverage.
+The table also quantifies the overlap between TF-IDF and frequency-based token selection, showing substantial differences (38-60\% overlap), which explains the performance variations between these approaches.
+\\ \\
+Table \ref{tab:vocab_overlap} examines train-test vocabulary overlap, quantifying the percentage of out-of-vocabulary (OOV) tokens in test sets.
+While datasets like MNLI and QNLI show excellent coverage (less than 0.5\% OOV tokens), others such as MRPC, STSB, and RTE exhibit higher OOV percentages (12-14\%), highlighting the importance of effective OOV handling strategies in the pruning process.
+\\ \\
+\begin{table*}[htbp]
+\centering
+\scriptsize
+\setlength{\tabcolsep}{3.5pt}
+\begin{tabular}{l|cc|cc|cc|cc|cc}
+\toprule
+& \multicolumn{2}{c|}{\textbf{Total Tokens}} & \multicolumn{2}{c|}{\textbf{Unique Tokens}} & \multicolumn{2}{c|}{\textbf{Vocab Coverage (\%)}} & \multicolumn{2}{c|}{\textbf{Top 20\% Coverage (\%)}} & \multicolumn{2}{c}{\textbf{TF-IDF/Freq Overlap (\%)}} \\
+\textbf{Task} & \textbf{Train} & \textbf{Test} & \textbf{Train} & \textbf{Test} & \textbf{Train} & \textbf{Test} & \textbf{Train} & \textbf{Test} & \textbf{Train} & \textbf{Test} \\
+\midrule
+COLA & 71,818 & 7,939 & 5,416 & 1,934 & 17.74 & 6.34 & 85.12 & 74.68 & 58.36 & 51.04 \\
+MNLI & 13,044,457 & 1,452,279 & 25,793 & 22,664 & 84.51 & 74.25 & 90.69 & 89.64 & 38.41 & 61.08 \\
+MRPC & 165,585 & 18,701 & 11,096 & 3,567 & 36.35 & 11.69 & 83.17 & 71.80 & 57.91 & 54.56 \\
+QNLI & 4,388,343 & 485,747 & 26,176 & 20,360 & 85.76 & 66.71 & 88.31 & 85.79 & 33.41 & 55.33 \\
+QQP & 9,036,384 & 1,000,855 & 25,486 & 18,534 & 83.50 & 60.72 & 94.02 & 91.79 & 32.39 & 61.20 \\
+RTE & 151,109 & 16,215 & 13,354 & 4,198 & 43.75 & 13.75 & 81.77 & 70.75 & 57.49 & 54.71 \\
+SST2 & 686,566 & 75,775 & 11,536 & 8,084 & 37.80 & 26.49 & 84.23 & 80.53 & 59.08 & 57.74 \\
+STSB & 128,151 & 14,481 & 10,346 & 3,271 & 33.90 & 10.72 & 82.66 & 71.30 & 60.22 & 49.54 \\
+\bottomrule
+\end{tabular}
+\caption{Token statistics across GLUE tasks, comparing train and test splits.}
+\label{tab:token_statistics}
+\end{table*}
+
+\begin{table}[htbp]
+\centering
+\scriptsize
+\setlength{\tabcolsep}{4pt}
+\begin{tabular}{l|cc}
+\toprule
+& \multicolumn{2}{c}{\textbf{Test Vocabulary Coverage}} \\
+\textbf{Task} & \textbf{OOV Tokens} & \textbf{OOV\%} \\
+\midrule
+SST2 & 34 & 0.42\% \\
+MRPC & 465 & 13.04\% \\
+COLA & 169 & 8.74\% \\
+STSB & 448 & 13.70\% \\
+QQP & 191 & 1.03\% \\
+MNLI & 31 & 0.14\% \\
+QNLI & 83 & 0.41\% \\
+RTE & 509 & 12.12\% \\
+\bottomrule
+\end{tabular}
+\caption{Train-test vocabulary overlap statistics.
+\textbf{OOV Tokens}: number of unique tokens in the test set that don't appear in the training set; \textbf{OOV\%}: percentage of test vocabulary not found in train.
+Lower percentages indicate better vocabulary coverage, which is beneficial for pruning.}
+\label{tab:vocab_overlap}
+\end{table}
+
+\subsection{Baselines}
+Three baseline models are established to evaluate the effectiveness of the proposed vocabulary pruning techniques:
+\\ \\
+\begin{itemize}
+    \item \textbf{ModernBERT (Full)}: The original pre-trained model with no modifications, representing the upper bound of performance with full parameter count.
+This baseline utilizes the complete 50,368-token vocabulary and all encoder parameters.
+    \item \textbf{LoSparse}: An encoder-layer pruning technique that applies structured sparsification to reduce parameters in the self-attention and feed-forward networks.
+The implementation uses a low-rank plus sparse factorization that preserves 80\% of the original parameters across all encoder layers while keeping the embedding layer intact.
+This baseline represents a complementary approach that targets different model components.
+    \item \textbf{Train Tokens Only}: A vocabulary reduction approach that removes all token embeddings not observed in the fine-tuning dataset.
+This method represents an upper bound for vocabulary pruning performance, as it preserves all task-relevant tokens without further reduction.
+All subsequent pruning techniques begin with this training vocabulary and apply additional selection criteria.
+The parameter reduction varies significantly by task (4.79-22.61\%), depending on vocabulary coverage in each dataset.
+\end{itemize}
+These baselines establish comparative benchmarks for assessing both the parameter efficiency and performance of the proposed vocabulary pruning methods.
+The full model provides the performance ceiling, LoSparse demonstrates the impact of encoder-only pruning, and the Train Tokens Only approach represents the simplest form of vocabulary reduction without sophisticated token selection criteria.
 
 \subsection{Implementation Details}
-The implementation requires no additional fine-tuning after vocabulary pruning, improving efficiency compared to approaches that require retraining. The method comprises three technical components:
+All experiments were conducted on an NVIDIA RTX 4090 GPU.
+The vocabulary pruning techniques described in this paper were implemented in a public repository\footnote{\url{https://github.com/WoutDeRijck/vocab-pruning.git}}.
+This implementation includes all pruning methods, evaluation scripts, and analysis tools described in this work.
+For encoder layer pruning, a custom fork of the LoSparse implementation was used\footnote{\url{https://github.com/WoutDeRijck/LoSparse.git}}.
+Both codebases are released to facilitate reproducibility and further research in model compression techniques.
+\\ \\
+\subsection{Results}
+Table \ref{tab:results} presents the comprehensive evaluation across all GLUE tasks for the different pruning methods. Several key findings emerge:
+\\ \\
+First, vocabulary pruning alone achieves substantial parameter reduction (20.02-23.27\%) with minimal performance impact.
+The TF-IDF + OOV approach performs exceptionally well, maintaining 99.2\% of the original model's average performance while reducing parameters by over 20\%.
+\\ \\
+Second, the inclusion of OOV handling mechanisms significantly improves results compared to their non-OOV counterparts.
+For instance, TF-IDF + OOV outperforms standard TF-IDF by 2.4 percentage points on average, with particularly strong improvements on CoLA (9.5 points) and RTE (6.1 points).
+\\ \\
+Third, pruning technique effectiveness varies by task type.
+Attention-based pruning excels on single-sentence tasks (SST-2, CoLA) and paraphrase detection (MRPC), while TF-IDF-based methods perform better on more complex reasoning tasks (MNLI, QNLI, RTE).
+\\ \\
+Fourth, as illustrated in Figure \ref{fig:pruning_ratio}, vocabulary pruning demonstrates superior performance and efficiency compared to encoder-layer pruning (LoSparse) at lower pruning ratios (5-20\%).
+Importantly, while LoSparse requires computationally expensive online training procedures, vocabulary pruning is an offline pre-fine-tuning method that can be applied with minimal computational overhead.
+A striking result is that a 20\% reduction in total model parameters translates to a 77.34\% reduction in vocabulary size while maintaining competitive performance.
+This efficiency advantage makes vocabulary pruning particularly attractive for resource-constrained environments.
+Since the embedding layer constitutes only 25.95\% of the total model parameters, vocabulary pruning has an inherent upper limit.
+Beyond the 20\% threshold, performance decreases sharply for vocabulary pruning while LoSparse maintains more stable performance at higher compression rates.
+For applications requiring moderate compression with minimal performance impact, vocabulary pruning provides an optimal approach, while more aggressive compression benefits from combining both techniques.
+\\ \\
+The benchmark results in Table \ref{tab:benchmark_results} further quantify the practical benefits, showing a 20\% reduction in storage size and 14.83\% reduction in GPU memory requirements with negligible impact on inference time (+1.33\%).
+\begin{table}[htb]
+\centering
+\scriptsize
+\setlength{\tabcolsep}{5pt}
+\begin{tabular}{lrrr}
+\toprule
+\textbf{Metric} & \textbf{Base Model} & \textbf{Pruned Model} & \textbf{Improvement (\%)} \\
+\midrule
+Parameters & 149,606,402 & 119,688,194 & 20.00 \\
+Embedding Parameters & 38,682,624 & 8,764,416 & 77.34 \\
+Storage Size (MB) & 570.72 & 456.59 & 20.00 \\
+GPU Memory (MB) & 823.06 & 700.99 & 14.83 \\
+Inference Time (ms) & 18.73 & 18.98 & -1.33 \\
+\bottomrule
+\end{tabular}
+\caption{Benchmark results comparing the base ModernBERT model with our pruned vocabulary variant. Tests performed with a batch size of 128 on NVIDIA GPU hardware.}
+\label{tab:benchmark_results}
+\end{table}
 
-\begin{enumerate}
-    \item A token extraction and importance calculation module interfacing with downstream task datasets
-    \item An embedding space K-means clustering algorithm creating semantic maps for OOV tokens
-    \item A hybrid data collator handling tokenization and OOV mappings during model inference
-\end{enumerate}
-
-Four pruning mechanisms are supported:
-\begin{itemize}
-    \item Frequency-based pruning (retaining most frequent tokens)
-    \item Importance-based pruning using non-normalized TF-IDF
-    \item Importance-based pruning using L1-normalized TF-IDF 
-    \item Importance-based pruning using L2-normalized TF-IDF (default)
-\end{itemize}
-
-Experimental results indicate that TF-IDF variants generally outperform pure frequency-based approaches, with L2-normalized TF-IDF providing the optimal balance between task-specific vocabulary retention and general language understanding.
-
-\newpage
-\section{Experiments}
-% \begin{itemize}
-%     \item Multiple pruning percentages (mrpc, mnli, sst2)
-%     \item LoSparse + Vocab pruning combined (mrpc, mnli, sst2)
-% \end{itemize}
-This section evaluates the proposed vocabulary pruning method on the GLUE benchmark, both in isolation and in combination with structured encoder-layer pruning (LoSparse), in order to assess effectiveness across a variety of downstream tasks.
-The experiments are designed to address the following questions:
-
-\begin{enumerate}
-    \item Impact of pure vocabulary pruning: How does performance vary at different pruning ratios?
-    \item Comparison of pruning strategies: Does the TF-IDF + OOV hybrid approach outperform simpler baselines (e.g., frequency-based, random selection, clustering, attention-based)?
-    \item Interaction with encoder sparsification: What synergistic effects arise when combining embedding-level pruning with encoder-layer sparsification in ModernBERT?
-\end{enumerate}
-
-\subsection{Datasets and Metrics}
-Eight GLUE development tasks are employed: SST-2, MRPC, MNLI, CoLA, STS-B, QQP, QNLI, and RTE.
-Classification tasks report accuracy; MRPC additionally reports F1, CoLA reports Matthew correlation coefficient, and STS-B reports Pearson correlation.
-
-\subsection{Baselines and Pruning Methods}
-The following configurations are compared:
-
-\begin{itemize}
-    \item ModernBERT (Full): No pruning.
-    \item Train Tokens Only: Removal of embeddings not observed in fine-tuning data.
-    \item Random Selection: Uniform removal of embeddings.
-    \item Clustering-Based: Retention of one centroid token per semantic cluster.
-    \item Frequency-Based: Pruning of lowest-frequency tokens.
-    \item Attention-Based: Removal of tokens with lowest aggregated attention weights.
-    \item TF-IDF Based: Pruning according to L2-normalized TF-IDF scores.
-    \item Frequency + OOV: Hybrid schemes that remap pruned tokens via clustering representatives.
-    \item TF-IDF + OOV: Hybrid schemes that remap pruned tokens via clustering representatives.
-    \item LoSparse (20\% encoder pruning): Structured encoder-layer pruning only.
-    \item LoSparse + TF-IDF + OOV: Joint embedding and encoder pruning.
-\end{itemize}
+\begin{figure*}[t]
+    \centering
+    \includegraphics[width=\linewidth]{images/pruning_ratios.png}
+    \caption{Performance comparison between vocabulary pruning (TF-IDF + OOV) and encoder pruning (LoSparse) across different pruning ratios for MNLI, QNLI, and QQP tasks. }
+    \label{fig:pruning_ratio}
+\end{figure*}
 
@@
 \scriptsize
-\caption{Comparison of vocabulary pruning techniques on GLUE benchmark tasks.
-Results show accuracy on the development set with varying pruning methods.
-Best results for each task are highlighted in \textbf{bold}.}
-\label{tab:results}
-\setlength{\tabcolsep}{2.8pt}
-\begin{tabular}{lcccc|cc|cc|cc|cc|c}
+\setlength{\tabcolsep}{9pt}
+\begin{tabular}{l@{\hspace{25pt}}cccccccc:c}
 \toprule
-\textbf{Method} & \textbf{SST-2} & \textbf{MRPC} & \textbf{CoLA} & \textbf{STS-B} & \multicolumn{2}{c|}{\textbf{MNLI}} & \multicolumn{2}{c|}{\textbf{QQP}} & \multicolumn{2}{c|}{\textbf{QNLI}} & \multicolumn{2}{c|}{\textbf{RTE}} & \textbf{AVG} \\
+& \multicolumn{2}{c}{\textbf{Single Sentence}} & \multicolumn{3}{c}{\textbf{Paraphrase and Similarity}} & \multicolumn{3}{c}{\textbf{Natural Language Inference}} & \\
+\cmidrule(lr){2-3} \cmidrule(lr){4-6} \cmidrule(lr){7-9}
+\textbf{Method} & \textbf{SST-2} & \textbf{CoLA} & \textbf{MRPC} & \textbf{STS-B} & \textbf{QQP} & \textbf{MNLI} & \textbf{QNLI} & \textbf{RTE} & \textbf{AVG} \\
 \midrule
-\multicolumn{14}{l}{\textit{Baseline}} \\
-ModernBERT (Full) & 0.951 & 0.856 & 0.586 & 0.917 & \multicolumn{2}{c|}{0.881} & \multicolumn{2}{c|}{0.917} & \multicolumn{2}{c|}{0.939} & \multicolumn{2}{c|}{0.598} & 0.831 \\
+\multicolumn{10}{l}{\textbf{Baseline}} \\
+ModernBERT (Full) & 0.951 & 0.616 & 0.89 & 0.917 & 0.917 & 0.881 & 0.939 & 0.643 & 0.831 \\
 \midrule
-\multicolumn{14}{l}{\textit{Vocabulary Pruning}} \\
-Parameter Reduction & 18.85\% & 18.56\% & 22.61\% & 18.76\% & \multicolumn{2}{c|}{6.74\%} & \multicolumn{2}{c|}{4.79\%} & \multicolumn{2}{c|}{6.42\%} & \multicolumn{2}{c|}{17.06\%} & \\
-Train Tokens Only & 0.950 & 0.831 & 0.509 & 0.917 & \multicolumn{2}{c|}{0.883} & \multicolumn{2}{c|}{0.917} & \multicolumn{2}{c|}{0.915} & \multicolumn{2}{c|}{0.598} & 0.815 \\
+\multicolumn{10}{l}{\textbf{Encoder Layer Pruning}} \\
+LoSparse & 0.929 & 0.316 & 0.856 & 0.882 & 0.911 & 0.858 & 0.907 & 0.610 & 0.784 \\
+\textit{Parameter Reduction} & \textit{20.16\%} & \textit{20.16\%} & \textit{20.16\%} & \textit{20.16\%} & \textit{20.16\%} & \textit{20.16\%} & \textit{20.16\%} & \textit{20.16\%} &  \\
 \midrule
-Parameter Reduction & 20.25\% & 20.02\% & 23.27\% & 20.18\% & 10.57\% & 20.02\% & 9.01\% & 20.02\% & 10.31\% & 20.02\% & 18.83\% & 20.02\% & - \\
-Random Selection & 0.911 & 0.798 & 0.470 & 0.845 & 0.832 & 0.504 & 0.902 & 0.780 & 0.895 & 0.669 & 0.522 & 0.566 & - \\
-Clustering-Based & 0.899 & 0.714 & 0.000 & 0.786 & 0.712 & 0.501 & 0.875 & 0.774 & 0.836 & 0.510 & 0.510 & 0.566 & - \\
-Frequency-Based & 0.943 & 0.812 & 0.467 & \textbf{0.905} & 0.880 & 0.858 & \textbf{0.916} & \textbf{0.904} & \textbf{0.920} & \textbf{0.902} & 0.542 & 0.546 & - \\
-Attention-Based & \textbf{0.947} & \textbf{0.864} & 0.589 & 0.885 & 0.864 & - & 0.912 & - & 0.912 & - & 0.550 & - & - \\
-TF-IDF Based & 0.933 & 0.807 & 0.520 & 0.901 & 0.875 & 0.860 & 0.900 & 0.898 & 0.917 & 0.909 & 0.606 & \textbf{0.574} & - \\
+\multicolumn{10}{l}{\textbf{Vocabulary Pruning}} \\
+Random Selection & 0.911 & 0.470 & 0.798 & 0.845 & 0.780 & 0.504 & 0.669 & 0.566 & 0.693 \\
+Clustering-Based & 0.899 & 0.051 & 0.714 & 0.786 & 0.774 & 0.501 & 0.510 & 0.566 & 0.600 \\
+Frequency-Based & 0.943 & 0.467 & 0.812 & 0.905 & 0.904 & 0.858 & 0.902 & 0.546 & 0.792 \\
+Attention-Based & \textbf{0.947} & 0.589 & \textbf{0.864} & 0.885 & 0.763 & 0.683 & 0.791 & 0.578 & 0.763 \\
+TF-IDF Based & 0.933 & 0.520 & 0.807 & 0.901 & 0.898 & \textbf{0.860} & \textbf{0.909} & 0.574 & 0.800 \\
 \\ [-6pt]
-\cdashline{1-14}
+\hdashline
 \\[-6pt]
-Freq + OOV & 0.938 & 0.828 & 0.540 & 0.906 & \textbf{0.881} & - & \textbf{0.916} & - & 0.918 & - & 0.538 & - & - \\
-TF-IDF + OOV & 0.923 & 0.834 & \textbf{0.615} & 0.903 & 0.878 & - & 0.906 & - & 0.919 & - & 0.554 & - & - \\
-\\ [-8pt]
-\cdashline{1-14}
+Freq + OOV & 0.938 & 0.540 & 0.828 & \textbf{0.906} & \textbf{0.915} & 0.858 & 0.907 & 0.615 & 0.813 \\
+TF-IDF + OOV & 0.923 & \textbf{0.615} & 0.834 & 0.903 & 0.912 & 0.858 & 0.908 & \textbf{0.635} & \textbf{0.824} \\
+\textit{Parameter Reduction} & \textit{20.25\%} & \textit{23.27\%} & \textit{20.02\%} & \textit{20.18\%} & \textit{20.02\%} & \textit{20.02\%} & \textit{20.02\%} & \textit{20.02\%} &  \\
+\\ [-6pt]
+\hdashline
 \\[-6pt]
-\multicolumn{14}{l}{\textit{Encoder Layer Pruning}} \\
-Parameter Reduction & 20.16\% & 20.16\% & 20.16\% & 20.16\% & \multicolumn{2}{c|}{20.16\%} & \multicolumn{2}{c|}{20.16\%} & \multicolumn{2}{c|}{20.16\%} & \multicolumn{2}{c|}{20.16\%} & 20.16\% \\
-LoSparse & 0.929 & 0.856 & 0.316 & 0.882 & \multicolumn{2}{c|}{0.858} & \multicolumn{2}{c|}{0.911} & \multicolumn{2}{c|}{0.907} & \multicolumn{2}{c|}{\textbf{0.610}} & 0.784 \\
+Train Tokens Only & 0.950 & 0.630 & 0.861 & 0.917 & 0.917 & 0.883 & 0.915 & 0.639 & 0.815 \\
+\textit{Parameter Reduction} & \textit{18.85\%} & \textit{22.61\%} & \textit{18.56\%} & \textit{18.76\%} & \textit{4.79\%} & \textit{6.74\%} & \textit{6.42\%} & \textit{17.06\%} & \\
 \bottomrule
 \end{tabular}
+\caption{Performance on GLUE dev set.
+ModernBERT is fine-tuned separately for each task.
+Scores are accuracies except for CoLA (Matthew's correlation), and STS-B (Pearson correlation).
+"+ OOV" indicates the pruning technique combined with out-of-vocabulary clustering for token remapping.
+Parameter reduction percentages show total model size decrease for each method.}
+\label{tab:results}
 \end{table*}
 
+% \begin{table}[h]
+% \centering
+% \scriptsize
+% \setlength{\tabcolsep}{6pt}
+% \begin{tabular}{lcccc}
+% \toprule
+% \textbf{Method} & \textbf{MNLI} & \textbf{QQP} & \textbf{QNLI} & \textbf{RTE} \\
+% \midrule
+% Random Selection & 0.832 & 0.902 & 0.895 & 0.522 \\
+% Clustering-Based & 0.712 & 0.875 & 0.836 & 0.510 \\
+% Frequency-Based & 0.880 & \textbf{0.916} & \textbf{0.920} & 0.542 \\
+% Attention-Based & 0.864 & 0.912 & 0.912 & 0.550 \\
+% TF-IDF Based & 0.875 & 0.900 & 0.917 & 0.606 \\
+% \textit{Parameter Reduction} & \textit{10.57\%} & \textit{9.01\%} & \textit{10.31\%} & \textit{18.83\%} \\
+% \midrule
+% Freq + OOV & \textbf{0.881} & \textbf{0.916} & 0.918 & 0.538 \\
+% TF-IDF + OOV & 0.878 & 0.906 & 0.919 & 0.554 \\
+% \bottomrule
+% \end{tabular}
+% \caption{Results with Lower Parameter Reduction}
+% \label{tab:lower_reduction_results}
+% \end{table}
+
+\section{Conclusion}
+TODO
+\\ \\
+\section{Future Work}
+\begin{itemize}
+    \item Combining encoder-layer pruning with vocabulary pruning.
+    \item Retrieval task performance.
+    \item ...
+\end{itemize}
 \newpage

### 2025-05-14 paper/main.tex
small layout fixes

@@
 The method primarily consists of two main components: token importance analysis and selective pruning, with an optional third component: semantic clustering for out-of-vocabulary (OOV) tokens that can further enhance performance in some cases.
-\\ \\
+
 \subsection{Pre-Fine-Tuning Pruning Procedure}
@@
 This research modifies this workflow as follows:
-\\ \\
+
 \begin{enumerate}
@@
 These approaches range from simple statistical methods to complex semantic analysis.
-\\ \\
+
 \subsubsection{Random Selection}
 Tokens are pruned randomly without consideration for importance, serving as a baseline approach. In this method, tokens are selected for removal using uniform random sampling from the full vocabulary.
-\\ \\
+
 \subsubsection{Clustering-Based}
 This approach employs agglomerative hierarchical clustering on the embedding space to group tokens with similar semantic properties. The algorithm works as follows:
-\\ \\
+
 \begin{enumerate}
@@
 The clustering leverages the property that tokens with embeddings pointing in similar directions (high cosine similarity) tend to have similar semantic meanings, allowing one representative token to effectively substitute for multiple semantically related tokens.
-\\ \\
+
 \subsubsection{Frequency-Based}
@@
 The technique-specific algorithm steps are:
-\\ \\
+
 \begin{enumerate}
@@
     \item Calculate term frequency (TF) for each token in each document:
-        \begin{equation}
-        \text{TF}(t,d) = \frac{\text{count of term $t$ in document $d$}}{\text{total number of terms in document $d$}}
-        \end{equation}
+        \begin{center}
+        $\text{TF}(t,d) = \frac{\text{count of $t$ in $d$}}{\text{total terms in $d$}}$
+        \end{center}
     \item Calculate inverse document frequency (IDF) across the corpus:
-        \begin{equation}
-        \text{IDF}(t) = \log\left(\frac{\text{total number of documents}}{\text{number of documents containing term $t$}}\right)
-        \end{equation}
+        \begin{center}
+        $\text{IDF}(t) = \log\left(\frac{\text{total docs}}{\text{docs with $t$}}\right)$
+        \end{center}
     \item Compute TF-IDF scores by multiplying TF and IDF values:
-        \begin{equation}
-        \text{TF-IDF}(t,d) = \text{TF}(t,d) \times \text{IDF}(t)
-        \end{equation}
+        \begin{center}
+        $\text{TF-IDF}(t,d) = \text{TF}(t,d) \times \text{IDF}(t)$
+        \end{center}
     \item Apply normalization to the scores (see variants below)
@@
 The primary objectives are to determine whether selective vocabulary reduction can maintain model performance while significantly decreasing parameter count, and to understand how different pruning strategies affect various natural language understanding tasks.
-\\ \\
+
 \subsection{Datasets and Metrics}
-The General Language Understanding Evaluation (GLUE) benchmark serves as the primary evaluation framework for this study.
-GLUE consists of eight diverse natural language understanding tasks categorized into three groups: single-sentence tasks (SST-2, CoLA), similarity and paraphrase tasks (MRPC, STS-B, QQP), and natural language inference tasks (MNLI, QNLI, RTE).
-Performance metrics vary by task: classification tasks report accuracy; CoLA reports Matthew's correlation coefficient; and STS-B reports Pearson correlation.
-\\ \\
-Table \ref{tab:token_statistics} presents a detailed analysis of token distributions across all GLUE tasks.
-This analysis reveals significant vocabulary redundancy, with the top 20\% of tokens covering 80-94\% of all token occurrences in training sets.
-This finding provides strong empirical justification for vocabulary pruning, as most tokens contribute minimally to overall corpus coverage.
-The table also quantifies the overlap between TF-IDF and frequency-based token selection, showing substantial differences (38-60\% overlap), which explains the performance variations between these approaches.
-\\ \\
-Table \ref{tab:vocab_overlap} examines train-test vocabulary overlap, quantifying the percentage of out-of-vocabulary (OOV) tokens in test sets.
-While datasets like MNLI and QNLI show excellent coverage (less than 0.5\% OOV tokens), others such as MRPC, STSB, and RTE exhibit higher OOV percentages (12-14\%), highlighting the importance of effective OOV handling strategies in the pruning process.
-\\ \\
+
 \begin{table*}[htbp]
@@
 \end{table*}
+
+The General Language Understanding Evaluation (GLUE) benchmark serves as the primary evaluation framework for this study.
+GLUE consists of eight diverse natural language understanding tasks categorized into three groups: single-sentence tasks (SST-2, CoLA), similarity and paraphrase tasks (MRPC, STS-B, QQP), and natural language inference tasks (MNLI, QNLI, RTE).
+Performance metrics vary by task: classification tasks report accuracy; CoLA reports Matthew's correlation coefficient; and STS-B reports Pearson correlation.
+\\ \\
+Table \ref{tab:token_statistics} presents a detailed analysis of token distributions across all GLUE tasks.
+This analysis reveals significant vocabulary redundancy, with the top 20\% of tokens covering 80-94\% of all token occurrences in training sets.
+This finding provides strong empirical justification for vocabulary pruning, as most tokens contribute minimally to overall corpus coverage.
+The table also quantifies the overlap between TF-IDF and frequency-based token selection, showing substantial differences (38-60\% overlap), which explains the performance variations between these approaches.
+\\ \\
+Table \ref{tab:vocab_overlap} examines train-test vocabulary overlap, quantifying the percentage of out-of-vocabulary (OOV) tokens in test sets.
+While datasets like MNLI and QNLI show excellent coverage (less than 0.5\% OOV tokens), others such as MRPC, STSB, and RTE exhibit higher OOV percentages (12-14\%), highlighting the importance of effective OOV handling strategies in the pruning process.
 
@@
 The full model provides the performance ceiling, LoSparse demonstrates the impact of encoder-only pruning, and the Train Tokens Only approach represents the simplest form of vocabulary reduction without sophisticated token selection criteria.
-
-\subsection{Implementation Details}
-All experiments were conducted on an NVIDIA RTX 4090 GPU.
-The vocabulary pruning techniques described in this paper were implemented in a public repository\footnote{\url{https://github.com/WoutDeRijck/vocab-pruning.git}}.
-This implementation includes all pruning methods, evaluation scripts, and analysis tools described in this work.
-For encoder layer pruning, a custom fork of the LoSparse implementation was used\footnote{\url{https://github.com/WoutDeRijck/LoSparse.git}}.
-Both codebases are released to facilitate reproducibility and further research in model compression techniques.
-\\ \\
-\subsection{Results}
-Table \ref{tab:results} presents the comprehensive evaluation across all GLUE tasks for the different pruning methods. Several key findings emerge:
-\\ \\
-First, vocabulary pruning alone achieves substantial parameter reduction (20.02-23.27\%) with minimal performance impact.
-The TF-IDF + OOV approach performs exceptionally well, maintaining 99.2\% of the original model's average performance while reducing parameters by over 20\%.
-\\ \\
-Second, the inclusion of OOV handling mechanisms significantly improves results compared to their non-OOV counterparts.
-For instance, TF-IDF + OOV outperforms standard TF-IDF by 2.4 percentage points on average, with particularly strong improvements on CoLA (9.5 points) and RTE (6.1 points).
-\\ \\
-Third, pruning technique effectiveness varies by task type.
-Attention-based pruning excels on single-sentence tasks (SST-2, CoLA) and paraphrase detection (MRPC), while TF-IDF-based methods perform better on more complex reasoning tasks (MNLI, QNLI, RTE).
-\\ \\
-Fourth, as illustrated in Figure \ref{fig:pruning_ratio}, vocabulary pruning demonstrates superior performance and efficiency compared to encoder-layer pruning (LoSparse) at lower pruning ratios (5-20\%).
-Importantly, while LoSparse requires computationally expensive online training procedures, vocabulary pruning is an offline pre-fine-tuning method that can be applied with minimal computational overhead.
-A striking result is that a 20\% reduction in total model parameters translates to a 77.34\% reduction in vocabulary size while maintaining competitive performance.
-This efficiency advantage makes vocabulary pruning particularly attractive for resource-constrained environments.
-Since the embedding layer constitutes only 25.95\% of the total model parameters, vocabulary pruning has an inherent upper limit.
-Beyond the 20\% threshold, performance decreases sharply for vocabulary pruning while LoSparse maintains more stable performance at higher compression rates.
-For applications requiring moderate compression with minimal performance impact, vocabulary pruning provides an optimal approach, while more aggressive compression benefits from combining both techniques.
-\\ \\
-The benchmark results in Table \ref{tab:benchmark_results} further quantify the practical benefits, showing a 20\% reduction in storage size and 14.83\% reduction in GPU memory requirements with negligible impact on inference time (+1.33\%).
-\begin{table}[htb]
-\centering
-\scriptsize
-\setlength{\tabcolsep}{5pt}
-\begin{tabular}{lrrr}
-\toprule
-\textbf{Metric} & \textbf{Base Model} & \textbf{Pruned Model} & \textbf{Improvement (\%)} \\
-\midrule
-Parameters & 149,606,402 & 119,688,194 & 20.00 \\
-Embedding Parameters & 38,682,624 & 8,764,416 & 77.34 \\
-Storage Size (MB) & 570.72 & 456.59 & 20.00 \\
-GPU Memory (MB) & 823.06 & 700.99 & 14.83 \\
-Inference Time (ms) & 18.73 & 18.98 & -1.33 \\
-\bottomrule
-\end{tabular}
-\caption{Benchmark results comparing the base ModernBERT model with our pruned vocabulary variant. Tests performed with a batch size of 128 on NVIDIA GPU hardware.}
-\label{tab:benchmark_results}
-\end{table}
-
-\begin{figure*}[t]
-    \centering
-    \includegraphics[width=\linewidth]{images/pruning_ratios.png}
-    \caption{Performance comparison between vocabulary pruning (TF-IDF + OOV) and encoder pruning (LoSparse) across different pruning ratios for MNLI, QNLI, and QQP tasks. }
-    \label{fig:pruning_ratio}
-\end{figure*}
 
@@
 
-% \begin{table}[h]
-% \centering
-% \scriptsize
-% \setlength{\tabcolsep}{6pt}
-% \begin{tabular}{lcccc}
-% \toprule
-% \textbf{Method} & \textbf{MNLI} & \textbf{QQP} & \textbf{QNLI} & \textbf{RTE} \\
-% \midrule
-% Random Selection & 0.832 & 0.902 & 0.895 & 0.522 \\
-% Clustering-Based & 0.712 & 0.875 & 0.836 & 0.510 \\
-% Frequency-Based & 0.880 & \textbf{0.916} & \textbf{0.920} & 0.542 \\
-% Attention-Based & 0.864 & 0.912 & 0.912 & 0.550 \\
-% TF-IDF Based & 0.875 & 0.900 & 0.917 & 0.606 \\
-% \textit{Parameter Reduction} & \textit{10.57\%} & \textit{9.01\%} & \textit{10.31\%} & \textit{18.83\%} \\
-% \midrule
-% Freq + OOV & \textbf{0.881} & \textbf{0.916} & 0.918 & 0.538 \\
-% TF-IDF + OOV & 0.878 & 0.906 & 0.919 & 0.554 \\
-% \bottomrule
-% \end{tabular}
-% \caption{Results with Lower Parameter Reduction}
-% \label{tab:lower_reduction_results}
-% \end{table}
+\subsection{Implementation Details}
+All experiments were conducted on an NVIDIA RTX 4090 GPU.
+The vocabulary pruning techniques described in this paper were implemented in a public repository\footnote{\url{https://github.com/WoutDeRijck/vocab-pruning.git}}.
+This implementation includes all pruning methods, evaluation scripts, and analysis tools described in this work.
+For encoder layer pruning, a custom fork of the LoSparse implementation was used\footnote{\url{https://github.com/WoutDeRijck/LoSparse.git}}.
+Both codebases are released to facilitate reproducibility and further research in model compression techniques.
+
+\subsection{Results}
+Table \ref{tab:results} presents the comprehensive evaluation across all GLUE tasks for the different pruning methods. Several key findings emerge:
+\\ \\
+First, vocabulary pruning alone achieves substantial parameter reduction (20.02-23.27\%) with minimal performance impact.
+The TF-IDF + OOV approach performs exceptionally well, maintaining 99.2\% of the original model's average performance while reducing parameters by over 20\%.
+\\ \\
+Second, the inclusion of OOV handling mechanisms significantly improves results compared to their non-OOV counterparts.
+For instance, TF-IDF + OOV outperforms standard TF-IDF by 2.4 percentage points on average, with particularly strong improvements on CoLA (9.5 points) and RTE (6.1 points).
+\\ \\
+Third, pruning technique effectiveness varies by task type.
+Attention-based pruning excels on single-sentence tasks (SST-2, CoLA) and paraphrase detection (MRPC), while TF-IDF-based methods perform better on more complex reasoning tasks (MNLI, QNLI, RTE).
+\\ \\
+Fourth, as illustrated in Figure \ref{fig:pruning_ratio}, vocabulary pruning demonstrates superior performance and efficiency compared to encoder-layer pruning (LoSparse) at lower pruning ratios (5-20\%).
+Importantly, while LoSparse requires computationally expensive online training procedures, vocabulary pruning is an offline pre-fine-tuning method that can be applied with minimal computational overhead.
+A striking result is that a 20\% reduction in total model parameters translates to a 77.34\% reduction in vocabulary size while maintaining competitive performance.
+This efficiency advantage makes vocabulary pruning particularly attractive for resource-constrained environments.
+Since the embedding layer constitutes only 25.95\% of the total model parameters, vocabulary pruning has an inherent upper limit.
+Beyond the 20\% threshold, performance decreases sharply for vocabulary pruning while LoSparse maintains more stable performance at higher compression rates.
+For applications requiring moderate compression with minimal performance impact, vocabulary pruning provides an optimal approach, while more aggressive compression benefits from combining both techniques.
+\\ \\
+The benchmark results in Table \ref{tab:benchmark_results} further quantify the practical benefits, showing a 20\% reduction in storage size and 14.83\% reduction in GPU memory requirements with negligible impact on inference time (+1.33\%).
+
+\begin{table}[htb]
+\centering
+\scriptsize
+\setlength{\tabcolsep}{3pt}
+\begin{tabular}{lrrr}
+\toprule
+\textbf{Metric} & \textbf{Base} & \textbf{Pruned} & \textbf{Impr.(\%)} \\
+\midrule
+Params (M) & 149.61 & 119.69 & 20.00 \\
+Embed. Params (M) & 38.68 & 8.76 & 77.34 \\
+Storage (MB) & 570.72 & 456.59 & 20.00 \\
+GPU Mem (MB) & 823.06 & 700.99 & 14.83 \\
+Infer. Time (ms) & 18.73 & 18.98 & -1.33 \\
+\bottomrule
+\end{tabular}
+\caption{Benchmark results comparing base ModernBERT with pruned variant. Tests with batch size 128 on NVIDIA RTX 4090.}
+\label{tab:benchmark_results}
+\end{table}
+
+\begin{figure*}[t]
+    \centering
+    \includegraphics[width=\linewidth]{images/pruning_ratios.png}
+    \caption{Performance comparison between vocabulary pruning (TF-IDF + OOV) and encoder pruning (LoSparse) across different pruning ratios for MNLI, QNLI, and QQP tasks. }
+    \label{fig:pruning_ratio}
+\end{figure*}
 
@@
 TODO
-\\ \\
+
 \section{Future Work}

### 2025-05-14 paper/main.tex
rephrase abstract and introduction

@@
 \usepackage{url}
+\usepackage{import}
 
@@
 \begin{abstract}
-Large language models like ModernBERT face deployment challenges due to their computational demands and memory footprint, limiting their use in resource-constrained environments.
-While various pruning techniques exist, most require resource-intensive fine-tuning, creating a bottleneck for efficient model adaptation.
-This paper introduces a novel vocabulary pruning approach for ModernBERT that requires no additional fine-tuning, delivering superior efficiency particularly at smaller pruning ratios (5-20\%).
-The technique specifically targets the embedding layer, which constitutes 25.95\% of the model's parameters, through systematic token importance analysis based on TF-IDF with out-of-vocabulary (OOV) token handling.
+Large language models like ModernBERT deliver impressive performance but face a critical deployment challenge: their substantial computational and memory demands restrict their use in resource-constrained environments.
+While various pruning techniques exist to address this bottleneck, most require expensive post-pruning fine-tuning—creating a paradoxical situation where model optimization itself demands significant resources.
+This paper introduces a novel vocabulary pruning approach for ModernBERT that eliminates the need for post-pruning recovery fine-tuning, delivering superior efficiency particularly at smaller pruning ratios (5-20\%).
+The proposed technique specifically targets the embedding layer—which constitutes 25.95\% of the model's parameters but remains surprisingly underexplored in compression research—through systematic token importance analysis based on TF-IDF with specialized out-of-vocabulary (OOV) token handling.
 Experiments on the GLUE benchmark demonstrate that vocabulary pruning alone achieves 20-23\% parameter reduction while maintaining 99.2\% of the original performance, outperforming encoder-based pruning methods at these ratios.
@@
 
-Large pre-trained transformer models such as BERT have ushered in a new era of performance across virtually every core NLP task---from sentiment analysis and natural language inference to question answering and summarization.
-Their success stems in large part from massive embedding layers and deep stacks of self-attention and feed-forward blocks, which together learn extraordinarily rich, contextualized token representations.
+Language models like ModernBERT have revolutionized natural language processing, achieving unprecedented performance across virtually every core NLP task—from sentiment analysis and natural language inference to question answering and summarization.
+However, this remarkable progress comes with a significant cost: a state-of-the-art model requiring multiple gigabytes of memory simply cannot run on the billions of smartphones, IoT devices, and edge computing platforms that could benefit most from NLP capabilities.
+This deployment gap—between what's technically possible in research environments and what's practically deployable in the real world—represents one of the most pressing challenges in making advanced AI accessible.
+The success of these models stems from massive embedding layers and deep stacks of self-attention and feed-forward blocks, which together learn extraordinarily rich, contextualized token representations.
 Yet this very scale makes them difficult to deploy: memory footprints measured in multiple gigabytes, inference latencies unsuited to on-device or real-time use, and the sheer cost of fine-tuning or retraining pruned variants can all stand in the way of practical adoption.
 \\ \\
-Pruning---selectively removing parameters deemed non-essential---has emerged as one of the most promising paths to slim down these models.
+Pruning—selectively removing parameters deemed non-essential—has emerged as one of the most promising paths to slim down these models.
 To date, however, the majority of work has focused on sparsifying or structurally pruning the encoder layers: zeroing out small weights in feed-forward networks, dropping entire attention heads, or even omitting full transformer blocks during inference.
-While these methods can yield substantial compression, they often require additional fine-tuning to recover lost accuracy, and they overlook a particularly ``heavy'' component of transformer architectures: the embedding layer.
+While these methods can yield substantial compression, they often require additional fine-tuning to recover lost accuracy, and they overlook a particularly "heavy" component of transformer architectures: the embedding layer.
+This tunnel vision in pruning research represents a missed opportunity for significant efficiency gains.
 \\ \\
 In BERT-style models, the token embedding matrix alone can account for 25.95\% of all parameters in ModernBERT, yet vocabulary pruning remains comparatively underexplored.
-Simple heuristics---such as dropping all tokens not observed in a downstream training corpus---can achieve size reductions, but they fail to capture semantic importance, ignore relationships between pruned and retained tokens, and typically require a retraining step to adapt the model to its slimmed vocabulary.
-As a result, embedding-level pruning often remains a ``bolt-on'' afterthought rather than an integral, efficient component of model compression.
+Simple heuristics—such as keeping only tokens observed in a downstream training corpus—provide a strong starting point and can achieve substantial size reductions with minimal performance impact.
+However, as the experiments in this paper demonstrate, these approaches don't fully capture semantic importance, miss opportunities to optimize token relationships, and often lack principled strategies for handling out-of-vocabulary tokens.
+While effective, they leave room for more sophisticated techniques that can push compression limits further while maintaining semantic integrity.
+This paper advances embedding-level pruning from a useful baseline to a central, integral component of model compression pipelines.
 \\ \\
-This paper presents a fine-tuning-free approach to vocabulary pruning in ModernBERT~\cite{modernbert2023} that directly addresses these gaps.
+This paper presents a vocabulary pruning approach for ModernBERT~\cite{modernbert2023} that operates as a pre-fine-tuning optimization step and directly addresses these gaps.
 The method proceeds in two complementary stages.
@@
 \\ \\
-In summary, our contributions are:
+In summary, the contributions of this paper are:
 
 \begin{itemize}
-    \item A novel, fine-tuning-free vocabulary pruning method that targets the otherwise-neglected embedding layer in transformer models.
+    \item A vocabulary pruning methodology that operates as a pre-fine-tuning optimization step, eliminating the need for post-pruning recovery fine-tuning typically associated with model compression.
 
-    \item A hybrid TF-IDF plus semantic clustering strategy for robust handling of pruned tokens, preserving semantic structure without [UNK] collapse.
+    \item A systematic comparison of token importance estimation techniques, revealing that TF-IDF with OOV handling provides the most robust performance across diverse NLP tasks.
 
-    \item Empirical evidence on GLUE that embedding pruning alone can yield $\sim$20\% size reduction with $<$1\% performance loss, and that combined embedding + encoder pruning achieves $>$35\% compression for under 2\% drop in accuracy.
+    \item Empirical evidence that vocabulary pruning outperforms encoder-based pruning at lower pruning ratios (5-20\%), offering a more efficient pathway for moderate model compression.
 
-    \item A modular pipeline compatible with any pre-trained encoder model and complementary to structured encoder-layer pruning methods.
+    \item Practical deployment benefits including 20\% reduction in storage size and 14.83\% reduction in GPU memory with no inference time penalty.
 \end{itemize}
@@
 Model compression strategies for LLMs can be categorized into four primary methods:
-\\ \\
+
 \begin{enumerate}
@@
 Early work by Abdaoui et al.~\cite{abdaoui2020load} showed that most parameters of multilingual BERT lie in the embedding layer, so trimming unused language tokens can dramatically cut size.
-In ``Load What You Need,'' they drop languages from mBERT's vocabulary and obtain up to a 45\% reduction in total parameters with comparable accuracy on XNLI.
+In "Load What You Need," they drop languages from mBERT's vocabulary and obtain up to a 45\% reduction in total parameters with comparable accuracy on XNLI.
 \\ \\
@@
 TextPruner by Shen et al.~\cite{shen2022textpruner} automates vocabulary pruning: it scans a corpus and removes any tokenizer token not present in the text, deleting its row in the embedding matrix.
-This simple ``drop unused tokens'' mode can cut model size and speed up training/inference on that domain.
+This simple "drop unused tokens" mode can cut model size and speed up training/inference on that domain.
 \\ \\
@@
 For example, after pruning, Nair et al.~\cite{nair2023blade} report retrieval effectiveness on par with larger baselines, while achieving much faster indexing.
-Dorkin et al.~\cite{dorkin2025estonian} explicitly note ``vocabulary pruning has no observable negative effect on the downstream task'' (Estonian NER).
+Dorkin et al.~\cite{dorkin2025estonian} explicitly note "vocabulary pruning has no observable negative effect on the downstream task" (Estonian NER).
 These findings suggest vocabulary pruning is an effective way to tailor large models to specific domains or languages.
@@
 \subsection{Encoder-Layer Pruning (Auxiliary Methods)}
-
-Although our focus is vocab/embedding pruning, many complementary pruning methods work at the encoder layers.
+Although this paper focuses on vocab/embedding pruning, many complementary pruning methods work at the encoder layers.
 A classic example is \textbf{attention-head pruning}: Michel et al.~\cite{michel2019sixteen} discovered that a large percentage of attention heads can be removed at test time without significantly impacting performance.
@@
 The high-level vocabulary pruning procedure, combining importance-based token pruning with optional semantic clustering for OOV token handling, is formalized in Algorithm~\ref{alg:hybrid_pruning}.
-\\ \\
+
 \begin{algorithm}[H]
@@
 
+\input{tables/token_statistics}
+
 \section{Experiments}
-In this section, the efficacy of vocabulary pruning is evaluated as a viable model compression technique for transformer-based language models.
-The primary objectives are to determine whether selective vocabulary reduction can maintain model performance while significantly decreasing parameter count, and to understand how different pruning strategies affect various natural language understanding tasks.
+Having established a theoretical foundation for the vocabulary pruning approach, this section presents a comprehensive evaluation designed to answer several key questions: Can selective vocabulary reduction maintain model performance while significantly decreasing parameter count?
+How do different pruning strategies compare on various natural language understanding tasks?
+Does the pre-fine-tuning approach that avoids post-pruning recovery fine-tuning deliver practical benefits in real-world deployment scenarios?
+Through careful experimentation across diverse tasks, the results demonstrate not only the theoretical soundness of the approach but also its practical advantages in creating more efficient transformer models.
 
 \subsection{Datasets and Metrics}
-
-\begin{table*}[htbp]
-\centering
-\scriptsize
-\setlength{\tabcolsep}{3.5pt}
-\begin{tabular}{l|cc|cc|cc|cc|cc}
-\toprule
-& \multicolumn{2}{c|}{\textbf{Total Tokens}} & \multicolumn{2}{c|}{\textbf{Unique Tokens}} & \multicolumn{2}{c|}{\textbf{Vocab Coverage (\%)}} & \multicolumn{2}{c|}{\textbf{Top 20\% Coverage (\%)}} & \multicolumn{2}{c}{\textbf{TF-IDF/Freq Overlap (\%)}} \\
-\textbf{Task} & \textbf{Train} & \textbf{Test} & \textbf{Train} & \textbf{Test} & \textbf{Train} & \textbf{Test} & \textbf{Train} & \textbf{Test} & \textbf{Train} & \textbf{Test} \\
-\midrule
-COLA & 71,818 & 7,939 & 5,416 & 1,934 & 17.74 & 6.34 & 85.12 & 74.68 & 58.36 & 51.04 \\
-MNLI & 13,044,457 & 1,452,279 & 25,793 & 22,664 & 84.51 & 74.25 & 90.69 & 89.64 & 38.41 & 61.08 \\
-MRPC & 165,585 & 18,701 & 11,096 & 3,567 & 36.35 & 11.69 & 83.17 & 71.80 & 57.91 & 54.56 \\
-QNLI & 4,388,343 & 485,747 & 26,176 & 20,360 & 85.76 & 66.71 & 88.31 & 85.79 & 33.41 & 55.33 \\
-QQP & 9,036,384 & 1,000,855 & 25,486 & 18,534 & 83.50 & 60.72 & 94.02 & 91.79 & 32.39 & 61.20 \\
-RTE & 151,109 & 16,215 & 13,354 & 4,198 & 43.75 & 13.75 & 81.77 & 70.75 & 57.49 & 54.71 \\
-SST2 & 686,566 & 75,775 & 11,536 & 8,084 & 37.80 & 26.49 & 84.23 & 80.53 & 59.08 & 57.74 \\
-STSB & 128,151 & 14,481 & 10,346 & 3,271 & 33.90 & 10.72 & 82.66 & 71.30 & 60.22 & 49.54 \\
-\bottomrule
-\end{tabular}
-\caption{Token statistics across GLUE tasks, comparing train and test splits.}
-\label{tab:token_statistics}
-\end{table*}
-
 The General Language Understanding Evaluation (GLUE) benchmark serves as the primary evaluation framework for this study.
@@
 
-\begin{table}[htbp]
-\centering
-\scriptsize
-\setlength{\tabcolsep}{4pt}
-\begin{tabular}{l|cc}
-\toprule
-& \multicolumn{2}{c}{\textbf{Test Vocabulary Coverage}} \\
-\textbf{Task} & \textbf{OOV Tokens} & \textbf{OOV\%} \\
-\midrule
-SST2 & 34 & 0.42\% \\
-MRPC & 465 & 13.04\% \\
-COLA & 169 & 8.74\% \\
-STSB & 448 & 13.70\% \\
-QQP & 191 & 1.03\% \\
-MNLI & 31 & 0.14\% \\
-QNLI & 83 & 0.41\% \\
-RTE & 509 & 12.12\% \\
-\bottomrule
-\end{tabular}
-\caption{Train-test vocabulary overlap statistics.
-\textbf{OOV Tokens}: number of unique tokens in the test set that don't appear in the training set; \textbf{OOV\%}: percentage of test vocabulary not found in train.
-Lower percentages indicate better vocabulary coverage, which is beneficial for pruning.}
-\label{tab:vocab_overlap}
-\end{table}
+\input{tables/vocab_overlap}
+
+\input{tables/results.tex}
 
@@
 Three baseline models are established to evaluate the effectiveness of the proposed vocabulary pruning techniques:
-\\ \\
+
 \begin{itemize}
@@
 
-\begin{table*}[h]
-\centering
-\scriptsize
-\setlength{\tabcolsep}{9pt}
-\begin{tabular}{l@{\hspace{25pt}}cccccccc:c}
-\toprule
-& \multicolumn{2}{c}{\textbf{Single Sentence}} & \multicolumn{3}{c}{\textbf{Paraphrase and Similarity}} & \multicolumn{3}{c}{\textbf{Natural Language Inference}} & \\
-\cmidrule(lr){2-3} \cmidrule(lr){4-6} \cmidrule(lr){7-9}
-\textbf{Method} & \textbf{SST-2} & \textbf{CoLA} & \textbf{MRPC} & \textbf{STS-B} & \textbf{QQP} & \textbf{MNLI} & \textbf{QNLI} & \textbf{RTE} & \textbf{AVG} \\
-\midrule
-\multicolumn{10}{l}{\textbf{Baseline}} \\
-ModernBERT (Full) & 0.951 & 0.616 & 0.89 & 0.917 & 0.917 & 0.881 & 0.939 & 0.643 & 0.831 \\
-\midrule
-\multicolumn{10}{l}{\textbf{Encoder Layer Pruning}} \\
-LoSparse & 0.929 & 0.316 & 0.856 & 0.882 & 0.911 & 0.858 & 0.907 & 0.610 & 0.784 \\
-\textit{Parameter Reduction} & \textit{20.16\%} & \textit{20.16\%} & \textit{20.16\%} & \textit{20.16\%} & \textit{20.16\%} & \textit{20.16\%} & \textit{20.16\%} & \textit{20.16\%} &  \\
-\midrule
-\multicolumn{10}{l}{\textbf{Vocabulary Pruning}} \\
-Random Selection & 0.911 & 0.470 & 0.798 & 0.845 & 0.780 & 0.504 & 0.669 & 0.566 & 0.693 \\
-Clustering-Based & 0.899 & 0.051 & 0.714 & 0.786 & 0.774 & 0.501 & 0.510 & 0.566 & 0.600 \\
-Frequency-Based & 0.943 & 0.467 & 0.812 & 0.905 & 0.904 & 0.858 & 0.902 & 0.546 & 0.792 \\
-Attention-Based & \textbf{0.947} & 0.589 & \textbf{0.864} & 0.885 & 0.763 & 0.683 & 0.791 & 0.578 & 0.763 \\
-TF-IDF Based & 0.933 & 0.520 & 0.807 & 0.901 & 0.898 & \textbf{0.860} & \textbf{0.909} & 0.574 & 0.800 \\
-\\ [-6pt]
-\hdashline
-\\[-6pt]
-Freq + OOV & 0.938 & 0.540 & 0.828 & \textbf{0.906} & \textbf{0.915} & 0.858 & 0.907 & 0.615 & 0.813 \\
-TF-IDF + OOV & 0.923 & \textbf{0.615} & 0.834 & 0.903 & 0.912 & 0.858 & 0.908 & \textbf{0.635} & \textbf{0.824} \\
-\textit{Parameter Reduction} & \textit{20.25\%} & \textit{23.27\%} & \textit{20.02\%} & \textit{20.18\%} & \textit{20.02\%} & \textit{20.02\%} & \textit{20.02\%} & \textit{20.02\%} &  \\
-\\ [-6pt]
-\hdashline
-\\[-6pt]
-Train Tokens Only & 0.950 & 0.630 & 0.861 & 0.917 & 0.917 & 0.883 & 0.915 & 0.639 & 0.815 \\
-\textit{Parameter Reduction} & \textit{18.85\%} & \textit{22.61\%} & \textit{18.56\%} & \textit{18.76\%} & \textit{4.79\%} & \textit{6.74\%} & \textit{6.42\%} & \textit{17.06\%} & \\
-\bottomrule
-\end{tabular}
-\caption{Performance on GLUE dev set.
-ModernBERT is fine-tuned separately for each task.
-Scores are accuracies except for CoLA (Matthew's correlation), and STS-B (Pearson correlation).
-"+ OOV" indicates the pruning technique combined with out-of-vocabulary clustering for token remapping.
-Parameter reduction percentages show total model size decrease for each method.}
-\label{tab:results}
-\end{table*}
-
 \subsection{Implementation Details}
@@
 Both codebases are released to facilitate reproducibility and further research in model compression techniques.
+
+\begin{figure*}[t]
+    \centering
+    \includegraphics[width=\linewidth]{images/pruning_ratios.png}
+    \caption{Performance comparison between vocabulary pruning (TF-IDF + OOV) and encoder pruning (LoSparse) across different pruning ratios for MNLI, QNLI, and QQP tasks. }
+    \label{fig:pruning_ratio}
+\end{figure*}
 
@@
 
-\begin{table}[htb]
-\centering
-\scriptsize
-\setlength{\tabcolsep}{3pt}
-\begin{tabular}{lrrr}
-\toprule
-\textbf{Metric} & \textbf{Base} & \textbf{Pruned} & \textbf{Impr.(\%)} \\
-\midrule
-Params (M) & 149.61 & 119.69 & 20.00 \\
-Embed. Params (M) & 38.68 & 8.76 & 77.34 \\
-Storage (MB) & 570.72 & 456.59 & 20.00 \\
-GPU Mem (MB) & 823.06 & 700.99 & 14.83 \\
-Infer. Time (ms) & 18.73 & 18.98 & -1.33 \\
-\bottomrule
-\end{tabular}
-\caption{Benchmark results comparing base ModernBERT with pruned variant. Tests with batch size 128 on NVIDIA RTX 4090.}
-\label{tab:benchmark_results}
-\end{table}
-
-\begin{figure*}[t]
-    \centering
-    \includegraphics[width=\linewidth]{images/pruning_ratios.png}
-    \caption{Performance comparison between vocabulary pruning (TF-IDF + OOV) and encoder pruning (LoSparse) across different pruning ratios for MNLI, QNLI, and QQP tasks. }
-    \label{fig:pruning_ratio}
-\end{figure*}
+\input{tables/benchmark_results}
 
 \section{Conclusion}
-TODO
+
+% This paper introduced a novel vocabulary pruning approach for transformer-based language models that eliminates the need for post-pruning recovery fine-tuning while delivering substantial parameter reduction with minimal performance impact.
+The work highlights the embedding layer as a critical yet often overlooked component for model compression, demonstrating that intelligent token selection can maintain 99.2\% of original model performance while reducing total parameters by 20-23\%.
+
+% The key contributions of this research are:
+
+% \begin{itemize}
+%     \item A vocabulary pruning methodology that operates as a pre-fine-tuning optimization step, eliminating the need for post-pruning recovery fine-tuning typically associated with model compression.
+
+%     \item A systematic comparison of token importance estimation techniques, revealing that TF-IDF with OOV handling provides the most robust performance across diverse NLP tasks.
+
+%     \item Empirical evidence that vocabulary pruning outperforms encoder-based pruning at lower pruning ratios (5-20\%), offering a more efficient pathway for moderate model compression.
+
+%     \item Practical deployment benefits including 20\% reduction in storage size and 14.83\% reduction in GPU memory with no inference time penalty.
+% \end{itemize}
+
+% The results challenge the conventional wisdom that encoder-layer pruning should be prioritized over vocabulary reduction.
+Instead, the findings demonstrate that a hybrid approach—focusing first on vocabulary pruning for initial efficiency gains and then applying encoder pruning for more aggressive compression—offers the most balanced pathway to efficient transformer models.
+
+% The broader implications of this work extend beyond the specific techniques presented.
+By eliminating the need for expensive post-pruning fine-tuning, the presented approach democratizes model compression, making it accessible even for organizations with limited computational resources.
+This could accelerate the adoption of transformer models in resource-constrained environments such as mobile devices, edge computing platforms, and regions with limited technological infrastructure.
+
+% Future work could explore dynamic vocabulary pruning that adapts to changing data distributions, application of these techniques to multilingual models where vocabulary redundancy is even more pronounced, and integration with quantization and knowledge distillation for more comprehensive compression pipelines.
+As large language models continue to grow in size and capability, efficient adaptation methods like vocabulary pruning will become increasingly critical for bridging the gap between research advances and practical deployment.
 
 \section{Future Work}
-\begin{itemize}
-    \item Combining encoder-layer pruning with vocabulary pruning.
-    \item Retrieval task performance.
-    \item ...
-\end{itemize}
+
+% While this paper demonstrates the effectiveness of vocabulary pruning as a standalone technique, several promising research directions could further extend its impact:
+
+% \begin{itemize}
+%     \item \textbf{Unified Compression Framework}: Developing an integrated approach that combines vocabulary pruning with encoder-layer compression techniques like LoSparse could provide a more comprehensive model compression solution.
+Preliminary experiments suggest these approaches are complementary, with vocabulary pruning excelling at lower compression ratios and encoder pruning maintaining better performance at higher ratios.
+
+%     \item \textbf{Task Adaptation Networks}: Instead of removing pruned token embeddings entirely, future work could explore low-dimensional adapter networks that efficiently transform a small set of base embeddings to approximate the full vocabulary, potentially offering better performance/size tradeoffs.
+
+%     \item \textbf{Retrieval Performance Analysis}: Evaluating vocabulary pruning on information retrieval tasks could reveal different optimal pruning strategies, as retrieval models may depend on different token importance distributions than classification tasks.
+
+%     \item \textbf{Cross-Lingual Applications}: Extending this approach to multilingual models presents an exciting opportunity, as these models typically have much larger vocabularies with significant redundancy across languages.
+Token importance may vary by language, suggesting language-specific pruning strategies.
+
+%     \item \textbf{Dynamic Vocabulary Adaptation}: Rather than static pruning, future systems could implement dynamic vocabulary management that adapts to shifting data distributions or task requirements during deployment.
+
+%     \item \textbf{Theoretical Foundations}: Developing formal methods to establish theoretical bounds on performance loss relative to vocabulary reduction could provide valuable insights for model design and compression strategies.
+% \end{itemize}
+
+% These directions collectively aim to further reduce the computational and memory requirements of large language models while maintaining their impressive capabilities, ultimately making state-of-the-art NLP more accessible across diverse computing environments.
 \newpage

### 2025-05-14 paper/tables/benchmark_results.tex
rephrase abstract and introduction

\begin{table}[htb]
\centering
\scriptsize
\setlength{\tabcolsep}{3pt}
\begin{tabular}{lrrr}
\toprule
\textbf{Metric} & \textbf{Base} & \textbf{Pruned} & \textbf{Impr.(\%)} \\
\midrule
Params (M) & 149.61 & 119.69 & 20.00 \\
Embed. Params (M) & 38.68 & 8.76 & 77.34 \\
Storage (MB) & 570.72 & 456.59 & 20.00 \\
GPU Mem (MB) & 823.06 & 700.99 & 14.83 \\
Infer. Time (ms) & 18.73 & 18.98 & -1.33 \\
\bottomrule
\end{tabular}
\caption{Benchmark results comparing base ModernBERT with pruned variant. Tests with batch size 128 on NVIDIA RTX 4090.}
\label{tab:benchmark_results}
\end{table}

### 2025-05-14 paper/tables/results.tex
rephrase abstract and introduction

\begin{table*}[h]
\centering
\scriptsize
\setlength{\tabcolsep}{9pt}
\begin{tabular}{l@{\hspace{25pt}}cccccccc:c}
\toprule
& \multicolumn{2}{c}{\textbf{Single Sentence}} & \multicolumn{3}{c}{\textbf{Paraphrase and Similarity}} & \multicolumn{3}{c}{\textbf{Natural Language Inference}} & \\
\cmidrule(lr){2-3} \cmidrule(lr){4-6} \cmidrule(lr){7-9}
\textbf{Method} & \textbf{SST-2} & \textbf{CoLA} & \textbf{MRPC} & \textbf{STS-B} & \textbf{QQP} & \textbf{MNLI} & \textbf{QNLI} & \textbf{RTE} & \textbf{AVG} \\
\midrule
\multicolumn{10}{l}{\textbf{Baseline}} \\
ModernBERT (Full) & 0.951 & 0.616 & 0.89 & 0.917 & 0.917 & 0.881 & 0.939 & 0.643 & 0.831 \\
\midrule
\multicolumn{10}{l}{\textbf{Encoder Layer Pruning}} \\
LoSparse & 0.929 & 0.316 & 0.856 & 0.882 & 0.911 & 0.858 & 0.907 & 0.610 & 0.784 \\
\textit{Parameter Reduction} & \textit{20.16\%} & \textit{20.16\%} & \textit{20.16\%} & \textit{20.16\%} & \textit{20.16\%} & \textit{20.16\%} & \textit{20.16\%} & \textit{20.16\%} &  \\
\midrule
\multicolumn{10}{l}{\textbf{Vocabulary Pruning}} \\
Random Selection & 0.911 & 0.470 & 0.798 & 0.845 & 0.780 & 0.504 & 0.669 & 0.566 & 0.693 \\
Clustering-Based & 0.899 & 0.051 & 0.714 & 0.786 & 0.774 & 0.501 & 0.510 & 0.566 & 0.600 \\
Frequency-Based & 0.943 & 0.467 & 0.812 & 0.905 & 0.904 & 0.858 & 0.902 & 0.546 & 0.792 \\
Attention-Based & \textbf{0.947} & 0.589 & \textbf{0.864} & 0.885 & 0.763 & 0.683 & 0.791 & 0.578 & 0.763 \\
TF-IDF Based & 0.933 & 0.520 & 0.807 & 0.901 & 0.898 & \textbf{0.860} & \textbf{0.909} & 0.574 & 0.800 \\
\\ [-6pt]
\hdashline
\\[-6pt]
Freq + OOV & 0.938 & 0.540 & 0.828 & \textbf{0.906} & \textbf{0.915} & 0.858 & 0.907 & 0.615 & 0.813 \\
TF-IDF + OOV & 0.923 & \textbf{0.615} & 0.834 & 0.903 & 0.912 & 0.858 & 0.908 & \textbf{0.635} & \textbf{0.824} \\
\textit{Parameter Reduction} & \textit{20.25\%} & \textit{23.27\%} & \textit{20.02\%} & \textit{20.18\%} & \textit{20.02\%} & \textit{20.02\%} & \textit{20.02\%} & \textit{20.02\%} &  \\
\\ [-6pt]
\hdashline
\\[-6pt]
Train Tokens Only & 0.950 & 0.630 & 0.861 & 0.917 & 0.917 & 0.883 & 0.915 & 0.639 & 0.815 \\
\textit{Parameter Reduction} & \textit{18.85\%} & \textit{22.61\%} & \textit{18.56\%} & \textit{18.76\%} & \textit{4.79\%} & \textit{6.74\%} & \textit{6.42\%} & \textit{17.06\%} & \\
\bottomrule
\end{tabular}
\caption{Performance on GLUE dev set. ModernBERT is fine-tuned separately for each task. Scores are accuracies except for CoLA (Matthew's correlation), and STS-B (Pearson correlation). "+ OOV" indicates the pruning technique combined with out-of-vocabulary clustering for token remapping. Parameter reduction percentages show total model size decrease for each method.}
\label{tab:results}
\end{table*}

### 2025-05-14 paper/tables/token_statistics.tex
rephrase abstract and introduction

\begin{table*}[htbp]
\centering
\scriptsize
\setlength{\tabcolsep}{3.5pt}
\begin{tabular}{l|cc|cc|cc|cc|cc}
\toprule
& \multicolumn{2}{c|}{\textbf{Total Tokens}} & \multicolumn{2}{c|}{\textbf{Unique Tokens}} & \multicolumn{2}{c|}{\textbf{Vocab Coverage (\%)}} & \multicolumn{2}{c|}{\textbf{Top 20\% Coverage (\%)}} & \multicolumn{2}{c}{\textbf{TF-IDF/Freq Overlap (\%)}} \\
\textbf{Task} & \textbf{Train} & \textbf{Test} & \textbf{Train} & \textbf{Test} & \textbf{Train} & \textbf{Test} & \textbf{Train} & \textbf{Test} & \textbf{Train} & \textbf{Test} \\
\midrule
COLA & 71,818 & 7,939 & 5,416 & 1,934 & 17.74 & 6.34 & 85.12 & 74.68 & 58.36 & 51.04 \\
MNLI & 13,044,457 & 1,452,279 & 25,793 & 22,664 & 84.51 & 74.25 & 90.69 & 89.64 & 38.41 & 61.08 \\
MRPC & 165,585 & 18,701 & 11,096 & 3,567 & 36.35 & 11.69 & 83.17 & 71.80 & 57.91 & 54.56 \\
QNLI & 4,388,343 & 485,747 & 26,176 & 20,360 & 85.76 & 66.71 & 88.31 & 85.79 & 33.41 & 55.33 \\
QQP & 9,036,384 & 1,000,855 & 25,486 & 18,534 & 83.50 & 60.72 & 94.02 & 91.79 & 32.39 & 61.20 \\
RTE & 151,109 & 16,215 & 13,354 & 4,198 & 43.75 & 13.75 & 81.77 & 70.75 & 57.49 & 54.71 \\
SST2 & 686,566 & 75,775 & 11,536 & 8,084 & 37.80 & 26.49 & 84.23 & 80.53 & 59.08 & 57.74 \\
STSB & 128,151 & 14,481 & 10,346 & 3,271 & 33.90 & 10.72 & 82.66 & 71.30 & 60.22 & 49.54 \\
\bottomrule
\end{tabular}
\caption{Token statistics across GLUE tasks, comparing train and test splits.}
\label{tab:token_statistics}
\end{table*}

### 2025-05-14 paper/tables/vocab_overlap.tex
rephrase abstract and introduction

\begin{table}[htbp]
\centering
\scriptsize
\setlength{\tabcolsep}{4pt}
\begin{tabular}{l|cc}
\toprule
& \multicolumn{2}{c}{\textbf{Test Vocabulary Coverage}} \\
\textbf{Task} & \textbf{OOV Tokens} & \textbf{OOV\%} \\
\midrule
SST2 & 34 & 0.42\% \\
MRPC & 465 & 13.04\% \\
COLA & 169 & 8.74\% \\
STSB & 448 & 13.70\% \\
QQP & 191 & 1.03\% \\
MNLI & 31 & 0.14\% \\
QNLI & 83 & 0.41\% \\
RTE & 509 & 12.12\% \\
\bottomrule
\end{tabular}
\caption{Train-test vocabulary overlap statistics. \textbf{OOV Tokens}: number of unique tokens in the test set that don't appear in the training set; \textbf{OOV\%}: percentage of test vocabulary not found in train. Lower percentages indicate better vocabulary coverage, which is beneficial for pruning.}
\label{tab:vocab_overlap}
\end{table}

### 2025-05-14 paper/main.tex
fix

@@
 The proposed technique specifically targets the embedding layer—which constitutes 25.95\% of the model's parameters but remains surprisingly underexplored in compression research—through systematic token importance analysis based on TF-IDF with specialized out-of-vocabulary (OOV) token handling.
-Experiments on the GLUE benchmark demonstrate that vocabulary pruning alone achieves 20-23\% parameter reduction while maintaining 99.2\% of the original performance, outperforming encoder-based pruning methods at these ratios.
+Experiments on the GLUE benchmark demonstrate that vocabulary pruning alone achieves 20-23\% parameter reduction while maintaining 97.6\% of the original performance, outperforming encoder-based pruning methods at these ratios.
 Benchmark measurements confirm practical benefits: 20\% reduction in storage size, about 15\% reduction in GPU memory, and no impact on inference time.
@@
 First, vocabulary pruning alone achieves substantial parameter reduction (20.02-23.27\%) with minimal performance impact.
-The TF-IDF + OOV approach performs exceptionally well, maintaining 99.2\% of the original model's average performance while reducing parameters by over 20\%.
+The TF-IDF + OOV approach performs exceptionally well, maintaining 97.6\% of the original model's average performance while reducing parameters by over 20\%.
 \\ \\
@@
 % This paper introduced a novel vocabulary pruning approach for transformer-based language models that eliminates the need for post-pruning recovery fine-tuning while delivering substantial parameter reduction with minimal performance impact.
-The work highlights the embedding layer as a critical yet often overlooked component for model compression, demonstrating that intelligent token selection can maintain 99.2\% of original model performance while reducing total parameters by 20-23\%.
+The work highlights the embedding layer as a critical yet often overlooked component for model compression, demonstrating that intelligent token selection can maintain 97.6\% of original model performance while reducing total parameters by 20-23\%.
 

### 2025-05-17 paper/main.tex
comments Karel

@@
 \usepackage{import}
+\usepackage{xcolor}
 
 \title{\huge \textbf{Words That Matter:\\ Vocabulary Pruning in ModernBERT}}
-\author{Wout De Rijck, Dr. Ir. Karel D'Oosterlinck, \\ Prof. Dr. Ir. Thomas Demeester, Prof. Dr. Ir. Chris Develder}
+\author{Wout De Rijck, Prof. Dr. Ir. Thomas Demeester, \\ Prof. Dr. Ir. Chris Develder, Dr. Ir. Karel D'Oosterlinck}
 \date{} % No date
@@
 \maketitle
-% Motivation / Context (1–2 sentences):
-% Start with the why. What's the broader problem or challenge? Why is this area important right now (e.g., limitations of current LLMs, safety, interpretability, efficiency)?
-
-% Gap / Problem Statement:
-% What specific problem or limitation does your paper address? This sets the stage for your contribution. It should hint at a shortcoming in current approaches without being overly critical.
-
-% Your Approach / Contribution:
-% What did you do? This is the heart of the abstract. It could be:
-
-% a new method or model
-
-% a novel dataset or benchmark
-
-% a unique analysis or framework
-
-% an improvement on performance, interpretability, etc.
 
 \begin{abstract}
-Large language models like ModernBERT deliver impressive performance but face a critical deployment challenge: their substantial computational and memory demands restrict their use in resource-constrained environments.
-While various pruning techniques exist to address this bottleneck, most require expensive post-pruning fine-tuning—creating a paradoxical situation where model optimization itself demands significant resources.
-This paper introduces a novel vocabulary pruning approach for ModernBERT that eliminates the need for post-pruning recovery fine-tuning, delivering superior efficiency particularly at smaller pruning ratios (5-20\%).
-The proposed technique specifically targets the embedding layer—which constitutes 25.95\% of the model's parameters but remains surprisingly underexplored in compression research—through systematic token importance analysis based on TF-IDF with specialized out-of-vocabulary (OOV) token handling.
-Experiments on the GLUE benchmark demonstrate that vocabulary pruning alone achieves 20-23\% parameter reduction while maintaining 97.6\% of the original performance, outperforming encoder-based pruning methods at these ratios.
-Benchmark measurements confirm practical benefits: 20\% reduction in storage size, about 15\% reduction in GPU memory, and no impact on inference time.
-This work establishes an efficient pathway for creating lightweight, task-specific transformer models through offline pre-fine-tuning optimizations rather than computationally expensive retraining.
+
+Large Language Models (LLMs) require substantial computational and memory resources, limiting their utility in resource constrained environments. 
+ModernBERT is a smaller encoder-only language model that excels at various tasks while being computationally highly efficient. 
+In this work, we study how much more ModernBERT can be compressed while retaining accuracy on any given task.
+Specifically, we introduce a series of very simple vocabulary pruning techniques that target the embedding layer.
+We compare the resulting accuracy with LoSparse, a state-of-the-art gradient-based pruning technique that targets the encoder layers.
+For the same reduction in total parameters of $\sim$20\%, our much simpler vocabulary pruning technique retained on average 97.6\% of ModernBERT's performance across various tasks, while LoSparse retained only 92.9\%.
+The strong performance of our simple technique indicates that task-specific pruning can meaningfully increase the efficiency of ModernBERT, an already highly efficient model.
+Additionally, our results suggest that state-of-the-art encoder-layer pruning can fall short of simple embedding-layer pruning.
+
 \end{abstract}
 
-% Key Results:
-% Give a quick summary of your main findings or improvements—include metrics or comparisons if possible, but keep it digestible.
-
-% Implications / Why It Matters:
-% What's the takeaway? How might this move the field forward, or be used by other researchers or practitioners?
-
-% Tone and Tease:
-% While all of the above should be there, the language should still invite curiosity. Think clear, precise, but also hook-y—like:
-% "We show that…", "Surprisingly, we find…", "This reveals a new direction for…"
-
 \section{Introduction}
-
-Language models like ModernBERT have revolutionized natural language processing, achieving unprecedented performance across virtually every core NLP task—from sentiment analysis and natural language inference to question answering and summarization.
-However, this remarkable progress comes with a significant cost: a state-of-the-art model requiring multiple gigabytes of memory simply cannot run on the billions of smartphones, IoT devices, and edge computing platforms that could benefit most from NLP capabilities.
-This deployment gap—between what's technically possible in research environments and what's practically deployable in the real world—represents one of the most pressing challenges in making advanced AI accessible.
-The success of these models stems from massive embedding layers and deep stacks of self-attention and feed-forward blocks, which together learn extraordinarily rich, contextualized token representations.
-Yet this very scale makes them difficult to deploy: memory footprints measured in multiple gigabytes, inference latencies unsuited to on-device or real-time use, and the sheer cost of fine-tuning or retraining pruned variants can all stand in the way of practical adoption.
+Efficiency is crucial for modern language models—every additional parameter not only increases storage and memory-bandwidth requirements but also drives up cloud-hosting bills and overall energy consumption.
+Efficiency can be assessed along several dimensions, including inference latency, throughput, energy consumption, parameter count, and memory usage.
+In this work, the focus is on pruning model weights to reduce the memory footprint, as fewer weights lead to lower storage and memory bandwidth requirements, resulting in faster model loading, reduced deployment costs, and improved accessibility for edge devices or resource-constrained environments.
 \\ \\
-Pruning—selectively removing parameters deemed non-essential—has emerged as one of the most promising paths to slim down these models.
-To date, however, the majority of work has focused on sparsifying or structurally pruning the encoder layers: zeroing out small weights in feed-forward networks, dropping entire attention heads, or even omitting full transformer blocks during inference.
-While these methods can yield substantial compression, they often require additional fine-tuning to recover lost accuracy, and they overlook a particularly "heavy" component of transformer architectures: the embedding layer.
-This tunnel vision in pruning research represents a missed opportunity for significant efficiency gains.
+\begin{figure}[t]
+\centering
+\includegraphics[width=\columnwidth]{images/performance_retention.png}
+\caption{Average Performance Retention Across All Tasks for different pruning methods.}
+\label{fig:pruning_retention}
+\end{figure}
+ModernBERT is one of the most efficient models from this perspective. 
+It is a highly optimized, encoder-only transformer that brings together multiple architectural and training innovations—such while maintaining strong downstream performance and minimizing inference cost shortly and the potential impact.
+Given these advances, an open question remains: Can ModernBERT's efficiency be pushed even further by pruning its weight parameters?
 \\ \\
-In BERT-style models, the token embedding matrix alone can account for 25.95\% of all parameters in ModernBERT, yet vocabulary pruning remains comparatively underexplored.
-Simple heuristics—such as keeping only tokens observed in a downstream training corpus—provide a strong starting point and can achieve substantial size reductions with minimal performance impact.
-However, as the experiments in this paper demonstrate, these approaches don't fully capture semantic importance, miss opportunities to optimize token relationships, and often lack principled strategies for handling out-of-vocabulary tokens.
-While effective, they leave room for more sophisticated techniques that can push compression limits further while maintaining semantic integrity.
-This paper advances embedding-level pruning from a useful baseline to a central, integral component of model compression pipelines.
+ModernBERT is not a fully general-purpose model, as it still requires fine-tuning for each specific downstream task.
+This work focuses on task-specific pruning, which leverages that adaptation phase to remove weights irrelevant to the target task, enabling efficient compression with minimal performance loss.
+Unlike general-purpose compression, task-specific pruning can be more precise, resulting in smaller models tailored to specific use cases.
 \\ \\
-This paper presents a vocabulary pruning approach for ModernBERT~\cite{modernbert2023} that operates as a pre-fine-tuning optimization step and directly addresses these gaps.
-The method proceeds in two complementary stages.
-First, a task-specific token importance analysis using TF-IDF scores---balancing how often a token appears in the target dataset against its informativeness across the corpus---identifies and retains only the most critical subword embeddings.
-Second, rather than relegating all pruned tokens to a generic [UNK] index, semantic clustering over the removed embeddings maps each out-of-vocabulary token to its nearest cluster representative.
-This preserves nuanced semantic relationships without expanding the model's vocabulary size.
-\\ \\
-Because this pruning pipeline operates purely as a pre-fine-tuning optimization step and introduces no additional gradient updates, it can be applied to any pre-trained transformer with minimal overhead.
-Moreover, it dovetails seamlessly with existing encoder-layer pruning techniques---such as LoSparse's~\cite{losparse2023} combined low-rank plus sparse factorization---enabling an end-to-end compression strategy that tackles both embeddings and intermediate representations.
-\\ \\
-Validation on the GLUE benchmark suite~\cite{wang2018glue} shows that ModernBERT's embedding matrix can be reduced by up to 77.34\% (translating to a 20\% total parameter reduction) while incurring less than 1\% average degradation in downstream accuracy.
-The approach demonstrates superior performance compared to encoder-layer pruning techniques at lower pruning ratios (5-20\%), with practical benefits including 20\% reduction in storage size and 14.83\% reduction in GPU memory requirements.
-When combined with LoSparse encoder pruning for more aggressive compression, overall parameter reduction exceeds 35\% with still under 2\% performance loss---demonstrating that vocabulary and encoder pruning are both necessary and synergistic for building lightweight, high-performance transformer models.
-\\ \\
-In summary, the contributions of this paper are:
-
-\begin{itemize}
-    \item A vocabulary pruning methodology that operates as a pre-fine-tuning optimization step, eliminating the need for post-pruning recovery fine-tuning typically associated with model compression.
-
-    \item A systematic comparison of token importance estimation techniques, revealing that TF-IDF with OOV handling provides the most robust performance across diverse NLP tasks.
-
-    \item Empirical evidence that vocabulary pruning outperforms encoder-based pruning at lower pruning ratios (5-20\%), offering a more efficient pathway for moderate model compression.
-
-    \item Practical deployment benefits including 20\% reduction in storage size and 14.83\% reduction in GPU memory with no inference time penalty.
-\end{itemize}
-By placing vocabulary pruning on equal footing with encoder-layer sparsification, Words That Matter opens a practical pathway to more compact, efficient, and deployable transformer models---bringing state-of-the-art NLP within reach of resource-constrained environments.
+This work targets the embedding layer for pruning, motivated by the observation that for many tasks, a small fraction of tokens accounts for the majority of token occurrences. 
+This redundancy allows for effective vocabulary pruning by removing unimportant embeddings with minimal impact on performance. 
+Simple frequency- and TF-IDF-based methods are used to identify and discard low-importance tokens. 
+The approach is compared to LoSparse, demonstrating that embedding-layer pruning can achieve strong efficiency gains with less complexity.
+\\
+The best-performing technique is TF-IDF-based pruning, which removes 77.34\% of the embedding layer while retaining 97.6\% of ModernBERT's original performance.
+This is a remarkable result, as it demonstrates that the vast majority of embeddings can be discarded without significantly affecting task accuracy.
+It highlights the extreme redundancy in the embedding layer and the effectiveness of simple, task-aware pruning strategies.
 
@@
     \item Quantization: Reduces the numerical precision of model weights, decreasing memory footprint while maintaining reasonable performance.
-
     \item Parameter pruning: Eliminates redundant or less important connections and neurons to create sparser models with fewer parameters.
-
     \item Low-rank approximation: Decomposes weight matrices into lower-dimensional representations that capture essential information while requiring fewer parameters.
-
     \item Knowledge distillation: Transfers knowledge from larger teacher models to smaller student models, enabling competitive performance with reduced architecture size.
@@
 
+% \subsection{Encoder-Layer Pruning}
+% Although this paper focuses on vocab/embedding pruning, many complementary pruning methods work at the encoder layers.
+A classic example is \textbf{attention-head pruning}: Michel et al.~\cite{michel2019sixteen} discovered that a large percentage of attention heads can be removed at test time without significantly impacting performance.
+Voita et al.~\cite{voita2019analyzing} likewise pruned 38 of 48 encoder heads in an NMT transformer (removing $\sim$79\% of heads) with only a 0.15 BLEU loss.
+% \\ \\
+% In practice, toolkit frameworks (like TextPruner) often support head and FFN pruning alongside vocab pruning.
+Other methods include \textit{movement pruning}~\cite{sanh2020movement} that gradually zeros out small weights during fine-tuning, or \textit{LayerDrop}~\cite{fan2020layerdrop} which stochastically drops whole layers during training.
+
+\subsection{Encoder-Layer Pruning}
+
+Encoder-layer pruning methods aim to remove redundant parameters within the transformer encoder to reduce model size while retaining accuracy.
+A representative example is \textbf{LoSparse}~\cite{li2023losparse}, which decomposes each weight matrix into a low-rank component and a sparse residual.
+The low-rank term captures the dominant shared subspace, while the sparse term accounts for residual variation.
+During fine-tuning, LoSparse uses first-order sensitivity scores—computed as the product of each weight and its gradient—to iteratively prune low-importance parameters.
+\\
+Several alternative encoder pruning techniques have been proposed.
+\textbf{Attention-head pruning}~\cite{michel2019sixteen} removes entire attention heads with minimal performance loss.
+\textbf{Movement pruning}~\cite{sanh2020movement} applies gradient-guided masking to gradually zero out low-importance weights during fine-tuning.
+\textbf{LayerDrop}~\cite{fan2020layerdrop} introduces structured dropout, allowing entire transformer layers to be stochastically removed during training and optionally pruned at inference.
+\\ \\
+These encoder-layer pruning methods differ in their granularity and strategy—targeting individual weights, attention heads, or full layers.
+Among them, LoSparse stands out as a strong baseline for encoder-level pruning.
+
+\subsection{Embedding-Layer Pruning and Compression}
+Pruning the embedding layer often means removing or compressing rows (token vectors) to shrink this heavy component.
+\textbf{Partial Embedding Matrix Adaptation (PEMA)} by Bousquet et al.~\cite{bousquet2023pema} is one systematic approach: they first note that fine-tuning datasets typically use only a small fraction of the full vocabulary.
+PEMA therefore builds a partial embedding matrix containing only tokens seen in the task data, training on that smaller matrix.
+This saves GPU memory during fine-tuning without altering task accuracy.
+After fine-tuning, the updated embeddings are merged back into the full matrix so the model structure remains unchanged.
+Experiments show large memory reductions: e.g., BERT$_{\text{BASE}}$'s embedding matrix can shrink by $\sim$47\% on average (and RoBERTa by $\sim$52\%) when only task-relevant tokens are kept.
+Multilingual models see even more savings (mBERT $\sim$44\% embedding reduction, XLM-R $\sim$64\%) because their vocabularies are huge.
+Crucially, PEMA reports no drop in downstream performance, making it a practical way to prune embeddings during training.
+\\ \\
+Another approach is \textbf{embedding compression} via factorization or hashing.
+Wang et al.~\cite{wang2023lighttoken} propose \textbf{LightToken}, which compresses the entire embedding matrix using low-rank SVD plus a binary residual autoencoder.
+They note token embeddings are highly redundant (the embedding takes $\sim$21--31\% of BERT/RoBERTa's size).
+LightToken first applies a rank-$k$ SVD to capture most variation, then learns compact hash codes to reconstruct the residual.
+In experiments on GLUE, LightToken achieves very high compression (e.g., 25$\times$ smaller embeddings) while preserving accuracy.
+For BERT$_{\text{BASE}}$, a 25$\times$ embedding compression yields an average GLUE score of 82.9 (baseline 83.0).
+\\ \\
+These methods show that even structured low-rank or quantized pruning of embeddings can yield massive space savings with minimal impact on performance.
+\\ \\
+Overall, embedding-layer pruning techniques exploit the fact that many token vectors (especially for rare subwords) contribute little to end-task accuracy.
+By either dropping unused rows (PEMA/TextPruner) or factoring and quantizing embeddings (LightToken), these methods reduce $\sim$25--50\% of model parameters in the embedding matrix, enabling lighter models in practice.
+
 \subsection{Vocabulary Pruning Techniques}
-
 Vocabulary pruning removes rarely used or task-irrelevant tokens to shrink the model's vocabulary (and its embedding matrix).
@@
 
-\subsection{Embedding-Layer Pruning and Compression}
-
-Pruning the embedding layer often means removing or compressing rows (token vectors) to shrink this heavy component.
-\textbf{Partial Embedding Matrix Adaptation (PEMA)} by Bousquet et al.~\cite{bousquet2023pema} is one systematic approach: they first note that fine-tuning datasets typically use only a small fraction of the full vocabulary.
-PEMA therefore builds a partial embedding matrix containing only tokens seen in the task data, training on that smaller matrix.
-This saves GPU memory during fine-tuning without altering task accuracy.
-After fine-tuning, the updated embeddings are merged back into the full matrix so the model structure remains unchanged.
-Experiments show large memory reductions: e.g., BERT$_{\text{BASE}}$'s embedding matrix can shrink by $\sim$47\% on average (and RoBERTa by $\sim$52\%) when only task-relevant tokens are kept.
-Multilingual models see even more savings (mBERT $\sim$44\% embedding reduction, XLM-R $\sim$64\%) because their vocabularies are huge.
-Crucially, PEMA reports no drop in downstream performance, making it a practical way to prune embeddings during training.
-\\ \\
-Another approach is \textbf{embedding compression} via factorization or hashing.
-Wang et al.~\cite{wang2023lighttoken} propose \textbf{LightToken}, which compresses the entire embedding matrix using low-rank SVD plus a binary residual autoencoder.
-They note token embeddings are highly redundant (the embedding takes $\sim$21--31\% of BERT/RoBERTa's size).
-LightToken first applies a rank-$k$ SVD to capture most variation, then learns compact hash codes to reconstruct the residual.
-In experiments on GLUE, LightToken achieves very high compression (e.g., 25$\times$ smaller embeddings) while preserving accuracy.
-For BERT$_{\text{BASE}}$, a 25$\times$ embedding compression yields an average GLUE score of 82.9 (baseline 83.0).
-\\ \\
-These methods show that even structured low-rank or quantized pruning of embeddings can yield massive space savings with minimal impact on performance.
-\\ \\
-Overall, embedding-layer pruning techniques exploit the fact that many token vectors (especially for rare subwords) contribute little to end-task accuracy.
-By either dropping unused rows (PEMA/TextPruner) or factoring and quantizing embeddings (LightToken), these methods reduce the $\sim$25--50\% of model parameters in the embedding matrix, enabling lighter models in practice.
-
-\subsection{Encoder-Layer Pruning (Auxiliary Methods)}
-Although this paper focuses on vocab/embedding pruning, many complementary pruning methods work at the encoder layers.
-A classic example is \textbf{attention-head pruning}: Michel et al.~\cite{michel2019sixteen} discovered that a large percentage of attention heads can be removed at test time without significantly impacting performance.
-Voita et al.~\cite{voita2019analyzing} likewise pruned 38 of 48 encoder heads in an NMT transformer (removing $\sim$79\% of heads) with only a 0.15 BLEU loss.
-\\ \\
-In practice, toolkit frameworks (like TextPruner) often support head and FFN pruning alongside vocab pruning.
-Other methods include \textit{movement pruning}~\cite{sanh2020movement} that gradually zeros out small weights during fine-tuning, or \textit{LayerDrop}~\cite{fan2020layerdrop} which stochastically drops whole layers during training.
-
 \section{Method}
-
 The proposed hybrid vocabulary pruning method for ModernBERT targets the embedding layer, which constitutes approximately 25\% of the model's parameters.
@@
 \\ \\
-The high-level vocabulary pruning procedure, combining importance-based token pruning with optional semantic clustering for OOV token handling, is formalized in Algorithm~\ref{alg:hybrid_pruning}.
+The high-level vocabulary pruning procedure, is formalized in Algorithm~\ref{alg:pre_ft_pruning}.
 
 \begin{algorithm}[H]
-\caption{Vocabulary Pruning Procedure}
-\label{alg:hybrid_pruning}
+\small
+\caption{Pre-Fine-Tuning Vocabulary Pruning}
+\label{alg:pre_ft_pruning}
 \begin{algorithmic}[1]
-\footnotesize
-\Require Pre-trained model $M$ with vocabulary $V_{orig}$, task dataset $D$, pruning ratio $p$
-\Ensure Pruned model $M'$ with reduced vocabulary
-
-\Function{ExtractTaskVocabulary}{$D, V_{orig}$}
-    \State $V_{task} \gets $ unique tokens in dataset $D$
-    \State $V_{special} \gets $ special tokens from $V_{orig}$ (e.g., [CLS], [SEP])
-    \State \Return $V_{special} \cup V_{task}$ \Comment{Initial vocabulary reduction}
-\EndFunction
-
-\Function{RankByImportance}{$V, D$}
-    \State Calculate importance scores for tokens in $V$ using selected method
-    \State Sort $V$ by descending importance score
-    \State \Return sorted token list
-\EndFunction
-
-\Function{PruneVocabulary}{$V_{ranked}, p, M$}
-    \State $k \gets \lfloor |V_{ranked}| \times (1 - p) \rfloor$ \Comment{Number of tokens to keep}
-    \State $V_{keep} \gets $ first $k$ tokens from $V_{ranked}$
-    \State $V_{prune} \gets V_{ranked} \setminus V_{keep}$ \Comment{Tokens to be pruned}
-
-    \If{using OOV handling}
-        \State Create mapping from pruned tokens $V_{prune}$ to appropriate representatives
-        \State \Return $V_{keep}$, mapping
-    \Else
-        \State \Return $V_{keep}$, null
-    \EndIf
-\EndFunction
-
-\Function{CreateReducedModel}{$M, V_{keep}, mapping$}
-    \State Extract embeddings for tokens in $V_{keep}$ from $M$
-    \State Create reduced embedding matrix $E'$
-    \State Update model with reduced embedding layer using $E'$
-    \If{mapping is not null}
-        \State Store mapping for OOV token handling during inference
-    \EndIf
-    \State \Return updated model
-\EndFunction
-
-\State // Main procedure
-\State $V_{base} \gets $ \Call{ExtractTaskVocabulary}{$D, V_{orig}$}
-\State $V_{ranked} \gets $ \Call{RankByImportance}{$V_{base}, D$}
-\State $V_{keep}$, mapping $\gets $ \Call{PruneVocabulary}{$V_{ranked}, p, M$}
-\State $M' \gets $ \Call{CreateReducedModel}{$M, V_{keep}$, mapping}
-
-\State \Return $M'$ \Comment{Model ready for fine-tuning}
+\Require Pretrained model $M$ with vocab $V$, dataset $D$, pruning ratio $p$
+\Ensure Pruned model $M'$
+\State $\text{stats} \gets \text{compute\_token\_statistics}(D)$
+\State $\text{scores} \gets \text{rank\_tokens}(\text{stats})$
+\State $V_{\text{keep}} \gets \text{special\_tokens}(M)\cup\text{top\_k\_tokens}(\text{scores}, 1 - p)$
+\State $M' \gets \text{rebuild\_model\_with\_vocab}(M, V_{\text{keep}})$
+\State \Return $M'$
 \end{algorithmic}
@@
 Frequency-based approaches naturally prioritize common words like articles and prepositions, while attention-based analysis highlights sentiment-laden terms that are semantically relevant to the classification task.
-\begin{figure*}[t]
-\centering
-\begin{minipage}{0.48\textwidth}
-  \centering
-  \includegraphics[width=\textwidth]{images/wordcloud_frequency.png}
-\end{minipage}%
-\hfill
-\begin{minipage}{0.48\textwidth}
-  \centering
-  \includegraphics[width=\textwidth]{images/wordcloud_attention.png}
-\end{minipage}
-\caption{Comparison of top-ranked tokens for SST-2 task using frequency-based (left) versus attention-based (right) importance metrics.}
-\label{fig:wordcloud-comparison}
-\end{figure*}
-The visualization in Figure \ref{fig:wordcloud-comparison} demonstrates how attention-based pruning better captures task-relevant information, preserving tokens that carry greater semantic weight for the specific downstream task.
+% \begin{figure*}[t]
+% \centering
+% \begin{minipage}{0.48\textwidth}
+%   \centering
+%   \includegraphics[width=\textwidth]{images/wordcloud_frequency.png}
+% \end{minipage}%
+% \hfill
+% \begin{minipage}{0.48\textwidth}
+%   \centering
+%   \includegraphics[width=\textwidth]{images/wordcloud_attention.png}
+% \end{minipage}
+% \caption{Comparison of top-ranked tokens for SST-2 task using frequency-based (left) versus attention-based (right) importance metrics.}
+% \label{fig:wordcloud-comparison}
+% \end{figure*}
+% The visualization in Figure \ref{fig:wordcloud-comparison} demonstrates how attention-based pruning better captures task-relevant information, preserving tokens that carry greater semantic weight for the specific downstream task.
 \\ \\
@@
 
-% While this paper demonstrates the effectiveness of vocabulary pruning as a standalone technique, several promising research directions could further extend its impact:
-
-% \begin{itemize}
-%     \item \textbf{Unified Compression Framework}: Developing an integrated approach that combines vocabulary pruning with encoder-layer compression techniques like LoSparse could provide a more comprehensive model compression solution.
+\section{Future Work}
+TODO (these are just things I thought about)
+\begin{itemize}
+    \item \textbf{Unified Compression Framework}: Developing an integrated approach that combines vocabulary pruning with encoder-layer compression techniques like LoSparse could provide a more comprehensive model compression solution.
 Preliminary experiments suggest these approaches are complementary, with vocabulary pruning excelling at lower compression ratios and encoder pruning maintaining better performance at higher ratios.
 
-%     \item \textbf{Task Adaptation Networks}: Instead of removing pruned token embeddings entirely, future work could explore low-dimensional adapter networks that efficiently transform a small set of base embeddings to approximate the full vocabulary, potentially offering better performance/size tradeoffs.
+    \item \textbf{Task Adaptation Networks}: Instead of removing pruned token embeddings entirely, future work could explore low-dimensional adapter networks that efficiently transform a small set of base embeddings to approximate the full vocabulary, potentially offering better performance/size tradeoffs.
 
-%     \item \textbf{Retrieval Performance Analysis}: Evaluating vocabulary pruning on information retrieval tasks could reveal different optimal pruning strategies, as retrieval models may depend on different token importance distributions than classification tasks.
+    \item \textbf{Retrieval Performance Analysis}: Evaluating vocabulary pruning on information retrieval tasks could reveal different optimal pruning strategies, as retrieval models may depend on different token importance distributions than classification tasks.
+\end{itemize}
 
-%     \item \textbf{Cross-Lingual Applications}: Extending this approach to multilingual models presents an exciting opportunity, as these models typically have much larger vocabularies with significant redundancy across languages.
-Token importance may vary by language, suggesting language-specific pruning strategies.
-
-%     \item \textbf{Dynamic Vocabulary Adaptation}: Rather than static pruning, future systems could implement dynamic vocabulary management that adapts to shifting data distributions or task requirements during deployment.
-
-%     \item \textbf{Theoretical Foundations}: Developing formal methods to establish theoretical bounds on performance loss relative to vocabulary reduction could provide valuable insights for model design and compression strategies.
-% \end{itemize}
-
-% These directions collectively aim to further reduce the computational and memory requirements of large language models while maintaining their impressive capabilities, ultimately making state-of-the-art NLP more accessible across diverse computing environments.
 \newpage
+\onecolumn
 \bibliographystyle{plain}

### 2025-05-17 examples/run_attention_pruning.py
learning rate cola

@@ -44,7 +44,7 @@ PRUNE_PERCENT = 20
 TASK_PARAMS = {
     "cola": {
         "batch_size": BATCH_SIZE, 
-        "learning_rate": 8e-5, 
+        "learning_rate": 5e-5, 
         "epochs": 8,
         "prune_percent": PRUNE_PERCENT,
         "weight_decay": 1e-6

### 2025-05-17 examples/run_clustering_pruning.py

@@ -37,7 +37,7 @@ PRUNE_PERCENT = 20
 TASK_PARAMS = {
     "cola": {
         "epochs": 8,
-        "learning_rate": 8e-5,
+        "learning_rate": 5e-5,
         "batch_size": BATCH_SIZE,
         "weight_decay": 1e-6,
         "prune_percent": PRUNE_PERCENT,

### 2025-05-17 examples/run_frequency_oov_pruning.py

@@ -38,7 +38,7 @@ NUM_CLUSTERS = 50
 TASK_PARAMS = {
     "cola": {
         "epochs": 8,
-        "learning_rate": 8e-5,
+        "learning_rate": 5e-5,
         "batch_size": BATCH_SIZE,
         "weight_decay": 1e-6,
         "prune_percent": PRUNE_PERCENT,

### 2025-05-17 examples/run_frequency_pruning.py

@@ -36,7 +36,7 @@ PRUNE_PERCENT = 20
 TASK_PARAMS = {
     "cola": {
         "epochs": 8,
-        "learning_rate": 8e-5,
+        "learning_rate": 5e-5,
         "batch_size": BATCH_SIZE,
         "weight_decay": 1e-6,
         "prune_percent": PRUNE_PERCENT,

### 2025-05-17 examples/run_importance_oov_pruning.py

@@ -48,7 +48,7 @@ IMPORTANCE_TYPE = 3
 TASK_PARAMS = {
     "cola": {
         "epochs": 8,
-        "learning_rate": 8e-5,
+        "learning_rate": 5e-5,
         "batch_size": BATCH_SIZE,
         "weight_decay": 1e-6,
         "prune_percent": PRUNE_PERCENT,

### 2025-05-17 examples/run_importance_pruning.py

@@ -52,7 +52,7 @@ IMPORTANCE_TYPE = 3
 TASK_PARAMS = {
     "cola": {
         "batch_size": BATCH_SIZE, 
-        "learning_rate": 8e-5, 
+        "learning_rate": 5e-5, 
         "epochs": 8,
         "prune_percent": PRUNE_PERCENT,
         "weight_decay": 1e-6,

### 2025-05-17 examples/run_random_pruning.py

@@ -38,7 +38,7 @@ PRUNE_PERCENT = 20
 TASK_PARAMS = {
     "cola": {
         "epochs": 8,
-        "learning_rate": 8e-5,
+        "learning_rate": 5e-5,
         "batch_size": BATCH_SIZE,
         "weight_decay": 1e-6,
         "prune_percent": PRUNE_PERCENT,

### 2025-05-17 examples/run_train_only_pruning.py

@@ -36,7 +36,7 @@ BATCH_SIZE = 128
 TASK_PARAMS = {
     "cola": {
         "epochs": 8,
-        "learning_rate": 8e-5,
+        "learning_rate": 5e-5,
         "batch_size": BATCH_SIZE,
         "weight_decay": 1e-6,
     },

### 2025-05-18 paper/main.tex
introduction improvement + make related works more concise

@@
 \title{\huge \textbf{Words That Matter:\\ Vocabulary Pruning in ModernBERT}}
-\author{Wout De Rijck, Prof. Dr. Ir. Thomas Demeester, \\ Prof. Dr. Ir. Chris Develder, Dr. Ir. Karel D'Oosterlinck}
+\author{Wout De Rijck, Prof. Dr. Ir. Thomas Demeester, \\ Prof. Dr. Ir. Chris Develder, Karel D'Oosterlinck}
 \date{} % No date
@@
 
-Large Language Models (LLMs) require substantial computational and memory resources, limiting their utility in resource constrained environments. 
+Large language models (LLMs) require substantial computational and memory resources, limiting their utility in resource-constrained environments. 
 ModernBERT is a smaller encoder-only language model that excels at various tasks while being computationally highly efficient. 
@@
 We compare the resulting accuracy with LoSparse, a state-of-the-art gradient-based pruning technique that targets the encoder layers.
-For the same reduction in total parameters of $\sim$20\%, our much simpler vocabulary pruning technique retained on average 97.6\% of ModernBERT's performance across various tasks, while LoSparse retained only 92.9\%.
+For parameter reductions up to $\sim$20\%, our much simpler vocabulary pruning technique outperforms LoSparse, retaining on average 97.6\% of ModernBERT's performance across various tasks, compared to LoSparse's 92.9\%.
 The strong performance of our simple technique indicates that task-specific pruning can meaningfully increase the efficiency of ModernBERT, an already highly efficient model.
@@
 \section{Introduction}
-Efficiency is crucial for modern language models—every additional parameter not only increases storage and memory-bandwidth requirements but also drives up cloud-hosting bills and overall energy consumption.
-Efficiency can be assessed along several dimensions, including inference latency, throughput, energy consumption, parameter count, and memory usage.
-In this work, the focus is on pruning model weights to reduce the memory footprint, as fewer weights lead to lower storage and memory bandwidth requirements, resulting in faster model loading, reduced deployment costs, and improved accessibility for edge devices or resource-constrained environments.
-\\ \\
+Language models need to be efficient for optimal deployment.
+Efficiency can be assessed in several dimensions, including inference latency, throughput, energy consumption, parameter count, and memory usage.
+In this work, we study how to optimally reduce the memory footprint of language models by removing unimportant vocabulary tokens from the embedding layer, resulting in faster model loading, reduced deployment costs, and improved accessibility for edge devices or resource-constrained environments.
+
 \begin{figure}[t]
@@
 \includegraphics[width=\columnwidth]{images/performance_retention.png}
-\caption{Average Performance Retention Across All Tasks for different pruning methods.}
+\caption{Average Performance Retention Across All Tasks for different pruning methods.
+Pruning ratio represents the fraction of total parameters removed from the model, limited to 0.25 as the embedding layer constitutes 25.95\% of all parameters.}
 \label{fig:pruning_retention}
 \end{figure}
-ModernBERT is one of the most efficient models from this perspective. 
-It is a highly optimized, encoder-only transformer that brings together multiple architectural and training innovations—such while maintaining strong downstream performance and minimizing inference cost shortly and the potential impact.
-Given these advances, an open question remains: Can ModernBERT's efficiency be pushed even further by pruning its weight parameters?
-\\ \\
-ModernBERT is not a fully general-purpose model, as it still requires fine-tuning for each specific downstream task.
-This work focuses on task-specific pruning, which leverages that adaptation phase to remove weights irrelevant to the target task, enabling efficient compression with minimal performance loss.
-Unlike general-purpose compression, task-specific pruning can be more precise, resulting in smaller models tailored to specific use cases.
-\\ \\
-This work targets the embedding layer for pruning, motivated by the observation that for many tasks, a small fraction of tokens accounts for the majority of token occurrences. 
-This redundancy allows for effective vocabulary pruning by removing unimportant embeddings with minimal impact on performance. 
-Simple frequency- and TF-IDF-based methods are used to identify and discard low-importance tokens. 
-The approach is compared to LoSparse, demonstrating that embedding-layer pruning can achieve strong efficiency gains with less complexity.
-\\
-The best-performing technique is TF-IDF-based pruning, which removes 77.34\% of the embedding layer while retaining 97.6\% of ModernBERT's original performance.
-This is a remarkable result, as it demonstrates that the vast majority of embeddings can be discarded without significantly affecting task accuracy.
-It highlights the extreme redundancy in the embedding layer and the effectiveness of simple, task-aware pruning strategies.
+
+From this perspective, ModernBERT stands out as one of the most lightweight and optimized encoder-only transformers available.
+It integrates several architectural and training innovations, successfully maintaining robust downstream performance while minimizing inference costs.
+Despite these advancements, a critical question remains open: Can we further enhance ModernBERT's efficiency by pruning parameters?
+
+This question is particularly significant for two reasons.
+First, ModernBERT's inherent efficiency and its strong potential for large-scale deployment mean that optimal pruning strategies on top of this model could yield exceptionally efficient models.
+Second, existing research into pruning techniques have predominantly targeted decoder-only LLMs \cite{namburi2023llm} due to their widespread popularity, leaving encoder-only models relatively underexplored.
+
+Specifically, we explore task-specific weight pruning, as ModernBERT is intended to be fine-tuned on specific tasks for optimal behavior (unlike more general-purpose decoder-only models that can utilize in-context learning).
+We hypothesize that simple task-specific pruning of the embedding matrix can be highly efficient, as many tokens can be irrelevant for any specific downstream task and the embedding layers counts for a significant portion ($\sim$25\% of the total parameters).
+
+We explore a series of simple vocabulary pruning heuristics on a range of Natural Language Processing tasks including linguistic acceptability, sentiment analysis, paraphrase detection, semantic similarity, question answering, and natural language inference, as covered by the GLUE benchmark.
+For any task, we first apply this pruning to the base ModernBERT model and subsequently fine-tune the pruned model.
+For any given pruning ratio, we measure the accuracy retention compared to simply fine-tuning the model.
+We compare these vocabulary pruning techniques with LoSparse, a state-of-the-art gradient-based pruning technique that targets the encoder layers.
+
+As shown in Figure \ref{fig:pruning_retention}, TF-IDF-based vocabulary pruning outperforms encoder-layer pruning (LoSparse) for compression ratios up to 20\%, maintaining 97.6\% of ModernBERT's original performance while removing 77.34\% of embedding parameters.
+This demonstrates that substantial efficiency gains can be achieved with virtually no performance loss using a simple, offline pruning technique.
+Beyond 20\% compression, performance drops sharply as we approach the embedding layer's parameter limit (25.95\% of total model parameters), while LoSparse maintains more stable performance at higher ratios.
+Unlike LoSparse, which requires computationally expensive gradient-based optimization, our vocabulary pruning operates as a straightforward pre-fine-tuning step with minimal overhead, making it particularly suitable for resource-constrained environments requiring moderate model compression.
 
@@
 
-% \subsection{Encoder-Layer Pruning}
-% Although this paper focuses on vocab/embedding pruning, many complementary pruning methods work at the encoder layers.
-A classic example is \textbf{attention-head pruning}: Michel et al.~\cite{michel2019sixteen} discovered that a large percentage of attention heads can be removed at test time without significantly impacting performance.
-Voita et al.~\cite{voita2019analyzing} likewise pruned 38 of 48 encoder heads in an NMT transformer (removing $\sim$79\% of heads) with only a 0.15 BLEU loss.
-% \\ \\
-% In practice, toolkit frameworks (like TextPruner) often support head and FFN pruning alongside vocab pruning.
-Other methods include \textit{movement pruning}~\cite{sanh2020movement} that gradually zeros out small weights during fine-tuning, or \textit{LayerDrop}~\cite{fan2020layerdrop} which stochastically drops whole layers during training.
-
 \subsection{Encoder-Layer Pruning}
@@
 During fine-tuning, LoSparse uses first-order sensitivity scores—computed as the product of each weight and its gradient—to iteratively prune low-importance parameters.
-\\
+
 Several alternative encoder pruning techniques have been proposed.
@@
 \textbf{LayerDrop}~\cite{fan2020layerdrop} introduces structured dropout, allowing entire transformer layers to be stochastically removed during training and optionally pruned at inference.
-\\ \\
+
 These encoder-layer pruning methods differ in their granularity and strategy—targeting individual weights, attention heads, or full layers.
@@
 
-\subsection{Embedding-Layer Pruning and Compression}
+\subsection{Embedding-Layer Compression}
 Pruning the embedding layer often means removing or compressing rows (token vectors) to shrink this heavy component.
-\textbf{Partial Embedding Matrix Adaptation (PEMA)} by Bousquet et al.~\cite{bousquet2023pema} is one systematic approach: they first note that fine-tuning datasets typically use only a small fraction of the full vocabulary.
-PEMA therefore builds a partial embedding matrix containing only tokens seen in the task data, training on that smaller matrix.
-This saves GPU memory during fine-tuning without altering task accuracy.
-After fine-tuning, the updated embeddings are merged back into the full matrix so the model structure remains unchanged.
-Experiments show large memory reductions: e.g., BERT$_{\text{BASE}}$'s embedding matrix can shrink by $\sim$47\% on average (and RoBERTa by $\sim$52\%) when only task-relevant tokens are kept.
-Multilingual models see even more savings (mBERT $\sim$44\% embedding reduction, XLM-R $\sim$64\%) because their vocabularies are huge.
-Crucially, PEMA reports no drop in downstream performance, making it a practical way to prune embeddings during training.
-\\ \\
+\textbf{Partial Embedding Matrix Adaptation (PEMA)}~\cite{bousquet2023pema} temporarily removes unused tokens during fine-tuning and restores them afterward.
+It builds a partial embedding matrix containing only task-relevant tokens, reducing GPU memory requirements without compromising accuracy.
+Experiments show significant memory reductions: BERT$_{\text{BASE}}$'s embedding matrix shrinks by $\sim$47\%.
+PEMA maintains full performance while making embedding pruning practical during training.
+
 Another approach is \textbf{embedding compression} via factorization or hashing.
-Wang et al.~\cite{wang2023lighttoken} propose \textbf{LightToken}, which compresses the entire embedding matrix using low-rank SVD plus a binary residual autoencoder.
-They note token embeddings are highly redundant (the embedding takes $\sim$21--31\% of BERT/RoBERTa's size).
-LightToken first applies a rank-$k$ SVD to capture most variation, then learns compact hash codes to reconstruct the residual.
-In experiments on GLUE, LightToken achieves very high compression (e.g., 25$\times$ smaller embeddings) while preserving accuracy.
-For BERT$_{\text{BASE}}$, a 25$\times$ embedding compression yields an average GLUE score of 82.9 (baseline 83.0).
-\\ \\
+\textbf{LightToken}~\cite{wang2023lighttoken} compresses the embedding matrix using low-rank SVD combined with a binary residual autoencoder.
+They note that token embeddings are highly redundant (comprising $\sim$21--31\% of BERT/RoBERTa's size).
+LightToken captures primary variations through rank-$k$ SVD, then uses compact hash codes for residuals.
+This achieves 25$\times$ smaller embeddings while maintaining accuracy—for BERT$_{\text{BASE}}$, it produces an average GLUE score of 82.9 versus the baseline 83.0.
+
 These methods show that even structured low-rank or quantized pruning of embeddings can yield massive space savings with minimal impact on performance.
-\\ \\
+
 Overall, embedding-layer pruning techniques exploit the fact that many token vectors (especially for rare subwords) contribute little to end-task accuracy.
@@
 
-\subsection{Vocabulary Pruning Techniques}
+\subsubsection{Vocabulary Pruning Techniques}
 Vocabulary pruning removes rarely used or task-irrelevant tokens to shrink the model's vocabulary (and its embedding matrix).
@@
 In "Load What You Need," they drop languages from mBERT's vocabulary and obtain up to a 45\% reduction in total parameters with comparable accuracy on XNLI.
-\\ \\
-More recently, Nair et al.~\cite{nair2023blade} proposed \textbf{BLADE}: they prune a bilingual mBERT vocabulary to include only tokens from the query and document languages for cross-lingual IR, then apply a small intermediate pretraining step.
-This pruned model (keeping only needed embeddings) reduced parameters by $\sim$36.5\% versus full mBERT and sped up training by $\sim$30\% and inference by $\sim$55\%, with minimal loss in retrieval quality.
-\\ \\
+
+More recently, Nair et al.~\cite{nair2023blade} proposed \textbf{BLADE}, extending the implicit pruning in \textbf{SPLADE-X}~\cite{formal2023spladex} (which restricted output to query-language tokens).
+BLADE explicitly builds a pruned bilingual model containing only tokens from query and document languages, reducing parameters by $\sim$36.5\% versus full mBERT while speeding up training by $\sim$30\% and inference by $\sim$55\%, with minimal retrieval quality loss.
+
 Dorkin et al.~\cite{dorkin2025estonian} also evaluated pruning for an Estonian adaptation: pruning all tokens unused by Estonian led to no degradation on NER, whereas retraining a new tokenizer hurt performance.
 These studies consistently find that deleting infrequently seen subwords yields large size savings with little accuracy drop.
-\\ \\
-TextPruner by Shen et al.~\cite{shen2022textpruner} automates vocabulary pruning: it scans a corpus and removes any tokenizer token not present in the text, deleting its row in the embedding matrix.
-This simple "drop unused tokens" mode can cut model size and speed up training/inference on that domain.
-\\ \\
-In sparse-retrieval models like SPLADE, vocabulary pruning is implicit.
-SPLADE-X~\cite{formal2023spladex} (multi-lingual) restricted output to query-language tokens only, effectively pruning all other vocab.
-BLADE extends this idea by explicitly building a pruned bilingual model with only the union of query and document subwords, discarding other embeddings entirely.
+
+\textbf{TextPruner} by Shen et al.~\cite{shen2022textpruner} uses a simple "drop unused tokens" technique: it scans a corpus and removes any tokenizer token not present in the text, deleting its row in the embedding matrix.
+This approach can cut model size and speed up training/inference on that domain.
 \\ \\
@@
 
+\newpage
 \section{Method}
@@
 The resulting pruned model maintains its standard inference pipeline while benefiting from reduced memory requirements.
-\\ \\
+
 The high-level vocabulary pruning procedure, is formalized in Algorithm~\ref{alg:pre_ft_pruning}.
@@
 Unlike frequency-based approaches that simply count occurrences, attention-based pruning preserves tokens that meaningfully contribute to the model's decision-making process.
-\\ \\
-As an illustrative example, the important tokens identified in the SST-2 dataset differ markedly between methods.
+
+For example, the important tokens identified in the SST-2 dataset differ markedly between methods.
 Frequency-based approaches naturally prioritize common words like articles and prepositions, while attention-based analysis highlights sentiment-laden terms that are semantically relevant to the classification task.
@@
 This method prioritizes tokens that appear frequently in specific documents but are rare across the corpus—capturing task-specific terminology while filtering out ubiquitous tokens that carry less semantic value.
-\\ \\
+
 The TF-IDF approach is particularly effective for tasks with distinct document categories (like sentiment classification or topic detection) where certain tokens strongly differentiate between classes.
@@
 The method is particularly effective for technical vocabularies where semantically related terms form natural clusters in the embedding space.
-\\ \\
+
 The number of clusters (k) provides a tunable parameter to balance compression rate against semantic precision—larger values of k preserve more semantic distinctiveness but reduce the overall pruning benefit.
@@
 \section{Experiments}
-Having established a theoretical foundation for the vocabulary pruning approach, this section presents a comprehensive evaluation designed to answer several key questions: Can selective vocabulary reduction maintain model performance while significantly decreasing parameter count?
+After developing several algorithmic approaches to vocabulary pruning, this section presents a comprehensive evaluation designed to answer several key questions: Can selective vocabulary reduction maintain model performance while significantly decreasing parameter count?
 How do different pruning strategies compare on various natural language understanding tasks?
 Does the pre-fine-tuning approach that avoids post-pruning recovery fine-tuning deliver practical benefits in real-world deployment scenarios?
-Through careful experimentation across diverse tasks, the results demonstrate not only the theoretical soundness of the approach but also its practical advantages in creating more efficient transformer models.
+Through careful experimentation across diverse tasks, the results demonstrate not only the effectiveness of these algorithms but also their practical advantages in creating more efficient transformer models.
 
@@
 GLUE consists of eight diverse natural language understanding tasks categorized into three groups: single-sentence tasks (SST-2, CoLA), similarity and paraphrase tasks (MRPC, STS-B, QQP), and natural language inference tasks (MNLI, QNLI, RTE).
-Performance metrics vary by task: classification tasks report accuracy; CoLA reports Matthew's correlation coefficient; and STS-B reports Pearson correlation.
-\\ \\
+Performance metrics vary by task: classification tasks report accuracy; CoLA reports Matthew's correlation coefficient, and STS-B reports Pearson correlation.
+
 Table \ref{tab:token_statistics} presents a detailed analysis of token distributions across all GLUE tasks.
@@
 The table also quantifies the overlap between TF-IDF and frequency-based token selection, showing substantial differences (38-60\% overlap), which explains the performance variations between these approaches.
-\\ \\
+
 Table \ref{tab:vocab_overlap} examines train-test vocabulary overlap, quantifying the percentage of out-of-vocabulary (OOV) tokens in test sets.
@@
 \subsection{Results}
-Table \ref{tab:results} presents the comprehensive evaluation across all GLUE tasks for the different pruning methods. Several key findings emerge:
-\\ \\
+Table \ref{tab:results} presents the comprehensive evaluation across all GLUE tasks for the different pruning methods. Several key findings emerge.
+
 First, vocabulary pruning alone achieves substantial parameter reduction (20.02-23.27\%) with minimal performance impact.
 The TF-IDF + OOV approach performs exceptionally well, maintaining 97.6\% of the original model's average performance while reducing parameters by over 20\%.
-\\ \\
+
 Second, the inclusion of OOV handling mechanisms significantly improves results compared to their non-OOV counterparts.
 For instance, TF-IDF + OOV outperforms standard TF-IDF by 2.4 percentage points on average, with particularly strong improvements on CoLA (9.5 points) and RTE (6.1 points).
-\\ \\
-Third, pruning technique effectiveness varies by task type.
+
+Third, the effectiveness of pruning varies across tasks.
 Attention-based pruning excels on single-sentence tasks (SST-2, CoLA) and paraphrase detection (MRPC), while TF-IDF-based methods perform better on more complex reasoning tasks (MNLI, QNLI, RTE).
-\\ \\
+
 Fourth, as illustrated in Figure \ref{fig:pruning_ratio}, vocabulary pruning demonstrates superior performance and efficiency compared to encoder-layer pruning (LoSparse) at lower pruning ratios (5-20\%).
@@
 Beyond the 20\% threshold, performance decreases sharply for vocabulary pruning while LoSparse maintains more stable performance at higher compression rates.
-For applications requiring moderate compression with minimal performance impact, vocabulary pruning provides an optimal approach, while more aggressive compression benefits from combining both techniques.
-\\ \\
+Vocabulary pruning provides an optimal approach for applications requiring moderate compression with minimal performance impact, while more aggressive compression could benefit from combining both techniques.
+
 The benchmark results in Table \ref{tab:benchmark_results} further quantify the practical benefits, showing a 20\% reduction in storage size and 14.83\% reduction in GPU memory requirements with negligible impact on inference time (+1.33\%).
@@
 \section{Future Work}
+TODO (these are just things I thought about)
 
-\section{Future Work}
-TODO (these are just things I thought about)
 \begin{itemize}
@@
     \item \textbf{Retrieval Performance Analysis}: Evaluating vocabulary pruning on information retrieval tasks could reveal different optimal pruning strategies, as retrieval models may depend on different token importance distributions than classification tasks.
+
 \end{itemize}
-
 \newpage
-\onecolumn
 \bibliographystyle{plain}

### 2025-05-18 paper/main.tex
next iteration, more concise

@@
 
-\title{\huge \textbf{Words That Matter:\\ Vocabulary Pruning in ModernBERT}}
-\author{Wout De Rijck, Prof. Dr. Ir. Thomas Demeester, \\ Prof. Dr. Ir. Chris Develder, Karel D'Oosterlinck}
+\setcounter{secnumdepth}{0}
+
+\title{\vspace{-1.5cm}\huge \textbf{Words That Matter:\\ Vocabulary Pruning in ModernBERT}}
+\author{
+Wout De Rijck\\[0.5em]
+\small{Supervisors:} \normalsize Prof. Dr. Ir. Thomas Demeester, \normalsize Prof. Dr. Ir. Chris Develder\\
+\small{Councellor:} \normalsize Dr. Ir. Karel D'Oosterlinck
+}
 \date{} % No date
@@
 
-\begin{abstract}
-
-Large language models (LLMs) require substantial computational and memory resources, limiting their utility in resource-constrained environments. 
+\textbf{Abstract:} Large language models (LLMs) require substantial computational and memory resources, limiting their utility in resource-constrained environments. 
 ModernBERT is a smaller encoder-only language model that excels at various tasks while being computationally highly efficient. 
@@
 
-\end{abstract}
+\textbf{Keywords:} vocabulary pruning, ModernBERT, encoder-only model, model compression, embedding-layer, task-specific pruning, TF-IDF
 
@@
 Most modern LLM compression approaches operate under post-training settings, eliminating the need for resource-intensive retraining.
-Model compression strategies for LLMs can be categorized into four primary methods:
+Model compression strategies for LLMs can be categorized into four primary methods \cite{wan2023efficient}:
 
@@
 
-\newpage
 \section{Method}
@@
 This approach operates on the principle that rarely used tokens contribute less to model performance.
-The technique-specific algorithm steps are:
 
-\begin{enumerate}
-    \item Count occurrences of each token in the task-specific dataset
-    \item Sort tokens by frequency (most common first)
-\end{enumerate}
-\subsubsection{Attention-Based}
-This method leverages the attention mechanism of transformer models to identify important tokens.
-The key insight is that tokens receiving higher attention during inference are those the model relies on most heavily when making predictions, providing a direct measure of token importance from the model's perspective.
-Unlike frequency-based approaches that simply count occurrences, attention-based pruning preserves tokens that meaningfully contribute to the model's decision-making process.
-
-For example, the important tokens identified in the SST-2 dataset differ markedly between methods.
+%For example, the important tokens identified in the SST-2 dataset differ markedly between methods.
 Frequency-based approaches naturally prioritize common words like articles and prepositions, while attention-based analysis highlights sentiment-laden terms that are semantically relevant to the classification task.
@@
 % The visualization in Figure \ref{fig:wordcloud-comparison} demonstrates how attention-based pruning better captures task-relevant information, preserving tokens that carry greater semantic weight for the specific downstream task.
+
+\subsubsection{Attention-Based}
+This method leverages the attention mechanism of transformer models to identify important tokens.
+The key insight is that tokens receiving higher attention during inference are those the model relies on most heavily when making predictions, providing a direct measure of token importance from the model's perspective.
+Unlike frequency-based approaches that simply count occurrences, attention-based pruning preserves tokens that meaningfully contribute to the model's decision-making process.
 \\ \\
@@
 \end{enumerate}
-For this approach, three normalization variants were implemented and evaluated:
-\begin{itemize}
-    \item Non-normalized TF-IDF: Raw scores that heavily prioritize rare but task-specific tokens
-    \item L1-normalized TF-IDF: Scores normalized by sum, balancing frequency and discriminative power
-    \item L2-normalized TF-IDF: Scores normalized by Euclidean norm, the standard approach in information retrieval that prevents outlier tokens from dominating
-\end{itemize}
+For this approach, three normalization variants were implemented: Non-normalized TF-IDF (raw scores), L1-normalized TF-IDF (scores normalized by sum), and L2-normalized TF-IDF (scores normalized by Euclidean norm, the standard approach in information retrieval that prevents outlier tokens from dominating).
+Only the L2-normalized variant was evaluated in our experiments.
 \subsection{Out-of-Vocabulary Token Handling}
@@
 \section{Conclusion}
+In this work, we designed and evaluated a series of vocabulary pruning techniques for the ModernBERT language model to reduce model size while maintaining task performance.
 
-% This paper introduced a novel vocabulary pruning approach for transformer-based language models that eliminates the need for post-pruning recovery fine-tuning while delivering substantial parameter reduction with minimal performance impact.
-The work highlights the embedding layer as a critical yet often overlooked component for model compression, demonstrating that intelligent token selection can maintain 97.6\% of original model performance while reducing total parameters by 20-23\%.
+We found that simple vocabulary pruning techniques outperform encoder-layer pruning for moderate compression ratios.
+Our TF-IDF-based method with OOV handling achieved 97.6\% of the original model's performance across GLUE tasks while reducing total parameters by 20.02\%, compared to LoSparse's 92.9\% retention at the same compression ratio.
+This resulted in practical benefits of 14.83\% reduction in GPU memory requirements and 20\% less storage space.
 
-% The key contributions of this research are:
-
-% \begin{itemize}
-%     \item A vocabulary pruning methodology that operates as a pre-fine-tuning optimization step, eliminating the need for post-pruning recovery fine-tuning typically associated with model compression.
-
-%     \item A systematic comparison of token importance estimation techniques, revealing that TF-IDF with OOV handling provides the most robust performance across diverse NLP tasks.
-
-%     \item Empirical evidence that vocabulary pruning outperforms encoder-based pruning at lower pruning ratios (5-20\%), offering a more efficient pathway for moderate model compression.
-
-%     \item Practical deployment benefits including 20\% reduction in storage size and 14.83\% reduction in GPU memory with no inference time penalty.
-% \end{itemize}
-
-% The results challenge the conventional wisdom that encoder-layer pruning should be prioritized over vocabulary reduction.
-Instead, the findings demonstrate that a hybrid approach—focusing first on vocabulary pruning for initial efficiency gains and then applying encoder pruning for more aggressive compression—offers the most balanced pathway to efficient transformer models.
-
-% The broader implications of this work extend beyond the specific techniques presented.
-By eliminating the need for expensive post-pruning fine-tuning, the presented approach democratizes model compression, making it accessible even for organizations with limited computational resources.
-This could accelerate the adoption of transformer models in resource-constrained environments such as mobile devices, edge computing platforms, and regions with limited technological infrastructure.
-
-% Future work could explore dynamic vocabulary pruning that adapts to changing data distributions, application of these techniques to multilingual models where vocabulary redundancy is even more pronounced, and integration with quantization and knowledge distillation for more comprehensive compression pipelines.
-As large language models continue to grow in size and capability, efficient adaptation methods like vocabulary pruning will become increasingly critical for bridging the gap between research advances and practical deployment.
+These results challenge the assumption that encoder pruning should precede vocabulary reduction and highlight the significant redundancy within embedding layers of pre-trained models, providing an efficient path to model compression for resource-constrained environments.
 
 \section{Future Work}
-TODO (these are just things I thought about)
+Future research could explore an integrated approach combining vocabulary pruning with encoder-layer compression techniques like LoSparse, as preliminary experiments suggest these approaches are complementary.
+We hypothesize that pruning up to a certain threshold using this combined approach will yield better results than using either vocabulary pruning or encoder-layer pruning alone.
 
-\begin{itemize}
-    \item \textbf{Unified Compression Framework}: Developing an integrated approach that combines vocabulary pruning with encoder-layer compression techniques like LoSparse could provide a more comprehensive model compression solution.
-Preliminary experiments suggest these approaches are complementary, with vocabulary pruning excelling at lower compression ratios and encoder pruning maintaining better performance at higher ratios.
+Additionally, evaluating vocabulary pruning on information retrieval tasks may reveal different optimal pruning strategies, as retrieval models likely depend on different token importance distributions than classification tasks.
 
-    \item \textbf{Task Adaptation Networks}: Instead of removing pruned token embeddings entirely, future work could explore low-dimensional adapter networks that efficiently transform a small set of base embeddings to approximate the full vocabulary, potentially offering better performance/size tradeoffs.
-
-    \item \textbf{Retrieval Performance Analysis}: Evaluating vocabulary pruning on information retrieval tasks could reveal different optimal pruning strategies, as retrieval models may depend on different token importance distributions than classification tasks.
-
-\end{itemize}
-\newpage
 \bibliographystyle{plain}

### 2025-05-19 paper/main.tex
make paper more compact

@@
 
-\title{\vspace{-1.5cm}\huge \textbf{Words That Matter:\\ Vocabulary Pruning in ModernBERT}}
+\title{\vspace{-1.5cm}\huge Words That Matter:\\ Vocabulary Pruning in ModernBERT}
 \author{
@@
 We compare the resulting accuracy with LoSparse, a state-of-the-art gradient-based pruning technique that targets the encoder layers.
-For parameter reductions up to $\sim$20\%, our much simpler vocabulary pruning technique outperforms LoSparse, retaining on average 97.6\% of ModernBERT's performance across various tasks, compared to LoSparse's 92.9\%.
+For parameter reductions up to $\sim$20\%, our much simpler vocabulary pruning technique outperforms LoSparse, retaining up to 97.6\% of ModernBERT's performance across various tasks, compared to LoSparse's 92.9\%.
 The strong performance of our simple technique indicates that task-specific pruning can meaningfully increase the efficiency of ModernBERT, an already highly efficient model.
@@
 It integrates several architectural and training innovations, successfully maintaining robust downstream performance while minimizing inference costs.
-Despite these advancements, a critical question remains open: Can we further enhance ModernBERT's efficiency by pruning parameters?
+Despite these advancements, a critical question remains open: can we further enhance ModernBERT's efficiency by pruning parameters?
 
@@
 As shown in Figure \ref{fig:pruning_retention}, TF-IDF-based vocabulary pruning outperforms encoder-layer pruning (LoSparse) for compression ratios up to 20\%, maintaining 97.6\% of ModernBERT's original performance while removing 77.34\% of embedding parameters.
-This demonstrates that substantial efficiency gains can be achieved with virtually no performance loss using a simple, offline pruning technique.
+This shows that substantial efficiency gains can be achieved with virtually no performance loss using a simple, offline pruning technique.
 Beyond 20\% compression, performance drops sharply as we approach the embedding layer's parameter limit (25.95\% of total model parameters), while LoSparse maintains more stable performance at higher ratios.
@@
 
+Our results indicate that the memory footprint of light-weight models, such as ModernBERT, can still be significantly improved.
+Additionally, the superior performance of our simple vocabulary pruning methods highlights the need to further develop pruning methods.
+We open-source all our code and pruned models.
+
 \section{Related work}
 Model compression techniques enhance the efficiency of large language models (LLMs) by reducing their size and computational requirements.
-Most modern LLM compression approaches operate under post-training settings, eliminating the need for resource-intensive retraining.
-Model compression strategies for LLMs can be categorized into four primary methods \cite{wan2023efficient}:
 
-\begin{enumerate}
-    \item Quantization: Reduces the numerical precision of model weights, decreasing memory footprint while maintaining reasonable performance.
-    \item Parameter pruning: Eliminates redundant or less important connections and neurons to create sparser models with fewer parameters.
-    \item Low-rank approximation: Decomposes weight matrices into lower-dimensional representations that capture essential information while requiring fewer parameters.
-    \item Knowledge distillation: Transfers knowledge from larger teacher models to smaller student models, enabling competitive performance with reduced architecture size.
-\end{enumerate}
-These approaches are complementary and can be combined to achieve optimal compression from different perspectives. 
-This work focuses on pruning techniques to reduce model size while preserving performance.
-For ModernBERT, an encoder-only model, this research examines specific pruning methods for encoder layers and separate techniques for the embedding matrix.
+These techniques can be broadly categorized into four primary methods:
+(i) \textbf{quantization}, which reduces the numerical precision of model weights, decreasing memory footprint;
+(ii) \textbf{parameter pruning}, which eliminates redundant or less important connections and neurons to create sparser models with fewer parameters;
+(iii) \textbf{low-rank approximation}, which decomposes weight matrices into lower-dimensional representations that capture essential information while requiring fewer parameters; and
+(iv) \textbf{knowledge distillation}, which transfers knowledge from larger teacher models to smaller student models, enabling competitive performance with reduced architecture size~\cite{wan2023efficient}.
 
-\subsection{Encoder-Layer Pruning}
+%can you give an example here?
+%Misschien niet nodig voor de paper maar heel goed om je expertise te tonen in de full thesis
+These approaches are complementary and can be combined to achieve optimal compression.
+This work focuses on structured parameter pruning techniques to reduce model size while preserving performance.
+The rest of this section is dedicated to enumerating the most important parameter pruning techniques.
 
-Encoder-layer pruning methods aim to remove redundant parameters within the transformer encoder to reduce model size while retaining accuracy.
-A representative example is \textbf{LoSparse}~\cite{li2023losparse}, which decomposes each weight matrix into a low-rank component and a sparse residual.
-The low-rank term captures the dominant shared subspace, while the sparse term accounts for residual variation.
-During fine-tuning, LoSparse uses first-order sensitivity scores—computed as the product of each weight and its gradient—to iteratively prune low-importance parameters.
+% In de thesis kan je veel content "winnen" door LoSparse volledig uit te leggen, kan dan ook beknopt in de paper komen.
+% Maar vooral belangrijk voor de thesis om de equations uit te schrijven als je content nodig hebt
+\paragraph{Encoder-Layer Pruning methods}~remove less critical parameters within the transformer encoder to reduce model size while maintaining acceptable accuracy.
+A representative state-of-the-art method is LoSparse~\cite{li2023losparse}, which decomposes each weight matrix into a low-rank component and a sparse residual.
+During fine-tuning, LoSparse calculates the importance of each weight by multiplying it with its gradient.
 
-Several alternative encoder pruning techniques have been proposed.
-\textbf{Attention-head pruning}~\cite{michel2019sixteen} removes entire attention heads with minimal performance loss.
-\textbf{Movement pruning}~\cite{sanh2020movement} applies gradient-guided masking to gradually zero out low-importance weights during fine-tuning.
-\textbf{LayerDrop}~\cite{fan2020layerdrop} introduces structured dropout, allowing entire transformer layers to be stochastically removed during training and optionally pruned at inference.
+Attention-head pruning~\cite{michel2019sixteen} removes entire attention heads.
+Movement pruning~\cite{sanh2020movement} applies gradient-guided masking to gradually zero out low-importance weights during fine-tuning.
+LayerDrop~\cite{fan2020layerdrop} introduces structured dropout, allowing entire transformer layers to be stochastically removed during training and optionally pruned at inference.
 
@@
 
-\subsection{Embedding-Layer Compression}
-Pruning the embedding layer often means removing or compressing rows (token vectors) to shrink this heavy component.
-\textbf{Partial Embedding Matrix Adaptation (PEMA)}~\cite{bousquet2023pema} temporarily removes unused tokens during fine-tuning and restores them afterward.
-It builds a partial embedding matrix containing only task-relevant tokens, reducing GPU memory requirements without compromising accuracy.
-Experiments show significant memory reductions: BERT$_{\text{BASE}}$'s embedding matrix shrinks by $\sim$47\%.
-PEMA maintains full performance while making embedding pruning practical during training.
+\paragraph{Embedding-Layer Pruning methods}
+target the embedding layer, which can comprise up to 20-30\% of a model's parameters.
+These approaches can be broadly categorized into two main strategies: (1) vocabulary pruning, which removes entire token entries, and (2) embedding compression, which retains all tokens but reduces their dimensional representation.
 
-Another approach is \textbf{embedding compression} via factorization or hashing.
-\textbf{LightToken}~\cite{wang2023lighttoken} compresses the embedding matrix using low-rank SVD combined with a binary residual autoencoder.
-They note that token embeddings are highly redundant (comprising $\sim$21--31\% of BERT/RoBERTa's size).
-LightToken captures primary variations through rank-$k$ SVD, then uses compact hash codes for residuals.
-This achieves 25$\times$ smaller embeddings while maintaining accuracy—for BERT$_{\text{BASE}}$, it produces an average GLUE score of 82.9 versus the baseline 83.0.
+\textbf{Embedding compression} reduces the memory footprint of the embedding matrix through dimensionality reduction techniques.
+LightToken~\cite{wang2023lighttoken} is a model- and task-agnostic framework for compressing the token embedding layers.
+It uses a combination of low-rank approximation, a residual binary autoencoder, and a novel compression loss to drastically reduce embedding size.
 
-These methods show that even structured low-rank or quantized pruning of embeddings can yield massive space savings with minimal impact on performance.
+\textbf{Vocabulary Pruning Techniques} are a specialized form of embedding pruning that remove rarely used or task-irrelevant tokens to shrink the model's vocabulary.
+These methods simply delete rows in the embedding matrix, making them straightforward to implement.
 
-Overall, embedding-layer pruning techniques exploit the fact that many token vectors (especially for rare subwords) contribute little to end-task accuracy.
-By either dropping unused rows (PEMA/TextPruner) or factoring and quantizing embeddings (LightToken), these methods reduce $\sim$25--50\% of model parameters in the embedding matrix, enabling lighter models in practice.
+The simplest approach is "drop unused tokens." Both TextPruner~\cite{shen2022textpruner} and Partial Embedding Matrix Adaptation (PEMA;~\cite{bousquet2023pema}) scan a corpus and remove tokenizer tokens not present in the target text.
+TextPruner permanently deletes these entries, while PEMA temporarily removes them during fine-tuning and restores them afterward.
+These methods can reduce BERT$_{\text{BASE}}$'s embedding matrix by up to 47\% without compromising performance.
 
-\subsubsection{Vocabulary Pruning Techniques}
-Vocabulary pruning removes rarely used or task-irrelevant tokens to shrink the model's vocabulary (and its embedding matrix).
-Early work by Abdaoui et al.~\cite{abdaoui2020load} showed that most parameters of multilingual BERT lie in the embedding layer, so trimming unused language tokens can dramatically cut size.
-In "Load What You Need," they drop languages from mBERT's vocabulary and obtain up to a 45\% reduction in total parameters with comparable accuracy on XNLI.
+For multilingual models, language-specific pruning has proven effective.
+Abdaoui et al.~\cite{abdaoui2020load} showed that trimming unused language tokens from mBERT can achieve a 45\% parameter reduction with comparable accuracy on XNLI.
+Similarly, BLADE~\cite{nair2023blade} builds pruned bilingual models containing only tokens from target languages, reducing parameters by $\sim$36.5\% versus full mBERT while speeding up training by $\sim$30\% and inference by $\sim$55\%.
 
-More recently, Nair et al.~\cite{nair2023blade} proposed \textbf{BLADE}, extending the implicit pruning in \textbf{SPLADE-X}~\cite{formal2023spladex} (which restricted output to query-language tokens).
-BLADE explicitly builds a pruned bilingual model containing only tokens from query and document languages, reducing parameters by $\sim$36.5\% versus full mBERT while speeding up training by $\sim$30\% and inference by $\sim$55\%, with minimal retrieval quality loss.
-
-Dorkin et al.~\cite{dorkin2025estonian} also evaluated pruning for an Estonian adaptation: pruning all tokens unused by Estonian led to no degradation on NER, whereas retraining a new tokenizer hurt performance.
-These studies consistently find that deleting infrequently seen subwords yields large size savings with little accuracy drop.
-
-\textbf{TextPruner} by Shen et al.~\cite{shen2022textpruner} uses a simple "drop unused tokens" technique: it scans a corpus and removes any tokenizer token not present in the text, deleting its row in the embedding matrix.
-This approach can cut model size and speed up training/inference on that domain.
-\\ \\
-Across tasks and languages, pruned-vocabulary models often match full models.
-For example, after pruning, Nair et al.~\cite{nair2023blade} report retrieval effectiveness on par with larger baselines, while achieving much faster indexing.
-Dorkin et al.~\cite{dorkin2025estonian} explicitly note "vocabulary pruning has no observable negative effect on the downstream task" (Estonian NER).
-These findings suggest vocabulary pruning is an effective way to tailor large models to specific domains or languages.
+These findings consistently demonstrate that vocabulary pruning is an effective way to tailor models to specific domains or languages with minimal performance impact.
 
 \section{Method}
-The proposed hybrid vocabulary pruning method for ModernBERT targets the embedding layer, which constitutes approximately 25\% of the model's parameters.
+We propose a task-specific vocabulary pruning method for ModernBERT that targets the embedding layer, which accounts for approximately 25\% of the model's parameters.
 This approach is based on the observation that tokens in a vocabulary have varying importance for downstream tasks, and that certain tokens can be selectively removed with minimal impact on performance.
@@
 In contrast to methods that require resource-intensive post-pruning fine-tuning, the proposed vocabulary pruning approach operates as a pre-fine-tuning offline optimization step in the model adaptation pipeline.
-The standard workflow for adapting ModernBERT to a downstream task involves taking the pre-trained base model and fine-tuning it directly on the task data.
-This research modifies this workflow as follows:
 
-\begin{enumerate}
-    \item Task data analysis: The downstream task dataset is processed to extract token statistics.
-    \item Token importance calculation: Based on the analysis, tokens are ranked by their importance using one of several methods (frequency, TF-IDF, attention patterns, etc.).
-    \item Vocabulary pruning: The embedding layer is pruned by removing less important tokens according to the calculated rankings.
-    \item Optional OOV handling: For pruned tokens, semantic clustering may be applied to create mappings to representative tokens.
-    \item Model fine-tuning: The pruned model is then fine-tuned on the downstream task using standard procedures.
-\end{enumerate}
-All pruning techniques follow these common steps:
-\begin{enumerate}
-    \item Always preserve special tokens ([CLS], [SEP], [PAD], etc.) regardless of their importance scores
-    \item Calculate token importance using the technique-specific method
-    \item Sort tokens by their importance scores in descending order
-    \item Keep the top $(100-p)\%$ highest-scoring tokens, where $p$ is the pruning percentage
-    \item Create a reduced embedding matrix using only the retained tokens
-\end{enumerate}
+Compared to the standard workflow for adapting ModernBERT to a downstream task, which involves taking the pre-trained base model and fine-tuning it directly on the task data, we first analyze the task-specific dataset to extract token statistics.
+Tokens are then ranked based on their importance and the least important tokens are pruned from the embedding layer.
+
+The pruning techniques follow the same procedure, formalized in Algorithm~\ref{alg:pre_ft_pruning}.
+The embedding matrix is then reconstructed using only the retained tokens, yielding a smaller model.
+Optional semantic clustering can be applied to map pruned tokens to semantically similar retained tokens,
+
 This approach front-loads the pruning work to the pre-fine-tuning phase, meaning the actual fine-tuning process remains unchanged and operates on a model that already has a reduced parameter count.
 The resulting pruned model maintains its standard inference pipeline while benefiting from reduced memory requirements.
-
-The high-level vocabulary pruning procedure, is formalized in Algorithm~\ref{alg:pre_ft_pruning}.
 
@@
 \begin{algorithmic}[1]
-\Require Pretrained model $M$ with vocab $V$, dataset $D$, pruning ratio $p$
+\Require Pretrained model $M$ with vocabulary $V$, downstream dataset $D$, pruning ratio $p$
 \Ensure Pruned model $M'$
@@
 \State $\text{scores} \gets \text{rank\_tokens}(\text{stats})$ 
-\State $V_{\text{keep}} \gets \text{special\_tokens}(M)\cup\text{top\_k\_tokens}(\text{scores}, 1 - p)$
+\State $V_{\text{special}} \gets \text{special\_tokens}(M)$ \Comment{([CLS], [SEP], etc.)}
+\State $V_{\text{keep}} \gets V_{\text{special}} \cup \text{top\_k\_tokens}(\text{scores}, 1 - p)$
 \State $M' \gets \text{rebuild\_model\_with\_vocab}(M, V_{\text{keep}})$
@@
 
+\input{tables/token_statistics}
 \subsection{Token Importance Analysis}
 A critical challenge in vocabulary pruning is determining which tokens to retain and which to remove.
-This research explores and compares several token importance estimation techniques, each with distinct theoretical foundations and practical implications for model performance.
+This research explores and compares several token importance estimation techniques, each with distinct approaches and practical implications for model performance.
 These approaches range from simple statistical methods to complex semantic analysis.
 
-\subsubsection{Random Selection}
-Tokens are pruned randomly without consideration for importance, serving as a baseline approach. In this method, tokens are selected for removal using uniform random sampling from the full vocabulary.
+\paragraph{Random Selection}
+prunes tokens randomly without consideration for importance, serving as a baseline approach. In this method, tokens are selected for removal using uniform random sampling from the full vocabulary.
 
-\subsubsection{Clustering-Based}
-This approach employs agglomerative hierarchical clustering on the embedding space to group tokens with similar semantic properties. The algorithm works as follows:
+\paragraph{Clustering-Based}
+preserves semantic diversity across the vocabulary while reducing redundancy by grouping similar tokens together and keeping only one representative from each group.
+It leverages the property that embeddings pointing in similar directions (high cosine similarity) typically have similar meanings.
 
-\begin{enumerate}
-    \item Each token's embedding vector is normalized to unit length (magnitude of 1), allowing the algorithm to compare tokens based solely on their semantic direction rather than magnitude.
-    \item Agglomerative clustering begins with each token as its own cluster, then iteratively merges the closest pairs of clusters.
-    \item The merging process uses average linkage, where the distance between clusters is the average distance between all pairs of embeddings across the two clusters.
-    \item The algorithm stops when the target number of clusters is reached (determined by the pruning percentage).
-    \item For each resulting cluster, the token closest to the centroid is identified and retained as the representative.
-    \item All other tokens in the cluster are pruned from the vocabulary.
-\end{enumerate}
-This method preserves semantic diversity across the vocabulary while reducing redundancy.
-By selecting representatives closest to cluster centroids, the approach ensures that the preserved tokens maximize coverage of the semantic space.
-The clustering leverages the property that tokens with embeddings pointing in similar directions (high cosine similarity) tend to have similar semantic meanings, allowing one representative token to effectively substitute for multiple semantically related tokens.
+The approach uses agglomerative hierarchical clustering on normalized embedding vectors to: 
+(i) \textit{group} semantically similar tokens into clusters, 
+(ii) \textit{identify} the token closest to each cluster's centroid as its representative, and
+(iii) \textit{retain} only these representatives while pruning all other tokens.
+This process ensures maximal semantic coverage with a reduced vocabulary size.
 
-\subsubsection{Frequency-Based}
-Tokens are ranked by their frequency in the target dataset, with least frequently used tokens pruned first.
-This approach operates on the principle that rarely used tokens contribute less to model performance.
+\paragraph{Frequency-Based}
+ranks tokens by their frequency in the target dataset, with least frequently used tokens pruned first.
+This approach operates on the heuristic that rarely used tokens contribute less to model performance.
+
+We explore two frequency-based ranking methods:
+
+\subparagraph{Simple Frequency} counts raw token occurrences in the training data, favoring common tokens regardless of their distribution across documents or examples.
+
+\subparagraph{TF-IDF} (Term Frequency-Inverse Document Frequency) balances token occurrence with discriminative power across documents.
+This method prioritizes tokens that appear frequently in specific documents but are rare across the corpus, capturing task-specific terminology while filtering out ubiquitous tokens that carry less semantic value.
+
+The TF-IDF approach operates by:
+(i) \textit{calculating} term frequency (TF) as the normalized count of a token in each document,
+(ii) \textit{computing} inverse document frequency (IDF) as the logarithm of the ratio between total documents and documents containing the token,
+(iii) \textit{multiplying} these values to get TF-IDF scores that highlight discriminative tokens, and
+(iv) \textit{normalizing} scores (using L2 normalization in our experiments) to prevent outlier tokens from dominating.
 
@@
 
-\subsubsection{Attention-Based}
-This method leverages the attention mechanism of transformer models to identify important tokens.
-The key insight is that tokens receiving higher attention during inference are those the model relies on most heavily when making predictions, providing a direct measure of token importance from the model's perspective.
-Unlike frequency-based approaches that simply count occurrences, attention-based pruning preserves tokens that meaningfully contribute to the model's decision-making process.
-\\ \\
-The technique-specific algorithm steps are:
-\begin{enumerate}
-    \item First fine-tune the base model on the target task to learn task-specific attention patterns
-    \item Process the task dataset through this fine-tuned model, capturing attention matrices from all layers and heads
-    \item For each token, aggregate the attention it receives across all contexts where it appears
-    \item Normalize attention scores by token frequency to avoid bias toward common tokens
-\end{enumerate}
-\subsubsection{TF-IDF Based}
-Tokens are ranked using Term Frequency-Inverse Document Frequency (TF-IDF) scores, which balance token occurrence frequency with discriminative power across documents.
-This method prioritizes tokens that appear frequently in specific documents but are rare across the corpus—capturing task-specific terminology while filtering out ubiquitous tokens that carry less semantic value.
+\paragraph{Attention-Based}
+leverages the attention mechanism of transformer models to identify potentially important tokens.
+The underlying principle is that tokens receiving higher attention during inference might be those the model relies on more when making predictions, offering a model-based perspective on token importance.
+While attention scores don't necessarily imply causal relevance, they provide a task-specific signal that differs from simple statistical measures.
+Attention-based pruning aims to preserve tokens that appear to participate more actively in the model's decision-making process.
 
-The TF-IDF approach is particularly effective for tasks with distinct document categories (like sentiment classification or topic detection) where certain tokens strongly differentiate between classes.
-Unlike simple frequency counts, TF-IDF prevents common but semantically shallow tokens (articles, conjunctions, etc.) from dominating the pruned vocabulary.
-\\ \\
-The technique-specific algorithm steps are:
-\begin{enumerate}
-    \item Tokenize all examples in the task dataset
-    \item Calculate term frequency (TF) for each token in each document:
-        \begin{center}
-        $\text{TF}(t,d) = \frac{\text{count of $t$ in $d$}}{\text{total terms in $d$}}$
-        \end{center}
-    \item Calculate inverse document frequency (IDF) across the corpus:
-        \begin{center}
-        $\text{IDF}(t) = \log\left(\frac{\text{total docs}}{\text{docs with $t$}}\right)$
-        \end{center}
-    \item Compute TF-IDF scores by multiplying TF and IDF values:
-        \begin{center}
-        $\text{TF-IDF}(t,d) = \text{TF}(t,d) \times \text{IDF}(t)$
-        \end{center}
-    \item Apply normalization to the scores (see variants below)
-\end{enumerate}
-For this approach, three normalization variants were implemented: Non-normalized TF-IDF (raw scores), L1-normalized TF-IDF (scores normalized by sum), and L2-normalized TF-IDF (scores normalized by Euclidean norm, the standard approach in information retrieval that prevents outlier tokens from dominating).
-Only the L2-normalized variant was evaluated in our experiments.
+The approach operates by:
+(i) \textit{fine-tuning} the base model on the target task to learn task-specific attention patterns,
+(ii) \textit{processing} the dataset through this model to capture attention matrices from all layers and heads,
+(iii) \textit{aggregating} the attention each token receives across all its occurrences, and
+(iv) \textit{normalizing} scores by token frequency to avoid bias toward common tokens.
+
 \subsection{Out-of-Vocabulary Token Handling}
-The handling of out-of-vocabulary (OOV) tokens presents an opportunity to preserve additional semantic information during vocabulary pruning.
-While standard approaches typically map all pruned tokens to a single UNK token, this approach utilizes the semantic properties of the embedding space to maintain some of the original meaning of pruned tokens.
-This can help mitigate performance degradation by retaining partial semantic value for tokens not included in the reduced vocabulary.
-\\ \\
-The clustering-based OOV handling process works as follows:
-\begin{enumerate}
-    \item Extract embeddings for all tokens marked for removal during pruning
-    \item Apply K-means clustering to these embeddings, grouping semantically similar tokens together
-    \item For each cluster, identify the token closest to the centroid as the representative
-    \item Create a mapping from each pruned token to its cluster representative
-    \item During inference, when an OOV token is encountered, it is dynamically mapped to its assigned representative rather than the generic UNK token
-\end{enumerate}
-This approach maintains semantic nuance by ensuring that pruned tokens are replaced by semantically similar alternatives rather than a single catchall token.
-For example, domain-specific terminology might be mapped to more common synonyms, or rare morphological variants might be mapped to their common forms (e.g., "stabilized" → "stable").
-The method is particularly effective for technical vocabularies where semantically related terms form natural clusters in the embedding space.
+Another question is what happens with tokens that are removed from the vocabulary. 
+One option is to map them to a single UNK token, effectively discarding their semantic information.
+Alternatively, semantic clustering can be used to maintain some of the original meaning of pruned tokens.
 
-The number of clusters (k) provides a tunable parameter to balance compression rate against semantic precision—larger values of k preserve more semantic distinctiveness but reduce the overall pruning benefit.
+The clustering-based OOV handling process extracts embeddings for all pruned tokens and applies K-means clustering to group semantically similar tokens together.
+From each cluster, the token closest to the centroid is selected as a representative.
+During inference, when an OOV token is encountered, it is mapped to its assigned representative rather than the generic UNK token.
+The number of clusters (k) offers a tunable parameter to balance compression rate against semantic precision.
+
+We will extend the frequency-based approaches with this clustering-based OOV handling.
 % TODO: Add experimental results and a plot
@@
 
-\input{tables/token_statistics}
-
 \section{Experiments}
-After developing several algorithmic approaches to vocabulary pruning, this section presents a comprehensive evaluation designed to answer several key questions: Can selective vocabulary reduction maintain model performance while significantly decreasing parameter count?
+After developing several algorithmic approaches to vocabulary pruning, this section presents a comprehensive evaluation designed to answer several key questions; can selective vocabulary reduction maintain model performance while significantly decreasing parameter count?
 How do different pruning strategies compare on various natural language understanding tasks?
@@
 
-\subsection{Datasets and Metrics}
-The General Language Understanding Evaluation (GLUE) benchmark serves as the primary evaluation framework for this study.
-GLUE consists of eight diverse natural language understanding tasks categorized into three groups: single-sentence tasks (SST-2, CoLA), similarity and paraphrase tasks (MRPC, STS-B, QQP), and natural language inference tasks (MNLI, QNLI, RTE).
-Performance metrics vary by task: classification tasks report accuracy; CoLA reports Matthew's correlation coefficient, and STS-B reports Pearson correlation.
+\input{tables/results.tex}
+\paragraph{Datasets and Metrics} 
+The General Language Understanding Evaluation (GLUE) benchmark\cite{wang2018glue} serves as our primary evaluation framework, comprising eight diverse NLP tasks: single-sentence tasks (SST-2, CoLA), similarity/paraphrase tasks (MRPC, STS-B, QQP), and inference tasks (MNLI, QNLI, RTE).
+We report task-appropriate metrics: accuracy for most classification tasks, Matthew's correlation for CoLA, and Pearson correlation for STS-B.
 
-Table \ref{tab:token_statistics} presents a detailed analysis of token distributions across all GLUE tasks.
-This analysis reveals significant vocabulary redundancy, with the top 20\% of tokens covering 80-94\% of all token occurrences in training sets.
-This finding provides strong empirical justification for vocabulary pruning, as most tokens contribute minimally to overall corpus coverage.
-The table also quantifies the overlap between TF-IDF and frequency-based token selection, showing substantial differences (38-60\% overlap), which explains the performance variations between these approaches.
+Table \ref{tab:token_statistics} reveals significant vocabulary redundancy across GLUE tasks, with the top 20\% of tokens covering 80-94\% of all occurrences in training sets—providing strong empirical justification for vocabulary pruning.
+The table also examines train-test vocabulary overlap, which varies significantly across tasks and underscores the importance of effective OOV handling strategies in the pruning process.
 
-Table \ref{tab:vocab_overlap} examines train-test vocabulary overlap, quantifying the percentage of out-of-vocabulary (OOV) tokens in test sets.
-While datasets like MNLI and QNLI show excellent coverage (less than 0.5\% OOV tokens), others such as MRPC, STSB, and RTE exhibit higher OOV percentages (12-14\%), highlighting the importance of effective OOV handling strategies in the pruning process.
+\paragraph{Baselines} 
+We establish three baseline models to evaluate our vocabulary pruning techniques: 
+(i) \textit{ModernBERT (Full)} serves as the upper performance bound, employing the complete 50,368-token vocabulary and all encoder parameters; 
+(ii) \textit{LoSparse} provides a complementary encoder-layer pruning baseline that uses low-rank plus sparse factorization to preserve 80\% of parameters in encoder layers while maintaining the full embedding layer;
+(iii) \textit{Train Tokens Only} represents a simple vocabulary reduction approach that removes embeddings not observed in fine-tuning data, achieving variable parameter reduction (4.79-22.61\%) depending on dataset vocabulary coverage.
+These baselines establish benchmarks for assessing both parameter efficiency and performance of our proposed methods.
 
-\input{tables/vocab_overlap}
-
-\input{tables/results.tex}
-
-\subsection{Baselines}
-Three baseline models are established to evaluate the effectiveness of the proposed vocabulary pruning techniques:
-
-\begin{itemize}
-    \item \textbf{ModernBERT (Full)}: The original pre-trained model with no modifications, representing the upper bound of performance with full parameter count.
-This baseline utilizes the complete 50,368-token vocabulary and all encoder parameters.
-    \item \textbf{LoSparse}: An encoder-layer pruning technique that applies structured sparsification to reduce parameters in the self-attention and feed-forward networks.
-The implementation uses a low-rank plus sparse factorization that preserves 80\% of the original parameters across all encoder layers while keeping the embedding layer intact.
-This baseline represents a complementary approach that targets different model components.
-    \item \textbf{Train Tokens Only}: A vocabulary reduction approach that removes all token embeddings not observed in the fine-tuning dataset.
-This method represents an upper bound for vocabulary pruning performance, as it preserves all task-relevant tokens without further reduction.
-All subsequent pruning techniques begin with this training vocabulary and apply additional selection criteria.
-The parameter reduction varies significantly by task (4.79-22.61\%), depending on vocabulary coverage in each dataset.
-\end{itemize}
-These baselines establish comparative benchmarks for assessing both the parameter efficiency and performance of the proposed vocabulary pruning methods.
-The full model provides the performance ceiling, LoSparse demonstrates the impact of encoder-only pruning, and the Train Tokens Only approach represents the simplest form of vocabulary reduction without sophisticated token selection criteria.
-
-\subsection{Implementation Details}
+\paragraph{Implementation Details}
 All experiments were conducted on an NVIDIA RTX 4090 GPU.
@@
 For encoder layer pruning, a custom fork of the LoSparse implementation was used\footnote{\url{https://github.com/WoutDeRijck/LoSparse.git}}.
-Both codebases are released to facilitate reproducibility and further research in model compression techniques.
 
@@
 
-\subsection{Results}
-Table \ref{tab:results} presents the comprehensive evaluation across all GLUE tasks for the different pruning methods. Several key findings emerge.
+\paragraph{Results}
+Table \ref{tab:results} presents a comprehensive evaluation across GLUE tasks for different the pruning methods.
 
-First, vocabulary pruning alone achieves substantial parameter reduction (20.02-23.27\%) with minimal performance impact.
-The TF-IDF + OOV approach performs exceptionally well, maintaining 97.6\% of the original model's average performance while reducing parameters by over 20\%.
+Vocabulary pruning achieves substantial parameter reduction (20.02-23.27\%) with minimal performance impact, with TF-IDF + OOV maintaining 97.6\% of the original model's performance while reducing parameters by over 20\%.
 
-Second, the inclusion of OOV handling mechanisms significantly improves results compared to their non-OOV counterparts.
-For instance, TF-IDF + OOV outperforms standard TF-IDF by 2.4 percentage points on average, with particularly strong improvements on CoLA (9.5 points) and RTE (6.1 points).
+OOV handling mechanisms boost results compared to their non-OOV counterparts (+2.4 percentage points on average).
 
-Third, the effectiveness of pruning varies across tasks.
-Attention-based pruning excels on single-sentence tasks (SST-2, CoLA) and paraphrase detection (MRPC), while TF-IDF-based methods perform better on more complex reasoning tasks (MNLI, QNLI, RTE).
+Pruning effectiveness varies by task: attention-based methods excel on single-sentence tasks and paraphrase detection, while TF-IDF-based methods perform better on complex reasoning tasks.
 
-Fourth, as illustrated in Figure \ref{fig:pruning_ratio}, vocabulary pruning demonstrates superior performance and efficiency compared to encoder-layer pruning (LoSparse) at lower pruning ratios (5-20\%).
-Importantly, while LoSparse requires computationally expensive online training procedures, vocabulary pruning is an offline pre-fine-tuning method that can be applied with minimal computational overhead.
-A striking result is that a 20\% reduction in total model parameters translates to a 77.34\% reduction in vocabulary size while maintaining competitive performance.
-This efficiency advantage makes vocabulary pruning particularly attractive for resource-constrained environments.
-Since the embedding layer constitutes only 25.95\% of the total model parameters, vocabulary pruning has an inherent upper limit.
-Beyond the 20\% threshold, performance decreases sharply for vocabulary pruning while LoSparse maintains more stable performance at higher compression rates.
-Vocabulary pruning provides an optimal approach for applications requiring moderate compression with minimal performance impact, while more aggressive compression could benefit from combining both techniques.
+As shown in Figure \ref{fig:pruning_ratio}, vocabulary pruning outperforms encoder-layer pruning (LoSparse) at lower pruning ratios (5-20\%), while requiring only minimal computational overhead as an offline pre-fine-tuning method.
+A 20\% reduction in total parameters (77.34\% of vocabulary size) maintains competitive performance, making this approach particularly valuable for resource-constrained environments.
+The embedding layer's 25.95\% parameter share imposes an upper limit on vocabulary pruning, with performance declining beyond the 20\% threshold while LoSparse maintains more stable performance at higher compression rates.
 
-The benchmark results in Table \ref{tab:benchmark_results} further quantify the practical benefits, showing a 20\% reduction in storage size and 14.83\% reduction in GPU memory requirements with negligible impact on inference time (+1.33\%).
+The benchmark results in Table \ref{tab:benchmark_results} show 20\% storage reduction and 14.83\% lower GPU memory use with negligible inference time impact (+1.33\%).
 
 \input{tables/benchmark_results}
-
 \section{Conclusion}

### 2025-05-19 paper/tables/token_statistics.tex
make paper more compact

@@
 \toprule
-& \multicolumn{2}{c|}{\textbf{Total Tokens}} & \multicolumn{2}{c|}{\textbf{Unique Tokens}} & \multicolumn{2}{c|}{\textbf{Vocab Coverage (\%)}} & \multicolumn{2}{c|}{\textbf{Top 20\% Coverage (\%)}} & \multicolumn{2}{c}{\textbf{TF-IDF/Freq Overlap (\%)}} \\
-\textbf{Task} & \textbf{Train} & \textbf{Test} & \textbf{Train} & \textbf{Test} & \textbf{Train} & \textbf{Test} & \textbf{Train} & \textbf{Test} & \textbf{Train} & \textbf{Test} \\
+& \multicolumn{2}{c|}{\textbf{Total Tokens}} & \multicolumn{2}{c|}{\textbf{Unique Tokens}} & \multicolumn{2}{c|}{\textbf{Vocab Coverage (\%)}} & \multicolumn{2}{c|}{\textbf{Top 20\% Coverage (\%)}} & \multicolumn{2}{c}{\textbf{Test Vocabulary Coverage}} \\
+\textbf{Task} & \textbf{Train} & \textbf{Test} & \textbf{Train} & \textbf{Test} & \textbf{Train} & \textbf{Test} & \textbf{Train} & \textbf{Test} & \textbf{OOV Tokens} & \textbf{OOV\%} \\
 \midrule
-COLA & 71,818 & 7,939 & 5,416 & 1,934 & 17.74 & 6.34 & 85.12 & 74.68 & 58.36 & 51.04 \\
-MNLI & 13,044,457 & 1,452,279 & 25,793 & 22,664 & 84.51 & 74.25 & 90.69 & 89.64 & 38.41 & 61.08 \\
-MRPC & 165,585 & 18,701 & 11,096 & 3,567 & 36.35 & 11.69 & 83.17 & 71.80 & 57.91 & 54.56 \\
-QNLI & 4,388,343 & 485,747 & 26,176 & 20,360 & 85.76 & 66.71 & 88.31 & 85.79 & 33.41 & 55.33 \\
-QQP & 9,036,384 & 1,000,855 & 25,486 & 18,534 & 83.50 & 60.72 & 94.02 & 91.79 & 32.39 & 61.20 \\
-RTE & 151,109 & 16,215 & 13,354 & 4,198 & 43.75 & 13.75 & 81.77 & 70.75 & 57.49 & 54.71 \\
-SST2 & 686,566 & 75,775 & 11,536 & 8,084 & 37.80 & 26.49 & 84.23 & 80.53 & 59.08 & 57.74 \\
-STSB & 128,151 & 14,481 & 10,346 & 3,271 & 33.90 & 10.72 & 82.66 & 71.30 & 60.22 & 49.54 \\
+COLA & 71,818 & 7,939 & 5,416 & 1,934 & 17.74 & 6.34 & 85.12 & 74.68 & 169 & 8.74\% \\
+MNLI & 13,044,457 & 1,452,279 & 25,793 & 22,664 & 84.51 & 74.25 & 90.69 & 89.64 & 31 & 0.14\% \\
+MRPC & 165,585 & 18,701 & 11,096 & 3,567 & 36.35 & 11.69 & 83.17 & 71.80 & 465 & 13.04\% \\
+QNLI & 4,388,343 & 485,747 & 26,176 & 20,360 & 85.76 & 66.71 & 88.31 & 85.79 & 83 & 0.41\% \\
+QQP & 9,036,384 & 1,000,855 & 25,486 & 18,534 & 83.50 & 60.72 & 94.02 & 91.79 & 191 & 1.03\% \\
+RTE & 151,109 & 16,215 & 13,354 & 4,198 & 43.75 & 13.75 & 81.77 & 70.75 & 509 & 12.12\% \\
+SST2 & 686,566 & 75,775 & 11,536 & 8,084 & 37.80 & 26.49 & 84.23 & 80.53 & 34 & 0.42\% \\
+STSB & 128,151 & 14,481 & 10,346 & 3,271 & 33.90 & 10.72 & 82.66 & 71.30 & 448 & 13.70\% \\
 \bottomrule
 \end{tabular}
-\caption{Token statistics across GLUE tasks, comparing train and test splits.}
+\caption{Token statistics across GLUE tasks, comparing train and test splits.
+OOV Tokens: number of unique tokens in the test set that do not appear in the training set; OOV\%: percentage of test vocabulary not found in train.}
 \label{tab:token_statistics}

### 2025-05-20 paper/main.tex
make it a lot more tight

@@
 Additionally, our results suggest that state-of-the-art encoder-layer pruning can fall short of simple embedding-layer pruning.
+We provide open-source implementations of all pruning methods and evaluation tools to support further research in this area.\footnote{\url{https://github.com/WoutDeRijck/vocab-pruning.git}}\footnote{\url{https://github.com/WoutDeRijck/LoSparse.git}}
 
@@
 
-From this perspective, ModernBERT stands out as one of the most lightweight and optimized encoder-only transformers available.
-It integrates several architectural and training innovations, successfully maintaining robust downstream performance while minimizing inference costs.
+ModernBERT stands out as one of the most lightweight and optimized encoder-only transformers available.
+It integrates several architectural and training innovations while maintaining robust downstream performance.
 Despite these advancements, a critical question remains open: can we further enhance ModernBERT's efficiency by pruning parameters?
 
-This question is particularly significant for two reasons.
-First, ModernBERT's inherent efficiency and its strong potential for large-scale deployment mean that optimal pruning strategies on top of this model could yield exceptionally efficient models.
-Second, existing research into pruning techniques have predominantly targeted decoder-only LLMs \cite{namburi2023llm} due to their widespread popularity, leaving encoder-only models relatively underexplored.
+This question is particularly interesting for two reasons.
+First, ModernBERT's inherent efficiency means that optimal pruning strategies could yield exceptionally efficient models.
+Second, existing research into pruning techniques have predominantly targeted decoder-only LLMs \cite{namburi2023llm}, leaving encoder-only models relatively underexplored.
 
-Specifically, we explore task-specific weight pruning, as ModernBERT is intended to be fine-tuned on specific tasks for optimal behavior (unlike more general-purpose decoder-only models that can utilize in-context learning).
+Specifically, we explore task-specific weight pruning, as ModernBERT is intended to be finetuned on specific tasks for optimal behavior (unlike more general-purpose decoder-only models that can utilize in-context learning).
 We hypothesize that simple task-specific pruning of the embedding matrix can be highly efficient, as many tokens can be irrelevant for any specific downstream task and the embedding layers counts for a significant portion ($\sim$25\% of the total parameters).
@@
 We explore a series of simple vocabulary pruning heuristics on a range of Natural Language Processing tasks including linguistic acceptability, sentiment analysis, paraphrase detection, semantic similarity, question answering, and natural language inference, as covered by the GLUE benchmark.
-For any task, we first apply this pruning to the base ModernBERT model and subsequently fine-tune the pruned model.
-For any given pruning ratio, we measure the accuracy retention compared to simply fine-tuning the model.
+For any task, we first apply vocabulary pruning to the base ModernBERT model and subsequently finetune the pruned model.
 We compare these vocabulary pruning techniques with LoSparse, a state-of-the-art gradient-based pruning technique that targets the encoder layers.
 
-As shown in Figure \ref{fig:pruning_retention}, TF-IDF-based vocabulary pruning outperforms encoder-layer pruning (LoSparse) for compression ratios up to 20\%, maintaining 97.6\% of ModernBERT's original performance while removing 77.34\% of embedding parameters.
+Our results show that TF-IDF-based vocabulary pruning outperforms encoder-layer pruning (LoSparse) for compression ratios up to 20\%, maintaining 97.6\% of ModernBERT's original performance while removing 77.34\% of embedding parameters (see Figure \ref{fig:pruning_retention}).
 This shows that substantial efficiency gains can be achieved with virtually no performance loss using a simple, offline pruning technique.
 Beyond 20\% compression, performance drops sharply as we approach the embedding layer's parameter limit (25.95\% of total model parameters), while LoSparse maintains more stable performance at higher ratios.
-Unlike LoSparse, which requires computationally expensive gradient-based optimization, our vocabulary pruning operates as a straightforward pre-fine-tuning step with minimal overhead, making it particularly suitable for resource-constrained environments requiring moderate model compression.
+Unlike LoSparse, which requires computationally expensive gradient-based optimization, our vocabulary pruning operates as a straightforward pre-finetuning step with minimal overhead.
 
@@
 Model compression techniques enhance the efficiency of large language models (LLMs) by reducing their size and computational requirements.
-
-These techniques can be broadly categorized into four primary methods:
-(i) \textbf{quantization}, which reduces the numerical precision of model weights, decreasing memory footprint;
+% 
+These techniques can be broadly categorized into four primary methods~\cite{wan2023efficient}:
+(i) \textbf{quantization}, which reduces the numerical precision of model weights;
 (ii) \textbf{parameter pruning}, which eliminates redundant or less important connections and neurons to create sparser models with fewer parameters;
-(iii) \textbf{low-rank approximation}, which decomposes weight matrices into lower-dimensional representations that capture essential information while requiring fewer parameters; and
-(iv) \textbf{knowledge distillation}, which transfers knowledge from larger teacher models to smaller student models, enabling competitive performance with reduced architecture size~\cite{wan2023efficient}.
+(iii) \textbf{low-rank approximation}, which decomposes weight matrices into lower-dimensional representations that capture essential information while requiring fewer parameters~\cite{Hu2021LoRA}; and
+(iv) \textbf{knowledge distillation} transfers knowledge from larger teacher models to smaller student models, enabling competitive performance with reduced architecture size.~\cite{Hinton2015Distillation}
 
@@
 %Misschien niet nodig voor de paper maar heel goed om je expertise te tonen in de full thesis
-These approaches are complementary and can be combined to achieve optimal compression.
-This work focuses on structured parameter pruning techniques to reduce model size while preserving performance.
+The model compression techniques described above are complementary and can be combined to achieve optimal compression.
+This work focuses on parameter pruning techniques to reduce model size while preserving performance.
 The rest of this section is dedicated to enumerating the most important parameter pruning techniques.
@@
 % Maar vooral belangrijk voor de thesis om de equations uit te schrijven als je content nodig hebt
-\paragraph{Encoder-Layer Pruning methods}~remove less critical parameters within the transformer encoder to reduce model size while maintaining acceptable accuracy.
+\textbf{Encoder-Layer Pruning methods} remove less critical parameters within the encoder to reduce model size while maintaining acceptable accuracy.
 A representative state-of-the-art method is LoSparse~\cite{li2023losparse}, which decomposes each weight matrix into a low-rank component and a sparse residual.
-During fine-tuning, LoSparse calculates the importance of each weight by multiplying it with its gradient.
+During finetuning, LoSparse calculates the importance of each weight by multiplying it by its gradient, which is resource-intensive.
 
-Attention-head pruning~\cite{michel2019sixteen} removes entire attention heads.
-Movement pruning~\cite{sanh2020movement} applies gradient-guided masking to gradually zero out low-importance weights during fine-tuning.
+Attention-head pruning~\cite{michel2019sixteen} removes entire attention heads, whereas LoSparse removes entire neurons (i.e., rows in weight matrices).
+Movement pruning~\cite{sanh2020movement} applies gradient-guided masking to gradually zero out low-importance weights during finetuning.
 LayerDrop~\cite{fan2020layerdrop} introduces structured dropout, allowing entire transformer layers to be stochastically removed during training and optionally pruned at inference.
@@
 These encoder-layer pruning methods differ in their granularity and strategy—targeting individual weights, attention heads, or full layers.
-Among them, LoSparse stands out as a strong baseline for encoder-level pruning.
+LoSparse stands out as a strong baseline for encoder-level pruning.
+~\cite{pmlr-v235-li24bi}~\cite{huang2023awesome}~\cite{ren2025llmcompression}
 
-\paragraph{Embedding-Layer Pruning methods}
-target the embedding layer, which can comprise up to 20-30\% of a model's parameters.
-These approaches can be broadly categorized into two main strategies: (1) vocabulary pruning, which removes entire token entries, and (2) embedding compression, which retains all tokens but reduces their dimensional representation.
+% \paragraph{Embedding-Layer Pruning methods}
+\textbf{Embedding-Layer Pruning methods} reduce the memory footprint of the embedding layer, which can comprise up to 20-30\% of a model's parameters.
+These methods can be categorized into two main strategies: embedding compression and vocabulary pruning.
 
-\textbf{Embedding compression} reduces the memory footprint of the embedding matrix through dimensionality reduction techniques.
+\textbf{Embedding compression} reduces the dimensional representation of the embeddings.
 LightToken~\cite{wang2023lighttoken} is a model- and task-agnostic framework for compressing the token embedding layers.
@@
 
-\textbf{Vocabulary Pruning Techniques} are a specialized form of embedding pruning that remove rarely used or task-irrelevant tokens to shrink the model's vocabulary.
-These methods simply delete rows in the embedding matrix, making them straightforward to implement.
+\textbf{Vocabulary Pruning Techniques} are a specialized form of embedding pruning that removes rarely used or task-irrelevant tokens.
 
-The simplest approach is "drop unused tokens." Both TextPruner~\cite{shen2022textpruner} and Partial Embedding Matrix Adaptation (PEMA;~\cite{bousquet2023pema}) scan a corpus and remove tokenizer tokens not present in the target text.
-TextPruner permanently deletes these entries, while PEMA temporarily removes them during fine-tuning and restores them afterward.
-These methods can reduce BERT$_{\text{BASE}}$'s embedding matrix by up to 47\% without compromising performance.
+The simplest approach is to drop unused tokens.
+TextPruner~\cite{shen2022textpruner} scans a corpus and removes tokenizer tokens not present in the target text.
+This can reduce BERT$_{\text{BASE}}$'s embedding matrix by up to 47\% without compromising performance.
 
-For multilingual models, language-specific pruning has proven effective.
+For multilingual models, most of the parameters are located in the embeddings layer, language-specific pruning has proven effective.
 Abdaoui et al.~\cite{abdaoui2020load} showed that trimming unused language tokens from mBERT can achieve a 45\% parameter reduction with comparable accuracy on XNLI.
@@
 
-These findings consistently demonstrate that vocabulary pruning is an effective way to tailor models to specific domains or languages with minimal performance impact.
+% These findings consistently demonstrate that vocabulary pruning is an effective way to tailor models to specific domains or languages with minimal performance impact.
+While simply removing embeddings for unseen (i.e., never‑used) tokens has proven to be effective, we show one can go substantially further by additionally ranking and pruning low‑utility tokens.
 
+TODO: on modernbert, important!
+
+To our knowledge, this is the first study to (i) rigorously evaluate a range of importance heuristics in a single framework, (ii) incorporate OOV‑aware mapping to preserve semantic fidelity of pruned tokens, and (iii) apply vocabulary pruning atop ModernBERT—a state‑of‑the‑art lightweight encoder—thereby demonstrating the practical benefits of embedding‑layer reduction in resource‑constrained settings.
+
+\newpage
 \section{Method}
@@
 This approach is based on the observation that tokens in a vocabulary have varying importance for downstream tasks, and that certain tokens can be selectively removed with minimal impact on performance.
-The method primarily consists of two main components: token importance analysis and selective pruning, with an optional third component: semantic clustering for out-of-vocabulary (OOV) tokens that can further enhance performance in some cases.
+We explore and compare several token importance estimation techniques, ranging from simple statistical methods to semantic analysis.
+The method consists of three components: token importance analysis, selective pruning, and an optional semantic clustering for out-of-vocabulary (OOV) tokens that can further enhance performance in some cases.
 
-\subsection{Pre-Fine-Tuning Pruning Procedure}
-In contrast to methods that require resource-intensive post-pruning fine-tuning, the proposed vocabulary pruning approach operates as a pre-fine-tuning offline optimization step in the model adaptation pipeline.
+In contrast to methods that require resource-intensive post-pruning finetuning, the proposed vocabulary pruning approach operates as a \textbf{pre-finetuning offline pruning} step in the model adaptation pipeline.
+Instead, it relies on either simple statistical measures or a single attention analysis pass, which does not require gradient calculations, resulting in minimal computational overhead before standard finetuning.
 
-Compared to the standard workflow for adapting ModernBERT to a downstream task, which involves taking the pre-trained base model and fine-tuning it directly on the task data, we first analyze the task-specific dataset to extract token statistics.
-Tokens are then ranked based on their importance and the least important tokens are pruned from the embedding layer.
-
-The pruning techniques follow the same procedure, formalized in Algorithm~\ref{alg:pre_ft_pruning}.
-The embedding matrix is then reconstructed using only the retained tokens, yielding a smaller model.
-Optional semantic clustering can be applied to map pruned tokens to semantically similar retained tokens,
-
-This approach front-loads the pruning work to the pre-fine-tuning phase, meaning the actual fine-tuning process remains unchanged and operates on a model that already has a reduced parameter count.
-The resulting pruned model maintains its standard inference pipeline while benefiting from reduced memory requirements.
+We compute token importance scores using either attention-based (Algorithm \ref{alg:pre_ft_pruning} lines 2-4) or statistical approaches (line 6) via \texttt{compute\_importance\_scores}, then sort and select the top $k$ tokens based on the pruning ratio $p$ (lines 7-9).
+The least important tokens are pruned from the embedding layer, and the embedding matrix is reconstructed using only the retained tokens via \texttt{rebuild\_model\_with\_vocab} (line 13), yielding a smaller model.
+To recover the loss of throwing away tokens, semantic clustering can be applied with \texttt{cluster\_pruned\_tokens} to map pruned tokens to semantically similar retained tokens (lines 10-12). 
+This approach front-loads the pruning work to the pre-finetuning phase, meaning the actual finetuning process remains unchanged and operates on an already-reduced model.
+With a pruning ratio of 0, our algorithm defaults to standard finetuning of the base model.
 
 \begin{algorithm}[H]
-\small
-\caption{Pre-Fine-Tuning Vocabulary Pruning}
+\footnotesize
+\caption{Pre-Finetuning Vocabulary Pruning}
 \label{alg:pre_ft_pruning}
 \begin{algorithmic}[1]
-\Require Pretrained model $M$ with vocabulary $V$, downstream dataset $D$, pruning ratio $p$
-\Ensure Pruned model $M'$
-\State $\text{stats} \gets \text{compute\_token\_statistics}(D)$
-\State $\text{scores} \gets \text{rank\_tokens}(\text{stats})$ 
-\State $V_{\text{special}} \gets \text{special\_tokens}(M)$ \Comment{([CLS], [SEP], etc.)}
-\State $V_{\text{keep}} \gets V_{\text{special}} \cup \text{top\_k\_tokens}(\text{scores}, 1 - p)$
-\State $M' \gets \text{rebuild\_model\_with\_vocab}(M, V_{\text{keep}})$
+\Require Model $M$ with vocab $V$, dataset $D$, pruning ratio $p$
+\Ensure Pruned model $M'$ with reduced vocabulary
+\State $V_s \gets \text{get\_special\_tokens}(M)$ \Comment{[CLS], [SEP]}
+\If{using attention scoring}
+    \State $M_\text{ft} \gets \text{finetune}(M, D)$ \Comment{Finetune model on task}
+    \State $s \gets \text{compute\_attention\_scores}(M_\text{ft}, D, V)$ 
+\Else
+    \State $s \gets \text{compute\_token\_scores}(D, V)$
+\EndIf
+\State $V_{\text{sorted}} \gets \text{sort}(V \setminus V_s, s, \text{desc})$
+\State $k \gets \lfloor (1-p) \cdot |V_{\text{sorted}}| \rfloor$ \Comment{Number of tokens to keep}
+\State $V_{\text{keep}} \gets V_s \cup V_{\text{sorted}}[1:k]$
+\If{using OOV handling}
+    \State $\text{clusters} \gets \text{cluster\_pruned\_tokens}(V \setminus V_{\text{keep}})$
+    \State $\text{oov\_map} \gets \text{map\_pruned\_to\_representatives}(\text{clusters})$
+\EndIf
+\State $M' \gets \text{rebuild\_model\_with\_vocab}(M, V_{\text{keep}}, \text{oov\_map})$
 \State \Return $M'$
@@
 
-\input{tables/token_statistics}
+\input{tables/results.tex}
 \subsection{Token Importance Analysis}
 A critical challenge in vocabulary pruning is determining which tokens to retain and which to remove.
-This research explores and compares several token importance estimation techniques, each with distinct approaches and practical implications for model performance.
-These approaches range from simple statistical methods to complex semantic analysis.
 
-\paragraph{Random Selection}
-prunes tokens randomly without consideration for importance, serving as a baseline approach. In this method, tokens are selected for removal using uniform random sampling from the full vocabulary.
+\textbf{Random Selection}
+prunes tokens randomly without consideration for importance, serving as a baseline approach. In this method, tokens are selected for removal using uniform random sampling from the vocabulary.
 
-\paragraph{Clustering-Based}
-preserves semantic diversity across the vocabulary while reducing redundancy by grouping similar tokens together and keeping only one representative from each group.
-It leverages the property that embeddings pointing in similar directions (high cosine similarity) typically have similar meanings.
-
-The approach uses agglomerative hierarchical clustering on normalized embedding vectors to: 
-(i) \textit{group} semantically similar tokens into clusters, 
-(ii) \textit{identify} the token closest to each cluster's centroid as its representative, and
-(iii) \textit{retain} only these representatives while pruning all other tokens.
+\textbf{Clustering-Based}
+preserves semantic diversity across the vocabulary by grouping similar tokens together and keeping only one representative from each group.
+It leverages the property that embeddings pointing in similar directions (high cosine similarity) typically have similar meanings\cite{mikolov2013efficient}.
 This process ensures maximal semantic coverage with a reduced vocabulary size.
 
-\paragraph{Frequency-Based}
-ranks tokens by their frequency in the target dataset, with least frequently used tokens pruned first.
-This approach operates on the heuristic that rarely used tokens contribute less to model performance.
+\textbf{Frequency-Based} ranks tokens by their frequency in the training dataset, pruning the least frequently used tokens first.
+This approach is grounded in the heuristic that infrequent tokens contribute less to model performance.~\cite {li2024enhancing}
 
-We explore two frequency-based ranking methods:
-
-\subparagraph{Simple Frequency} counts raw token occurrences in the training data, favoring common tokens regardless of their distribution across documents or examples.
-
-\subparagraph{TF-IDF} (Term Frequency-Inverse Document Frequency) balances token occurrence with discriminative power across documents.
+We explore two frequency-based ranking methods. 
+\textit{Simple Frequency} counts raw token occurrences in the training data, favoring common tokens.
+\textit{TF-IDF} (Term Frequency-Inverse Document Frequency) balances token occurrence with discriminative power across documents.
 This method prioritizes tokens that appear frequently in specific documents but are rare across the corpus, capturing task-specific terminology while filtering out ubiquitous tokens that carry less semantic value.
-
-The TF-IDF approach operates by:
-(i) \textit{calculating} term frequency (TF) as the normalized count of a token in each document,
-(ii) \textit{computing} inverse document frequency (IDF) as the logarithm of the ratio between total documents and documents containing the token,
-(iii) \textit{multiplying} these values to get TF-IDF scores that highlight discriminative tokens, and
-(iv) \textit{normalizing} scores (using L2 normalization in our experiments) to prevent outlier tokens from dominating.
 
@@
 
-\paragraph{Attention-Based}
+\textbf{Attention-Based}
 leverages the attention mechanism of transformer models to identify potentially important tokens.
-The underlying principle is that tokens receiving higher attention during inference might be those the model relies on more when making predictions, offering a model-based perspective on token importance.
-While attention scores don't necessarily imply causal relevance, they provide a task-specific signal that differs from simple statistical measures.
-Attention-based pruning aims to preserve tokens that appear to participate more actively in the model's decision-making process.
-
-The approach operates by:
-(i) \textit{fine-tuning} the base model on the target task to learn task-specific attention patterns,
-(ii) \textit{processing} the dataset through this model to capture attention matrices from all layers and heads,
-(iii) \textit{aggregating} the attention each token receives across all its occurrences, and
-(iv) \textit{normalizing} scores by token frequency to avoid bias toward common tokens.
+The underlying principle is that tokens receiving higher attention during inference might be those the model relies on more when making predictions, offering a model-based perspective on token importance\cite{guo2024attention}.
+While attention scores do not necessarily imply causal relevance, they provide a task-specific signal that differs from simple statistical measures.
+The approach first finetunes the base model on the target task to learn task-specific attention patterns, then processes the dataset through this model to capture attention matrices from all layers and heads.
+It aggregates the attention each token receives across all its occurrences, and finally normalizes scores by token frequency to avoid bias toward common tokens.
 
 \subsection{Out-of-Vocabulary Token Handling}
-Another question is what happens with tokens that are removed from the vocabulary. 
-One option is to map them to a single UNK token, effectively discarding their semantic information.
+Vocabulary pruning raises an important question: how should the model process tokens that were removed from its vocabulary?
+Handling these out-of-vocabulary (OOV) tokens can recover performance lost due to pruning.
+One option is to map all pruned tokens to a single UNK token, discarding their semantic information.
 Alternatively, semantic clustering can be used to maintain some of the original meaning of pruned tokens.
 
-The clustering-based OOV handling process extracts embeddings for all pruned tokens and applies K-means clustering to group semantically similar tokens together.
+The clustering-based Out-Of-Vocabulary (OOV) handling process extracts embeddings for all pruned tokens and applies K-means clustering to group semantically similar tokens together.
 From each cluster, the token closest to the centroid is selected as a representative.
-During inference, when an OOV token is encountered, it is mapped to its assigned representative rather than the generic UNK token.
+During inference, when an OOV token is encountered, it is mapped to its assigned representative.
 The number of clusters (k) offers a tunable parameter to balance compression rate against semantic precision.
+While this approach adds a few more tokens back into the vocabulary, the intuition is that it can recover most of the performance lost during pruning.
 
-We will extend the frequency-based approaches with this clustering-based OOV handling.
+After evaluating the different token importance metrics, we apply clustering-based OOV handling to the best-performing approach to further improve the results.
 % TODO: Add experimental results and a plot
 %Experimental results show that even a modest number of clusters (50-100) significantly improves model performance compared to a single UNK mapping approach.
-
-\section{Experiments}
-After developing several algorithmic approaches to vocabulary pruning, this section presents a comprehensive evaluation designed to answer several key questions; can selective vocabulary reduction maintain model performance while significantly decreasing parameter count?
-How do different pruning strategies compare on various natural language understanding tasks?
-Does the pre-fine-tuning approach that avoids post-pruning recovery fine-tuning deliver practical benefits in real-world deployment scenarios?
-Through careful experimentation across diverse tasks, the results demonstrate not only the effectiveness of these algorithms but also their practical advantages in creating more efficient transformer models.
-
-\input{tables/results.tex}
-\paragraph{Datasets and Metrics} 
-The General Language Understanding Evaluation (GLUE) benchmark\cite{wang2018glue} serves as our primary evaluation framework, comprising eight diverse NLP tasks: single-sentence tasks (SST-2, CoLA), similarity/paraphrase tasks (MRPC, STS-B, QQP), and inference tasks (MNLI, QNLI, RTE).
-We report task-appropriate metrics: accuracy for most classification tasks, Matthew's correlation for CoLA, and Pearson correlation for STS-B.
-
-Table \ref{tab:token_statistics} reveals significant vocabulary redundancy across GLUE tasks, with the top 20\% of tokens covering 80-94\% of all occurrences in training sets—providing strong empirical justification for vocabulary pruning.
-The table also examines train-test vocabulary overlap, which varies significantly across tasks and underscores the importance of effective OOV handling strategies in the pruning process.
-
-\paragraph{Baselines} 
-We establish three baseline models to evaluate our vocabulary pruning techniques: 
-(i) \textit{ModernBERT (Full)} serves as the upper performance bound, employing the complete 50,368-token vocabulary and all encoder parameters; 
-(ii) \textit{LoSparse} provides a complementary encoder-layer pruning baseline that uses low-rank plus sparse factorization to preserve 80\% of parameters in encoder layers while maintaining the full embedding layer;
-(iii) \textit{Train Tokens Only} represents a simple vocabulary reduction approach that removes embeddings not observed in fine-tuning data, achieving variable parameter reduction (4.79-22.61\%) depending on dataset vocabulary coverage.
-These baselines establish benchmarks for assessing both parameter efficiency and performance of our proposed methods.
-
-\paragraph{Implementation Details}
-All experiments were conducted on an NVIDIA RTX 4090 GPU.
-The vocabulary pruning techniques described in this paper were implemented in a public repository\footnote{\url{https://github.com/WoutDeRijck/vocab-pruning.git}}.
-This implementation includes all pruning methods, evaluation scripts, and analysis tools described in this work.
-For encoder layer pruning, a custom fork of the LoSparse implementation was used\footnote{\url{https://github.com/WoutDeRijck/LoSparse.git}}.
 
@@
     \includegraphics[width=\linewidth]{images/pruning_ratios.png}
-    \caption{Performance comparison between vocabulary pruning (TF-IDF + OOV) and encoder pruning (LoSparse) across different pruning ratios for MNLI, QNLI, and QQP tasks. }
+    \caption{Performance comparison between vocabulary pruning (TF-IDF + OOV) and encoder pruning (LoSparse).}
     \label{fig:pruning_ratio}
@@
 
-\paragraph{Results}
-Table \ref{tab:results} presents a comprehensive evaluation across GLUE tasks for different the pruning methods.
+\section{Experiments}
+We conducted a comprehensive evaluation across multiple natural language understanding tasks to assess the effectiveness of our vocabulary pruning methods compared to baseline approaches.
+
+\textbf{Datasets and Metrics} 
+The General Language Understanding Evaluation (GLUE) benchmark\cite{wang2018glue} serves as our primary evaluation framework, comprising eight diverse NLP tasks: single-sentence tasks (SST-2, CoLA), similarity/paraphrase tasks (MRPC, STS-B, QQP), and inference tasks (MNLI, QNLI, RTE).
+We report task-appropriate metrics: accuracy for most classification tasks, Matthew's correlation for CoLA, and Pearson correlation for STS-B.
+
+\textbf{Baselines} 
+We establish three baseline models to evaluate our vocabulary pruning techniques: 
+(i) \textit{ModernBERT (Full)} serves as the upper performance bound, employing the complete 50,368-token vocabulary and all encoder parameters; 
+(ii) \textit{LoSparse} provides a complementary encoder-layer pruning baseline that uses low-rank plus sparse factorization to preserve 80\% of parameters in encoder layers while maintaining the full embedding layer;
+(iii) \textit{Train Tokens Only} represents a simple vocabulary reduction approach that removes embeddings not observed in finetuning data, achieving variable parameter reduction (4.79-22.61\%) depending on dataset vocabulary coverage.
+These baselines establish benchmarks for assessing both parameter efficiency and performance of our proposed methods.
+
+\textbf{Results}
+Table \ref{tab:results} presents a comprehensive evaluation across GLUE tasks for different pruning methods.
 
 Vocabulary pruning achieves substantial parameter reduction (20.02-23.27\%) with minimal performance impact, with TF-IDF + OOV maintaining 97.6\% of the original model's performance while reducing parameters by over 20\%.
-
 OOV handling mechanisms boost results compared to their non-OOV counterparts (+2.4 percentage points on average).
@@
 Pruning effectiveness varies by task: attention-based methods excel on single-sentence tasks and paraphrase detection, while TF-IDF-based methods perform better on complex reasoning tasks.
+As shown in Figure \ref{fig:pruning_ratio}, vocabulary pruning outperforms encoder-layer pruning (LoSparse) at lower pruning ratios (5-20\%), while only using simple statistics before finetuning.
+A 20\% reduction in total parameters (77.34\% of vocabulary size) maintains competitive performance, making this approach particularly valuable for resource-constrained environments.
 
-As shown in Figure \ref{fig:pruning_ratio}, vocabulary pruning outperforms encoder-layer pruning (LoSparse) at lower pruning ratios (5-20\%), while requiring only minimal computational overhead as an offline pre-fine-tuning method.
-A 20\% reduction in total parameters (77.34\% of vocabulary size) maintains competitive performance, making this approach particularly valuable for resource-constrained environments.
+The efficacy of vocabulary pruning is explained by the significant redundancy revealed in Table \ref{tab:token_statistics}, where the top 20\% of tokens cover 80-94\% of all occurrences in training sets across GLUE tasks.
+This redundancy provides strong empirical justification for our approach.
+The table also highlights varying train-test vocabulary overlap across tasks, underscoring why effective OOV handling strategies significantly improve performance.
+
 The embedding layer's 25.95\% parameter share imposes an upper limit on vocabulary pruning, with performance declining beyond the 20\% threshold while LoSparse maintains more stable performance at higher compression rates.
 
-The benchmark results in Table \ref{tab:benchmark_results} show 20\% storage reduction and 14.83\% lower GPU memory use with negligible inference time impact (+1.33\%).
+Practically, the benchmark results in Table \ref{tab:benchmark_results} demonstrate that our approach delivers 20\% storage reduction and 14.83\% lower GPU memory use with negligible inference time impact (+1.33\%).
 
+\input{tables/token_statistics}
 \input{tables/benchmark_results}
+
+\newpage
 \section{Conclusion}
@@
 
+\newpage
 \bibliographystyle{plain}

### 2025-05-20 paper/tables/token_statistics.tex
make it a lot more tight

@@
-\begin{table*}[htbp]
+\begin{table}[htbp]
 \centering
 \scriptsize
-\setlength{\tabcolsep}{3.5pt}
-\begin{tabular}{l|cc|cc|cc|cc|cc}
+\setlength{\tabcolsep}{3pt}
+\begin{tabular}{l|cc|cc|cc|c}
 \toprule
-& \multicolumn{2}{c|}{\textbf{Total Tokens}} & \multicolumn{2}{c|}{\textbf{Unique Tokens}} & \multicolumn{2}{c|}{\textbf{Vocab Coverage (\%)}} & \multicolumn{2}{c|}{\textbf{Top 20\% Coverage (\%)}} & \multicolumn{2}{c}{\textbf{Test Vocabulary Coverage}} \\
-\textbf{Task} & \textbf{Train} & \textbf{Test} & \textbf{Train} & \textbf{Test} & \textbf{Train} & \textbf{Test} & \textbf{Train} & \textbf{Test} & \textbf{OOV Tokens} & \textbf{OOV\%} \\
+\textbf{Task} & \multicolumn{2}{c|}{\textbf{Unique Tokens}} & \multicolumn{2}{c|}{\textbf{Vocab Cov. (\%)}} & \multicolumn{2}{c|}{\textbf{Top 20\%}} & \textbf{OOV} \\
+ & \textbf{Train} & \textbf{Test} & \textbf{Train} & \textbf{Test} & \textbf{Train} & \textbf{Test} & \textbf{\%} \\
 \midrule
-COLA & 71,818 & 7,939 & 5,416 & 1,934 & 17.74 & 6.34 & 85.12 & 74.68 & 169 & 8.74\% \\
-MNLI & 13,044,457 & 1,452,279 & 25,793 & 22,664 & 84.51 & 74.25 & 90.69 & 89.64 & 31 & 0.14\% \\
-MRPC & 165,585 & 18,701 & 11,096 & 3,567 & 36.35 & 11.69 & 83.17 & 71.80 & 465 & 13.04\% \\
-QNLI & 4,388,343 & 485,747 & 26,176 & 20,360 & 85.76 & 66.71 & 88.31 & 85.79 & 83 & 0.41\% \\
-QQP & 9,036,384 & 1,000,855 & 25,486 & 18,534 & 83.50 & 60.72 & 94.02 & 91.79 & 191 & 1.03\% \\
-RTE & 151,109 & 16,215 & 13,354 & 4,198 & 43.75 & 13.75 & 81.77 & 70.75 & 509 & 12.12\% \\
-SST2 & 686,566 & 75,775 & 11,536 & 8,084 & 37.80 & 26.49 & 84.23 & 80.53 & 34 & 0.42\% \\
-STSB & 128,151 & 14,481 & 10,346 & 3,271 & 33.90 & 10.72 & 82.66 & 71.30 & 448 & 13.70\% \\
+COLA & 5,416 & 1,934 & 17.74 & 6.34 & 85.12 & 74.68 & 8.74 \\
+MNLI & 25,793 & 22,664 & 84.51 & 74.25 & 90.69 & 89.64 & 0.14 \\
+MRPC & 11,096 & 3,567 & 36.35 & 11.69 & 83.17 & 71.80 & 13.04 \\
+QNLI & 26,176 & 20,360 & 85.76 & 66.71 & 88.31 & 85.79 & 0.41 \\
+QQP & 25,486 & 18,534 & 83.50 & 60.72 & 94.02 & 91.79 & 1.03 \\
+RTE & 13,354 & 4,198 & 43.75 & 13.75 & 81.77 & 70.75 & 12.12 \\
+SST2 & 11,536 & 8,084 & 37.80 & 26.49 & 84.23 & 80.53 & 0.42 \\
+STSB & 10,346 & 3,271 & 33.90 & 10.72 & 82.66 & 71.30 & 13.70 \\
 \bottomrule
 \end{tabular}
-\caption{Token statistics across GLUE tasks, comparing train and test splits.
-OOV Tokens: number of unique tokens in the test set that do not appear in the training set; OOV\%: percentage of test vocabulary not found in train.}
+\caption{Token statistics across GLUE tasks, comparing train and test splits. OOV\%: percentage of test vocabulary not found in train.}
 \label{tab:token_statistics}
-\end{table*}
+\end{table}

### 2025-05-22 paper/main.tex
final changes paper

@@
 \includegraphics[width=\columnwidth]{images/performance_retention.png}
-\caption{Average Performance Retention Across All Tasks for different pruning methods.
+\caption{Average performance retention across all GLUE tasks.
+Our vocabulary pruning removes tokens based on their importance using TF-IDF, with Out-of-Vocabulary (OOV) handling that maps removed tokens to semantically similar retained ones.
+LoSparse is a gradient-based pruning technique that targets encoder layers through low-rank plus sparse factorization.
 Pruning ratio represents the fraction of total parameters removed from the model, limited to 0.25 as the embedding layer constitutes 25.95\% of all parameters.}
@@
 
-Our results indicate that the memory footprint of light-weight models, such as ModernBERT, can still be significantly improved.
-Additionally, the superior performance of our simple vocabulary pruning methods highlights the need to further develop pruning methods.
-We open-source all our code and pruned models.
+Our results indicate that the memory footprint of light-weight models, such as ModernBERT, can still be significantly improved.  We open-source all our code and pruned models.
 
@@
 
-TODO: on modernbert, important!
-
 To our knowledge, this is the first study to (i) rigorously evaluate a range of importance heuristics in a single framework, (ii) incorporate OOV‑aware mapping to preserve semantic fidelity of pruned tokens, and (iii) apply vocabulary pruning atop ModernBERT—a state‑of‑the‑art lightweight encoder—thereby demonstrating the practical benefits of embedding‑layer reduction in resource‑constrained settings.
@@
 
-We compute token importance scores using either attention-based (Algorithm \ref{alg:pre_ft_pruning} lines 2-4) or statistical approaches (line 6) via \texttt{compute\_importance\_scores}, then sort and select the top $k$ tokens based on the pruning ratio $p$ (lines 7-9).
+We compute token importance scores using either attention-based (Algorithm \ref{alg:pre_ft_pruning}, lines 2-4) or statistical approaches (line 6) via \texttt{compute\_importance\_scores}, then sort and select the top $k$ tokens based on the pruning ratio $p$ (lines 7-9).
 The least important tokens are pruned from the embedding layer, and the embedding matrix is reconstructed using only the retained tokens via \texttt{rebuild\_model\_with\_vocab} (line 13), yielding a smaller model.
@@
 A critical challenge in vocabulary pruning is determining which tokens to retain and which to remove.
+Our approach first retains only tokens that appear in the training dataset, removing those never encountered during training.
+From this reduced set, we then apply more targeted pruning techniques to further decrease vocabulary size.
 
@@
     \includegraphics[width=\linewidth]{images/pruning_ratios.png}
-    \caption{Performance comparison between vocabulary pruning (TF-IDF + OOV) and encoder pruning (LoSparse).}
+    \caption{Performance comparison between vocabulary pruning (TF-IDF + OOV) and encoder pruning (LoSparse).
+Vocabulary pruning outperforms LoSparse at lower pruning ratios (5-20\%), but performance drops sharply beyond 20\% as the embedding layer's parameter limit is reached, while LoSparse maintains more stable performance at higher compression rates.}
     \label{fig:pruning_ratio}
@@
 \section{Experiments}
-We conducted a comprehensive evaluation across multiple natural language understanding tasks to assess the effectiveness of our vocabulary pruning methods compared to baseline approaches.
+We evaluate our vocabulary pruning methods across multiple natural language understanding tasks to assess the effectiveness of our vocabulary pruning methods and compare them to baseline approaches.
 
-\textbf{Datasets and Metrics} 
+\textbf{For datasets and metrics,} 
 The General Language Understanding Evaluation (GLUE) benchmark\cite{wang2018glue} serves as our primary evaluation framework, comprising eight diverse NLP tasks: single-sentence tasks (SST-2, CoLA), similarity/paraphrase tasks (MRPC, STS-B, QQP), and inference tasks (MNLI, QNLI, RTE).
@@
 
-\textbf{Baselines} 
-We establish three baseline models to evaluate our vocabulary pruning techniques: 
-(i) \textit{ModernBERT (Full)} serves as the upper performance bound, employing the complete 50,368-token vocabulary and all encoder parameters; 
-(ii) \textit{LoSparse} provides a complementary encoder-layer pruning baseline that uses low-rank plus sparse factorization to preserve 80\% of parameters in encoder layers while maintaining the full embedding layer;
-(iii) \textit{Train Tokens Only} represents a simple vocabulary reduction approach that removes embeddings not observed in finetuning data, achieving variable parameter reduction (4.79-22.61\%) depending on dataset vocabulary coverage.
+\textbf{As baselines,} 
+we establish three baseline models to evaluate our vocabulary pruning techniques: 
+(i) \textit{ModernBERT (Full)} serves as the upper performance bound, employing the complete 50,368-token vocabulary and all encoder parameters, finetuned on each task; 
+(ii) \textit{LoSparse} provides a complementary encoder-layer pruning baseline that uses low-rank and sparse factorization to preserve 80\% of parameters in encoder layers while maintaining the full embedding layer;
+(iii) \textit{Train Tokens Only} represents a simple vocabulary reduction approach that removes tokens not observed in finetuning data, achieving variable parameter reduction (4.79-22.61\%) depending on dataset vocabulary coverage.
 These baselines establish benchmarks for assessing both parameter efficiency and performance of our proposed methods.
 
-\textbf{Results}
+\textbf{For results,}
 Table \ref{tab:results} presents a comprehensive evaluation across GLUE tasks for different pruning methods. 
+% 
+Vocabulary pruning achieves substantial parameter reduction (20.02-23.27\%) with minimal performance impact, with TF-IDF + OOV maintaining 97.6\% of the original model's performance while reducing parameters by over 20\%.
+OOV handling mechanisms boost results compared to their non-OOV counterparts (+1.9 percentage points on average).
 
-Vocabulary pruning achieves substantial parameter reduction (20.02-23.27\%) with minimal performance impact, with TF-IDF + OOV maintaining 97.6\% of the original model's performance while reducing parameters by over 20\%.
-OOV handling mechanisms boost results compared to their non-OOV counterparts (+2.4 percentage points on average).
-
-Pruning effectiveness varies by task: attention-based methods excel on single-sentence tasks and paraphrase detection, while TF-IDF-based methods perform better on complex reasoning tasks.
+% Pruning effectiveness varies by task: attention-based methods excel on single-sentence tasks and paraphrase detection, while TF-IDF-based methods perform better on complex reasoning tasks. 
+Pruning effectiveness varies by task: attention-based methods excel on single-sentence tasks and paraphrase detection, while TFIDF-based methods perform better on complex reasoning tasks.
+We hypothesize this is because simpler tasks rely on specific sentiment-related tokens that receive high attention during inference, while complex reasoning requires a broader set of discriminative tokens that TF-IDF effectively identifies across documents.
 As shown in Figure \ref{fig:pruning_ratio}, vocabulary pruning outperforms encoder-layer pruning (LoSparse) at lower pruning ratios (5-20\%), while only using simple statistics before finetuning.
 A 20\% reduction in total parameters (77.34\% of vocabulary size) maintains competitive performance, making this approach particularly valuable for resource-constrained environments.
+Vocabulary pruning outperforms encoder-layer pruning (LoSparse) at lower pruning ratios (5-20\%), while only using simple statistics before fine-tuning, as shown in Figure \ref{fig:pruning_ratio}.
 
-The efficacy of vocabulary pruning is explained by the significant redundancy revealed in Table \ref{tab:token_statistics}, where the top 20\% of tokens cover 80-94\% of all occurrences in training sets across GLUE tasks.
+The efficacy of vocabulary pruning can be explained by the token-distribution, 20\% of tokens account for 80-94\% of all occurrences in training sets across GLUE tasks, as shown in Table \ref{tab:token_statistics}.
 This redundancy provides strong empirical justification for our approach.
-The table also highlights varying train-test vocabulary overlap across tasks, underscoring why effective OOV handling strategies significantly improve performance.
+
+The table also highlights varying train-test vocabulary overlap across tasks, with OOV percentages ranging from 0.14\% to 13.70\%.
+These high OOV rates for some tasks explain why effective OOV handling improves performance by providing semantic mappings for unseen tokens rather than discarding their meaning.
 
@@
 
-\newpage
+% \newpage
 \section{Conclusion}
-In this work, we designed and evaluated a series of vocabulary pruning techniques for the ModernBERT language model to reduce model size while maintaining task performance.
-
-We found that simple vocabulary pruning techniques outperform encoder-layer pruning for moderate compression ratios.
+We find that vocabulary pruning strategies can outperform more advanced encoder-layer pruning methods, for pruning ratios up to 20\% on the ModernBERT encoder model.
 Our TF-IDF-based method with OOV handling achieved 97.6\% of the original model's performance across GLUE tasks while reducing total parameters by 20.02\%, compared to LoSparse's 92.9\% retention at the same compression ratio.
@@
 
-These results challenge the assumption that encoder pruning should precede vocabulary reduction and highlight the significant redundancy within embedding layers of pre-trained models, providing an efficient path to model compression for resource-constrained environments.
+These results challenge the typical prioritization of encoder-layer pruning over vocabulary reduction and highlight the significant redundancy within embedding layers of pre-trained models, providing an efficient path to model compression for resource-constrained environments.
 
 \section{Future Work}
-Future research could explore an integrated approach combining vocabulary pruning with encoder-layer compression techniques like LoSparse, as preliminary experiments suggest these approaches are complementary.
-We hypothesize that pruning up to a certain threshold using this combined approach will yield better results than using either vocabulary pruning or encoder-layer pruning alone.
+Future research could explore an integrated approach combining vocabulary pruning with encoder-layer compression techniques like LoSparse.
+Since vocabulary pruning cannot exceed 25\% of total parameters (the embedding layer's size), and our results in Figure \ref{fig:pruning_retention} show vocabulary pruning outperforms LoSparse up to 20\%, we hypothesize that applying LoSparse on top of 20\% vocabulary pruning would yield better results than using LoSparse alone at 25\% compression.
 
-Additionally, evaluating vocabulary pruning on information retrieval tasks may reveal different optimal pruning strategies, as retrieval models likely depend on different token importance distributions than classification tasks.
+% Additionally, evaluating vocabulary pruning on information retrieval tasks may reveal different optimal pruning strategies, as retrieval models likely depend on different token importance distributions than classification tasks.
 
-\newpage
+% \newpage
 \bibliographystyle{plain}

### 2025-05-22 paper/tables/results.tex
fix
introduction improvement + make related works more concise
final changes paper

@@
 \multicolumn{10}{l}{\textbf{Baseline}} \\
-ModernBERT (Full) & 0.951 & 0.616 & 0.89 & 0.917 & 0.917 & 0.881 & 0.939 & 0.643 & 0.831 \\
+ModernBERT (Full) & 0.951 & 0.632 & 0.89 & 0.917 & 0.917 & 0.881 & 0.939 & 0.643 & 0.846 \\
 \midrule
 \multicolumn{10}{l}{\textbf{Encoder Layer Pruning}} \\
-LoSparse & 0.929 & 0.316 & 0.856 & 0.882 & 0.911 & 0.858 & 0.907 & 0.610 & 0.784 \\
+LoSparse & 0.929 & 0.525 & 0.856 & 0.882 & 0.911 & 0.858 & 0.907 & 0.610 & 0.810 \\
 \textit{Parameter Reduction} & \textit{20.16\%} & \textit{20.16\%} & \textit{20.16\%} & \textit{20.16\%} & \textit{20.16\%} & \textit{20.16\%} & \textit{20.16\%} & \textit{20.16\%} &  \\
@@
 Attention-Based & \textbf{0.947} & 0.589 & \textbf{0.864} & 0.885 & 0.763 & 0.683 & 0.791 & 0.578 & 0.763 \\
-TF-IDF Based & 0.933 & 0.520 & 0.807 & 0.901 & 0.898 & \textbf{0.860} & \textbf{0.909} & 0.574 & 0.800 \\
+TF-IDF Based & 0.933 & 0.520 & 0.807 & 0.901 & 0.898 & \textbf{0.860} & \textbf{0.909} & 0.610 & 0.805 \\
 \\ [-6pt]
@@
 \\[-6pt]
-Train Tokens Only & 0.950 & 0.630 & 0.861 & 0.917 & 0.917 & 0.883 & 0.915 & 0.639 & 0.815 \\
+Train Tokens Only & 0.950 & 0.630 & 0.861 & 0.917 & 0.917 & 0.883 & 0.915 & 0.639 & 0.839 \\
 \textit{Parameter Reduction} & \textit{18.85\%} & \textit{22.61\%} & \textit{18.56\%} & \textit{18.76\%} & \textit{4.79\%} & \textit{6.74\%} & \textit{6.42\%} & \textit{17.06\%} & \\

### 2025-05-23 paper/tables/benchmark_results.tex
make paper more compact
update tables

@@
 \end{tabular}
-\caption{Benchmark results comparing base ModernBERT with pruned variant. Tests with batch size 128 on NVIDIA RTX 4090.}
+% \caption{Benchmark results comparing base ModernBERT with pruned variant.}
+% \caption{Benchmark results comparing base ModernBERT with pruned variant during inference on an RTX 4090.}
+\caption{Benchmark results comparing base ModernBERT with pruned variant (TF-IDF + OOV) during inference on an RTX 4090.
+Our method achieves 20\% reduction in total parameters (77.34\% of embedding parameters) and a 14.83\% reduction in GPU memory with negligible impact on inference time.}
+
 \label{tab:benchmark_results}

### 2025-05-23 paper/tables/results.tex
update tables

@@
 \setlength{\tabcolsep}{9pt}
-\begin{tabular}{l@{\hspace{25pt}}cccccccc:c}
+\begin{tabular}{l@{\hspace{25pt}}ccccccccc}
 \toprule
-& \multicolumn{2}{c}{\textbf{Single Sentence}} & \multicolumn{3}{c}{\textbf{Paraphrase and Similarity}} & \multicolumn{3}{c}{\textbf{Natural Language Inference}} & \\
+& \multicolumn{2}{c}{\textbf{Single Sentence}} & \multicolumn{3}{c}{\textbf{Paraphrase and Similarity}} & \multicolumn{3}{c}{\textbf{Natural Language Inference}} & \multirow{2}{*}{\makebox[-20pt][c]{\vrule width 0.5pt height 175pt}\hspace{35pt}} \\
 \cmidrule(lr){2-3} \cmidrule(lr){4-6} \cmidrule(lr){7-9}
@@
 \midrule
-\multicolumn{10}{l}{\textbf{Encoder Layer Pruning}} \\
-LoSparse & 0.929 & 0.525 & 0.856 & 0.882 & 0.911 & 0.858 & 0.907 & 0.610 & 0.810 \\
-\textit{Parameter Reduction} & \textit{20.16\%} & \textit{20.16\%} & \textit{20.16\%} & \textit{20.16\%} & \textit{20.16\%} & \textit{20.16\%} & \textit{20.16\%} & \textit{20.16\%} &  \\
-\midrule
 \multicolumn{10}{l}{\textbf{Vocabulary Pruning}} \\
+Train Tokens Only & 0.950 & 0.630 & 0.861 & 0.917 & 0.917 & 0.883 & 0.915 & 0.639 & 0.839 \\
+\textit{Parameter Reduction} & \textit{18.85\%} & \textit{22.61\%} & \textit{18.56\%} & \textit{18.76\%} & \textit{4.79\%} & \textit{6.74\%} & \textit{6.42\%} & \textit{17.06\%} & \textit{14.22\%} \\
+\\ [-6pt]
+\hdashline
+\\[-6pt]
 Random Selection & 0.911 & 0.470 & 0.798 & 0.845 & 0.780 & 0.504 & 0.669 & 0.566 & 0.693 \\
 Clustering-Based & 0.899 & 0.051 & 0.714 & 0.786 & 0.774 & 0.501 & 0.510 & 0.566 & 0.600 \\
-Frequency-Based & 0.943 & 0.467 & 0.812 & 0.905 & 0.904 & 0.858 & 0.902 & 0.546 & 0.792 \\
 Attention-Based & \textbf{0.947} & 0.589 & \textbf{0.864} & 0.885 & 0.763 & 0.683 & 0.791 & 0.578 & 0.763 \\
+Simple Frequency-Based & 0.943 & 0.467 & 0.812 & 0.905 & 0.904 & 0.858 & 0.902 & 0.546 & 0.792 \\
 TF-IDF Based & 0.933 & 0.520 & 0.807 & 0.901 & 0.898 & \textbf{0.860} & \textbf{0.909} & 0.610 & 0.805 \\
@@
 \\[-6pt]
-Freq + OOV & 0.938 & 0.540 & 0.828 & \textbf{0.906} & \textbf{0.915} & 0.858 & 0.907 & 0.615 & 0.813 \\
+Simple Frequency + OOV & 0.938 & 0.540 & 0.828 & \textbf{0.906} & \textbf{0.915} & 0.858 & 0.907 & 0.615 & 0.813 \\
 TF-IDF + OOV & 0.923 & \textbf{0.615} & 0.834 & 0.903 & 0.912 & 0.858 & 0.908 & \textbf{0.635} & \textbf{0.824} \\
-\textit{Parameter Reduction} & \textit{20.25\%} & \textit{23.27\%} & \textit{20.02\%} & \textit{20.18\%} & \textit{20.02\%} & \textit{20.02\%} & \textit{20.02\%} & \textit{20.02\%} &  \\
-\\ [-6pt]
-\hdashline
-\\[-6pt]
-Train Tokens Only & 0.950 & 0.630 & 0.861 & 0.917 & 0.917 & 0.883 & 0.915 & 0.639 & 0.839 \\
-\textit{Parameter Reduction} & \textit{18.85\%} & \textit{22.61\%} & \textit{18.56\%} & \textit{18.76\%} & \textit{4.79\%} & \textit{6.74\%} & \textit{6.42\%} & \textit{17.06\%} & \\
+\textit{Parameter Reduction} & \textit{20.25\%} & \textit{23.27\%} & \textit{20.02\%} & \textit{20.18\%} & \textit{20.02\%} & \textit{20.02\%} & \textit{20.02\%} & \textit{20.02\%} & \textit{20.48\%} \\
+\midrule
+\multicolumn{10}{l}{\textbf{Encoder Layer Pruning}} \\
+LoSparse & 0.929 & 0.525 & 0.856 & 0.882 & 0.911 & 0.858 & 0.907 & 0.610 & 0.810 \\
+\textit{Parameter Reduction} & \textit{20.16\%} & \textit{20.16\%} & \textit{20.16\%} & \textit{20.16\%} & \textit{20.16\%} & \textit{20.16\%} & \textit{20.16\%} & \textit{20.16\%} & \textit{20.16\%} \\
 \bottomrule
@@
 Scores are accuracies except for CoLA (Matthew's correlation), and STS-B (Pearson correlation).
-"+ OOV" indicates the pruning technique combined with out-of-vocabulary clustering for token remapping.
-Parameter reduction percentages show total model size decrease for each method.}
+Notation "+ OOV" indicates pruning with out-of-vocabulary clustering.
+At 20\% parameter reduction, TF-IDF + OOV maintains 97.6\% of original performance and outperforms LoSparse (0.824 vs 0.810 average score).
+OOV handling improves results by 1.9 percentage points on average.}
+
 \label{tab:results}

### 2025-06-01 paper/main.tex
feedback + citations

@@
 
-\setcounter{secnumdepth}{0}
+% \setcounter{secnumdepth}{0}
 
@@
 We compare the resulting accuracy with LoSparse, a state-of-the-art gradient-based pruning technique that targets the encoder layers.
-For parameter reductions up to $\sim$20\%, our much simpler vocabulary pruning technique outperforms LoSparse, retaining up to 97.6\% of ModernBERT's performance across various tasks, compared to LoSparse's 92.9\%.
+For parameter reductions up to $\sim$20\%, our much simpler vocabulary pruning technique outperforms LoSparse, retaining up to 97.6\% of ModernBERT's performance across various tasks, compared to LoSparse's 96.2\%.
 The strong performance of our simple technique indicates that task-specific pruning can meaningfully increase the efficiency of ModernBERT, an already highly efficient model.
 Additionally, our results suggest that state-of-the-art encoder-layer pruning can fall short of simple embedding-layer pruning.
-We provide open-source implementations of all pruning methods and evaluation tools to support further research in this area.\footnote{\url{https://github.com/WoutDeRijck/vocab-pruning.git}}\footnote{\url{https://github.com/WoutDeRijck/LoSparse.git}}
+We provide open-source implementations of all pruning methods and evaluation tools to support further research in this area. (\url{https://github.com/WoutDeRijck/vocab-pruning.git})
 
@@
 \includegraphics[width=\columnwidth]{images/performance_retention.png}
-\caption{Average performance retention across all GLUE tasks.
-Our vocabulary pruning removes tokens based on their importance using TF-IDF, with Out-of-Vocabulary (OOV) handling that maps removed tokens to semantically similar retained ones.
-LoSparse is a gradient-based pruning technique that targets encoder layers through low-rank plus sparse factorization.
-Pruning ratio represents the fraction of total parameters removed from the model, limited to 0.25 as the embedding layer constitutes 25.95\% of all parameters.}
+\caption{Simple vocabulary pruning techniques retain \raisebox{0.2ex}{\scalebox{0.8}{$>$}}97\% of ModernBERT's performance while removing $20\%$ of total parameters across eight GLUE benchmark tasks.
+Our vocabulary pruning removes tokens based on their importance using \textit{TF-IDF}, with Out-of-Vocabulary (OOV) handling that maps removed tokens to semantically similar retained ones, substantially outperforming LoSparse---a state-of-the-art gradient-based encoder pruning method.
+Pruning ratio represents the fraction of total parameters removed from the model, limited to $0.25$ since the embedding layer constitutes $25.95\%$ of all model parameters, representing the theoretical maximum for vocabulary-only pruning.}
 \label{fig:pruning_retention}
@@
 It integrates several architectural and training innovations while maintaining robust downstream performance.
-Despite these advancements, a critical question remains open: can we further enhance ModernBERT's efficiency by pruning parameters?
-
-This question is particularly interesting for two reasons.
-First, ModernBERT's inherent efficiency means that optimal pruning strategies could yield exceptionally efficient models.
-Second, existing research into pruning techniques have predominantly targeted decoder-only LLMs \cite{namburi2023llm}, leaving encoder-only models relatively underexplored.
+Despite these advancements, we investigate whether ModernBERT's efficiency can be further enhanced through parameter pruning.
+Given ModernBERT's inherent efficiency, optimal pruning strategies could yield exceptionally efficient models.
 
 Specifically, we explore task-specific weight pruning, as ModernBERT is intended to be finetuned on specific tasks for optimal behavior (unlike more general-purpose decoder-only models that can utilize in-context learning).
-We hypothesize that simple task-specific pruning of the embedding matrix can be highly efficient, as many tokens can be irrelevant for any specific downstream task and the embedding layers counts for a significant portion ($\sim$25\% of the total parameters).
+We hypothesize that simple task-specific pruning of the embedding matrix can be highly efficient, as many tokens can be irrelevant for any specific downstream task, and the embedding layers counts for a significant portion ($\sim$25\%) of the total parameters.
 
-We explore a series of simple vocabulary pruning heuristics on a range of Natural Language Processing tasks including linguistic acceptability, sentiment analysis, paraphrase detection, semantic similarity, question answering, and natural language inference, as covered by the GLUE benchmark.
+We explore a series of simple vocabulary pruning heuristics on a range of Natural Language Processing tasks, including linguistic acceptability, sentiment analysis, paraphrase detection, semantic similarity, question answering, and natural language inference, as covered by the GLUE benchmark \cite{wang2018glue}.
 For any task, we first apply vocabulary pruning to the base ModernBERT model and subsequently finetune the pruned model.
-We compare these vocabulary pruning techniques with LoSparse, a state-of-the-art gradient-based pruning technique that targets the encoder layers.
+We compare these vocabulary pruning techniques with LoSparse, a state-of-the-art gradient-based pruning technique that targets the encoder layers \cite{li2023losparse}.
 
-Our results show that TF-IDF-based vocabulary pruning outperforms encoder-layer pruning (LoSparse) for compression ratios up to 20\%, maintaining 97.6\% of ModernBERT's original performance while removing 77.34\% of embedding parameters (see Figure \ref{fig:pruning_retention}).
+Our results show that TFIDF-based vocabulary pruning (with frequencies computed from the target task's training dataset) outperforms encoder-layer pruning (LoSparse) for compression ratios up to 20\%, maintaining 97.6\% of ModernBERT's original performance while removing 77.34\% of embedding parameters (see Figure \ref{fig:pruning_retention}).
 This shows that substantial efficiency gains can be achieved with virtually no performance loss using a simple, offline pruning technique.
-Beyond 20\% compression, performance drops sharply as we approach the embedding layer's parameter limit (25.95\% of total model parameters), while LoSparse maintains more stable performance at higher ratios.
-Unlike LoSparse, which requires computationally expensive gradient-based optimization, our vocabulary pruning operates as a straightforward pre-finetuning step with minimal overhead.
+Beyond 20\% compression, performance drops sharply as we approach the embedding layer's parameter limit (25.95\% of total model parameters),
+% while LoSparse maintains more stable performance at higher ratios.
+while LoSparse maintains more stable performance at higher ratios, since it can continue pruning across the larger encoder layers (74.05\% of total parameters). 
+Unlike LoSparse, which requires computationally expensive gradient-based optimization, our vocabulary pruning operates as a pre-finetuning step with minimal overhead.
 
@@
 (iii) \textbf{low-rank approximation}, which decomposes weight matrices into lower-dimensional representations that capture essential information while requiring fewer parameters~\cite{Hu2021LoRA}; and
-(iv) \textbf{knowledge distillation} transfers knowledge from larger teacher models to smaller student models, enabling competitive performance with reduced architecture size.~\cite{Hinton2015Distillation}
+(iv) \textbf{knowledge distillation} transfers knowledge from larger teacher models to smaller student models, enabling competitive performance with reduced architecture size ~\cite{Hinton2015Distillation}.
 
@@
 The model compression techniques described above are complementary and can be combined to achieve optimal compression.
-This work focuses on parameter pruning techniques to reduce model size while preserving performance.
-The rest of this section is dedicated to enumerating the most important parameter pruning techniques.
+This work focuses on parameter pruning techniques to reduce model size while preserving performance. The rest of this section gives an overview of the most important parameter pruning techniques.
 
@@
 These encoder-layer pruning methods differ in their granularity and strategy—targeting individual weights, attention heads, or full layers.
-LoSparse stands out as a strong baseline for encoder-level pruning.
-~\cite{pmlr-v235-li24bi}~\cite{huang2023awesome}~\cite{ren2025llmcompression}
+LoSparse stands out as a strong baseline for encoder-level pruning~\cite{pmlr-v235-li24bi}~\cite{huang2023awesome}~\cite{ren2025llmcompression}.
 
@@
 
-\textbf{Embedding compression} reduces the dimensional representation of the embeddings.
+\textit{\textbf{Embedding compression}} reduces the dimensional representation of the embeddings.
 LightToken~\cite{wang2023lighttoken} is a model- and task-agnostic framework for compressing the token embedding layers.
@@
 
-\textbf{Vocabulary Pruning Techniques} are a specialized form of embedding pruning that removes rarely used or task-irrelevant tokens.
+\textit{\textbf{Vocabulary Pruning Techniques}} are a specialized form of embedding pruning that removes rarely used or task-irrelevant tokens.
 
@@
 
-For multilingual models, most of the parameters are located in the embeddings layer, language-specific pruning has proven effective.
+For multilingual models, most of the parameters are located in the embedding layer, language-specific pruning has proven effective.
 Abdaoui et al.~\cite{abdaoui2020load} showed that trimming unused language tokens from mBERT can achieve a 45\% parameter reduction with comparable accuracy on XNLI.
@@
 \newpage
-\section{Method}
-We propose a task-specific vocabulary pruning method for ModernBERT that targets the embedding layer, which accounts for approximately 25\% of the model's parameters.
+% \section{Method}
+% \section{Token-level encoder compression}
+\section{\fontsize{13}{15.6}\selectfont Task-Specific Vocabulary Pruning}
+Our vocabulary pruning method is straightforward: we first remove all tokens that do not appear in our training dataset, then we eliminate an additional percentage of the remaining tokens based on importance scores.
+Optionally, we cluster removed tokens into groups and add the cluster centroids back into the vocabulary to handle out-of-vocabulary (OOV) cases during inference.
+
+We apply this method to ModernBERT, where the embedding layer accounts for approximately 25\% of the model's parameters.
 This approach is based on the observation that tokens in a vocabulary have varying importance for downstream tasks, and that certain tokens can be selectively removed with minimal impact on performance.
-We explore and compare several token importance estimation techniques, ranging from simple statistical methods to semantic analysis.
-The method consists of three components: token importance analysis, selective pruning, and an optional semantic clustering for out-of-vocabulary (OOV) tokens that can further enhance performance in some cases.
 
 In contrast to methods that require resource-intensive post-pruning finetuning, the proposed vocabulary pruning approach operates as a \textbf{pre-finetuning offline pruning} step in the model adaptation pipeline.
-Instead, it relies on either simple statistical measures or a single attention analysis pass, which does not require gradient calculations, resulting in minimal computational overhead before standard finetuning.
+It requires only the training dataset to determine token importance and prune the vocabulary, without needing gradients from finetuning, allowing the resulting pruned model to be finetuned with minimal computational overhead.
 
 We compute token importance scores using either attention-based (Algorithm \ref{alg:pre_ft_pruning}, lines 2-4) or statistical approaches (line 6) via \texttt{compute\_importance\_scores}, then sort and select the top $k$ tokens based on the pruning ratio $p$ (lines 7-9).
-The least important tokens are pruned from the embedding layer, and the embedding matrix is reconstructed using only the retained tokens via \texttt{rebuild\_model\_with\_vocab} (line 13), yielding a smaller model.
+The least important tokens are pruned from the embedding layer, and the embedding matrix is reconstructed using only the retained tokens (\texttt{rebuild\_model\_with\_vocab}, line 13), yielding a smaller model.
 To recover the loss of throwing away tokens, semantic clustering can be applied with \texttt{cluster\_pruned\_tokens} to map pruned tokens to semantically similar retained tokens (lines 10-12). 
@@
 A critical challenge in vocabulary pruning is determining which tokens to retain and which to remove.
-Our approach first retains only tokens that appear in the training dataset, removing those never encountered during training.
-From this reduced set, we then apply more targeted pruning techniques to further decrease vocabulary size.
+We explore several token importance estimation techniques, ranging from simple statistical methods to semantic analysis.
 
@@
 \textbf{Frequency-Based} ranks tokens by their frequency in the training dataset, pruning the least frequently used tokens first.
-This approach is grounded in the heuristic that infrequent tokens contribute less to model performance.~\cite {li2024enhancing}
+This approach is grounded in the heuristic that infrequent tokens contribute less to model performance ~\cite {li2024enhancing}.
 
@@
 
+% The clustering-based Out-Of-Vocabulary (OOV) handling process extracts embeddings for all pruned tokens and applies K-means clustering to group semantically similar tokens together.
+From each cluster, the token closest to the centroid is selected as a representative.
+During inference, when an OOV token is encountered, it is mapped to its assigned representative.
 The clustering-based Out-Of-Vocabulary (OOV) handling process extracts embeddings for all pruned tokens and applies K-means clustering to group semantically similar tokens together.
 From each cluster, the token closest to the centroid is selected as a representative.
+While alternative clustering methods (e.g., DBSCAN \cite{wang2019dbscan}) and representative selection strategies (e.g., averaging embeddings) could be explored, we focus on this approach for its simplicity and effectiveness.
 During inference, when an OOV token is encountered, it is mapped to its assigned representative.
@@
 After evaluating the different token importance metrics, we apply clustering-based OOV handling to the best-performing approach to further improve the results.
-% TODO: Add experimental results and a plot
-%Experimental results show that even a modest number of clusters (50-100) significantly improves model performance compared to a single UNK mapping approach.
 
@@
     \includegraphics[width=\linewidth]{images/pruning_ratios.png}
-    \caption{Performance comparison between vocabulary pruning (TF-IDF + OOV) and encoder pruning (LoSparse).
+    \caption{Performance comparison between vocabulary pruning (\textit{TF-IDF} with and without OOV) and encoder pruning (LoSparse).
 Vocabulary pruning outperforms LoSparse at lower pruning ratios (5-20\%), but performance drops sharply beyond 20\% as the embedding layer's parameter limit is reached, while LoSparse maintains more stable performance at higher compression rates.}
@@
 \section{Experiments}
-We evaluate our vocabulary pruning methods across multiple natural language understanding tasks to assess the effectiveness of our vocabulary pruning methods and compare them to baseline approaches.
+We evaluate our vocabulary pruning methods across multiple natural language understanding tasks to assess their effectiveness. Additionally, we compare them to baseline approaches.
 
 \textbf{For datasets and metrics,}
-The General Language Understanding Evaluation (GLUE) benchmark\cite{wang2018glue} serves as our primary evaluation framework, comprising eight diverse NLP tasks: single-sentence tasks (SST-2, CoLA), similarity/paraphrase tasks (MRPC, STS-B, QQP), and inference tasks (MNLI, QNLI, RTE).
+The General Language Understanding Evaluation (GLUE) benchmark\cite{wang2018glue} serves as our primary evaluation framework, comprising eight diverse NLP tasks: single-sentence tasks (SST-2\cite{socher2013recursive}, CoLA\cite{warstadt2019neural}), similarity/paraphrase tasks (MRPC\cite{dolan2005automatically}, STS-B\cite{cer2017semeval}, QQP\cite{qqp}), and inference tasks (MNLI\cite{williams2018broad}, QNLI\cite{rajpurkar2016squad}, RTE\cite{dagan2005pascal}).
 We report task-appropriate metrics: accuracy for most classification tasks, Matthew's correlation for CoLA, and Pearson correlation for STS-B.
@@
 \textbf{As baselines,} 
-we establish three baseline models to evaluate our vocabulary pruning techniques: 
-(i) \textit{ModernBERT (Full)} serves as the upper performance bound, employing the complete 50,368-token vocabulary and all encoder parameters, finetuned on each task; 
+we establish three baselines to evaluate our vocabulary pruning techniques: 
+(i) \textit{ModernBERT (Full)} serves as the upper performance bound, using the complete 50,368-token vocabulary and all encoder parameters, finetuned on the training dataset of each task; 
 (ii) \textit{LoSparse} provides a complementary encoder-layer pruning baseline that uses low-rank and sparse factorization to preserve 80\% of parameters in encoder layers while maintaining the full embedding layer;
-(iii) \textit{Train Tokens Only} represents a simple vocabulary reduction approach that removes tokens not observed in finetuning data, achieving variable parameter reduction (4.79-22.61\%) depending on dataset vocabulary coverage.
-These baselines establish benchmarks for assessing both parameter efficiency and performance of our proposed methods.
+(iii) \textit{Train Tokens Only} represents a simple vocabulary reduction approach that removes tokens not observed in finetuning data, achieving variable parameter reduction (4.79-22.61\%) depending on the dataset's vocabulary coverage.
 
-\textbf{For results,}
-Table \ref{tab:results} presents a comprehensive evaluation across GLUE tasks for different pruning methods. 
+\textbf{The results in}
+Table \ref{tab:results} present a comprehensive evaluation across GLUE tasks for different pruning methods. 
 % 
-Vocabulary pruning achieves substantial parameter reduction (20.02-23.27\%) with minimal performance impact, with TF-IDF + OOV maintaining 97.6\% of the original model's performance while reducing parameters by over 20\%.
-OOV handling mechanisms boost results compared to their non-OOV counterparts (+1.9 percentage points on average).
+Vocabulary pruning achieves substantial parameter reduction (20.02-23.27\%) with minimal performance impact, with \textit{TF-IDF + OOV} maintaining 97.6\% of the original model's performance while reducing parameters by over 20\%.
+OOV handling mechanisms boost results compared to their non-OOV counterparts (+2.0 percentage points on average).
 
-% Pruning effectiveness varies by task: attention-based methods excel on single-sentence tasks and paraphrase detection, while TF-IDF-based methods perform better on complex reasoning tasks. 
-Pruning effectiveness varies by task: attention-based methods excel on single-sentence tasks and paraphrase detection, while TFIDF-based methods perform better on complex reasoning tasks.
-We hypothesize this is because simpler tasks rely on specific sentiment-related tokens that receive high attention during inference, while complex reasoning requires a broader set of discriminative tokens that TF-IDF effectively identifies across documents.
-As shown in Figure \ref{fig:pruning_ratio}, vocabulary pruning outperforms encoder-layer pruning (LoSparse) at lower pruning ratios (5-20\%), while only using simple statistics before finetuning.
-A 20\% reduction in total parameters (77.34\% of vocabulary size) maintains competitive performance, making this approach particularly valuable for resource-constrained environments.
-Vocabulary pruning outperforms encoder-layer pruning (LoSparse) at lower pruning ratios (5-20\%), while only using simple statistics before fine-tuning, as shown in Figure \ref{fig:pruning_ratio}.
+The substantial performance degradation observed with random selection (0.693 average vs 0.846 baseline) demonstrates that strategic token selection is crucial for vocabulary pruning effectiveness.
 
-The efficacy of vocabulary pruning can be explained by the token-distribution, 20\% of tokens account for 80-94\% of all occurrences in training sets across GLUE tasks, as shown in Table \ref{tab:token_statistics}.
+Vocabulary pruning outperforms encoder-layer pruning (LoSparse) at lower pruning ratios (5-20\%), while only using simple statistics before finetuning, as shown in Figure \ref{fig:pruning_ratio}.
+Specifically, a 20\% reduction in total parameters (77.34\% of vocabulary size) maintains competitive performance.
+
+The efficacy of vocabulary pruning can be explained by the token-distribution: 20\% of tokens account for 80-94\% of all occurrences in training sets across GLUE tasks, as shown in Table \ref{tab:token_statistics}.
 This redundancy provides strong empirical justification for our approach.
 
-The table also highlights varying train-test vocabulary overlap across tasks, with OOV percentages ranging from 0.14\% to 13.70\%.
-These high OOV rates for some tasks explain why effective OOV handling improves performance by providing semantic mappings for unseen tokens rather than discarding their meaning.
+The table also highlights varying train-test vocabulary overlap across tasks, with unseen token percentages ranging from 0.14\% to 13.70\%.
+These high rates of unseen tokens for some tasks explain why effective OOV handling improves performance by providing semantic mappings for unseen tokens rather than discarding their meaning.
 
@@
 
-Practically, the benchmark results in Table \ref{tab:benchmark_results} demonstrate that our approach delivers 20\% storage reduction and 14.83\% lower GPU memory use with negligible inference time impact (+1.33\%).
+To assess the practical efficiency improvements, we benchmark the computational and memory requirements of our vocabulary pruning approach.
+The benchmark results in Table \ref{tab:benchmark_results} demonstrate that our approach delivers 20\% storage reduction and 14.83\% lower GPU memory use with negligible inference time impact (+1.33\%).
 
@@
 \input{tables/benchmark_results}
-
 % \newpage
@@
 We find that vocabulary pruning strategies can outperform more advanced encoder-layer pruning methods, for pruning ratios up to 20\% on the ModernBERT encoder model.
-Our TF-IDF-based method with OOV handling achieved 97.6\% of the original model's performance across GLUE tasks while reducing total parameters by 20.02\%, compared to LoSparse's 92.9\% retention at the same compression ratio.
+Our \textit{TFIDF-based} method with OOV handling achieved 97.6\% of the original model's performance across GLUE tasks while reducing total parameters by 20.02\%, compared to LoSparse's 96.2\% retention at the same compression ratio.
 This resulted in practical benefits of 14.83\% reduction in GPU memory requirements and 20\% less storage space.
@@
 
-\section{Future Work}
-Future research could explore an integrated approach combining vocabulary pruning with encoder-layer compression techniques like LoSparse.
-Since vocabulary pruning cannot exceed 25\% of total parameters (the embedding layer's size), and our results in Figure \ref{fig:pruning_retention} show vocabulary pruning outperforms LoSparse up to 20\%, we hypothesize that applying LoSparse on top of 20\% vocabulary pruning would yield better results than using LoSparse alone at 25\% compression.
+\section{Future Work and Limitations}
+Several limitations constrain our approach. 
+Our clustering-based methods for both token importance estimation and OOV handling rely on K-means clustering, while alternative methods such as DBSCAN~\cite{wang2019dbscan} might better capture complex semantic relationships in embedding spaces without requiring pre-specified cluster numbers.
+Similarly, our attention-based token importance method relies solely on attention scores, but recent work suggests this may be incomplete, as investigation into value vector norms reveals notably non-uniform patterns~\cite{guo2024attention}.
+Future work should explore Value-Aware Token Pruning (VATP) approaches that combine attention scores with the $\ell_1$ norm of value vectors for more comprehensive token importance evaluation.
 
-% Additionally, evaluating vocabulary pruning on information retrieval tasks may reveal different optimal pruning strategies, as retrieval models likely depend on different token importance distributions than classification tasks.
+Additionally, this work focuses exclusively on encoder-only models where embedding layers constitute approximately 25\% of parameters.
+In contrast, decoder-only models and encoder-decoder architectures typically have much larger transformer blocks relative to their embedding layers, meaning vocabulary pruning would yield substantially smaller compression benefits as embeddings represent only a small fraction of total parameters.
+This embedding layer constraint fundamentally limits our vocabulary pruning to 25\% total parameter reduction, beyond which performance degrades sharply.
+However, future research could explore combining vocabulary pruning with encoder-layer compression techniques like LoSparse to achieve higher compression ratios, applying LoSparse to the remaining 75\% of encoder parameters on top of optimal vocabulary pruning to potentially exceed either method's individual performance.
 
-% \newpage
-\bibliographystyle{plain}
+Finally, our approach is inherently task-specific, requiring separate pruning and finetuning for each target task, which increases deployment complexity compared to general-purpose compression methods, though it enables more aggressive compression aligned with ModernBERT's intended use case.
+
+{\footnotesize
+\bibliographystyle{abbrv}
 \bibliography{references}
+}
 

### 2025-06-01 paper/tables/results.tex
feedback + citations

@@
 Simple Frequency-Based & 0.943 & 0.467 & 0.812 & 0.905 & 0.904 & 0.858 & 0.902 & 0.546 & 0.792 \\
-TF-IDF Based & 0.933 & 0.520 & 0.807 & 0.901 & 0.898 & \textbf{0.860} & \textbf{0.909} & 0.610 & 0.805 \\
+TF-IDF Based & 0.933 & 0.520 & 0.807 & 0.901 & 0.898 & \textbf{0.860} & 0.909 & 0.610 & 0.805 \\
 \\ [-6pt]
@@
 Simple Frequency + OOV & 0.938 & 0.540 & 0.828 & \textbf{0.906} & \textbf{0.915} & 0.858 & 0.907 & 0.615 & 0.813 \\
-TF-IDF + OOV & 0.923 & \textbf{0.615} & 0.834 & 0.903 & 0.912 & 0.858 & 0.908 & \textbf{0.635} & \textbf{0.824} \\
+TF-IDF + OOV & 0.934 & \textbf{0.615} & 0.834 & 0.903 & 0.912 & 0.858 & \textbf{0.910} & \textbf{0.635} & \textbf{0.825} \\
 \textit{Parameter Reduction} & \textit{20.25\%} & \textit{23.27\%} & \textit{20.02\%} & \textit{20.18\%} & \textit{20.02\%} & \textit{20.02\%} & \textit{20.02\%} & \textit{20.02\%} & \textit{20.48\%} \\
@@
 \multicolumn{10}{l}{\textbf{Encoder Layer Pruning}} \\
-LoSparse & 0.929 & 0.525 & 0.856 & 0.882 & 0.911 & 0.858 & 0.907 & 0.610 & 0.810 \\
+LoSparse & 0.935 & 0.525 & 0.856 & 0.887 & 0.915 & 0.871 & 0.906 & 0.614 & 0.814 \\
 \textit{Parameter Reduction} & \textit{20.16\%} & \textit{20.16\%} & \textit{20.16\%} & \textit{20.16\%} & \textit{20.16\%} & \textit{20.16\%} & \textit{20.16\%} & \textit{20.16\%} & \textit{20.16\%} \\

### 2025-06-03 examples/run_clustering_pruning.py
add dbscan as clustering method

@@ -176,7 +176,7 @@ def parse_args():
         "--clustering_method", 
         type=str, 
         default=None,
-        choices=["agglomerative", "kmeans"],
+        choices=["agglomerative", "kmeans", "dbscan"],
         help="Override the default clustering method for all tasks"
     )
     

### 2025-06-03 main.py

@@ -135,7 +135,7 @@ def parse_args():
         "--clustering_method", 
         type=str, 
         default="agglomerative",
-        choices=["agglomerative", "kmeans"],
+        choices=["agglomerative", "kmeans", "dbscan"],
         help="Clustering algorithm (for clustering method)"
     )
     

### 2025-06-03 pruning/clustering.py

@@ -7,8 +7,9 @@ This module implements vocabulary pruning based on token embedding clustering.
 import logging
 import numpy as np
 import torch.nn as nn
-from sklearn.cluster import KMeans, AgglomerativeClustering
+from sklearn.cluster import KMeans, AgglomerativeClustering, DBSCAN
 from scipy.spatial.distance import pdist, squareform
+from scipy.cluster.hierarchy import linkage, fcluster
 from transformers import AutoTokenizer
 
 from .base import get_dataset_vocabulary, create_reduced_embeddings
@@ -27,7 +28,7 @@ def cluster_embeddings(token_ids, model, prune_percent, clustering_method="agglo
         token_ids: List of token IDs to consider for clustering
         model: Model whose embeddings will be used
         prune_percent: Percentage of vocabulary to prune through clustering
-        clustering_method: Method to use for clustering (agglomerative or kmeans)
+        clustering_method: Method to use for clustering (agglomerative, kmeans, or dbscan)
         param_based: If True, prune based on parameter percentage rather than token percentage
         total_params: Total parameters in the model (needed for parameter-based pruning)
         total_vocab_size: Total size of the original vocabulary
@@ -104,6 +105,119 @@ def cluster_embeddings(token_ids, model, prune_percent, clustering_method="agglo
         # Use K-means clustering
         clustering = KMeans(n_clusters=n_clusters, random_state=42)
         cluster_labels = clustering.fit_predict(normalized_embeddings)
+    elif clustering_method == "dbscan":
+        # Use DBSCAN clustering with post-processing to achieve target pruning
+        logger.info("Using DBSCAN clustering with post-processing to achieve target pruning ratio...")
+        
+        # Start with a reasonable eps value for initial clustering
+        initial_eps = 0.3
+        dbscan = DBSCAN(eps=initial_eps, min_samples=2, metric='cosine')
+        cluster_labels = dbscan.fit_predict(normalized_embeddings)
+        
+        # Handle noise points by assigning each to its own cluster
+        noise_points = np.where(cluster_labels == -1)[0]
+        if len(noise_points) > 0:
+            max_cluster = max(cluster_labels) if len(cluster_labels[cluster_labels != -1]) > 0 else -1
+            for i, noise_idx in enumerate(noise_points):
+                cluster_labels[noise_idx] = max_cluster + 1 + i
+        
+        initial_clusters = len(set(cluster_labels))
[…]

### 2025-06-04 paper/main.tex
dbscan results clustering

@@
 Language models need to be efficient for optimal deployment.
-Efficiency can be assessed in several dimensions, including inference latency, throughput, energy consumption, parameter count, and memory usage.
+Efficiency can be assessed in several dimensions, including inference latency, throughput, energy consumption, parameter count, and memory usage\cite{yuan2025efficientllm}.
 In this work, we study how to optimally reduce the memory footprint of language models by removing unimportant vocabulary tokens from the embedding layer, resulting in faster model loading, reduced deployment costs, and improved accessibility for edge devices or resource-constrained environments.
@@
 
-ModernBERT stands out as one of the most lightweight and optimized encoder-only transformers available.
+ModernBERT \cite{warner2024modernbert} stands out as one of the most lightweight and optimized encoder-only transformers available.
 It integrates several architectural and training innovations while maintaining robust downstream performance.
@@
 Specifically, we explore task-specific weight pruning, as ModernBERT is intended to be finetuned on specific tasks for optimal behavior (unlike more general-purpose decoder-only models that can utilize in-context learning).
-We hypothesize that simple task-specific pruning of the embedding matrix can be highly efficient, as many tokens can be irrelevant for any specific downstream task, and the embedding layers counts for a significant portion ($\sim$25\%) of the total parameters.
+We hypothesize that simple task-specific pruning of the embedding matrix can be highly efficient, as many tokens can be irrelevant for any specific downstream task, and the embedding layers counts for a significant portion ($\sim$25\%) of the total parameters \cite{yang2022taskspecific}.
 
@@
 For any task, we first apply vocabulary pruning to the base ModernBERT model and subsequently finetune the pruned model.
-We compare these vocabulary pruning techniques with LoSparse, a state-of-the-art gradient-based pruning technique that targets the encoder layers \cite{li2023losparse}.
+We compare these vocabulary pruning techniques with LoSparse \cite{li2023losparse}, a state-of-the-art gradient-based pruning technique that targets the encoder layers.
 
@@
 
-Our results indicate that the memory footprint of light-weight models, such as ModernBERT, can still be significantly improved.  We open-source all our code and pruned models.
+Our results indicate that the memory footprint of light-weight models, such as ModernBERT, can still be significantly improved.  We open-source all our code.
 
@@
 % 
-These techniques can be broadly categorized into four primary methods~\cite{wan2023efficient}:
-(i) \textbf{quantization}, which reduces the numerical precision of model weights;
-(ii) \textbf{parameter pruning}, which eliminates redundant or less important connections and neurons to create sparser models with fewer parameters;
+These techniques can be broadly categorized into four primary methods~\cite{wan2023efficient, park2024comprehensive, cheng2017survey, mishra2020survey, xailient2021compression}:
+(i) \textbf{quantization}, which reduces the numerical precision of model weights \cite{gholami2021survey};
+(ii) \textbf{parameter pruning}, which eliminates redundant or less important connections and neurons to create sparser models with fewer parameters \cite{cheng2023survey};
 (iii) \textbf{low-rank approximation}, which decomposes weight matrices into lower-dimensional representations that capture essential information while requiring fewer parameters~\cite{Hu2021LoRA}; and
@@
 %Misschien niet nodig voor de paper maar heel goed om je expertise te tonen in de full thesis
-The model compression techniques described above are complementary and can be combined to achieve optimal compression.
+The model compression techniques described above are complementary and can be combined to achieve optimal compression \cite{muralidharan2024compact, han2015deep, choukroun2019low}.
 This work focuses on parameter pruning techniques to reduce model size while preserving performance. The rest of this section gives an overview of the most important parameter pruning techniques.
@@
 These encoder-layer pruning methods differ in their granularity and strategy—targeting individual weights, attention heads, or full layers.
-LoSparse stands out as a strong baseline for encoder-level pruning~\cite{pmlr-v235-li24bi}~\cite{huang2023awesome}~\cite{ren2025llmcompression}.
+LoSparse stands out as a strong baseline for encoder-level pruning~\cite{pmlr-v235-li24bi, huang2023awesome, ren2025llmcompression}.
 
@@
 
-To our knowledge, this is the first study to (i) rigorously evaluate a range of importance heuristics in a single framework, (ii) incorporate OOV‑aware mapping to preserve semantic fidelity of pruned tokens, and (iii) apply vocabulary pruning atop ModernBERT—a state‑of‑the‑art lightweight encoder—thereby demonstrating the practical benefits of embedding‑layer reduction in resource‑constrained settings.
+To our knowledge, this is the first study to (i) rigorously evaluate a range of importance heuristics in a single framework, (ii) incorporate OOV‑aware mapping to preserve semantic fidelity of pruned tokens, and (iii) apply vocabulary pruning atop ModernBERT thereby demonstrating the practical benefits of embedding‑layer reduction in resource‑constrained settings.
 
@@
 \textbf{Clustering-Based}
-preserves semantic diversity across the vocabulary by grouping similar tokens together and keeping only one representative from each group.
-It leverages the property that embeddings pointing in similar directions (high cosine similarity) typically have similar meanings\cite{mikolov2013efficient}.
-This process ensures maximal semantic coverage with a reduced vocabulary size.
+preserves semantic diversity across the vocabulary using DBSCAN~\cite{wang2019dbscan} clustering to group similar token embeddings and keeping one representative per cluster.
+This leverages the property that embeddings with high cosine similarity typically have similar meanings\cite{mikolov2013efficient}, ensuring maximal semantic coverage with reduced vocabulary size.
 
@@
 
-%For example, the important tokens identified in the SST-2 dataset differ markedly between methods.
-Frequency-based approaches naturally prioritize common words like articles and prepositions, while attention-based analysis highlights sentiment-laden terms that are semantically relevant to the classification task.
-% \begin{figure*}[t]
-% \centering
-% \begin{minipage}{0.48\textwidth}
-%   \centering
-%   \includegraphics[width=\textwidth]{images/wordcloud_frequency.png}
-% \end{minipage}%
-% \hfill
-% \begin{minipage}{0.48\textwidth}
-%   \centering
-%   \includegraphics[width=\textwidth]{images/wordcloud_attention.png}
-% \end{minipage}
-% \caption{Comparison of top-ranked tokens for SST-2 task using frequency-based (left) versus attention-based (right) importance metrics.}
-% \label{fig:wordcloud-comparison}
-% \end{figure*}
-% The visualization in Figure \ref{fig:wordcloud-comparison} demonstrates how attention-based pruning better captures task-relevant information, preserving tokens that carry greater semantic weight for the specific downstream task.
-
 \textbf{Attention-Based}
 leverages the attention mechanism of transformer models to identify potentially important tokens.
-The underlying principle is that tokens receiving higher attention during inference might be those the model relies on more when making predictions, offering a model-based perspective on token importance\cite{guo2024attention}.
+The underlying principle is that tokens receiving higher attention during inference might be those the model relies on more when making predictions, offering a model-based perspective on token importance\cite{guo2024attention, kim2022learned, goyal2020powerbert, chen2021earlybert}.
 While attention scores do not necessarily imply causal relevance, they provide a task-specific signal that differs from simple statistical measures.
@@
 From each cluster, the token closest to the centroid is selected as a representative.
-While alternative clustering methods (e.g., DBSCAN \cite{wang2019dbscan}) and representative selection strategies (e.g., averaging embeddings) could be explored, we focus on this approach for its simplicity and effectiveness.
+While alternative clustering methods (e.g., DBSCAN) and representative selection strategies (e.g., averaging embeddings) could be explored, we focus on this approach for its simplicity and effectiveness.
 During inference, when an OOV token is encountered, it is mapped to its assigned representative.
@@
 (i) \textit{ModernBERT (Full)} serves as the upper performance bound, using the complete 50,368-token vocabulary and all encoder parameters, finetuned on the training dataset of each task; 
-(ii) \textit{LoSparse} provides a complementary encoder-layer pruning baseline that uses low-rank and sparse factorization to preserve 80\% of parameters in encoder layers while maintaining the full embedding layer;
+(ii) \textit{LoSparse} provides an encoder-layer pruning baseline that uses low-rank and sparse factorization to preserve 80\% of parameters in encoder layers while maintaining the full embedding layer;
 (iii) \textit{Train Tokens Only} represents a simple vocabulary reduction approach that removes tokens not observed in finetuning data, achieving variable parameter reduction (4.79-22.61\%) depending on the dataset's vocabulary coverage.
@@
 \section{Future Work and Limitations}
-Several limitations constrain our approach. 
-Our clustering-based methods for both token importance estimation and OOV handling rely on K-means clustering, while alternative methods such as DBSCAN~\cite{wang2019dbscan} might better capture complex semantic relationships in embedding spaces without requiring pre-specified cluster numbers.
+While our methods show promise, several limitations remain.
+Our clustering-based method for OOV handling relies on K-means clustering, while alternative methods such as DBSCAN~\cite{wang2019dbscan} might better capture complex semantic relationships in embedding spaces without requiring pre-specified cluster numbers.
 Similarly, our attention-based token importance method relies solely on attention scores, but recent work suggests this may be incomplete, as investigation into value vector norms reveals notably non-uniform patterns~\cite{guo2024attention}.

### 2025-06-04 paper/tables/results.tex
dbscan results clustering

@@
 Random Selection & 0.911 & 0.470 & 0.798 & 0.845 & 0.780 & 0.504 & 0.669 & 0.566 & 0.693 \\
-Clustering-Based & 0.899 & 0.051 & 0.714 & 0.786 & 0.774 & 0.501 & 0.510 & 0.566 & 0.600 \\
+Clustering-Based & 0.894 & 0.188 & 0.752 & 0.696 & 0.871 & 0.683 & 0.833 & 0.566 & 0.685 \\
 Attention-Based & \textbf{0.947} & 0.589 & \textbf{0.864} & 0.885 & 0.763 & 0.683 & 0.791 & 0.578 & 0.763 \\
@@
 Notation "+ OOV" indicates pruning with out-of-vocabulary clustering.
-At 20\% parameter reduction, TF-IDF + OOV maintains 97.6\% of original performance and outperforms LoSparse (0.824 vs 0.810 average score).
-OOV handling improves results by 1.9 percentage points on average.}
+At 20\% parameter reduction, TF-IDF + OOV maintains 97.6\% of the original performance and outperforms LoSparse (0.825 vs 0.810 average score).
+OOV handling improves results by 2.0 percentage points on average.}
 

### 2025-06-05 paper/tables/results.tex
more clear table

@@
 Simple Frequency-Based & 0.943 & 0.467 & 0.812 & 0.905 & 0.904 & 0.858 & 0.902 & 0.546 & 0.792 \\
-TF-IDF Based & 0.933 & 0.520 & 0.807 & 0.901 & 0.898 & \textbf{0.860} & 0.909 & 0.610 & 0.805 \\
+TF-IDF Based & 0.933 & 0.520 & 0.807 & 0.901 & 0.898 & 0.860 & 0.909 & 0.610 & 0.805 \\
 \\ [-6pt]
@@
 \multicolumn{10}{l}{\textbf{Encoder Layer Pruning}} \\
-LoSparse & 0.935 & 0.525 & 0.856 & 0.887 & 0.915 & 0.871 & 0.906 & 0.614 & 0.814 \\
+LoSparse & 0.935 & 0.525 & 0.856 & 0.887 & \textbf{0.915} & \textbf{0.871} & 0.906 & 0.614 & 0.814 \\
 \textit{Parameter Reduction} & \textit{20.16\%} & \textit{20.16\%} & \textit{20.16\%} & \textit{20.16\%} & \textit{20.16\%} & \textit{20.16\%} & \textit{20.16\%} & \textit{20.16\%} & \textit{20.16\%} \\
